### 2019-12-11 README.Rmd
first commit

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# berlin.trees

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to download or install

You can download the compendium as a zip from from this URL: </archive/master.zip>

Or you can install this compendium as an R package, berlin.trees, from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
remotes::install_github("/")
```

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2019-12-11 R/functions.R

@@ -1,0 +1,319 @@
+## Functions for reading/accessing data
+
+
+#' Download Trees from daten.berlin.de
+#'
+#'
+#' @details Downloads tree data from WFS Service for all packages containing 'baum',
+#' excluding the Baumscheiben data set.
+#'
+#' @export
+#' @import ckanr
+#'
+
+download_berlin_trees <- function(){
+
+
+    # test
+
+
+
+    # set base url
+    ckanr_setup(url = "https://datenregister.berlin.de")
+
+    # pull all packages from api, limit to 2500 (currentl ~ 2100 pckgs)
+    p_list <- package_list(as = "table", url = "https://datenregister.berlin.de/", limit = 2500)
+
+
+    # get tree-related data sets
+    baum_index <- grep("baum", p_list)
+    baum_packages <- p_list[baum_index]
+
+    package_drop <- grep("baumscheibe", baum_packages)
+    baum_packages <- baum_packages[-package_drop]
+
+    # download meta-data for resources in each package
+    package_results <- lapply(baum_packages, package_show)
+
+    # extract only resources with WFS Service data
+    wfs_subset <- purrr::map(package_results, "resources") %>%
+        purrr::map_depth(2, ~purrr::has_element(.x, "WFS Service"))
+
+    wfs_baum_resources <- purrr::map(package_results, "resources") %>%
+        purrr::flatten() %>%
+        purrr::keep(.p = unlist(wfs_subset))
+
+
+    # get names from pertinent packages,
+    # could call API again, but potentially costly
+    wfs_baum_packages <- baum_packages[unlist(
+        purrr::map(wfs_subset,
+                   function(x) any(x == TRUE)))]
+
+
+    # TEST IF COLLECTING RIGHT DATA
+    # identical(wfs_baum_packages %>%
+    #     purrr::map(package_show) %>%
+    #     purrr::map("id"),
+    #
+    #     wfs_baum_resources %>%
+    #         purrr::map("package_id") )
+    # TEST TRUE
+
+
+
+    # get urls for WFS service
+    wfs_urls <- wfs_baum_resources %>%
+        purrr::map("id") %>%
+        purrr::map(resource_show) %>%
+        purrr::map_chr("url")
+
+
+    # set up paths to write results to disk
+    json_paths <- wfs_baum_packages %>%
+        purrr::map(package_show) %>%
+        purrr::map_chr("title") %>%
+        purrr::map_chr(function(x){
+
+            file.path(here::here(),
[…]

### 2019-12-11 R/plan.R

@@ -1,0 +1,40 @@
+library(drake)
+
+
+plan <- drake_plan(
+    download_data = target(berlin.trees::download_berlin_trees()),
+
+    berlin_polygons = target(berlin.trees::get_berlin_polygons_as_sf(),
+                             trigger = trigger(condition = redownload)),
+
+    tree_data_in_lists = berlin.trees::load_downloaded_data_to_lists(download_data),
+
+
+    cropped_data_set = berlin.trees::crop_data_with_bbox(tree_data_in_lists,
+                                                         bounding_box),
+
+
+    bounding_box = berlin.trees::make_bbox(52.083962, 52.847599,
+                                           12.712024, 14.238359,
+                                           "greater_berlin",
+                                           crs = 4326),
+
+    full_data_set = berlin.trees::bind_rows_sf(cropped_data_set),
+
+    full_data_set_cleaned = berlin.trees::clean_genus(full_data_set),
+
+
+    overview_map = berlin.trees::make_overview_map(full_data_set_cleaned,
+                                                   berlin_polygons)
+
+
+    # data = raw_data %>%
+    #     mutate(Species = forcats::fct_inorder(Species)),
+    # hist = create_plot(data),
+    # fit = lm(Sepal.Width ~ Petal.Width + Species, data),
+    # report = rmarkdown::render(
+    #     knitr_in("report.Rmd"),
+    #     output_file = file_out("report.html"),
+    #     quiet = TRUE
+    # )
+)

### 2019-12-11 R/utils.R

@@ -1,0 +1,12 @@
+# # util functions
+#
+#
+# #' Pipe
+# #'
+# #' Put description here
+# #'
+# #' @importFrom magrittr %>%
+# #' @name %>%
+# #' @rdname pipe
+# #' @export
+# NULL

### 2019-12-11 _drake.R

@@ -1,0 +1,21 @@
+# This file serves the r_*() functions (e.g. r_make()) documented at
+# https://books.ropensci.org/drake/projects.html#safer-interactivity # nolint
+# and
+# https://docs.ropensci.org/drake/reference/r_make.html
+
+# Load your packages and supporting functions into your session.
+# If you use supporting scripts like the ones below,
+# you will need to supply them yourself. Examples:
+# https://github.com/wlandau/drake-examples/tree/master/main/R
+# source("R/packages.R")  # Load your packages, e.g. library(drake).
+# source("R/functions.R") # Define your custom code as a bunch of functions.
+# devtools::install(upgrade = FALSE)
+
+drake::expose_imports("berlin.trees",
+                      character_only = TRUE)
+redownload <- FALSE
+
+source("R/plan.R")      # Create your drake plan.
+# _drake.R must end with a call to drake_config().
+# The arguments to drake_config() are basically the same as those to make().
+drake_config(plan)

### 2019-12-11 _setup.R

@@ -1,0 +1,30 @@
+# usethis::use_package("ckanr")
+# usethis::use_package("ggplot2")
+# usethis::use_package("dplyr")
+# usethis::use_package("xml2")
+# usethis::use_package("httr")
+# usethis::use_package("purrr")
+# usethis::use_package("stringr")
+# usethis::use_package("sf")
+# usethis::use_package("future.callr")
+# usethis::use_package("drake")
+# usethis::use_package("furrr")
+usethis::use_package("tools")
+
+## packages
+
+library(drake)
+library(future.callr)
+future::plan(future.callr::callr)
+
+# clean()
+#
+devtools::install(upgrade = FALSE)
+drake::r_outdated(source = './_drake.R')
+# redownload <- FALSE
+
+drake::r_make()
+
+
+source("R/plan.R")      # Create your drake plan.
+drake::vis_drake_graph(config = drake_config(plan))

### 2019-12-11 analysis/scripts/01_get_tree_data.R

@@ -1,0 +1,185 @@
+library(httr)
+library(xml2)
+library(dplyr)
+library(ckanr)
+library(ggplot2)
+
+
+
+# set base url
+ckanr_setup(url = "https://datenregister.berlin.de")
+
+# pull all packages from api, limit to 2500 (currentl ~ 2100 pckgs)
+p_list <- package_list(as = "table", url = "https://datenregister.berlin.de/", limit = 2500)
+
+
+# check for tree-related data sets
+-
+
+# extract target urls for first packages (street trees)
+anlagen_id <- package_results[[1]]$resources[[2]]$id
+anlagen_url <- resource_show(id = anlagen_id)$url
+
+# request data via GET (REST API)
+
+
+json_path <- file.path(here::here(),
+                       "analysis",
+                       "data",
+                       "raw_data",
+                       paste0(package_results[[1]]$title,
+                              ".json"))
+
+
+# n <- 10
+query <- list(service = "WFS",
+              request = "GetFeature",
+              version = "2.0.0",
+              TypeNames = "s_wfs_baumbestand_an",
+              # count = n,
+              outputFormat = 'application/geo+json')
+
+
+# extract content
+
+httr::GET(anlagen_url, query = query, httr::write_disk(json_path, overwrite = TRUE))
+trees_sf <- sf::read_sf(json_path)
+
+## Alternative approach via loading data as geojson string directly into memory
+# result <- httr::GET(anlagen_url, query = query, httr::write_disk(json_path, overwrite = TRUE))
+# tree_data <- httr::content(result, type = 'text')
+# trees_sf <- sf::read_sf(tree_data)
+
+
+
+
+
+
+
+# get berlin district shapes
+berlin_poly <-  sf::read_sf("https://raw.githubusercontent.com/m-hoerz/berlin-shapes/master/berliner-bezirke.geojson")
+
+
+many_trees.plot <- trees_sf %>%
+
+    mutate(STAMMUMFG = as.numeric(STAMMUMFG),
+           GATTUNG = forcats::fct_infreq(as.factor(GATTUNG)) ,
+           gattung_short = forcats::fct_lump(GATTUNG,10) %>% forcats::fct_rev(),
+           bezirk_num = as.numeric(as.factor(BEZIRK))) %>%
+    filter(STAMMUMFG > 62.83185) %>%
+    ggplot() +
+    ggplot2::geom_sf(inherit.aes = FALSE,
+                     data = berlin_poly,
+                     fill = "transparent",
[…]

### 2019-12-11 analysis/scripts/02_get_tree_data_full.R

@@ -1,0 +1,114 @@
+
+
+
+# set base url
+ckanr_setup(url = "https://datenregister.berlin.de")
+
+# pull all packages from api, limit to 2500 (currentl ~ 2100 pckgs)
+p_list <- package_list(as = "table", url = "https://datenregister.berlin.de/", limit = 2500)
+
+
+# get tree-related data sets
+baum_index <- grep("baum", p_list)
+baum_packages <- p_list[baum_index]
+
+package_drop <- grep("baumscheibe", baum_packages)
+baum_packages <- baum_packages[-package_drop]
+
+# download meta-data for resources in each package
+package_results <- lapply(baum_packages, package_show)
+
+# extract only resources with WFS Service data
+wfs_subset <- purrr::map(package_results, "resources") %>%
+    purrr::map_depth(2, ~purrr::has_element(.x, "WFS Service"))
+
+wfs_baum_resources <- purrr::map(package_results, "resources") %>%
+    purrr::flatten() %>%
+   purrr::keep(.p = unlist(wfs_subset))
+
+
+# get names from pertinent packages,
+# could call API again, but potentially costly
+wfs_baum_packages <- baum_packages[unlist(
+    purrr::map(wfs_subset,
+               function(x) any(x == TRUE)))]
+
+
+# TEST IF COLLECTING RIGHT DATA
+# identical(wfs_baum_packages %>%
+#     purrr::map(package_show) %>%
+#     purrr::map("id"),
+#
+#     wfs_baum_resources %>%
+#         purrr::map("package_id") )
+# TEST TRUE
+
+
+
+# get urls for WFS service
+wfs_urls <- wfs_baum_resources %>%
+            purrr::map("id") %>%
+    purrr::map(resource_show) %>%
+    purrr::map_chr("url")
+
+
+# set up paths to write results to disk
+json_paths <- wfs_baum_packages %>%
+        purrr::map(package_show) %>%
+        purrr::map_chr("title") %>%
+    purrr::map_chr(function(x){
+
+        file.path(here::here(),
+                  "analysis",
+                  "data",
+                  "raw_data",
+                  paste0(x,
+                         ".json"))
+    })
+
+
+# get feature names for GET request via WFS urls
+feature_names <- purrr::map(wfs_urls, function(x){
+
+    capabilities <- httr::GET(x)
+    xml_capabilities <- content(capabilities, type = "text/xml")
+
+    feature_name <- xml_contents(
+        xml_child(
+            xml_child(
[…]

### 2019-12-11 make.R

@@ -1,0 +1,22 @@
+# Walkthrough: https://books.ropensci.org/drake/walkthrough.html
+# Download the code: drake_example("main") # nolint
+
+# Load your packages and supporting functions into your session.
+# If you use supporting scripts like the ones below,
+# you will need to supply them yourself. Examples:
+# https://github.com/wlandau/drake-examples/tree/master/main/R
+# source("R/packages.R")  # Load your packages, e.g. library(drake).
+# source("R/functions.R") # Define your custom code as a bunch of functions.
+# source("R/plan.R")      # Create your drake plan.
+# devtools::install(upgrade = FALSE)
+
+drake::expose_imports("berlin.trees",
+                      character_only = TRUE)
+redownload <- FALSE
+
+source("R/plan.R")
+
+# Call make() to run your work.
+# Your targets will be stored in a hidden .drake/ cache,
+# and you can read them back into memory with loadd() and readd().
+# make(plan)

### 2020-01-09 analysis/paper/paper.Rmd
cont work

---
title: "Exploratory overview of Berlin city trees"
author:
  - Alexander Hurley:
      email: hurley@gfz-potsdam.de
      institute: [GFZ]
      correspondence: true
  - Ingo Heinrich:
      email: heinrich@gfz-potsdam.de
      institute: [GFZ]
      correspondence: false
institute:
  - GFZ: Geoforschungszentrum Potsdam
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes

bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Quick overview of publicly available tree records.
keywords: |
  Uferbaeume; Anlagenbaeume; Strassenhighlights
highlights: |
  We will see. 
---

<!--     bookdown::word_document2: -->
<!--       fig_caption: yes -->
<!--       reference_docx: "../templates/template.docx" # Insert path for the DOCX file -->
<!--       pandoc_args: -->
<!--       - --lua-filter=../templates/scholarly-metadata.lua -->
<!--       - --lua-filter=../templates/author-info-blocks.lua -->
<!--       - --lua-filter=../templates/pagebreak.lua -->

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

```{r setup, echo = FALSE, cache = FALSE, include = FALSE}
knitr::opts_knit$set(root.dir = here::here())

knitr::opts_chunk$set(
  fig.align = "center",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  # fig.path = "../figures/",
  fig.path = paste0(file.path(here::here(), "analysis", "figures"), "/"),
  dpi = 300
)

library(drake)
library(ggplot2)
```

# Introduction

Here is a citation [@Marwick2017]

# Background

# Methods

# Results

Some results

```{r fig-tree-overview-map, fig.cap="A plot of random numbers",fig.width = 15, fig.height = 8, out.width="100%"}
readd(plot_overview_map)
# plot_overview_map

```

```{r fig-tree-count-map, fig.cap="A plot of random numbers",fig.width = 12, fig.height = 12, out.width="100%"}
readd(plot_count_map)
# plot_count_map

# plot(1:10)
```

<!-- Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot -->

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->
\newpage

# References 
<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
<div id="refs"></div>

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2020-01-09 R/functions.R

@@ -1,4 +1,4 @@
-## Functions for reading/accessing data
+# Functions for reading, accessing and wrangling data --------------------
 
 
 #' Download Trees from daten.berlin.de
@@ -158,6 +158,53 @@ load_downloaded_data_to_lists <- function(download_data){
 }
 
 
+#' Bind rows of sf tibbles to single data set
+#'
+#' @param sf_list list containing simple feature tibbles
+#'
+#' @return single sf-tibble
+#' @export
+#'
+bind_rows_sf <- function(sf_list){
+    # bind data without geometry
+    sf_df  <- lapply(sf_list, function(x) sf::st_set_geometry(x, NULL)) %>%
+        dplyr::bind_rows()
+
+    # make data frame and set geometry based on original sfcs
+    d_sf  <-  sf::st_sf(sf_df, geometry = purrr::map(sf_list, "geometry") %>%
+                            do.call(c, .))
+
+    return(d_sf)
+
+}
+
+
+#' Clean Feature Meta Data
+#'
+#' @param sf_data sf-tibble of Berlin City trees
+#'
+#' @description Formats upper/lower case and removes "" as genera; still need to check if other
+#' problematic values exist or can be recovered from other meta data.
+#'
+#' @return sf tibble with cleaned genus names
+#' @export
+#'
+clean_genus <- function(sf_data){
+
+    sf_data$GATTUNG <- tools::toTitleCase(tolower(sf_data$GATTUNG))
+    sf_data$GATTUNG[sf_data$GATTUNG==""] <- NA
+
+    return(sf_data)
+
+}
+
+
+
+# Spatial -----------------------------------------------------------------
+
+
+
+
 #' Define bounding box as area of interest
 #'
 #' @param lat_min numeric
@@ -198,43 +245,6 @@ make_bbox <- function(lat_min, lat_max, lon_min, lon_max, feature_name, crs = 43
 
 
 
-
-#' Bind rows of sf tibbles to single data set
-#'
-#' @param sf_list list containing simple feature tibbles
-#'
-#' @return single sf-tibble
-#' @export
-#'
-bind_rows_sf <- function(sf_list){
-    # bind data without geometry
-    sf_df  <- lapply(sf_list, function(x) sf::st_set_geometry(x, NULL)) %>%
-        dplyr::bind_rows()
-
-    # make data frame and set geometry based on original sfcs
[…]

### 2020-01-09 R/plan.R

@@ -2,39 +2,63 @@ library(drake)
 
 
 plan <- drake_plan(
-    download_data = target(berlin.trees::download_berlin_trees()),
 
-    berlin_polygons = target(berlin.trees::get_berlin_polygons_as_sf(),
-                             trigger = trigger(condition = redownload)),
-
-    tree_data_in_lists = berlin.trees::load_downloaded_data_to_lists(download_data),
+    # getting data
 
 
-    cropped_data_set = berlin.trees::crop_data_with_bbox(tree_data_in_lists,
-                                                         bounding_box),
 
+    ## Spatial Ancillary
+
+    ### Berlin districts and BBoxx
+    berlin_polygons = target(berlin.trees::get_berlin_polygons_as_sf()),
 
     bounding_box = berlin.trees::make_bbox(52.083962, 52.847599,
                                            12.712024, 14.238359,
                                            "greater_berlin",
                                            crs = 4326),
 
+    ### Berlin UHI gridded data
+
+
+
+    uhi_stacks = berlin.trees::get_uhi_rasters(file.path(here::here(), "analysis", "data", "raw_data", "UHI_explorer")),
+
+
+
+    ## Berlin trees
+    download_data = target(berlin.trees::download_berlin_trees()),
+
+    ### Cleaning
+    tree_data_in_lists = berlin.trees::load_downloaded_data_to_lists(download_data),
+
+
+    cropped_data_set = berlin.trees::crop_data_with_bbox(tree_data_in_lists,
+                                                         bounding_box),
     full_data_set = berlin.trees::bind_rows_sf(cropped_data_set),
 
-    full_data_set_cleaned = berlin.trees::clean_genus(full_data_set),
+    full_data_set_clean = target(berlin.trees::clean_genus(full_data_set),
+                                 format = "rds"),
 
 
-    overview_map = berlin.trees::make_overview_map(full_data_set_cleaned,
-                                                   berlin_polygons)
+    # Plotting
+    plot_overview_map = berlin.trees::make_overview_map(full_data_set_clean,
+                                                        berlin_polygons),
+
+    plot_tree_sums_bar = berlin.trees::tree_sums_bar_plot(full_data_set_clean),
+
+    plot_count_map = berlin.trees::tree_count_map(full_data_set_clean,
+                                                  berlin_polygons),
 
 
-    # data = raw_data %>%
[…]

### 2020-01-09 R/utils.R

@@ -1,12 +1,7 @@
-# # util functions
-#
-#
-# #' Pipe
-# #'
-# #' Put description here
-# #'
-# #' @importFrom magrittr %>%
-# #' @name %>%
-# #' @rdname pipe
-# #' @export
-# NULL
+#' re-export magrittr pipe operator
+#'
+#' @importFrom magrittr %>%
+#' @name %>%
+#' @rdname pipe
+#' @export
+NULL

### 2020-01-09 _drake.R

@@ -10,12 +10,16 @@
 # source("R/packages.R")  # Load your packages, e.g. library(drake).
 # source("R/functions.R") # Define your custom code as a bunch of functions.
 # devtools::install(upgrade = FALSE)
+# drake::expose_imports("bookdown",
+#                       character_only = TRUE)
 
 drake::expose_imports("berlin.trees",
                       character_only = TRUE)
-redownload <- FALSE
+# redownload <- FALSE
 
 source("R/plan.R")      # Create your drake plan.
 # _drake.R must end with a call to drake_config().
 # The arguments to drake_config() are basically the same as those to make().
+# drake::expose_imports("berlin.trees",
+                      # character_only = TRUE)
 drake_config(plan)

### 2020-01-09 _setup.R

@@ -9,7 +9,12 @@
 # usethis::use_package("future.callr")
 # usethis::use_package("drake")
 # usethis::use_package("furrr")
-usethis::use_package("tools")
+# usethis::use_package("tools")
+# usethis::use_package("forcats")
+# usethis::use_package("tibble")
+# usethis::use_package("rmarkdown")
+# usethis::use_package("raster")
+# usethis::use_package("magrittr")
 
 ## packages
 
@@ -19,7 +24,9 @@ future::plan(future.callr::callr)
 
 # clean()
 #
-devtools::install(upgrade = FALSE)
+devtools::install(upgrade = TRUE)
+
+
 drake::r_outdated(source = './_drake.R')
 # redownload <- FALSE
 

### 2020-01-09 analysis/scripts/03_dump.R

@@ -1,0 +1,165 @@
+get_uhi_hist_data <- function(uhi_stack_list, sf_data){
+
+
+    # brief checks
+    if (!is(sf_data, "sf")) {
+        stop("input data is not in sf")
+    }
+    # if (!is(uhi_stack_list, "RasterLayer")) {
+    #     stop("input raster is not a RasterLayer")
+    # }
+
+
+    # LOGIC
+
+
+
+    ## raster stats
+
+    stat_funs <- c("median", "mean", "sd")
+
+    ### cycle into list containing the 4 (total) stacks (2x summer, 2x winter)
+    raster_stats <- lapply(stat_funs,
+                           function(stats) {
+
+                               # into list (summer vs. winter)
+                               lapply(uhi_stack_list,
+                                      function(stacks) {
+                                          # into stacks (day vs. night)
+                                          lapply(stacks,
+                                                 raster::cellStats, stats)
+                                      })
+                           }) %>%
+        setNames(stat_funs)
+
+
+    ## tree uhi exposure
+
+    ### prepare tree data
+
+    sf_data <- sf_data %>%
+        dplyr::filter(!is.na(GATTUNG)) %>%
+        dplyr::mutate(STAMMUMFG = as.numeric(STAMMUMFG),
+                      # GATTUNG = forcats::fct_infreq(as.factor(GATTUNG)),
+
+
+                      gattung_short = forcats::fct_lump(as.factor(GATTUNG),11, other_level = "Other") %>%
+                          as.character(),
+                      gattung_short =   ifelse(GATTUNG == "Pinus", "Pinus", gattung_short),
+                      gattung_short = forcats::fct_infreq(as.factor(gattung_short)),
+                      gattung_short = forcats::fct_relevel(gattung_short, "Other", after = Inf),
+
+
+                      bezirk_num = as.numeric(as.factor(BEZIRK)))
+
+
+
+    ### re-project raster to UTM of sf data
+
+    uhi_stack_list <- lapply(uhi_stack_list,
+                          function(stack)
+                              lapply(stack, function(x)  {
+
+                                  raster::projectRaster(x,
+                                                        crs =  sf::st_crs(sf_data)$proj4string)
+
+                              }  ))
+
+
+    extract_vals <- lapply(uhi_stack_list,
+                           function(stacks){
[…]

### 2020-01-09 make.R

@@ -12,11 +12,11 @@
 
 drake::expose_imports("berlin.trees",
                       character_only = TRUE)
-redownload <- FALSE
+# redownload <- FALSE
 
 source("R/plan.R")
 
 # Call make() to run your work.
 # Your targets will be stored in a hidden .drake/ cache,
 # and you can read them back into memory with loadd() and readd().
-# make(plan)
+make(plan)

### 2020-01-13 analysis/paper/paper.Rmd
cont work

@@
 
+```{r fig-bar-tree, fig.cap="A plot of random numbers",fig.width = 15, fig.height = 8, out.width="100%"}
+
+# loadd(full_data_set_clean)
+# loadd(plot_overview_map)
+# print(class(full_data_set_clean))
+readd(plot_tree_sums_bar)
+# plot_overview_map
+
+```
+
 ```{r fig-tree-overview-map, fig.cap="A plot of random numbers",fig.width = 15, fig.height = 8, out.width="100%"}
+
+# loadd(full_data_set_clean)
+# loadd(plot_overview_map)
+# print(class(full_data_set_clean))
 readd(plot_overview_map)

### 2020-01-13 R/functions.R

@@ -189,10 +189,33 @@ bind_rows_sf <- function(sf_list){
 #' @return sf tibble with cleaned genus names
 #' @export
 #'
-clean_genus <- function(sf_data){
+clean_data <- function(sf_data){
 
     sf_data$GATTUNG <- tools::toTitleCase(tolower(sf_data$GATTUNG))
     sf_data$GATTUNG[sf_data$GATTUNG==""] <- NA
+
+
+    too_old_idx <- which(as.numeric(sf_data$STANDALTER) > 1500)
+
+
+    sf_data[too_old_idx, c("STANDALTER", "PFLANZJAHR")] <-
+        sf_data[too_old_idx, rev(c("STANDALTER", "PFLANZJAHR"))]
+
+
+    sf_data <- sf_data %>%
+        dplyr::filter(!is.na(GATTUNG)) %>%
+        dplyr::mutate(STAMMUMFG = as.numeric(STAMMUMFG),
+                      # GATTUNG = forcats::fct_infreq(as.factor(GATTUNG)),
+
+
+                      gattung_short = forcats::fct_lump(as.factor(GATTUNG),11, other_level = "Other") %>%
+                          as.character(),
+                      gattung_short =  ifelse(GATTUNG == "Pinus", "Pinus", gattung_short),
+                      gattung_short = forcats::fct_infreq(as.factor(gattung_short)),
+                      gattung_short = forcats::fct_relevel(gattung_short, "Other", after = Inf),
+
+
+                      bezirk_num = as.numeric(as.factor(BEZIRK)))
 
     return(sf_data)
 
@@ -392,7 +415,6 @@ get_uhi_rasters <- function(path){
 #'
 #' @export
 #'
-#' @examples
 add_uhi_hist_data <- function(uhi_stack_list, sf_data){
 
 
@@ -409,43 +431,24 @@ add_uhi_hist_data <- function(uhi_stack_list, sf_data){
 
 
 
-    ## raster stats
-
-    stat_funs <- c("median", "mean", "sd")
-
-    ### cycle into list containing the 4 (total) stacks (2x summer, 2x winter)
-    raster_stats <- lapply(stat_funs,
-                           function(stats) {
-
-                               # into list (summer vs. winter)
-                               lapply(uhi_stack_list,
-                                      function(stacks) {
-                                          # into stacks (day vs. night)
-                                          lapply(stacks,
-                                                 raster::cellStats, stats)
-                                      })
[…]

### 2020-01-13 R/plan.R

@@ -36,7 +36,7 @@ plan <- drake_plan(
                                                          bounding_box),
     full_data_set = berlin.trees::bind_rows_sf(cropped_data_set),
 
-    full_data_set_clean = target(berlin.trees::clean_genus(full_data_set),
+    full_data_set_clean = target(berlin.trees::clean_data(full_data_set),
                                  format = "rds"),
 
 

### 2020-01-13 analysis/scripts/03_dump.R

@@ -1,112 +1,34 @@
-get_uhi_hist_data <- function(uhi_stack_list, sf_data){
-
-
-    # brief checks
-    if (!is(sf_data, "sf")) {
-        stop("input data is not in sf")
-    }
-    # if (!is(uhi_stack_list, "RasterLayer")) {
-    #     stop("input raster is not a RasterLayer")
-    # }
-
-
-    # LOGIC
-
-
-
-    ## raster stats
-
-    stat_funs <- c("median", "mean", "sd")
-
-    ### cycle into list containing the 4 (total) stacks (2x summer, 2x winter)
-    raster_stats <- lapply(stat_funs,
-                           function(stats) {
-
-                               # into list (summer vs. winter)
-                               lapply(uhi_stack_list,
-                                      function(stacks) {
-                                          # into stacks (day vs. night)
-                                          lapply(stacks,
-                                                 raster::cellStats, stats)
-                                      })
-                           }) %>%
-        setNames(stat_funs)
-
-
-    ## tree uhi exposure
-
-    ### prepare tree data
-
-    sf_data <- sf_data %>%
-        dplyr::filter(!is.na(GATTUNG)) %>%
-        dplyr::mutate(STAMMUMFG = as.numeric(STAMMUMFG),
-                      # GATTUNG = forcats::fct_infreq(as.factor(GATTUNG)),
-
-
-                      gattung_short = forcats::fct_lump(as.factor(GATTUNG),11, other_level = "Other") %>%
-                          as.character(),
-                      gattung_short =   ifelse(GATTUNG == "Pinus", "Pinus", gattung_short),
-                      gattung_short = forcats::fct_infreq(as.factor(gattung_short)),
-                      gattung_short = forcats::fct_relevel(gattung_short, "Other", after = Inf),
-
-
-                      bezirk_num = as.numeric(as.factor(BEZIRK)))
-
-
-
-    ### re-project raster to UTM of sf data
-
-    uhi_stack_list <- lapply(uhi_stack_list,
-                          function(stack)
-                              lapply(stack, function(x)  {
-
-                                  raster::projectRaster(x,
-                                                        crs =  sf::st_crs(sf_data)$proj4string)
-
-                              }  ))
-
-
-    extract_vals <- lapply(uhi_stack_list,
-                           function(stacks){
-                               lapply(stacks,
-                                      raster::extract,
[…]

### 2020-02-14 R/functions.R
updated figures

@@ -799,7 +799,12 @@ tree_sums_bar_plot <- function(sf_data){
               axis.text.y = element_text(face = "italic")) +
 
 
-        scale_fill_brewer(type = "seq", palette = "Greens")
+    ggplot2::scale_fill_brewer(type = "qual",
+                                palette = "Set2",
+                                direction = +1,
+                                labels = c("Riparian",
+                                           "Street",
+                                           "Park"))
 
 }
 
@@ -978,8 +983,18 @@ dens_plot_trees <- function(sf_data,
 
         # scales
         xlim(c(-5, 6.5)) +
-        scale_color_brewer(type = "qual", palette = "Set2") +
-        scale_fill_brewer(type = "qual", palette = "Set2") +
+        scale_color_brewer(type = "qual",
+                           palette = "Set2",
+                           labels = c("Riparian",
+                                      "Street",
+                                      "Park"),
+                           direction = +1) +
+        scale_fill_brewer(type = "qual",
+                          palette = "Set2",
+                          labels = c("Riparian",
+                                     "Street",
+                                     "Park"),
+                          direction = +1) +
 
         # theming
         guides(color = FALSE) +
@@ -1016,7 +1031,10 @@ dens_plot_trees <- function(sf_data,
 
                                             scale_alpha_manual(values = c(0.3,
                                                                           0.95)) +
-                                            scale_fill_brewer(type = "qual", palette = "Set2") +
+                                            scale_fill_brewer(type = "qual", palette = "Set2",
+                                                              labels = c("Riparian",
+                                                                         "Street",
+                                                                         "Park")) +
 
 
                                             scale_y_continuous(breaks = c(0, 150000),
@@ -1031,7 +1049,8 @@ dens_plot_trees <- function(sf_data,
                                                   panel.grid = element_blank(),
                                                   axis.ticks.length = unit(2, "mm"),
                                                   axis.ticks = element_line(color = "gray10"),
-                                                  axis.ticks.y = element_blank())
+                                                  axis.ticks.y = element_blank(),
+                                                  strip.text = element_text(face = "italic"))
                                     ),
 
 
@@ -1105,7 +1124,10 @@ make_ranef_plot <- function(model_out,
[…]

### 2020-02-14 R/plan.R

@@ -118,22 +118,19 @@ plan <- drake_plan(
     plot_LME_age = berlin.trees::make_ranef_plot(model_out = model_res,
                     model_name = "heat_age_RIspecies_RSspecies_RIprovenance",
                     df = full_data_set_clean,
-                    n_top_species = 3)
+                    n_top_species = 3),
 
 
 
 
-
-# ,
-# ART_BOT %in% top_species$ART_BOT[top_species$n > 150]
     # Reporting
-    # paper = rmarkdown::render(
-    #     knitr_in("./analysis/paper/paper.Rmd"),
-    #     output_file = file_out(file.path(here::here(), "paper_knit.html")),
-    #     # output_file = file_out(file.path("./paper_knit.html")),
-    #     output_format = bookdown::html_document2(),
-    #     quiet = TRUE
-    # )
+    paper = rmarkdown::render(
+        knitr_in("./analysis/paper/paper.Rmd"),
+        output_file = file_out(file.path(here::here(), "paper_knit.html")),
+        # output_file = file_out(file.path("./paper_knit.html")),
+        output_format = bookdown::html_document2(),
+        quiet = TRUE
+    )
 
 )
 

### 2020-02-14 analysis/scripts/04_urban_tree_db_EDA.R

@@ -1,0 +1,30 @@
+# EDA for  Urban tree database -------------------------------------------
+
+# # for source
+# https://data.nal.usda.gov/dataset/urban-tree-database
+# # for overview
+# https://www.fs.usda.gov/rds/archive/catalog/RDS-2016-0005
+# # for meta
+# https://www.fs.usda.gov/rds/archive/products/RDS-2016-0005/_fileindex_RDS-2016-0005.html
+# # for download
+# https://www.fs.usda.gov/rds/archive/products/RDS-2016-0005/RDS-2016-0005.zip
+
+
+# download ----------------------------------------------------------------
+
+
+
+url <- "https://www.fs.usda.gov/rds/archive/products/RDS-2016-0005/RDS-2016-0005.zip"
+
+
+udb <- curl::curl_download(url,
+    destfile = tempfile(fileext = ".zip")
+)
+
+td <- tempdir()
+
+udb_unzip <- unzip(udb, exdir = td)
+
+
+udb_trees <- udb_unzip[grepl("Raw_tree_data", x = udb_unzip)]
+udb_data <- read.csv(udb_trees, header = TRUE, stringsAsFactors = FALSE)

### 2020-02-14 analysis/scripts/05_berlin_trees_GAMM_EDA.R

@@ -1,0 +1,9 @@
+
+# EDA for GAMM application ------------------------------------------------
+
+
+library(drake)
+loadd(model_df)
+
+
+tilia <- model_df

### 2020-03-05 analysis/paper/paper.Rmd
updated figures
fixing onedrive issues

--- 2020-01-09 analysis/paper/paper.Rmd
@@
 ---
-title: "Exploratory overview of Berlin city trees"
+title: "Species-specific effects of the Urban Heat Island on tree growth across Berlin"
 author:
@@
       fig_caption: yes
+      toc: yes
+    bookdown::pdf_document2:
+      fig_caption: yes
+      toc: yes
 
-bibliography: references.bib
+bibliography: uhi_references.bib
 csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
 abstract: |
-  Quick overview of publicly available tree records.
+  **This document serves as a brief overview and outline** .
 keywords: |
-  Uferbaeume; Anlagenbaeume; Strassenhighlights
+  Uferbaeume; Anlagenbaeume; Strassenbaeume; 
 highlights: |
@@
 library(ggplot2)
+library(patchwork)
 ```
@@
 
-```{r fig-tree-overview-map, fig.cap="A plot of random numbers",fig.width = 15, fig.height = 8, out.width="100%"}
-readd(plot_overview_map)
+```{r fig-tree-overview-map, fig.cap="A plot of random numbers",fig.width = 15, fig.height = 8, out.width="100%", include=TRUE, eval=TRUE}
+
+# loadd(full_data_set_clean)
+# loadd(plot_overview_map)
+# print(class(full_data_set_clean))
+
+# knitr::include_graphics(drake::file_in("./analysis/figures/map_01_overview.png"))
+
+loadd(plot_overview_map)
+knitr::include_graphics(file.path(here::here(), "analysis", "figures", "map_01_overview.png"))
+# knitr::include_graphics("../figures/map_01_overview.png")
+# print(dir(path = "./analysis/figures"  ))
+# readd(plot_overview_map) + 
+#      +
+#     labs(caption = NULL)
+
 # plot_overview_map
@@
 
-```{r fig-tree-count-map, fig.cap="A plot of random numbers",fig.width = 12, fig.height = 12, out.width="100%"}
-readd(plot_count_map)
+```{r fig-uhi, fig.cap="A plot of random numbers",fig.width = 15, fig.height = 8, out.width="100%", include=TRUE, eval=TRUE}
+
+loadd(plot_uhi_map)
+knitr::include_graphics(file.path(here::here(),
+                                                 "analysis", 
+                                                 "figures", 
+                                                 "map_03_uhi.png"))
+# knitr::include_graphics(drake::file_in(file.path(here::here(),
+#                                                  "analysis", 
+#                                                  "figures", 
+#                                                  "map_03_uhi.png")))
+
+```
+
+```{r fig-tree-count-map, fig.cap="A plot of random numbers",fig.width = 12, fig.height = 14, out.width="100%"}
+
+loadd(plot_count_map)
+knitr::include_graphics(file.path(here::here(), "analysis", "figures", "map_02_tree_sums_standardized.png"))
+
+# readd(plot_count_map) + 
+#   labs(caption = NULL) +
+#   theme(legend.position = c(0.65, 0.1),
+#         legend.key.width = unit(3, units = "cm")) +
+#   guides(color = guide_colorbar(title.position = "top"))
+
 # plot_count_map
+
+# plot(1:10)
+```
+
+```{r fig-count-bars, fig.cap="A plot of random numbers",fig.width = 6, fig.height = 6, out.width="50%", include = FALSE, eval = FALSE}
+loadd(plot_tree_sums_bar)
+knitr::include_graphics(file.path(here::here(), "analysis", "figures", "plot_01_tree_sums_bar.png"))
+
+# plot(1:10)
+```
+
+```{r fig-density, fig.cap="A plot of random numbers",fig.width = 12, fig.height = 12, out.width="100%"}
+loadd(plot_density)
+knitr::include_graphics(file.path(here::here(), "analysis", "figures", "plot_02_genus_UHI_dens.png"))
+
+# plot(1:10)
+```
+
+```{r fig-lmestat, fig.cap="A plot of random numbers",fig.width = 20, fig.height = 12, out.width="100%"}
+loadd(plot_LME_age)
+knitr::include_graphics(file.path(here::here(), "analysis", "figures", "plot_03_ranef_species_dbh_uhi.png"))
 

### 2020-03-05 R/plan.R

@@ -44,7 +44,7 @@ plan <- drake_plan(
 
 
     extract_uhi_values_to_list = berlin.trees::add_uhi_hist_data(uhi_stack_list = uhi_stacks,
-                                                      sf_data = full_data_set_clean[, ]),
+                                                                 sf_data = full_data_set_clean[, ]),
 
 
 
@@ -84,41 +84,79 @@ plan <- drake_plan(
 
     model_res = berlin.trees::apply_models(df = model_df %>%
                                                dplyr::filter(STANDALTER < 350 &
-                                                          krone_m < 50 &
-                                                          dbh_cm < 600),
-                             model_list = model_list,
-                             n_top_species = 3,
-                             min_individuals = 150),
+                                                                 krone_m < 50 &
+                                                                 dbh_cm < 600),
+                                           model_list = model_list,
+                                           n_top_species = 3,
+                                           min_individuals = 150),
 
 
 
 
     # Plotting
     plot_overview_map = berlin.trees::make_overview_map(full_data_set_clean,
-                                                        berlin_polygons),
+                                                        berlin_polygons,
+                                                        file = drake::file_out("./analysis/figures/map_01_overview.png"),
+                                                        height = 4,
+                                                        width = 12,
+                                                        dpi = 300),
 
-    plot_tree_sums_bar = berlin.trees::tree_sums_bar_plot(full_data_set_clean),
 
     plot_count_map = berlin.trees::tree_count_map(full_data_set_clean,
-                                                  berlin_polygons),
+                                                  berlin_polygons,file = drake::file_out("./analysis/figures/map_02_tree_sums_standardized.png"),
+                                                  height = 12,
+                                                  width = 12,
+                                                  dpi = 300),
 
+    plot_uhi_map = berlin.trees::make_uhi_plot(uhi_stacks = uhi_stacks,
+                                               berlin_poly = berlin_polygons,
+                                               base_size = 18,
+                                               file = drake::file_out("./analysis/figures/map_03_uhi.png"),
+                                               height = 6,
+                                               width = 6,
+                                               dpi = 300),
+
+
+
+
+    plot_tree_sums_bar = berlin.trees::tree_sums_bar_plot(full_data_set_clean,
+                                                          file = drake::file_out("./analysis/figures/plot_01_tree_sums_bar.png"),
+                                                          base_size = 18,
+                                                          height = 12,
[…]

### 2020-03-05 _setup.R

@@ -1,22 +1,26 @@
-# usethis::use_package("ckanr")
-# usethis::use_package("ggplot2")
-# usethis::use_package("dplyr")
-# usethis::use_package("xml2")
-# usethis::use_package("httr")
-# usethis::use_package("purrr")
-# usethis::use_package("stringr")
-# usethis::use_package("sf")
-# usethis::use_package("future.callr")
-# usethis::use_package("drake")
-# usethis::use_package("furrr")
-# usethis::use_package("tools")
-# usethis::use_package("forcats")
-# usethis::use_package("tibble")
-# usethis::use_package("rmarkdown")
-# usethis::use_package("raster")
-# usethis::use_package("magrittr")
-# usethis::use_package("lme4")
-# usethis::use_package("rlang")
+usethis::use_package("ckanr")
+usethis::use_package("ggplot2")
+usethis::use_package("dplyr")
+usethis::use_package("xml2")
+usethis::use_package("httr")
+usethis::use_package("purrr")
+usethis::use_package("stringr")
+usethis::use_package("sf")
+usethis::use_package("future.callr")
+usethis::use_package("drake")
+usethis::use_package("furrr")
+usethis::use_package("tools")
+usethis::use_package("forcats")
+usethis::use_package("tibble")
+usethis::use_package("rmarkdown")
+usethis::use_package("raster")
+usethis::use_package("magrittr")
+usethis::use_package("lme4")
+usethis::use_package("rlang")
+usethis::use_package("stars")
+usethis::use_package("extrafont")
+usethis::use_package("scales")
+usethis::use_package("raster")
 
 ## packages
 
@@ -36,4 +40,4 @@ drake::r_make()
 
 
 source("R/plan.R")      # Create your drake plan.
-drake::vis_drake_graph(config = drake_config(plan))
+drake::vis_drake_graph(plan, targets_only = TRUE)

### 2020-03-06 analysis/paper/paper.Rmd
updated work

--- 2020-01-09 analysis/paper/paper.Rmd
@@
 ---
-title: "Exploratory overview of Berlin city trees"
+title: "Species-specific effects of the Urban Heat Island on tree growth across Berlin"
 author:
-  - Alexander Hurley:
-      email: hurley@gfz-potsdam.de
-      institute: [GFZ]
-      correspondence: true
-  - Ingo Heinrich:
-      email: heinrich@gfz-potsdam.de
-      institute: [GFZ]
-      correspondence: false
-institute:
-  - GFZ: Geoforschungszentrum Potsdam
+  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^a$
+  - $^a$ *Joint Geoforschungszentrum Potsdam, Section 4.3, Germany*
+  - "* Corresponding author: hurley@gfz-potsdam.de"
 date: "`r format(Sys.time(), '%d %B, %Y')`"
 output: 
-    bookdown::html_document2:
+    bookdown::word_document2:
       fig_caption: yes
-
-bibliography: references.bib
+      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      pandoc_args:
+      - --lua-filter=../templates/scholarly-metadata.lua
+      - --lua-filter=../templates/author-info-blocks.lua
+      - --lua-filter=../templates/pagebreak.lua
+bibliography: uhi_references.bib
 csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
 abstract: |
-  Quick overview of publicly available tree records.
+  **This document serves as a brief overview and outline** .
 keywords: |
-  Uferbaeume; Anlagenbaeume; Strassenhighlights
+  Uferbaeume; Anlagenbaeume; Strassenbaeume; 
 highlights: |
@@
 
-<!--     bookdown::word_document2: -->
+<!--     bookdown::html_document2: -->
 <!--       fig_caption: yes -->
-<!--       reference_docx: "../templates/template.docx" # Insert path for the DOCX file -->
-<!--       pandoc_args: -->
-<!--       - --lua-filter=../templates/scholarly-metadata.lua -->
-<!--       - --lua-filter=../templates/author-info-blocks.lua -->
-<!--       - --lua-filter=../templates/pagebreak.lua -->
+<!--       toc: yes -->
+<!--     bookdown::pdf_document2: -->
+<!--       fig_caption: yes -->
+<!--       toc: yes -->
 
@@
 ```{r setup, echo = FALSE, cache = FALSE, include = FALSE}
-knitr::opts_knit$set(root.dir = here::here())
+knitr::opts_knit$set(root.dir = here::here(),
+                     base.dir = here::here())
 
@@
 
+pgbreak <- function() {
+  if (grepl("docx", out_format)) {
+    return("##### pagebreak")
+  } else if (is_latex_output()) {
+    return("\\newpage")
+  }
+}
+
 library(drake)
-library(ggplot2)
+# library(ggplot2)
+# library(patchwork)
 ```
@@
 
-Here is a citation [@Marwick2017]
+Berlin is subject to the most intense Urban Heat Island in Germany due to its large extent and development intensity.
+Here, substantially higher temperatures have been observed for urban as compared to rural areas, with differences of up to 12 K during day-time and 6 K on average for night-time temperatures (2001-2010).
+
+Consequently, urban green (infrastructure) systems are  subjected to increased heat more frequently than rural ecosystems, potentially affecting  their process dynamics positively or adversely. 
+Their performance and health, however, is closely tied to local energy budgets [@grimmond1996 ; @hertel2019], which in turn are decisive for controlling human wellbeing [e.g.
+@maras2016a], amongst other factors.
+Assessing the effect of increased temperatures on green infrastructure, as part of the urban landscape, is therefore instrumental for understanding, and ultimately mitigating, the potential impact of future warming and (continued) development  on increasingly urban societies [@norton2015].
+
+Trees, in particular, provide shading as well as transpirative cooling in their vicinity [e.g.
+@oke1982], and therefore can reduce ambient temperatures, infrastructure power-consumption and (human) thermal discomfort  [e.g.
+@gulyas2006; @akbari2001; @mayer1987; @hoyano1988]; simultaneously, they provide numerous other environmental, cultural and psychological services and/or benefits [see @tzoulas2007 for review].
+
+Tree growth dynamics in urban areas are modulated by air temperature, amongst other factors, with a tendency for enhanced growth rates (compared to rural individuals) as reported by @pretzsch2017. 
+Hence,  temperature increases may also allow using recent tree growth dynamics and trajectories as a proxy for on-going and future warming, providing an additional line of evidence to support the growing knowledge base on future climate-vegetation dynamics.
+For Berlin, @dahlhausen2018, showed divergent patterns for young and old trees dependent on urban development intensity for *Tilia cordata* Mill, the most abundant tree of the city, but overall reported an increased growth rate which they attributed to the urban heat island effect (UHI).
+Similarly, @moser-reischl2019a showed positive associations between air temperature and radial growth for two species commonly selected by urban planners (*T.
+cordata*, *Rubinia pseudoacacia*) in Munich.
+By contrast, @gillner2014 highlight decreased growth for *Acer* species  (*A.
+platanoides* and *pseudoplatanus*), *Platanus x hispanica* and *Quercus rubra* with higher summer temperatures of the preceding year, especially when compounded with drought, in another German metropolis (Dresden).
+
+These results indicate a high degree of complexity in effect sizes and directions specific to species and locality.
+Tree growth in urban settings is affected by myriad processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc.
+[@pauleit2002].
+Therefore, studies reliant on labor-intensive sampling (e.g.
+based on increment cores; dendroecological methods), must carefully address confounding factors to ensure effect sizes are estimated adequately.
+While multiple (locally) important species have been assessed along urban and UHI gradients, studies are  limited logistically by sampling effort, reducing sample size and coverage (i.e.
+genera and/or species, and space), which typically hinders highly-resolved predictions of effect sizes across entire urban areas and tree stocks.
+To complement detailed dendroecological analyses in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individual trees provided by the Berlin Senate Administration (Senatsverwaltung).
+This data set contains information on location, species, trunk diameter (at breast height; DBH), and height, amongst others. 
+In a space-for-time substitution, the absolute growth potential of species can assessed across the entire urban area, and related to effects of the UHI, as well as other location-specific factors, such as road type, development intensity, available soil volume, etc.
 
@@
 
-```{r fig-tree-overview-map, fig.cap="A plot of random numbers",fig.width = 15, fig.height = 8, out.width="100%"}
-readd(plot_overview_map)
+<!-- ```{r fig-bar-tree, fig.cap="A plot of random numbers",fig.width = 15, fig.height = 8, out.width="100%", include=FALSE, eval=FALSE} -->
+
+<!-- readd(plot_tree_sums_bar) -->
+
+<!-- ``` -->
+
+```{r fig-tree-overview-map, fig.cap="A plot of random numbers",fig.width = 15, fig.height = 8, out.width="100%", include=TRUE, eval=TRUE}
+
+# loadd(full_data_set_clean)
+# loadd(plot_overview_map)
+# print(class(full_data_set_clean))
+
+knitr::include_graphics(here::here("analysis/figures/map_01_overview.png"))
+
+# readd(plot_overview_map) + 
+#      +
+#     labs(caption = NULL)
+
 # plot_overview_map
@@
 
-```{r fig-tree-count-map, fig.cap="A plot of random numbers",fig.width = 12, fig.height = 12, out.width="100%"}
-readd(plot_count_map)
+```{r fig-tree-count-map, fig.cap="A plot of random numbers",fig.width = 12, fig.height = 14, out.width="100%"}
+
+knitr::include_graphics(here::here("analysis/figures/map_02_tree_sums_standardized.png"))
+
 # plot_count_map
@@
 
-<!-- Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot -->
+```{r fig-uhi-map, fig.cap="A plot of random numbers",fig.width = 12, fig.height = 14, out.width="50%"}
+
+knitr::include_graphics(here::here("analysis/figures/map_03_uhi.png"))
+
+# plot_count_map
+
+# plot(1:10)
+```
+
+```{r fig-density, fig.cap="A plot of random numbers",fig.width = 12, fig.height = 12, out.width="100%"}
+knitr::include_graphics(here::here("analysis/figures/plot_02_genus_UHI_dens.png"))
+
+# plot_count_map
+
+# plot(1:10)
+```
+
+```{r fig-lmestat, fig.cap="A plot of random numbers",fig.width = 20, fig.height = 12, out.width="100%"}
+knitr::include_graphics(here::here("analysis/figures/plot_03_ranef_species_dbh_uhi.png"))
+# plot_count_map
+
+# plot(1:10)
+```
+
+Figure \@ref(fig:lmestat) shows how we can have a caption and cross-reference for a plot
 

### 2020-03-06 R/functions.R

@@ -935,15 +935,16 @@ tree_count_map <- function(sf_data,
             scale_x_continuous(breaks = seq(13, 14, 0.2)) +
             scale_y_continuous(breaks = seq(52.3, 53.7, 0.1)) +
             scale_fill_viridis_c() +
-            guides(fill = guide_colorbar(barwidth = grid::unit(10, units = "cm"),
-                                         barheight = grid::unit(.45, units = "cm"))) +
+            guides(fill = guide_colorbar(barwidth = grid::unit(6, units = "cm"),
+                                         barheight = grid::unit(.45, units = "cm"),
+                                         ticks.colour = "gray80")) +
 
 
             theme_minimal(base_size = base_size, base_family = "Roboto Condensed") +
 
             theme(axis.title = element_blank(),
                   strip.text = element_text(face = c("bold.italic")),
-                  legend.position = "bottom",
+                  legend.position = c(0.6, 0.11),
                   legend.direction = "horizontal",
                   legend.title = element_text(vjust = 1))
 
@@ -1021,7 +1022,8 @@ make_uhi_plot <- function(uhi_stacks,
                                       rescaler = mid_rescaler(),
                                       na.value = "transparent")   +
 
-        ggplot2::labs(fill = expression(Summer~day-time~UHI~(degree*C)),
+        ggplot2::labs(fill = expression(atop(Summer~day-time,
+                                             UHI~(degree*C))),
                       x = NULL,
                       # title = "Estimate of Urban Heat Loading",
                       y = NULL) +
@@ -1030,9 +1032,10 @@ make_uhi_plot <- function(uhi_stacks,
         ggplot2::theme_minimal(base_family = "Roboto Condensed",
                                base_size = base_size) +
         ggplot2::theme(legend.direction = "horizontal",
-                       legend.position = c(0.8, 0.95)) +
-        ggplot2::guides(fill = guide_colorbar(barwidth = unit(5, "cm"),
-                                              title.vjust = 1))
+                       legend.position = c(0.6, 0.95)) +
+        ggplot2::guides(fill = guide_colorbar(barwidth = unit(3.5, "cm"),
+                                              title.vjust = 1,
+                                              ticks.colour = "gray30"))
 
 
     ggplot2::ggsave(filename = file,

### 2020-03-06 R/plan.R

@@ -57,20 +57,20 @@ plan <- drake_plan(
         # heat_RIspecies_provenance = function(x) lme4::lmer(dbh_cm ~ day_2007 + (1 | ART_BOT) + provenance , data = x) ,
         # heat_RIspecies_RSspecies = function(x) lme4::lmer(dbh_cm ~ day_2007 + (1 + day_2007 | ART_BOT) , data = x)  ,
         # heat_RIspecies_RSspecies_provenance = function(x) lme4::lmer(dbh_cm ~ day_2007 + (1 + day_2007 | ART_BOT) + provenance , data = x)  ,
-        heat_RIspecies_RSspecies_RIprovenance = function(x) lme4::lmer(dbh_cm ~ day_2007 + provenance + (1 + day_2007 | provenance : ART_BOT)  , data = x)  ,
+        # heat_RIspecies_RSspecies_RIprovenance = function(x) lme4::lmer(dbh_cm ~ day_2007 + provenance + (1 + day_2007 | provenance : ART_BOT)  , data = x)  ,
         # heat_age = function(x) lme4::lmer(dbh_cm ~ day_2007 + STANDALTER, data = x)  ,
         # heat_age_species = function(x) lme4::lmer(dbh_cm ~ day_2007 + STANDALTER + ART_BOT, data = x)  ,
         # heat_age_RIspecies = function(x) lme4::lmer(dbh_cm ~ day_2007 + STANDALTER + (1 | ART_BOT), data = x)  ,
         # heat_age_RIspecies_provenance = function(x) lme4::lmer(dbh_cm ~ day_2007 + STANDALTER + (1 | ART_BOT) + provenance, data = x)  ,
         # heat_age_RIspecies_RSspecies = function(x) lme4::lmer(dbh_cm ~ day_2007 + STANDALTER + (1 + day_2007 | ART_BOT), data = x)  ,
-        heat_age_RIspecies_RSspecies_provenance = function(x) lme4::lmer(dbh_cm ~ day_2007 + STANDALTER + provenance + (1 + day_2007 | ART_BOT), data = x)  ,
-        heat_age_RIspecies_RSspecies_RIprovenance = function(x) lme4::lmer(dbh_cm ~ day_2007 + STANDALTER + provenance + (1 + day_2007 | provenance : ART_BOT), data = x, control = lme4::lmerControl(optimizer ="Nelder_Mead"))  ,
[…]

### 2020-03-06 _setup.R

@@ -1,4 +1,5 @@
 
+usethis::use_package("rgdal")
 # usethis::use_package("ckanr")
 # usethis::use_package("ggplot2")
 # usethis::use_package("dplyr")
@@ -42,3 +43,4 @@ drake::r_make()
 
 source("R/plan.R")      # Create your drake plan.
 drake::vis_drake_graph(plan, targets_only = TRUE)
+drake::sankey_drake_graph(plan, targets_only = TRUE)

### 2020-03-06 analysis/paper/paper.Rmd
need to update bib

@@
       - --lua-filter=../templates/pagebreak.lua
+    bookdown::html_document2:
+      fig_caption: yes
+      toc: yes
+    bookdown::pdf_document2:
+      fig_caption: yes
+      toc: yes
 bibliography: uhi_references.bib
@@
 ---
-
-<!--     bookdown::html_document2: -->
-<!--       fig_caption: yes -->
-<!--       toc: yes -->
-<!--     bookdown::pdf_document2: -->
-<!--       fig_caption: yes -->
-<!--       toc: yes -->
 
@@
 
-Berlin is subject to the most intense Urban Heat Island in Germany due to its large extent and development intensity.
-Here, substantially higher temperatures have been observed for urban as compared to rural areas, with differences of up to 12 K during day-time and 6 K on average for night-time temperatures (2001-2010).
+Berlin is subject to the most intense Urban Heat Island in Germany due to its large extent and development intensity [@kuttler2015].
+Here, substantially higher temperatures have been observed for urban as compared to rural areas, with differences of up to 12 K during day-time and 6 K on average for night-time temperatures [2001-2010, @fenner2014].
 
@@
 Tree growth in urban settings is affected by myriad processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc.
-[@pauleit2002].
-Therefore, studies reliant on labor-intensive sampling (e.g.
-based on increment cores; dendroecological methods), must carefully address confounding factors to ensure effect sizes are estimated adequately.
-While multiple (locally) important species have been assessed along urban and UHI gradients, studies are  limited logistically by sampling effort, reducing sample size and coverage (i.e.
-genera and/or species, and space), which typically hinders highly-resolved predictions of effect sizes across entire urban areas and tree stocks.
+[@quigley2004; @randrup2001; @pauleit2002].
+While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e.
+genera and/or species, and space).
+This typically hinders the extrapolation from individual sampling sites toward prediction of effect sizes across entire urban areas and tree stocks. 
+<!-- Therefore, studies reliant on labor-intensive sampling, must carefully address confounding factors to ensure effect sizes are estimated adequately. -->
+
 To complement detailed dendroecological analyses in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individual trees provided by the Berlin Senate Administration (Senatsverwaltung).

### 2020-03-07 analysis/paper/paper.Rmd
start of day

@@
   - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^a$
-  - $^a$ *Joint Geoforschungszentrum Potsdam, Section 4.3, Germany*
+  - $^a$ *Geoforschungszentrum Potsdam, Section 4.3, Germany*
   - "* Corresponding author: hurley@gfz-potsdam.de"
@@
 Their performance and health, however, is closely tied to local energy budgets [@grimmond1996 ; @hertel2019], which in turn are decisive for controlling human wellbeing [e.g.
-@maras2016a], amongst other factors.
-Assessing the effect of increased temperatures on green infrastructure, as part of the urban landscape, is therefore instrumental for understanding, and ultimately mitigating, the potential impact of future warming and (continued) development  on increasingly urban societies [@norton2015].
+@maras2016], amongst other factors.
+Assessing the effect of increased temperatures on green infrastructure, as part of the urban landscape, is therefore instrumental for understanding, and ultimately mitigating, the potential impact of future warming and (continued) development  on increasingly urban societies [@norton2015 ].
 
-Trees, in particular, provide shading as well as transpirative cooling in their vicinity [e.g.
-@oke1982], and therefore can reduce ambient temperatures, infrastructure power-consumption and (human) thermal discomfort  [e.g.
+Trees, in particular, provide shading as well as transpirative cooling in their vicinity [@gillner2015;@oke1982], and therefore can reduce ambient temperatures, infrastructure power-consumption and (human) thermal discomfort  [e.g.
 @gulyas2006; @akbari2001; @mayer1987; @hoyano1988]; simultaneously, they provide numerous other environmental, cultural and psychological services and/or benefits [see @tzoulas2007 for review].
 
-Tree growth dynamics in urban areas are modulated by air temperature, amongst other factors, with a tendency for enhanced growth rates (compared to rural individuals) as reported by @pretzsch2017. 
-Hence,  temperature increases may also allow using recent tree growth dynamics and trajectories as a proxy for on-going and future warming, providing an additional line of evidence to support the growing knowledge base on future climate-vegetation dynamics.
+Tree growth dynamics in urban areas are modulated by air temperature, amongst other factors, with a tendency for enhanced growth rates (compared to rural individuals) as reported by @zhao2016 and @pretzsch2017.
+Hence,  temperature increases may also allow using recent tree growth dynamics and trajectories as a proxy for on-going and future warming, providing an additional line of evidence to support the growing knowledge base on future climate- and human-vegetation and dynamics [@zhao2016].
 For Berlin, @dahlhausen2018, showed divergent patterns for young and old trees dependent on urban development intensity for *Tilia cordata* Mill, the most abundant tree of the city, but overall reported an increased growth rate which they attributed to the urban heat island effect (UHI).
@@
 platanoides* and *pseudoplatanus*), *Platanus x hispanica* and *Quercus rubra* with higher summer temperatures of the preceding year, especially when compounded with drought, in another German metropolis (Dresden).
+**In the future.. drought, both soil moisture and atmospheric drought.. potential issue for street trees .. [@brune2016]
+??**
+[@zhao2016] and [@jia2018] direction.....
 
 These results indicate a high degree of complexity in effect sizes and directions specific to species and locality.
-Tree growth in urban settings is affected by myriad processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc.
-[@quigley2004; @randrup2001; @pauleit2002].
+Tree growth in urban settings is affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc.
+[@rhoades1999; @quigley2004; @randrup2001; @pauleit2002].
 While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e.
@@
 To complement detailed dendroecological analyses in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individual trees provided by the Berlin Senate Administration (Senatsverwaltung).
-This data set contains information on location, species, trunk diameter (at breast height; DBH), and height, amongst others. 
-In a space-for-time substitution, the absolute growth potential of species can assessed across the entire urban area, and related to effects of the UHI, as well as other location-specific factors, such as road type, development intensity, available soil volume, etc.
+This data set contains information on location, species, trunk diameter (at breast height; DBH), and height, amongst other variables. 
+In a space-for-time substitution, the absolute growth potential of species can be assessed across the entire urban area, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.
+[@quigley2004] .. inferred absolute growth potential for species across demographic strata
 

### 2020-03-07 analysis/paper/paper.Rmd
updated intro

@@
 
+out_format <- knitr::opts_knit$get("rmarkdown.pandoc.to")
+
 pgbreak <- function() {
@@
     return("##### pagebreak")
-  } else if (is_latex_output()) {
+  } else if (knitr::is_latex_output()) {
     return("\\newpage")
@@
 
-Berlin is subject to the most intense Urban Heat Island in Germany due to its large extent and development intensity [@kuttler2015].
-Here, substantially higher temperatures have been observed for urban as compared to rural areas, with differences of up to 12 K during day-time and 6 K on average for night-time temperatures [2001-2010, @fenner2014].
-
-Consequently, urban green (infrastructure) systems are  subjected to increased heat more frequently than rural ecosystems, potentially affecting  their process dynamics positively or adversely. 
+Berlin features the most intense Urban Heat Island (UHI) in Germany due to its large extent and development intensity [@kuttler2015], with temperature increases of up to $12~K$ during day-time and $6~K$ on average for night-times [2001-2010, @fenner2014] in urban $vs.$ rural areas.
+Consequently, urban green (infrastructure) systems are subjected to increased heat more frequently, potentially affecting their process dynamics - either positively or adversely. 
 Their performance and health, however, is closely tied to local energy budgets [@grimmond1996 ; @hertel2019], which in turn are decisive for controlling human wellbeing [e.g.
 @maras2016], amongst other factors.
-Assessing the effect of increased temperatures on green infrastructure, as part of the urban landscape, is therefore instrumental for understanding, and ultimately mitigating, the potential impact of future warming and (continued) development  on increasingly urban societies [@norton2015 ].
+Assessing the effect of increased temperatures on green infrastructure, as part of the urban landscape, is therefore instrumental for understanding, and ultimately mitigating, the potential impact of future warming on increasingly urban societies [@norton2015].
 
-Trees, in particular, provide shading as well as transpirative cooling in their vicinity [@gillner2015;@oke1982], and therefore can reduce ambient temperatures, infrastructure power-consumption and (human) thermal discomfort  [e.g.
-@gulyas2006; @akbari2001; @mayer1987; @hoyano1988]; simultaneously, they provide numerous other environmental, cultural and psychological services and/or benefits [see @tzoulas2007 for review].
+Trees, in particular, provide shading as well as transpirative cooling in their vicinity [@gillner2015;@oke1982;@endlicher2016], and therefore can reduce ambient temperatures, infrastructure power-consumption and (human) thermal discomfort  [e.g.
+@gulyas2006; @akbari2001; @mayer1987; @hoyano1988];
+simultaneously, they provide numerous other environmental, cultural and psychological services and/or benefits [see @tzoulas2007 for review].
+Further, recent tree growth dynamics as a proxy for on-going and future warming may provide an additional line of evidence to support the growing knowledge base on future climate-vegetation and dynamics [@zhao2016] and may aid in mitigation and adaptation efforts [@pretzsch2017].
 
-Tree growth dynamics in urban areas are modulated by air temperature, amongst other factors, with a tendency for enhanced growth rates (compared to rural individuals) as reported by @zhao2016 and @pretzsch2017.
-Hence,  temperature increases may also allow using recent tree growth dynamics and trajectories as a proxy for on-going and future warming, providing an additional line of evidence to support the growing knowledge base on future climate- and human-vegetation and dynamics [@zhao2016].
-For Berlin, @dahlhausen2018, showed divergent patterns for young and old trees dependent on urban development intensity for *Tilia cordata* Mill, the most abundant tree of the city, but overall reported an increased growth rate which they attributed to the urban heat island effect (UHI).
+Trees and green infrastructure in urban areas show a tendency for enhanced growth rates and/or productivity compared to rural counterparts [@jia2018; @pretzsch2017], yet feature a broad range of effect size ranges and, in some cases, directions specific to species and locality.
+ @zhao2016 .. remote sensing.. increased growth rates with higher degree of urbanity for X Chinese cities.
+For Berlin, @dahlhausen2018, showed divergent patterns for intermiedate and highly urbanized environments for *Tilia cordata* Mill (using growth increments), the most abundant tree of the city, with increased growth rates in the more intensely developed areas and attributed this to the UHI effect.
+<!-- For Berlin, @dahlhausen2018, showed divergent patterns for young and old trees dependent on urban development intensity for *Tilia cordata* Mill, the most abundant tree of the city, but overall reported an increased growth rate which they attributed to the UHI effect.
+-->
 Similarly, @moser-reischl2019a showed positive associations between air temperature and radial growth for two species commonly selected by urban planners (*T.
@@
 platanoides* and *pseudoplatanus*), *Platanus x hispanica* and *Quercus rubra* with higher summer temperatures of the preceding year, especially when compounded with drought, in another German metropolis (Dresden).
+Differences in growth trends may result from contrasting intrinsic inter- and intra-specific characteristics, but are indeed affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc.
+[@rhoades1999; @quigley2004; @randrup2001; @pauleit2002].
 **In the future.. drought, both soil moisture and atmospheric drought.. potential issue for street trees .. [@brune2016]
 ??**
-[@zhao2016] and [@jia2018] direction.....
 
-These results indicate a high degree of complexity in effect sizes and directions specific to species and locality.
-Tree growth in urban settings is affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc.
+Conditions affecting tree growth can vary greatly within urban areas or regions, and need to be accounted for when establishing relationships with pertinent drivers, such as the UHI effect.
+This typically complicates the extrapolation from individual sampling sites toward predicting effect sizes across entire urban areas and tree stocks, especially for studies reliant on labour-intensive methods limited logistically by sampling effort (reduced sample size, species and spatial coverage).
+<!-- While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e.
+genera and/or species, and space).
+-->
+<!-- However, urban tree growth in urban settings is affected by various processes, which  -->
+<!-- However, growth in urban settings is affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc.
 [@rhoades1999; @quigley2004; @randrup2001; @pauleit2002].
-While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e.
+-->
+<!-- While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e.
 genera and/or species, and space).
-This typically hinders the extrapolation from individual sampling sites toward prediction of effect sizes across entire urban areas and tree stocks. 
+-->
+<!-- This typically hinders the extrapolation from individual sampling sites toward prediction of effect sizes across entire urban areas and tree stocks.  -->
 <!-- Therefore, studies reliant on labor-intensive sampling, must carefully address confounding factors to ensure effect sizes are estimated adequately. -->
+To complement detailed dendroecological analyses in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individual trees provided by the Berlin Senate Administration (Senatsverwaltung).
+This data set contains information on location, species, trunk diameter (at breast height; $DBH$), and height, amongst other variables. 
+In a space-for-time substitution, the absolute growth potential of species can be assessed across the entire urban area, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.
 
-To complement detailed dendroecological analyses in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individual trees provided by the Berlin Senate Administration (Senatsverwaltung).
-This data set contains information on location, species, trunk diameter (at breast height; DBH), and height, amongst other variables. 
-In a space-for-time substitution, the absolute growth potential of species can be assessed across the entire urban area, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.
 [@quigley2004] .. inferred absolute growth potential for species across demographic strata
+
+Note [@gregg2003].. rather impeded growth adjacent to intense urban clusters; but, here focused on intra-city patterns.
 
@@
 
-# Methods
+# Proposed methods and data requirements
 
-# Results
+# Preliminary results
 
@@
 
-# Discussion
-
-# Conclusion
+# Outlook
 
@@
 <!-- The following line inserts a page break  -->
-\newpage
+`r pgbreak()`
 
@@
 
-\newpage
+`r pgbreak()`
 

### 2020-03-07 analysis/paper/paper.Rmd
cont writing

@@
 simultaneously, they provide numerous other environmental, cultural and psychological services and/or benefits [see @tzoulas2007 for review].
-Further, recent tree growth dynamics as a proxy for on-going and future warming may provide an additional line of evidence to support the growing knowledge base on future climate-vegetation and dynamics [@zhao2016] and may aid in mitigation and adaptation efforts [@pretzsch2017].
+Further, recent tree growth dynamics as a proxy for on-going and future warming may provide an additional line of evidence to support the growing knowledge base on future climate-vegetation dynamics [@zhao2016] and may aid in mitigation and adaptation efforts [@pretzsch2017;@brune2016].
 
 Trees and green infrastructure in urban areas show a tendency for enhanced growth rates and/or productivity compared to rural counterparts [@jia2018; @pretzsch2017], yet feature a broad range of effect size ranges and, in some cases, directions specific to species and locality.
- @zhao2016 .. remote sensing.. increased growth rates with higher degree of urbanity for X Chinese cities.
-For Berlin, @dahlhausen2018, showed divergent patterns for intermiedate and highly urbanized environments for *Tilia cordata* Mill (using growth increments), the most abundant tree of the city, with increased growth rates in the more intensely developed areas and attributed this to the UHI effect.
+@zhao2016 showed that growth rates increased within urban clusters as urbanization intensifies using remotely sensed vegetation indices.
+Similarly, for Berlin, @dahlhausen2018, identified positive growth modulation in highly urbanized environments for *Tilia cordata* Mill (using growth increments), the most abundant tree of the city, which they attributed to the UHI effect, while intermiedate development intensity showed indications of being least favorable for tree growth.
 <!-- For Berlin, @dahlhausen2018, showed divergent patterns for young and old trees dependent on urban development intensity for *Tilia cordata* Mill, the most abundant tree of the city, but overall reported an increased growth rate which they attributed to the UHI effect.
 -->
-Similarly, @moser-reischl2019a showed positive associations between air temperature and radial growth for two species commonly selected by urban planners (*T.
+Further, @moser-reischl2019a identified positive associations between air temperature and radial growth for two species commonly selected by urban planners (*T.
 cordata*, *Rubinia pseudoacacia*) in Munich.
@@
 [@rhoades1999; @quigley2004; @randrup2001; @pauleit2002].
-**In the future.. drought, both soil moisture and atmospheric drought.. potential issue for street trees .. [@brune2016]
-??**
+Under climate change, atmospheric drought will likely be compounded with high temperatures more frequently, adding further stress to current urban disturbance regimes [@roloff2009].
 
 Conditions affecting tree growth can vary greatly within urban areas or regions, and need to be accounted for when establishing relationships with pertinent drivers, such as the UHI effect.
-This typically complicates the extrapolation from individual sampling sites toward predicting effect sizes across entire urban areas and tree stocks, especially for studies reliant on labour-intensive methods limited logistically by sampling effort (reduced sample size, species and spatial coverage).
+This typically complicates the extrapolation from individual sampling sites toward predicting effect sizes across entire urban areas and tree stocks.
+This is especially the case for studies reliant on labour-intensive methods which are limited logistically by sampling effort, reducing sample sizes, as well as species and spatial coverage.
 <!-- While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e.
@@
 <!-- Therefore, studies reliant on labor-intensive sampling, must carefully address confounding factors to ensure effect sizes are estimated adequately. -->
-To complement detailed dendroecological analyses in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individual trees provided by the Berlin Senate Administration (Senatsverwaltung).
-This data set contains information on location, species, trunk diameter (at breast height; $DBH$), and height, amongst other variables. 
-In a space-for-time substitution, the absolute growth potential of species can be assessed across the entire urban area, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.
 
-[@quigley2004] .. inferred absolute growth potential for species across demographic strata
+To complement detailed dendroecological analyses of climate-growth relationships in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individuals provided by the Berlin Senate Administration (Senatsverwaltung).
+This data set contains information on location, species, trunk diameter (at breast height; $DBH$), and height, amongst other variables for street and park trees.
+In a space-for-time substitution, growth of individual species can be assessed across the entire cite of Berlin, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.
+Similar applications are found in @quigley2004, who inferred absolute growth potential for species across successional groups, and between rural and urban species, yet lacked spatially-explicit effect size estimates or predictions of maximum potential.
+Similarly, @pretzscho2017 applied linear hierarchical models to infer growth modulation for different cities, time periods and locations (urban $vs.$ rural).
+By contrast, we propose applying a statistical model (see Section$~$\@ref(sec:methods)) that is spatially explicite, while allowing to account for the nested nature of the data set.
+That is, 
+This also allows to infer the absolute growth potential of species given, for example, a specific location or UHI magnitude, as
 
-Note [@gregg2003].. rather impeded growth adjacent to intense urban clusters; but, here focused on intra-city patterns.
+We acknolwedge results from [@gregg2003], which indicated impeded growth adjacent to intense urban clusters, rather than enhancement within them.
+However
 

### 2020-03-07 analysis/paper/paper.Rmd
methods section

@@
 
-Trees and green infrastructure in urban areas show a tendency for enhanced growth rates and/or productivity compared to rural counterparts [@jia2018; @pretzsch2017], yet feature a broad range of effect size ranges and, in some cases, directions specific to species and locality.
+Trees and green infrastructure in urban areas show a tendency for enhanced growth rates and/or productivity compared to rural counterparts [@jia2018; @pretzsch2017], yet feature a broad range of effect size ranges and, in some cases, signs specific to species and locality.
 @zhao2016 showed that growth rates increased within urban clusters as urbanization intensifies using remotely sensed vegetation indices.
-Similarly, for Berlin, @dahlhausen2018, identified positive growth modulation in highly urbanized environments for *Tilia cordata* Mill (using growth increments), the most abundant tree of the city, which they attributed to the UHI effect, while intermiedate development intensity showed indications of being least favorable for tree growth.
+Similarly, for Berlin, @dahlhausen2018, identified positive growth modulation in highly urbanized environments for *Tilia cordata* Mill (using growth increments), the most abundant tree of the city, which they attributed to the UHI effect, while intermediate development intensity showed indications of being least favorable for tree growth.
 <!-- For Berlin, @dahlhausen2018, showed divergent patterns for young and old trees dependent on urban development intensity for *Tilia cordata* Mill, the most abundant tree of the city, but overall reported an increased growth rate which they attributed to the UHI effect.
@@
 [@rhoades1999; @quigley2004; @randrup2001; @pauleit2002].
-Under climate change, atmospheric drought will likely be compounded with high temperatures more frequently, adding further stress to current urban disturbance regimes [@roloff2009].
+Under climate change, atmospheric drought will likely be compounded with high temperatures - and intensified UHIs - more frequently, adding further stress to current urban disturbance regimes [@roloff2009].
 
@@
 <!-- Therefore, studies reliant on labor-intensive sampling, must carefully address confounding factors to ensure effect sizes are estimated adequately. -->
-
 To complement detailed dendroecological analyses of climate-growth relationships in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individuals provided by the Berlin Senate Administration (Senatsverwaltung).
@@
 Similar applications are found in @quigley2004, who inferred absolute growth potential for species across successional groups, and between rural and urban species, yet lacked spatially-explicit effect size estimates or predictions of maximum potential.
-Similarly, @pretzscho2017 applied linear hierarchical models to infer growth modulation for different cities, time periods and locations (urban $vs.$ rural).
-By contrast, we propose applying a statistical model (see Section$~$\@ref(sec:methods)) that is spatially explicite, while allowing to account for the nested nature of the data set.
-That is, 
-This also allows to infer the absolute growth potential of species given, for example, a specific location or UHI magnitude, as
+Similarly, @pretzsch2017 applied linear hierarchical models to infer growth modulation for different cities, time periods and locations (urban $vs.$ rural).
+By contrast, we propose applying a statistical model that is spatially explicit, while also allowing to account for the nested nature of the data set (e.g.
+streets and districts) as well as other pertinent factors (hierarchical, generalized additive model, see Section$~$\@ref(sec:methods)).
+This also allows to infer the absolute growth potential of species given, for example, a specific location, age or UHI magnitude.
 
 We acknolwedge results from [@gregg2003], which indicated impeded growth adjacent to intense urban clusters, rather than enhancement within them.
-However
-
-# Background
 
 # Proposed methods and data requirements 
+\label{sec:methods}
+
+The proposed statistical method is from the class of hierarchical, generalized additive models (GAM, or GAMM for mixed models/hierarchical models).
+In these models combinations of continuous and categorical predictor variables can be summed to estimate a response. 
+In particular, continuous variables that are linear, as well as non-linearly related to the response can be represented by applying a transfer function, typically termed "smoothing function" [@wood2017]; these are constructed using a number of base functions of varying complexity and form, which provides a high degree of flexibility, ideal for fitting ecosystem dynamics which are rarely linear [@pedersen2019], or correctly represented with deterministic functional forms (e.g.
+quadratic equations).
 

### 2020-03-07 analysis/paper/paper.Rmd
eod

@@
 @zhao2016 showed that growth rates increased within urban clusters as urbanization intensifies using remotely sensed vegetation indices.
-Similarly, for Berlin, @dahlhausen2018, identified positive growth modulation in highly urbanized environments for *Tilia cordata* Mill (using growth increments), the most abundant tree of the city, which they attributed to the UHI effect, while intermediate development intensity showed indications of being least favorable for tree growth.
+Similarly, for Berlin, @dahlhausen2018, identified positive growth modulation in highly urbanized environments (using growth increments) for *Tilia cordata* Mill, the most abundant tree of the city, which they attributed to the UHI effect, while intermediate development intensity showed indications of being least favorable for tree growth.
 <!-- For Berlin, @dahlhausen2018, showed divergent patterns for young and old trees dependent on urban development intensity for *Tilia cordata* Mill, the most abundant tree of the city, but overall reported an increased growth rate which they attributed to the UHI effect.
@@
 platanoides* and *pseudoplatanus*), *Platanus x hispanica* and *Quercus rubra* with higher summer temperatures of the preceding year, especially when compounded with drought, in another German metropolis (Dresden).
-Differences in growth trends may result from contrasting intrinsic inter- and intra-specific characteristics, but are indeed affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc.
+Differences in growth trends may result from contrasting species-specific characteristics, but are indeed affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc.
 [@rhoades1999; @quigley2004; @randrup2001; @pauleit2002].
@@
 In these models combinations of continuous and categorical predictor variables can be summed to estimate a response. 
-In particular, continuous variables that are linear, as well as non-linearly related to the response can be represented by applying a transfer function, typically termed "smoothing function" [@wood2017]; these are constructed using a number of base functions of varying complexity and form, which provides a high degree of flexibility, ideal for fitting ecosystem dynamics which are rarely linear [@pedersen2019], or correctly represented with deterministic functional forms (e.g.
+In particular, continuous variables that are linearly, as well as non-linearly related to the response can be represented by applying a transfer function, typically termed "smoothing function" [@wood2017]; these are constructed using a number of base functions of varying complexity and form, which provides a high degree of flexibility, ideal for fitting ecosystem dynamics which are rarely linear [@pedersen2019], or correctly represented with deterministic functional forms (e.g.
 quadratic equations).

### 2020-03-09 analysis/paper/paper.Rmd
updated methods brief

@@
       fig_caption: yes
-      toc: yes
+      toc: true
     bookdown::pdf_document2:
@@
 library(drake)
+library(kableExtra)
 # library(ggplot2)
@@
 In a space-for-time substitution, growth of individual species can be assessed across the entire cite of Berlin, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.
-Similar applications are found in @quigley2004, who inferred absolute growth potential for species across successional groups, and between rural and urban species, yet lacked spatially-explicit effect size estimates or predictions of maximum potential.
-Similarly, @pretzsch2017 applied linear hierarchical models to infer growth modulation for different cities, time periods and locations (urban $vs.$ rural).
-By contrast, we propose applying a statistical model that is spatially explicit, while also allowing to account for the nested nature of the data set (e.g.
+Comparable applications are found, for example, in @quigley2004 and @pretzsch2017.
+The former  inferred absolute growth potential for species across successional groups, and between rural and urban species, yet lacked spatially-explicit effect size estimates or predictions of maximum potential;
+@pretzsch2017 applied linear hierarchical models to infer growth modulation for different cities, time periods and locations (urban $vs.$ rural).
+
+**By contrast, we propose applying a statistical model that is spatially explicit, while also allowing to account for the nested nature of the data set (e.g.
 streets and districts) as well as other pertinent factors (hierarchical, generalized additive model, see Section$~$\@ref(sec:methods)).
-This also allows to infer the absolute growth potential of species given, for example, a specific location, age or UHI magnitude.
+This also allows to infer the absolute growth potential of species given, for example, a specific location, age or UHI magnitude.**
 
-We acknolwedge results from [@gregg2003], which indicated impeded growth adjacent to intense urban clusters, rather than enhancement within them.
+<!-- We acknolwedge results from [@gregg2003], which indicated impeded growth adjacent to intense urban clusters, rather than enhancement within them. -->
 
-# Proposed methods and data requirements 
-\label{sec:methods}
+# Proposed methods and data requirements {#sec:methods}
+
+## General analyses
 
@@
 quadratic equations).
+In general, a GAM can be written as:
+
+\begin{equation}
+E (Y)~=~g^{-1}\left( \beta_0 + \sum_{i = 1}^{n} f_i (x_i) \right),
+\end{equation}
+
+and
+
+\begin{equation}
+y~=~E (Y) + \epsilon,
+\end{equation}
+
+where $Y$ is taken from an appropriate distribution and corresponding link function $g$, $\beta_0$ is the intercept and $f_i$ represents a smooth function of a predictor [@pedersen2019], and $\epsilon \sim \mathcal{N}(0, \sigma ^2)$.
+Note, that $f_i$ consists of a smooth (e.g. spline) constructed via basis functions of different form and complexity, multiplied by a coeffecient:
+
+\begin{equation}
+f_i(x_i)~=~\sum_{k = 1}^{K} \beta_{i, k} b_{i,k}(x_i).
+\end{equation}
+
+Nested data structures (e.g.
+due to similar road [type]) can be accounted for by introducing random effects [@wood2017], while spatial dependence between observations can be included by constructing smoothing functions with e.g.
+northings and eastings, as for example done in [@augustin2009].
+Ultimately, the implementation of a such a GAMM will alow for establishing continuous prediction surfaces of growth potential (approximated via $DBH$) for individual species across urban areas (including parks) of Berlin.
+
+Currently, $DBH$ has been modelled using a hierarchical linear model (linear mixed effects model) with `lme4` [@bates2015] in @rcoreteam2020.
+The general form of this model is:
+
+\begin{equation}
+Y_{i,j} = (\beta_0 + b_{0,i,j}) + (\beta_1 + b_{1,i,j}) \cdot x_i  + \epsilon_{i,j},
+\end{equation}
+where $\beta_0$ is the intercept with its random component $b_0$, and $\beta_1$ the slope with its random component $b_1$.
+The random errors are assumed i.i.d. and distributed as $b \sim \mathcal{N}(0, \tau ^2)$. 
+The model for which results are presented in Figure$~$\@ref(fig:fig-lmestat) estimates $DBH$ from tree age and the local UHI intensity as continuous covariates with random slopes and intercepts for each species;
+note, that for computational efficiency each genera was modelled seperately.
+Further, models were only established for species with at least 150 individuals.
+
+## Available and required data
+
+Table$~$\@ref(tab:tab-data-req) provides a list with currently accessible/available data, including information on (desired) resolution, and sources.
+
+```{r tab-data-req,echo=FALSE}
+
+dat_req <- read.csv(here::here("analysis", "data", "tables", "01_data_requirements.csv"), header = TRUE)
+
+if (grepl("docx", out_format) ||
+    knitr::is_latex_output() ) {
+  table_dat_req <- dat_req %>% 
+  select(-owned) %>% 
+  knitr::kable(booktabs = TRUE,
+
+             caption = "Data requirements for analysis, including currently available/accessed and required/desired data.",
+             caption.short = "Data requirements for analysis",
+             col.names = c("Subject/Relevance",
+                          "Desc.",
+                          "Type",
+                          "Obsv. (n)",
+                          "Resolution (m)",
+                          "Source")) %>% 
+  kableExtra::kable_styling(full_width = TRUE, 
+                            font_size = 10) %>% 
+  kableExtra::row_spec(0, bold = TRUE) %>% 
+  kableExtra::pack_rows("available or accessed", 1, 4) %>%
+  kableExtra::pack_rows("required and/or desired", 5, 8) %>% kableExtra::landscape()
+
+} else {
+  table_dat_req <- dat_req %>% 
+  select(-owned) %>% 
+  knitr::kable(booktabs = TRUE,
+               format = "html",
+               caption = "Data requirements for analysis, including currently available/accessed and required/desired data.",
+             caption.short = "Data requirements for analysis",
+             col.names = c("Subject/Relevance",
+                          "Desc.",
+                          "Type",
+                          "Obsv. (n)",
+                          "Resolution (m)",
+                          "Source")) %>% 
+  kableExtra::kable_styling(full_width = TRUE, 
+                            font_size = 10) %>% 
+  kableExtra::row_spec(0, bold = TRUE) %>% 
+  kableExtra::pack_rows("available or accessed", 1, 4) %>%
+  kableExtra::pack_rows("required and/or desired", 5, 8)
+
+}
+
+ table_dat_req
+
+```
 
@@
 
-```{r fig-tree-overview-map, fig.cap="A plot of random numbers",fig.width = 15, fig.height = 8, out.width="100%", include=TRUE, eval=TRUE}
+```{r fig-tree-overview-map, fig.cap="Individual tree locations for three categories available in Berlin Senate urban tree data set.
+Note, that for each category 7000 observations were subsampled from the available pool to facilitate visualization.",fig.width = 15, fig.height = 8, out.width="100%", include=TRUE, eval=TRUE}
 
@@
 
-```{r fig-tree-count-map, fig.cap="A plot of random numbers",fig.width = 12, fig.height = 14, out.width="100%"}
+```{r fig-tree-count-map, fig.cap="Gridded counts for the 11 most frequent genera, as well as *Pinus* and remaining genera.
+Note, that counts are standardized to unity for individual genera.",fig.width = 12, fig.height = 14, out.width="100%"}
 
@@
 
-```{r fig-uhi-map, fig.cap="A plot of random numbers",fig.width = 12, fig.height = 14, out.width="50%"}
+(ref:chakraborty2019) [@chakraborty2019]
+
+```{r fig-uhi-map, fig.cap="Estimate of UHI intensity based on the algorithm in (ref:chakraborty2019), comparing urban with rural pixels within the greater metropolitan cluster.
+Presented values are averaged over the summer of 2007.",fig.width = 12, fig.height = 14, out.width="50%"}
 
@@
 
-```{r fig-density, fig.cap="A plot of random numbers",fig.width = 12, fig.height = 12, out.width="100%"}
+```{r fig-density, fig.cap="Empirical density distribution of all individuals within the presented genera along the UHI continuum.
+UHI intensities were extracted for each tree location, and the distribution hence represents the first detailed overview of the exposure of Berlin's trees to urban heat loading.
+Insets represent corresponding tree totals.",fig.width = 12, fig.height = 12, out.width="100%"}
 knitr::include_graphics(here::here("analysis/figures/plot_02_genus_UHI_dens.png"))
@@
 
-```{r fig-lmestat, fig.cap="A plot of random numbers",fig.width = 20, fig.height = 12, out.width="100%"}
+```{r fig-lmestat, fig.cap="Impact of UHI loading on tree diameter (\\(DBH\\)), accounting for age and inter-specific differences from the linear mixed model (via random slopes and intercepts).
+Line-ranges are standard errors of predicted effect sizes (i.e.
+slopes).
+Differences between street and park trees are considerable for some species, and may be due to local clustering and/or spatial under-representation across the UHI continuum.
+Further investigations need to address the degree of spatial autocorrelation and account for it where required in linear mixed models, and with smoothing interactions in a GAMM implementation.",fig.width = 20, fig.height = 12, out.width="100%"}
 knitr::include_graphics(here::here("analysis/figures/plot_03_ranef_species_dbh_uhi.png"))
@@
 
-Figure \@ref(fig:lmestat) shows how we can have a caption and cross-reference for a plot
+Figure \@ref(fig:fig-lmestat) shows how we can have a caption and cross-reference for a plot
 
 # Outlook
+
+We seek to build upon and improve the current analysis by:
+
+- incorporating more pertinent covariates as dependent variables in the linear mixed model
+- testing multiple model structures with formal model selection procedures
+- checking model resiudals for spatial auto-correlation and accounting for it where necessary to ensure unbiased estimates of effect sizes
+- repeating the above with a hierarchical GAM (i.e. GAMM) to allow for:  
+  * estimating continuous prediction surfaces for UHI impacts on individual species' growth (similar to results in Figure$~$\@ref(fig:fig-lmestat)) under recent conditions
+  * estimating absolute, species-specific growth potential under increased temperatures and UHI loading under climate change, ideally based on simulations (otherwise step-wise increases based on RCP scenarios) for the key species.
+  * assess potential age-dependent UHI impacts on individual species.
 

### 2020-03-20 analysis/paper/paper.Rmd
cont. writing

@@
 abstract: |
-  **This document serves as a brief overview and outline** .
+  **This document outlines the rational for an analysis tree growth (potential) and its relationship with the Urban Heat Island (UHI) effect in Berlin.
+It introduces preliminary results and provides an outlook for up-coming and potential work.** .
 keywords: |
@@
 <!-- Therefore, studies reliant on labor-intensive sampling, must carefully address confounding factors to ensure effect sizes are estimated adequately. -->
-To complement detailed dendroecological analyses of climate-growth relationships in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individuals provided by the Berlin Senate Administration (Senatsverwaltung).
+
+<!-- Detailed, but spatially-limited dendroecological analyses of climate-growth relationships in Berlin for key species can be enhanced by inferring growth modulation from a large data set in excess of 650000 individuals provided by the Berlin Senate Administration (Senatsverwaltung).
+-->
+
+To complement (existing) detailed dendroecological analyses of climate-growth relationships in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individuals provided by the Berlin Senate Administration (Senatsverwaltung).
 This data set contains information on location, species, trunk diameter (at breast height; $DBH$), and height, amongst other variables for street and park trees.
@@
 This also allows to infer the absolute growth potential of species given, for example, a specific location, age or UHI magnitude.**
+Tree-level growth data, however, is paramount in validating such relationships, and its inclusion in the model would also allow incorporating effects of varying climate.
 
@@
 E (Y)~=~g^{-1}\left( \beta_0 + \sum_{i = 1}^{n} f_i (x_i) \right),
+(\#eq:gam-general)
 \end{equation}
@@
 y~=~E (Y) + \epsilon,
+(\#eq:gam-expectation)
 \end{equation}
@@
 f_i(x_i)~=~\sum_{k = 1}^{K} \beta_{i, k} b_{i,k}(x_i).
+(\#eq:gam-smoothers)
 \end{equation}
@@
 Y_{i,j} = (\beta_0 + b_{0,i,j}) + (\beta_1 + b_{1,i,j}) \cdot x_i  + \epsilon_{i,j},
+(\#eq:lme)
 \end{equation}
@@
   kableExtra::kable_styling(full_width = TRUE, 
-                            font_size = 10) %>% 
-  kableExtra::row_spec(0, bold = TRUE) %>% 
+                            font_size = 11) %>% 
+  kableExtra::row_spec(0, bold = TRUE, font_size = 12) %>% 
+  kableExtra::row_spec(9,italic =  TRUE) %>% 
   kableExtra::pack_rows("available or accessed", 1, 4) %>%
-  kableExtra::pack_rows("required and/or desired", 5, 8) %>% kableExtra::landscape()
+  kableExtra::pack_rows("required and/or desired", 5, 9) %>% kableExtra::landscape()
 
@@
   kableExtra::pack_rows("available or accessed", 1, 4) %>%
-  kableExtra::pack_rows("required and/or desired", 5, 8)
+  kableExtra::pack_rows("required and/or desired", 5, 9)
 
@@
 
-Some results
+Tree locations are clustered and follow spatial structures based on their category, i.e. riparian, street and park trees (Fig.$~$\@ref(fig:fig-tree-overview-map)).
+Planting in space and time shows species specific patterns (by districts), often related to major events, such as the start and end of wars, as well as the German separation (not shown here).
+Table$~$\@ref(tab:tab-age) shows the binned distribution of generae across age classes.
+
+```{r tab-age}
+
+loadd(age_tables)
+
+age_tables <- age_tables %>% 
+  dplyr::arrange(desc(n_total))
+colnames(age_tables)[c(1,7,8)] <- c("Genera", "Total (n)", "Missing (n)")
+
+age_tables <- bind_rows(age_tables[age_tables$Genera!="Other",],
+                        age_tables[age_tables$Genera=="Other",])
+
+add_row <- as.data.frame(matrix(c(NA, colSums(age_tables[,-1])), nrow = 1))
+add_row[1,1] <- "Marg. Totals"
+colnames(add_row) <- colnames(age_tables)
+
+age_tables <- bind_rows(age_tables,
+                    add_row)
+
+if (grepl("docx", out_format) ||
+    knitr::is_latex_output() ) {
+  table_dat_req_kable <- age_tables %>% 
+  knitr::kable(booktabs = TRUE,
+
+             caption = "Trees",
+             caption.short = "Trees"
+             # col.names = c("Subject/Relevance",
+             #              "Desc.",
+             #              "Type",
+             #              "Obsv. (n)",
+             #              "Resolution (m)",
+             #              "Source")
+             ) %>% 
+  kableExtra::row_spec(13, hline_after = TRUE)
+
+} else {
+table_dat_req_kable <- age_tables %>% 
+  knitr::kable(booktabs = TRUE,
+               format = "html",
+               caption = "Trees",
+             caption.short = "Trees"
+             # col.names = c("Subject/Relevance",
+             #              "Desc.",
+             #              "Type",
+             #              "Obsv. (n)",
+             #              "Resolution (m)",
+             #              "Source")
+             ) %>% 
+  kableExtra::row_spec(13, extra_css = "border-bottom: 1px solid")
+
+# %>% 
+#   kableExtra::kable_styling(full_width = TRUE, 
+#                             font_size = 10) %>% 
+#   kableExtra::column_spec(1, italic = TRUE) %>%
+#   kableExtra::row_spec(0, bold = TRUE) %>% 
+#     kableExtra::row_spec(13, hline_after = TRUE)
+
+}
+
+   table_dat_req_kable %>% 
+  kableExtra::kable_styling(full_width = TRUE, 
+                            font_size = 11) %>% 
+  kableExtra::column_spec(1, italic = TRUE) %>%  
+  kableExtra::row_spec(0, italic = FALSE, bold = TRUE, font_size = 12)
+
+```
 
@@
 
+The distribution of the UHI effect is highly irregular and clustered in space (Fig.$~$\@ref(fig:fig-uhi-map)), and also shows variability through time (data not shown, refer to the [urban heat island explorer](https://yceo.users.earthengine.app/view/uhimap)).
+
 (ref:chakraborty2019) [@chakraborty2019]
@@
 
+The exposure to increased heat-loading of individual genera (and consequently species) is highly uneven throughout the city (Fig.$~$\@ref(fig:fig-density)).
+Street and park trees of most genera are clustered in urban areas with intermediate to high UHI loading, while riparian trees, and some street and park trees of other genera tend to be spread more evenly across Berlin's UHI range.
+
 ```{r fig-density, fig.cap="Empirical density distribution of all individuals within the presented genera along the UHI continuum.
 UHI intensities were extracted for each tree location, and the distribution hence represents the first detailed overview of the exposure of Berlin's trees to urban heat loading.
+The black line represents the density across all three categories.
 Insets represent corresponding tree totals.",fig.width = 12, fig.height = 12, out.width="100%"}
@@
 ```
+
+The effect of UHI loading on absolute growth potential varies between genera and species (Fig.$~$\@ref(fig:fig-lmestat)).
+Most notably, *Quercus*, the 3rd-most frequent genera, shows decreased absolute growth with increased UHI,
+while the most frequent genera, *Tilia* clearly features enhanced growth.
+Note, that the estimated effect sizes here are linear.
+Temperature may excert a non-linear control on absolute growth, however, and hence, applying a  method able to capture may result in altered presented effect sizes.
+Additionally, if temperatures increase in the future under climate warming current future increases in under climate warming, any non-linear effects may become more enhanced, stressing the need for a more flexible model fit and structure.
+
+alter the (preliminary) results substantially
 
@@
   * estimating continuous prediction surfaces for UHI impacts on individual species' growth (similar to results in Figure$~$\@ref(fig:fig-lmestat)) under recent conditions
-  * estimating absolute, species-specific growth potential under increased temperatures and UHI loading under climate change, ideally based on simulations (otherwise step-wise increases based on RCP scenarios) for the key species.
+  * estimating absolute, species-specific growth potential under increased temperatures and UHI loading under climate change, ideally based on climate simulations (otherwise step-wise increases based on RCP scenarios) for the key species.
+Note, that  complications presented by 'out-of-sample' predictions will be addressed.
   * assess potential age-dependent UHI impacts on individual species.

### 2020-03-20 R/functions.R

@@ -1341,6 +1341,49 @@ make_ranef_plot <- function(model_out,
 }
 
 
+# Tables ------------------------------------------------------------------
+
+#' Generate table of genera age distribution
+#'
+#' @param df Berlin trees data frame
+#' @param max_age numeric, last cut
+#' @param break_interval numeric, size of intervals, note, max_age/break_intervals must be an integer.
+#'
+#' @return
+#' @export
+#'
+#' @import data.table
+#'
+make_age_table <- function(df, max_age = 120, break_interval = 20){
+
+
+    # make df with age breaks
+    age_df <- data.table(df)[ , .(STANDALTER, gattung_short)][
+        , age_group := cut(STANDALTER, breaks = seq(0, max_age, break_interval), right = FALSE)]
+
+    species_totals <- age_df[ ,.(n_total = .N), by = gattung_short]
+
+    cols <- c(1,3:(3+ max_age/break_interval),2)
+
+
+    age_df_class <- dcast(age_df,
+                          gattung_short ~ age_group,
+                          value.var = "STANDALTER",
+                          fun.aggregate = length)[species_totals,
+                                                  on = "gattung_short"][ ,..cols]
+
+    setnames(age_df_class, old = "NA", new = "missing")
+
+
+
+
+    # age_df_perc <- age_df_class[, 2:8 ]/species_totals[,n_total]
+    # age_df_perc[, gattung_short := as.character(age_df$gattung_short)]
+
+    return(age_df_class)
+
+}
+
 
 
 

### 2020-03-20 R/plan.R

@@ -157,6 +157,15 @@ plan <- drake_plan(
                                                  dpi = 300),
 
 
+    # tables
+
+
+    age_tables = berlin.trees::make_age_table(df = model_df,
+                                              max_age = 150,
+                                              break_interval = 30),
+
+
+
 
 
 
@@ -172,13 +181,13 @@ plan <- drake_plan(
         output_format = bookdown::html_document2(),
         quiet = TRUE
     ),
-    paper_word = rmarkdown::render(
-        knitr_in("./analysis/paper/paper.Rmd"),
-        output_file = file_out(file.path(here::here(), "paper_knit.docx")),
-        # output_file = file_out(file.path("./paper_knit.html")),
-        output_format = bookdown::word_document2(),
-        quiet = TRUE
-    ),
+    # paper_word = rmarkdown::render(
+    #     knitr_in("./analysis/paper/paper.Rmd"),
+    #     output_file = file_out(file.path(here::here(), "paper_knit.docx")),
+    #     # output_file = file_out(file.path("./paper_knit.html")),
+    #     output_format = bookdown::word_document2(),
+    #     quiet = TRUE
+    # ),
     paper_pdf = rmarkdown::render(
         knitr_in("./analysis/paper/paper.Rmd"),
         output_file = file_out(file.path(here::here(), "paper_knit.pdf")),

### 2020-03-20 _setup.R

@@ -23,6 +23,7 @@ usethis::use_package("rgdal")
 # usethis::use_package("extrafont")
 # usethis::use_package("scales")
 # usethis::use_package("raster")
+# usethis::use_package("data.table")
 
 ## packages
 

### 2020-03-22 analysis/paper/paper.Rmd
compartmentalizing chapter et chambers

@@
 abstract: |
-  **This document outlines the rational for an analysis tree growth (potential) and its relationship with the Urban Heat Island (UHI) effect in Berlin.
-It introduces preliminary results and provides an outlook for up-coming and potential work.** .
+  **This document outlines the rationale for an analysis of tree growth (potential) and its relationship with the Urban Heat Island (UHI) effect in Berlin using an extensive, publicly available data set.
+It introduces preliminary results and provides an outlook for up-coming and potential work.**
 keywords: |
@@
 
-Table$~$\@ref(tab:tab-data-req) provides a list with currently accessible/available data, including information on (desired) resolution, and sources.
+Additional remote sensing data for local context, higher resolution UHI data, and most important, temporally-resolved tree growth data are still required and/or would greatly improve the impact and confidence in conclusions.
+See Tab.$~$\@ref(tab:tab-data-req) for details.
 
@@
              caption.short = "Data requirements for analysis",
-             col.names = c("Subject/Relevance",
+             col.names = c("Subject / Relevance",
                           "Desc.",
@@
 
-Figure \@ref(fig:fig-lmestat) shows how we can have a caption and cross-reference for a plot
-
 # Outlook

### 2020-03-23 analysis/paper/paper.Rmd
updated tables and text

@@
 To complement (existing) detailed dendroecological analyses of climate-growth relationships in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individuals provided by the Berlin Senate Administration (Senatsverwaltung).
-This data set contains information on location, species, trunk diameter (at breast height; $DBH$), and height, amongst other variables for street and park trees.
+This data set contains information on location, species, trunk diameter (at breast height; $DBH$), and height, amongst other variables for street and park trees (.
 In a space-for-time substitution, growth of individual species can be assessed across the entire cite of Berlin, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.

### 2020-03-23 R/functions.R

@@ -1343,6 +1343,33 @@ make_ranef_plot <- function(model_out,
 
 # Tables ------------------------------------------------------------------
 
+#' Make overview table
+#'
+#' @param df berlin trees df
+#'
+#' @return data frame ready for kabling
+#' @export
+#' @import data.table
+#'
+make_overview_table <- function(df){
+
+    full_data_set_clean <- data.table(df)
+    counts_full <- full_data_set_clean[,.N, by = provenance]
+    counts_with_age_size <- full_data_set_clean[!is.na(dbh_cm) & !is.na(STANDALTER),
+                                                .N,
+                                                by = provenance]
+
+    counts_full <- cbind(counts_full, counts_with_age_size[,2])
+    counts_full[3,3] <- NA
+    counts_full[,1] <- c("Park", "Street", "Riparian")
+    colnames(counts_full) <- c(" ", "N", "N_{age}")
+    counts_full[order(-N)]
+
+    return(counts_full)
+}
+
+
+
 #' Generate table of genera age distribution
 #'
 #' @param df Berlin trees data frame

### 2020-03-23 R/plan.R

@@ -9,7 +9,7 @@ plan <- drake_plan(
 
 
 
-    ## Spatial Ancillary
+    ## Spatial Ancillary -----------------------------------
 
     ### Berlin districts and BBoxx
     berlin_polygons = target(berlin.trees::get_berlin_polygons_as_sf()),
@@ -19,12 +19,15 @@ plan <- drake_plan(
                                            "greater_berlin",
                                            crs = 4326),
 
-    ### Berlin UHI gridded data
+    ### Berlin UHI gridded data -------------------------------
 
 
 
     uhi_stacks = berlin.trees::get_uhi_rasters(file.path(here::here(), "analysis", "data", "raw_data", "UHI_explorer")),
 
+
+
+    # Berlin trees data -------------------------------------------------------
 
 
 
@@ -46,6 +49,9 @@ plan <- drake_plan(
     extract_uhi_values_to_list = berlin.trees::add_uhi_hist_data(uhi_stack_list = uhi_stacks,
                                                                  sf_data = full_data_set_clean[, ]),
 
+
+
+     # Model -------------------------------------------------------------------
 
 
 
@@ -93,7 +99,7 @@ plan <- drake_plan(
 
 
 
-    # Plotting
+    # Plotting --------------------------------
     plot_overview_map = berlin.trees::make_overview_map(full_data_set_clean,
                                                         berlin_polygons,
                                                         file = drake::file_out("./analysis/figures/map_01_overview.png"),
@@ -157,7 +163,9 @@ plan <- drake_plan(
                                                  dpi = 300),
 
 
-    # tables
+    # tables ------------------------------------------
+
+    overview_table = berlin.trees::make_overview_table(full_data_set_clean),
 
 
     age_tables = berlin.trees::make_age_table(df = model_df,

### 2020-03-25 analysis/paper/paper.Rmd
cont work

--- 2020-03-22 analysis/paper/paper.Rmd
@@
 author:
-  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^a$
+  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^a$, ...
   - $^a$ *Geoforschungszentrum Potsdam, Section 4.3, Germany*
@@
       toc: yes
+header-includes:
+- \usepackage{booktabs}
+- \usepackage{longtable}
+- \usepackage{array}
+- \usepackage{multirow}
+- \usepackage{wrapfig}
+- \usepackage{float}
+- \usepackage{pdflscape}
+- \usepackage{tabu}
+- \usepackage{threeparttable}
+- \usepackage[normalem]{ulem}
 bibliography: uhi_references.bib
@@
 To complement (existing) detailed dendroecological analyses of climate-growth relationships in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individuals provided by the Berlin Senate Administration (Senatsverwaltung).
-This data set contains information on location, species, trunk diameter (at breast height; $DBH$), and height, amongst other variables for street and park trees.
+This data set contains information on location, species, trunk diameter (at breast height; $DBH$; see Tab.$~$\@ref(tab:tab-tree-overview)), and height, amongst other variables for street and park trees.
+
+```{r tab-tree-overview}
+loadd(overview_table)
+
+caption_ov <- "Available records by categoryin entire data set (n), and those with age and $DBH$ entries (n$_{full}$)."
+caption_ov_short <- "Available records in berlin tree database."
+
+if (grepl("docx", out_format) ||
+    knitr::is_latex_output() ) {
+  table_ov <- overview_table %>% 
+  knitr::kable(booktabs = TRUE,
+
+             caption = caption_ov,
+             caption.short = caption_ov_short,
+             format = "latex",
+             # col.names = c("Subject/Relevance",
+             #              "Desc.",
+             #              "Type",
+             #              "Obsv. (n)",
+             #              "Resolution (m)",
+             #              "Source")
+             escape = FALSE)
+
+} else {
+table_ov <- overview_table %>% 
+  knitr::kable(booktabs = TRUE,
+               format = "html",
+
+             caption = caption_ov,
+             caption.short = caption_ov_short,
+             # col.names = c("Subject/Relevance",
+             #              "Desc.",
+             #              "Type",
+             #              "Obsv. (n)",
+             #              "Resolution (m)",
+             #              "Source")
+             escape = FALSE) 
+}
+
+ table_ov %>% 
+  kableExtra::kable_styling(full_width = TRUE, 
+                            font_size = 11,
+                            latex_options = c("hold_position")) %>% 
+  kableExtra::row_spec(0, bold = TRUE, font_size = 12)
+
+```
+
 In a space-for-time substitution, growth of individual species can be assessed across the entire cite of Berlin, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.
@@
 The former  inferred absolute growth potential for species across successional groups, and between rural and urban species, yet lacked spatially-explicit effect size estimates or predictions of maximum potential;
-@pretzsch2017 applied linear hierarchical models to infer growth modulation for different cities, time periods and locations (urban $vs.$ rural).
+@pretzsch2017 applied linear hierarchical models to infer growth modulation for different cities, time periods and urban $vs.$ rural locations while accounting for stand-level variability.
 
-**By contrast, we propose applying a statistical model that is spatially explicit, while also allowing to account for the nested nature of the data set (e.g.
-streets and districts) as well as other pertinent factors (hierarchical, generalized additive model, see Section$~$\@ref(sec:methods)).
-This also allows to infer the absolute growth potential of species given, for example, a specific location, age or UHI magnitude.**
-Tree-level growth data, however, is paramount in validating such relationships, and its inclusion in the model would also allow incorporating effects of varying climate.
+**By contrast, we propose applying a statistical model that is fully spatially explicit, while also allowing to account for the nested nature of the data set (e.g.
+streets and districts) as well as other pertinent factors using hierarchical, generalized additive models (see Section$~$\@ref(sec:methods).
+This also allows to infer the absolute growth potential of a species given, for example, a specific location, age or UHI magnitude.**
+Tree-level growth data, however, is paramount in validating such relationships, and its inclusion in the model would also allow incorporating effects of varying climate over time.
 
@@
 
-Currently, $DBH$ has been modelled using a hierarchical linear model (linear mixed effects model) with `lme4` [@bates2015] in @rcoreteam2020.
+Currently, $DBH$ has been modelled using a hierarchical linear model (linear mixed effects model) with `lme4` [@bates2015] in @rcoreteam2020 (see Section$~$\@ref(sec:results)).
 The general form of this model is:
@@
 note, that for computational efficiency each genera was modelled seperately.
-Further, models were only established for species with at least 150 individuals.
+Further, models were established for the three most abundant species per genera with at least 1000 individuals.
+
+We caution, that 1) the presented model results are preliminary, 2) initial diagnostics indicate that modification of the current model structure (e.g.
+by accounting for individual-level variability) may be required and 3) that further, time-intensive data cleaning is necessary.
+First tests, however, showed that while the magnitude of effects changes with different model structures, their direction appears stable.
 
@@
 
-Additional remote sensing data for local context, higher resolution UHI data, and most important, temporally-resolved tree growth data are still required and/or would greatly improve the impact and confidence in conclusions.
+Additional remote sensing data for local context, higher resolution UHI data, and most important, temporally-resolved tree growth data are still required and/or would greatly improve the impact and confidence in effect size estimates.
 See Tab.$~$\@ref(tab:tab-data-req) for details.
@@
 dat_req <- read.csv(here::here("analysis", "data", "tables", "01_data_requirements.csv"), header = TRUE)
+
+caption_data_req <- "Data requirements for analysis, including currently available/accessed and required/desired data."
+caption_data_req_short <- "Data requirements for analysis."
 
@@
 
-             caption = "Data requirements for analysis, including currently available/accessed and required/desired data.",
-             caption.short = "Data requirements for analysis",
+             caption = caption_data_req,
+             caption.short = caption_data_req_short,
+             format = "latex",
              col.names = c("Subject / Relevance",
@@
 
-# Preliminary results
+# Preliminary results {#sec:results}
 
-Tree locations are clustered and follow spatial structures based on their category, i.e. riparian, street and park trees (Fig.$~$\@ref(fig:fig-tree-overview-map)).
-Planting in space and time shows species specific patterns (by districts), often related to major events, such as the start and end of wars, as well as the German separation (not shown here).
-Table$~$\@ref(tab:tab-age) shows the binned distribution of generae across age classes.
+Tree locations are clustered and structured based on their category, i.e. riparian, street and park trees (Fig.$~$\@ref(fig:fig-tree-overview-map)).
+Planting in space and time shows species-specific patterns (by districts), often related to major events, such as the start and end of armed and/or political conflict.
+Table$~$\@ref(tab:tab-age) shows the binned distribution of genera across age classes.
 
@@
   dplyr::arrange(desc(n_total))
-colnames(age_tables)[c(1,7,8)] <- c("Genera", "Total (n)", "Missing (n)")
+colnames(age_tables)[c(1,7,8, 9)] <- c("Genera", "150+", "Total (n)", "Missing (n)")
 
@@
 
+caption_trees <- "Binned age-distribution for street and park trees. Missing values feature no"
+caption_trees_short <- "Binned age-distribution for street and park trees."
+
 if (grepl("docx", out_format) ||
@@
   knitr::kable(booktabs = TRUE,
-
-             caption = "Trees",
-             caption.short = "Trees"
+               format = "latex",
+             caption = caption_trees,
+             caption.short = caption_trees_short
              # col.names = c("Subject/Relevance",
@@
                format = "html",
-               caption = "Trees",
-             caption.short = "Trees"
+              caption = caption_trees,
+             caption.short = caption_trees_short
              # col.names = c("Subject/Relevance",
@@
 UHI intensities were extracted for each tree location, and the distribution hence represents the first detailed overview of the exposure of Berlin's trees to urban heat loading.
-The black line represents the density across all three categories.
-Insets represent corresponding tree totals.",fig.width = 12, fig.height = 12, out.width="100%"}
+The black line is the density across all three categories.
+Insets show corresponding tree totals.",fig.width = 12, fig.height = 12, out.width="100%"}
 knitr::include_graphics(here::here("analysis/figures/plot_02_genus_UHI_dens.png"))
@@
 
-The effect of UHI loading on absolute growth potential varies between genera and species (Fig.$~$\@ref(fig:fig-lmestat)).
+The effect of UHI loading on absolute growth potential varies between genera and species (Fig.$~$\@ref(fig:fig-lmestat)); **note, that these results are preliminary and should be considered as template for future outputs, rather used for any inference**.
 Most notably, *Quercus*, the 3rd-most frequent genera, shows decreased absolute growth with increased UHI,
-while the most frequent genera, *Tilia* clearly features enhanced growth.
-Note, that the estimated effect sizes here are linear.
-Temperature may excert a non-linear control on absolute growth, however, and hence, applying a  method able to capture may result in altered presented effect sizes.
+while the most frequent genera, *Tilia* features contrasting relationships with UHI at species level.
+The estimated effect sizes presented here are linear.
+However, temperature may excert a non-linear control on absolute growth and, hence, applying a  method able to capture may result in altered presented effect sizes.
 Additionally, if temperatures increase in the future under climate warming current future increases in under climate warming, any non-linear effects may become more enhanced, stressing the need for a more flexible model fit and structure.
-
-alter the (preliminary) results substantially
 
@@
 
-- incorporating more pertinent covariates as dependent variables in the linear mixed model
+- validating the database with independent observations
+- incorporating more pertinent covariates as dependent variables in the linear mixed model, especially incremental tree growth
 - testing multiple model structures with formal model selection procedures
@@
 
-# Acknowledgements
+<!-- # Acknowledgements -->
 
-<!-- The following line inserts a page break  -->
-`r pgbreak()`
+<!-- <!-- The following line inserts a page break  --> 
+<!-- `r pgbreak()` -->
 

### 2020-03-25 R/functions.R

@@ -665,7 +665,7 @@ apply_models <- function(df = test_df,
                                    .data = test_set)
 
 
-    return(model_out)
+    return(list(model = model_out, test_data = test_set))
 
 
 
@@ -1355,15 +1355,26 @@ make_overview_table <- function(df){
 
     full_data_set_clean <- data.table(df)
     counts_full <- full_data_set_clean[,.N, by = provenance]
-    counts_with_age_size <- full_data_set_clean[!is.na(dbh_cm) & !is.na(STANDALTER),
-                                                .N,
-                                                by = provenance]
+    # counts_with_age_size <- full_data_set_clean[!is.na(dbh_cm) & !is.na(STANDALTER),
+    #                                             .N,
+    #                                             by = provenance]
+
+
+
+    counts_with_age_size <- full_data_set_clean[,full := {!is.na(STANDALTER) &
+            STANDALTER > 0 &
+            !is.na(dbh_cm)}]
+
+
+    counts_with_age_size <- counts_with_age_size[ , .(n_available = sum(full)), by = provenance]
+
+
+
 
     counts_full <- cbind(counts_full, counts_with_age_size[,2])
-    counts_full[3,3] <- NA
     counts_full[,1] <- c("Park", "Street", "Riparian")
-    colnames(counts_full) <- c(" ", "N", "N_{age}")
-    counts_full[order(-N)]
+    colnames(counts_full) <- c("Category", "n", "n$_{full}$")
+    counts_full[order(-n)]
 
     return(counts_full)
 }
@@ -1386,11 +1397,11 @@ make_age_table <- function(df, max_age = 120, break_interval = 20){
 
     # make df with age breaks
     age_df <- data.table(df)[ , .(STANDALTER, gattung_short)][
-        , age_group := cut(STANDALTER, breaks = seq(0, max_age, break_interval), right = FALSE)]
+        , age_group := cut(STANDALTER, breaks = c(seq(0, max_age, break_interval), max(STANDALTER, na.rm = TRUE)), right = TRUE)]
 
     species_totals <- age_df[ ,.(n_total = .N), by = gattung_short]
 
-    cols <- c(1,3:(3+ max_age/break_interval),2)
+    cols <- c(1,3:(4+ max_age/break_interval),2)
 
 
     age_df_class <- dcast(age_df,

### 2020-03-25 R/plan.R

@@ -91,10 +91,11 @@ plan <- drake_plan(
     model_res = berlin.trees::apply_models(df = model_df %>%
                                                dplyr::filter(STANDALTER < 350 &
                                                                  krone_m < 50 &
-                                                                 dbh_cm < 600),
+                                                                 dbh_cm < 300,
+                                                             dbh_cm > 20),
                                            model_list = model_list,
                                            n_top_species = 3,
-                                           min_individuals = 150),
+                                           min_individuals = 1000),
 
 
 
@@ -152,9 +153,9 @@ plan <- drake_plan(
     #                 ),
 
 
-    plot_LME_age = berlin.trees::make_ranef_plot(model_out = model_res,
+    plot_LME_age = berlin.trees::make_ranef_plot(model_out = model_res$model,
                                                  model_name = "heat_age_RIspecies_RSspecies_RIprovenance",
-                                                 df = full_data_set_clean,
+                                                 df = model_res$test_data,
                                                  n_top_species = 3,
                                                  base_size = 18,
                                                  file = drake::file_out("./analysis/figures/plot_03_ranef_species_dbh_uhi.png"),
@@ -189,13 +190,13 @@ plan <- drake_plan(
         output_format = bookdown::html_document2(),
         quiet = TRUE
     ),
-    # paper_word = rmarkdown::render(
-    #     knitr_in("./analysis/paper/paper.Rmd"),
-    #     output_file = file_out(file.path(here::here(), "paper_knit.docx")),
-    #     # output_file = file_out(file.path("./paper_knit.html")),
-    #     output_format = bookdown::word_document2(),
-    #     quiet = TRUE
-    # ),
+    paper_word = rmarkdown::render(
+        knitr_in("./analysis/paper/paper.Rmd"),
+        output_file = file_out(file.path(here::here(), "paper_knit.docx")),
+        # output_file = file_out(file.path("./paper_knit.html")),
+        output_format = bookdown::word_document2(),
+        quiet = TRUE
+    ),
     paper_pdf = rmarkdown::render(
         knitr_in("./analysis/paper/paper.Rmd"),
         output_file = file_out(file.path(here::here(), "paper_knit.pdf")),

### 2020-03-25 analysis/scripts/03_tree_dbh_eda_cleaning.R

@@ -1,0 +1,103 @@
+library(ggplot2)
+library(drake)
+library(data.table)
+loadd(full_data_set_clean)
+
+full_data_set_clean <- full_data_set_clean %>%
+    mutate(STANDALTER = as.numeric(as.character(STANDALTER)))
+
+min_dbh <- 1
+max_dbh <- 400
+max_age <- 350
+min_age <- 15
+
+filter_expression <- expression(dbh_cm > min_dbh &
+                                    dbh_cm < max_dbh &
+                                    STANDALTER < max_age &
+                                    STANDALTER > min_age)
+
+top_spec <- full_data_set_clean %>%
+    as.data.frame() %>%
+    select(ART_BOT, gattung_short, dbh_cm, STANDALTER) %>%
+    filter(eval(filter_expression)) %>%
+    dplyr::count(ART_BOT) %>%
+
+    arrange(-n) %>%
+    filter(n > 1000) %>%
+    mutate(gattung_short = stringr::str_extract(ART_BOT, "^([\\w]+)"))
+
+
+bezirke <- unique(full_data_set_clean$BEZIRK)
+
+
+purrr::walk(bezirke,
+            function(bz){
+
+                g_plot <- full_data_set_clean %>%
+                    filter(eval(filter_expression),
+                           ART_BOT %in% top_spec$ART_BOT,
+                           BEZIRK == bz) %>%
+                    ggplot(aes(x = STANDALTER,
+                               y = dbh_cm,
+                               color = NAMENR)) +
+                    geom_point(alpha = 0.3) +
+                    geom_smooth(method = "lm") +
+
+                    # ggforce::facet_grid_paginate(~ART_BOT, ncol = 5, nrow = 5)
+                    facet_wrap(~ART_BOT, nrow = 10, ncol = 6, scales = "free")
+
+                # np <- ggforce::n_pages(g_plot)
+
+                ggsave(paste0("./analysis/figures/eda/00_eda_dbh_age_", bz, ".png"),
+                       g_plot,
+                       width = 100,
+                       height = 100,
+                       units = "cm",
+                       dpi = 300)
+
+
+
+            })
+
+g_plot <- full_data_set_clean %>%
+    filter(eval(filter_expression),
+           ART_BOT %in% top_spec$ART_BOT) %>%
+    ggplot(aes(x = STANDALTER,
[…]

### 2020-03-27 analysis/paper/paper.Rmd
cont writing

@@
 
+<style>
+body {
+text-align: justify}
+</style>
+
 <!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
@@
 
-# Introduction
+# Introduction {#sec:intro}
 
@@
 
-To complement (existing) detailed dendroecological analyses of climate-growth relationships in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individuals provided by the Berlin Senate Administration (Senatsverwaltung).
-This data set contains information on location, species, trunk diameter (at breast height; $DBH$; see Tab.$~$\@ref(tab:tab-tree-overview)), and height, amongst other variables for street and park trees.
+To complement (existing) detailed dendroecological analyses of climate-growth relationships in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individuals covering 94 genera and at least 600 species and/or cultivars provided by the Berlin Senate Administration (Senatsverwaltung).
+This data set contains information on location, species, trunk diameter (at breast height; $DBH$; see Tab.$~$\@ref(tab:tab-tree-overview)), and height, amongst other variables for the majority of street and park trees.
 
@@
 
-caption_ov <- "Available records by categoryin entire data set (n), and those with age and $DBH$ entries (n$_{full}$)."
-caption_ov_short <- "Available records in berlin tree database."
+caption_ov <- "Available records by category in entire data set (n), and those with age and $DBH$ entries (n$_{full}$)."
+caption_ov_short <- "Available records in Berlin tree database."
 
@@
 
-In a space-for-time substitution, growth of individual species can be assessed across the entire cite of Berlin, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.
+In a space-for-time substitution, growth of individual species can be assessed across the entire city of Berlin, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.
 Comparable applications are found, for example, in @quigley2004 and @pretzsch2017.
-The former  inferred absolute growth potential for species across successional groups, and between rural and urban species, yet lacked spatially-explicit effect size estimates or predictions of maximum potential;
-@pretzsch2017 applied linear hierarchical models to infer growth modulation for different cities, time periods and urban $vs.$ rural locations while accounting for stand-level variability.
+The former  inferred absolute growth potential for species across successional groups (early, mid, late stage), and between rural and urban conspecifics, yet lacked spatially-explicit effect size estimates across the urban-rural space and was limited to comparatively small sample sizes per group ($n_{total}~=~230$ divided in 15 species, 3 groups and 2 locations).
+@pretzsch2017 applied linear hierarchical models to infer growth modulation on annual basis for different time periods and urban $vs.$ rural locations while accounting for stand-level variability;
+however, for Berlin only 145 individuals of one species (*T. cordata*) were assessed.
+As mentioned previously, climate-growth relationships can vary substantially between species, and in fact, @quigley2004 and @pretzsch2017 report contrasting results regarding average tree diameter, i.e.
+smaller or larger for urban $vs.$ rural trees of same age.
+Consequently, this variability of effect sizes and directions calls for a more comprehensive assessment across species and with greater spatial coverage throughout Berlin.
 
-**By contrast, we propose applying a statistical model that is fully spatially explicit, while also allowing to account for the nested nature of the data set (e.g.
-streets and districts) as well as other pertinent factors using hierarchical, generalized additive models (see Section$~$\@ref(sec:methods).
-This also allows to infer the absolute growth potential of a species given, for example, a specific location, age or UHI magnitude.**
-Tree-level growth data, however, is paramount in validating such relationships, and its inclusion in the model would also allow incorporating effects of varying climate over time.
+**We therefore propose applying a statistical model that is fully spatially explicit, while also allowing to account for the nested nature of the data set (e.g.
+streets and districts) as well as other pertinent factors using hierarchical, generalized additive models (see Section$~$\@ref(sec:methods)).
+As a result, the absolute growth potential of a species can be inferred given, for example, a specific location, age or UHI magnitude. 
+Further, the impact of UHI loading can be predicted for a single species across all of Berlin as a continuous surface.**
+The inclusion of independent, tree-level growth data, however, is paramount as it allows:
+
+1) validating the publicly-available Senate dataset ($DBH$) while providing a more reliable estimate of tree age,  
+2) applying the model with annual basal area as a response (enabling incorporating effects of varying climate/UHI intensities over time), and  
+3) comparing $DBH$ (Senate data) and basal area-derived models for a subset of locations and species, increasing confidence in Berlin-wide predictions.
 
@@
 
-We caution, that 1) the presented model results are preliminary, 2) initial diagnostics indicate that modification of the current model structure (e.g.
+<!-- We caution, that 1) the presented model results are preliminary, 2) initial diagnostics indicate that modification of the current model structure (e.g.
 by accounting for individual-level variability) may be required and 3) that further, time-intensive data cleaning is necessary.
-First tests, however, showed that while the magnitude of effects changes with different model structures, their direction appears stable.
+-->
+<!-- First tests, however, showed that while the magnitude of effects changes with different model structures, their direction appears stable. -->
 
@@
 
-Additional remote sensing data for local context, higher resolution UHI data, and most important, temporally-resolved tree growth data are still required and/or would greatly improve the impact and confidence in effect size estimates.
-See Tab.$~$\@ref(tab:tab-data-req) for details.
+Additional remote sensing data for local context, higher resolution UHI data, and most important, temporally-resolved tree growth data are still required and/or would greatly improve the confidence in effect size estimates (see Section$~$\@ref(sec:intro)).
+See Tab.$~$\@ref(tab:tab-data-req) for details on data sets.
 
@@
 
-caption_trees <- "Binned age-distribution for street and park trees. Missing values feature no"
-caption_trees_short <- "Binned age-distribution for street and park trees."
+caption_trees <- "Binned age-distribution for trees in Berlin dataset, and entries missing age information."
+caption_trees_short <- "Binned age-distribution for trees in Berlin dataset."
 
@@
 
-The effect of UHI loading on absolute growth potential varies between genera and species (Fig.$~$\@ref(fig:fig-lmestat)); **note, that these results are preliminary and should be considered as template for future outputs, rather used for any inference**.
-Most notably, *Quercus*, the 3rd-most frequent genera, shows decreased absolute growth with increased UHI,
-while the most frequent genera, *Tilia* features contrasting relationships with UHI at species level.
+**Note, that results below are preliminary and should be considered as a template for future outputs, rather than used for inference.**
+The effect of UHI loading on absolute growth potential varies between genera and species (Fig.$~$\@ref(fig:fig-lmestat)).
+Most notably, *Quercus*, the 3rd-most frequent genera, shows decreased absolute growth with increasing UHI loading,
+while the most frequent genera, *Tilia*, features contrasting relationships between species.
 The estimated effect sizes presented here are linear.
-However, temperature may excert a non-linear control on absolute growth and, hence, applying a  method able to capture may result in altered presented effect sizes.
-Additionally, if temperatures increase in the future under climate warming current future increases in under climate warming, any non-linear effects may become more enhanced, stressing the need for a more flexible model fit and structure.
+However, temperature may excert a non-linear control on absolute growth and, hence, applying a  method able to capture such dynamics may result in somewhat different effect sizes / behavior.
+Additionally, if temperatures increase in the future under climate warming, any non-linear effects may become more enhanced, stressing the need for a more flexible model fit and structure (i.e.
+using GAMM over linear models-).
 
@@
 - validating the database with independent observations
-- incorporating more pertinent covariates as dependent variables in the linear mixed model, especially incremental tree growth
+- incorporating more pertinent covariates as dependent variables in the linear mixed model  
 - testing multiple model structures with formal model selection procedures
@@
   * assess potential age-dependent UHI impacts on individual species.  
+  * repeating the above (GAMM) with total and incremental basal area as responses
 

### 2020-03-31 analysis/paper/paper.Rmd
updated paper

@@
 where $Y$ is taken from an appropriate distribution and corresponding link function $g$, $\beta_0$ is the intercept and $f_i$ represents a smooth function of a predictor [@pedersen2019], and $\epsilon \sim \mathcal{N}(0, \sigma ^2)$.
-Note, that $f_i$ consists of a smooth (e.g. spline) constructed via basis functions of different form and complexity, multiplied by a coeffecient:
+Note, that $f_i$ consists of a smooth (e.g. spline) constructed via basis functions of different form and complexity, multiplied by a coefficient:
 
@@
 northings and eastings, as for example done in [@augustin2009].
-Ultimately, the implementation of a such a GAMM will alow for establishing continuous prediction surfaces of growth potential (approximated via $DBH$) for individual species across urban areas (including parks) of Berlin.
+Ultimately, the implementation of a such a GAMM will allow for establishing continuous prediction surfaces of growth potential (approximated via $DBH$) for individual species across urban areas (including parks) of Berlin.
 
-Currently, $DBH$ has been modelled using a hierarchical linear model (linear mixed effects model) with `lme4` [@bates2015] in @rcoreteam2020 (see Section$~$\@ref(sec:results)).
+Currently, $DBH$ has been modeled using a hierarchical linear model (linear mixed effects model) with `lme4` [@bates2015] in @rcoreteam2020 (see Section$~$\@ref(sec:results)).
 The general form of this model is:
@@
 The model for which results are presented in Figure$~$\@ref(fig:fig-lmestat) estimates $DBH$ from tree age and the local UHI intensity as continuous covariates with random slopes and intercepts for each species;
-note, that for computational efficiency each genera was modelled seperately.
+note, that for computational efficiency each genera was modeled separately.
 Further, models were established for the three most abundant species per genera with at least 1000 individuals.
@@
 
-caption_trees <- "Binned age-distribution for trees in Berlin dataset, and entries missing age information."
-caption_trees_short <- "Binned age-distribution for trees in Berlin dataset."
+caption_trees <- "Binned age-distribution for trees in Berlin data set, and entries missing age information."
+caption_trees_short <- "Binned age-distribution for trees in Berlin data set."
 
@@
 The estimated effect sizes presented here are linear.
-However, temperature may excert a non-linear control on absolute growth and, hence, applying a  method able to capture such dynamics may result in somewhat different effect sizes / behavior.
+However, temperature may exert a non-linear control on absolute growth and, hence, applying a  method able to capture such dynamics may result in somewhat different effect sizes / behavior.
 Additionally, if temperatures increase in the future under climate warming, any non-linear effects may become more enhanced, stressing the need for a more flexible model fit and structure (i.e.
@@
 - testing multiple model structures with formal model selection procedures
-- checking model resiudals for spatial auto-correlation and accounting for it where necessary to ensure unbiased estimates of effect sizes
+- checking model residuals for spatial auto-correlation and accounting for it where necessary to ensure unbiased estimates of effect sizes
 - repeating the above with a hierarchical GAM (i.e. GAMM) to allow for:  

### 2020-04-20 README.Rmd
updated readme

@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)
+## Purpose
 
-This repository contains the data and code for our paper:
+This repository contains an analysis of tree growth (potential) and its relationship with the Urban Heat Island (UHI) effect in Berlin using an extensive, publicly available data set. 
+For some details of the (on-going) research, download this [file](./analysis/paper/paper.html)
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
+## Approach
 
-Our pre-print is online here:
+The folder structure follows the principal of a [research compendium](https://research-compendium.science/) following concepts developed by [Ben Marwick and others](https://www.tandfonline.com/doi/abs/10.1080/00031305.2017.1375986?journalCode=utas20).
+That means all analyses and the final write-up are organized in a (fairly) coherent manner, encouraing reproducibility.
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+I chose to package up the analyses in an `R` package to facilitate reproducibility, as dependencies are more easily tracked, and analyses steps can be broken down into functions that can be tested, adjusted when necessary, and most important, re-used.
 
-### How to cite
+<!-- [![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio) -->
 
-Please cite this compendium as:
+<!-- This repository contains the data and code for our paper: -->
 
-> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+<!-- > Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx> -->
 
-### How to download or install
+<!-- Our pre-print is online here: -->
 
-You can download the compendium as a zip from from this URL: </archive/master.zip>
+<!-- > Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx> -->
 
-Or you can install this compendium as an R package, berlin.trees, from GitHub with:
+<!-- ### How to cite -->
 
-```{r gh-installation, eval = FALSE}
-# install.packages("devtools")
-remotes::install_github("/")
-```
+<!-- Please cite this compendium as: -->
 
-### Licenses
+<!-- > Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx> -->
 
-**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)
+<!-- ### How to download or install -->
 
-**Code :** See the [DESCRIPTION](DESCRIPTION) file
+<!-- You can download the compendium as a zip from from this URL: </archive/master.zip> -->
 
-**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse
+<!-- Or you can install this compendium as an R package, berlin.trees, from GitHub with: -->
 
-### Contributions
+<!-- ```{r gh-installation, eval = FALSE} -->
+<!-- # install.packages("devtools") -->
+<!-- remotes::install_github("/") -->
+<!-- ``` -->
 
-We welcome contributions from everyone.
+<!-- ### Licenses -->
+
+<!-- **Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/) -->
+
+<!-- **Code :** See the [DESCRIPTION](DESCRIPTION) file -->
+
+<!-- **Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse -->
+
+<!-- ### Contributions -->
+
+<!-- We welcome contributions from everyone.
 Before you get started, please see our [contributor guidelines](CONTRIBUTING.md).
@@
 By participating in this project you agree to abide by its terms.
+-->

### 2020-04-20 README.Rmd
code review setup

@@
 
+ <img src="analysis/figures/map_03_uhi.png" align="right" height = 275px, width = 275px />
+
 ## Purpose
@@
 This repository contains an analysis of tree growth (potential) and its relationship with the Urban Heat Island (UHI) effect in Berlin using an extensive, publicly available data set. 
-For some details of the (on-going) research, download this [file](./analysis/paper/paper.html)
+For some details of the (on-going) research, download this [file](./analysis/paper/paper.html) and have a wee read if you like.
+Having a look at this first (mainly figures) will help contextualize some of the workflow/code/results, I think!
 
@@
 
-The folder structure follows the principal of a [research compendium](https://research-compendium.science/) following concepts developed by [Ben Marwick and others](https://www.tandfonline.com/doi/abs/10.1080/00031305.2017.1375986?journalCode=utas20).
+The folder structure follows the principal of a [**research compendium**](https://research-compendium.science/) following concepts developed by [Ben Marwick and others](https://www.tandfonline.com/doi/abs/10.1080/00031305.2017.1375986?journalCode=utas20).
 That means all analyses and the final write-up are organized in a (fairly) coherent manner, encouraing reproducibility.
@@
 I chose to package up the analyses in an `R` package to facilitate reproducibility, as dependencies are more easily tracked, and analyses steps can be broken down into functions that can be tested, adjusted when necessary, and most important, re-used.
+A brief overview and benefits of that approach are detailed in my [blog post here](https://the-hull.github.io/reproducibility_compendium/analyses/00_write-up.html).
+
+A workflow-management software [**`drake`**](https://github.com/ropensci/drake) and approach is also something I've relied on heavily here, as the analysis is quickly growing.
+While it takes time to get used to, and to implement, it has already paid off massively when I had to go back and change parameters/inputs.
+`drake` is extensively documented ([see book](https://books.ropensci.org/drake/)), and I highly recommend it.
+
+In brief:
+
+- Research compendium for coherent structure (based on `R` package; recognizable by others)
+- Packaged analyses steps/functions into `R package` (`berlin.trees`; call as `berlin.trees::my_function()` in scripts)
+- employed `drake` for managing analyses workflow
+
+Below is a brief explanation of the repository structure, and important files, as well as what drake does and how it relates to the analysis.
+
+### Repository structure
+
+The repo structure follows that of a typical `R` package.
+There are:
+
+- high-level files (this readme, `NAMESPACE`, `DESCRIPTION`) relating to typical `R` package development
+- high-level files relating to `drake` and its implementation (`make.R`, `_drake.R`)
+- high-level files that I use to get going with a package/compendium (`.Rprofile`, `_setup.R`)
+
+- `R` folder for functions
+- `man` folder for the documentation (i.e. help files) of the functions that pop-up in the Help viewer in `RStudio` (think of future you!)
+- `analysis` that containts figures and the manuscript as a Rmarkdown / latex document
+
+- output papers from the "raw" (uncompiled) manuscript (`paper_knit.xyz`)
+
+### Using `drake`
+
+`drake` uses an analyses plan (a set of instructions) to generate results. 
+The great thing about `drake` is that it knows which instructions depends on another (i.e. download data before plotting it, than put it in a manuscript).
+The set of instructions are located in `R/plan.R`, and outline all the steps that I take to generate results, figures and rendering the final manuscript.
+
+The files `make.R` and `_drake.R` both achieve the same thing, but in different contexts (typically difference is running the pipeline on a local `R` install vs. run on a server). 
+If `plan.R` is the brain, the above files are the muscles making things move. 
+Note, that they are generated automatically by `drake`, so you don't need to worry that much about what to put in them (`drake::use_drake()`).
+
+The file `_setup.R` is typially what I work in to start the pipeline, or add libraries to my analysis, and it contains the function `drake::r_make()` that starts of the pipeline!
+
+#### The `plan.R`
+
+`plan.R` is structured following the entire analyses from downloading tabular and spatial data, cleaning it, producing stats and plots, and then compiling the manuscript.
+I typically seperate indivudal steps with `# Step -----------` as that gives you a neat outline in `RStudio` (if click the lines symbol above the scroll-bar in the script window).
+Every analyses step is named, and that name can then be used in a sub-sequent step (this is how drake manages dependencies).
+For example, see this taken directly from the documentation for a fairly self-explanatory application:
+
+![](https://camo.githubusercontent.com/e9cd15743788ea28d336848828f458aad86f1ad9/68747470733a2f2f646f63732e726f70656e7363692e6f72672f6472616b652f7265666572656e63652f666967757265732f67726170682e706e67)
+
+`drake` caches (intermediate) results and keeps track of any changes to them, and tells you which parts steps will be re-executed next time to finish of the pipeline! 
+Super cool. 
+Any intermediate result can, for example, be inspected using `drake::readd(X)` or `drake::loadd()`, and this is also how you include results in e.g. an `Rmarkdown` report.
+
+### Workflow
+
+Note, that I've added some descriptions to the steps in `plan.R`, but typically I rely on the function help that I generate in `R/functions.R`. 
+The workflow I apply there is typically iterative, so I play around in a few dump/script files (`analyses/scripts/...`) and then refactor the code for a specific use into a function.
+As I've bundled the analysis into a package, any (properly formatted) documentation in the functions is then accessible using `berlin.trees::my_function()`, and should I forget what a function does, I put my cursor on the function name and hit `F1` - the help then appears in the `RStudio` viewer.
+I try to use verbs and wordy function names so I hope that most things would be self-explanatory (at least for myself).
+This also works for documenting data sets (e.g. try `?iris` in your `R` console) - and can be super helpful!
+
+Once functions are coded and the plan configured, I typically navigate to the file `_setup.R` (either using the shortcud `CTRL+.` or clicking) to run `drake::r_make()` and a few other things, or I just execute the latter command directly in the console.
+`drake` then let's you know what's up!
+
+### Code
+
+As mentioned before, I typically break down analyses into many steps, and use wordy names to indicate what they do. 
+I haven't implemented formal unit testing here, but I'm quite thorough with checking outputs (especially in exploratory analyses), and doing sanity checks on outputs.
+With the break-down approach, things tend to stay fairly straightforward for every step, and I only comment code where entirely necessary (for myself so far).
+I do try to keep the functions organized (using `# Section ----------` - ), so that an outline appears in `RStudio`:
+
+![](./functions.png)
+
+That really helps navigating!
+
+### Review / Feedback
+
+As Tom said, don't spend too much time going through individual things!
+I think having a look at the structure, and whether it's "straightforward" to follow along would be most useful, especially given that there are two potentially new concepts here (`R package` development and `drake`)
+Thanks for having a go!!
 

### 2020-04-20 R/plan.R

@@ -23,6 +23,9 @@ plan <- drake_plan(
 
 
 
+    # loads the raster data from file (not uploaded to github due to size,
+    # but see source in manuscript)
+
     uhi_stacks = berlin.trees::get_uhi_rasters(file.path(here::here(), "analysis", "data", "raw_data", "UHI_explorer")),
 
 
@@ -31,21 +34,26 @@ plan <- drake_plan(
 
 
 
-    ## Berlin trees
+    ## Berlin trees (download Senate tree data set from WFS using API)
     download_data = target(berlin.trees::download_berlin_trees()),
 
     ### Cleaning
+
+    # Load spatial-features data sets of all Berlin trees
     tree_data_in_lists = berlin.trees::load_downloaded_data_to_lists(download_data),
 
 
     cropped_data_set = berlin.trees::crop_data_with_bbox(tree_data_in_lists,
                                                          bounding_box),
+
+    # Bind rows of sf tibbles to single data set (currently in lists)
     full_data_set = berlin.trees::bind_rows_sf(cropped_data_set),
 
+    # Clean Feature Meta Data
     full_data_set_clean = target(berlin.trees::clean_data(full_data_set),
                                  format = "rds"),
 
-
+    # Add UHI data from RasterLayer stack to sf data frame
     extract_uhi_values_to_list = berlin.trees::add_uhi_hist_data(uhi_stack_list = uhi_stacks,
                                                                  sf_data = full_data_set_clean[, ]),
 
@@ -55,7 +63,7 @@ plan <- drake_plan(
 
 
 
-
+    # This is a list of models I apply simulatenously for later evaluation
     model_list = list(
         # null = function(x) lme4::lmer(dbh_cm ~ 1, data = x),
         # heat_only = function(x) lme4::lmer(dbh_cm ~ day_2007, data = x)  ,
@@ -81,13 +89,17 @@ plan <- drake_plan(
     ),
 
 
+    # Generate data set for stat-models
+    # This step just adds the UHI temperature data to the data set
+    # and adjusts a bit
     model_df = make_test_data_set(full_df = full_data_set_clean,
                                   extract_uhi = extract_uhi_values_to_list),
 
 
 
 
-
+   # here I apply all lme4 models not commented-out above,
+   # and also specify the data which I want to exclude
     model_res = berlin.trees::apply_models(df = model_df %>%
                                                dplyr::filter(STANDALTER < 350 &
[…]

### 2020-10-22 analysis/paper/paper.Rmd
stopped using package format explored LCZ from WUDAPT
re-ran analyses

@@
   table_dat_req <- dat_req %>% 
-  select(-owned) %>% 
+  dplyr::select(-owned) %>% 
   knitr::kable(booktabs = TRUE,

### 2020-10-22 R/functions.R

@@ -376,7 +376,7 @@ add_full_df_idx_to_baumscheiben <- function(bms,
 #'
 #' @param fulldf sf_df, all berlin trees
 #' @param bms sf_df, Baumscheiben polygons and their area
-#' @param max_dist numeric, how far from polygon centroid can tree lie and still be assigned?
+#' @param max_dist_m numeric, how far from polygon centroid can tree lie and still be assigned?
 #'
 #' @return fulldf with additional columns: "baumsch_dist_m",
 #'  "baumsch_elem_nr",
@@ -387,7 +387,7 @@ add_full_df_idx_to_baumscheiben <- function(bms,
 #' @import sf
 #' units
 #'
-add_baumscheiben_flaeche <- function(fulldf, bms, max_dist = 10){
+add_baumscheiben_flaeche <- function(fulldf, bms, max_dist_m = 10){
 
     nearest_idx <- sf::st_nearest_feature(fulldf, bms)
 
@@ -401,11 +401,11 @@ add_baumscheiben_flaeche <- function(fulldf, bms, max_dist = 10){
     fulldf <- cbind(fulldf,
                     bms)
 
-    fulldf$baumsch_flaeche_m2 <- ifelse(fulldf$baumsch_dist <= units::set_units(max_dist, m),
+    fulldf$baumsch_flaeche_m2 <- ifelse(fulldf$baumsch_dist <= units::set_units(max_dist_m, m),
                                         fulldf$baumsch_flaeche_m2,
                                         NA)
 
-    return(full_df)
+    return(fulldf)
 
 }
 
@@ -421,7 +421,12 @@ add_baumscheiben_flaeche <- function(fulldf, bms, max_dist = 10){
 #'
 bind_rows_sf <- function(sf_list){
     # bind data without geometry
-    sf_df  <- lapply(sf_list, function(x) sf::st_set_geometry(x, NULL)) %>%
+    sf_df  <- lapply(sf_list, function(x){
+
+        sf::st_set_geometry(x, NULL) %>%
+           setNames(toupper(colnames(x)))
+
+    }) %>%
         dplyr::bind_rows()
 
     # make data frame and set geometry based on original sfcs
@@ -444,6 +449,12 @@ bind_rows_sf <- function(sf_list){
 #' @export
 #'
 clean_data <- function(sf_data){
+
+    # colnames(sf_data)[which(colnames(sf_data) != "geometry")] <- toupper(colnames(sf_data)[which(colnames(sf_data) != "geometry")])
+
[…]

### 2020-10-22 R/packages.R

@@ -1,0 +1,27 @@
+library(bookdown)
+library(here)
+library(ckanr)
+library(ggplot2)
+library(dplyr)
+library(xml2)
+library(httr)
+library(purrr)
+library(stringr)
+library(sf)
+library(extrafont)
+library(future.callr)
+library(drake)
+library(furrr)
+library(tools)
+library(forcats)
+library(tibble)
+library(rmarkdown)
+library(magrittr)
+library(raster)
+library(lme4)
+library(rlang)
+library(stats)
+library(stars)
+library(scales)
+library(rgdal)
+library(data.table)

### 2020-10-22 R/plan.R

@@ -1,9 +1,10 @@
 library(drake)
-library(future.callr)
+# library(future.callr)
 library(dplyr)
 library(future)
-# future::plan(future.callr::callr)
-options(future.globals.maxSize = 600 * 1024 ^ 2)
+future::plan(future.callr::callr)
+# future::plan(future::multiprocess)
+# options(future.globals.maxSize = 600 * 1024 ^ 2)
 
 
 plan <- drake_plan(
@@ -15,9 +16,9 @@ plan <- drake_plan(
     ## Spatial Ancillary -----------------------------------
 
     ### Berlin districts and BBoxx
-    berlin_polygons = target(berlin.trees::get_berlin_polygons_as_sf()),
+    berlin_polygons = target(get_berlin_polygons_as_sf()),
 
-    bounding_box = berlin.trees::make_bbox(52.083962, 52.847599,
+    bounding_box = make_bbox(52.083962, 52.847599,
                                            12.712024, 14.238359,
                                            "greater_berlin",
                                            crs = 4326),
@@ -29,7 +30,7 @@ plan <- drake_plan(
     # loads the raster data from file (not uploaded to github due to size,
     # but see source in manuscript)
 
-    uhi_stacks = berlin.trees::get_uhi_rasters(file.path(here::here(), "analysis", "data", "raw_data", "UHI_explorer")),
+    uhi_stacks = get_uhi_rasters(file.path(here::here(), "analysis", "data", "raw_data", "UHI_explorer")),
 
 
 
@@ -38,41 +39,41 @@ plan <- drake_plan(
 
 
     ## Berlin trees (download Senate tree data set from WFS using API)
-    download_data = target(berlin.trees::download_berlin_trees()),
+    download_data = target(download_berlin_trees()),
 
     # Berlin Baumscheiben
-    download_data_baumscheiben = berlin.trees::download_berlin_baumscheiben(),
+    download_data_baumscheiben = download_berlin_baumscheiben(),
 
     ### Cleaning
 
     # Load spatial-features data sets of all Berlin trees
-    tree_data_in_lists = berlin.trees::load_downloaded_data_to_lists(download_data),
+    tree_data_in_lists = load_downloaded_data_to_lists(download_data),
 
 
     # Load spatial-features data sets of all Berlin baumscheiben
-    baumscheiben_in_lists = berlin.trees::load_downloaded_data_to_lists(download_data_baumscheiben),
[…]

### 2020-10-22 _drake.R

@@ -7,8 +7,8 @@
 # If you use supporting scripts like the ones below,
 # you will need to supply them yourself. Examples:
 # https://github.com/wlandau/drake-examples/tree/master/main/R
-# source("R/packages.R")  # Load your packages, e.g. library(drake).
-# source("R/functions.R") # Define your custom code as a bunch of functions.
+source("R/packages.R")  # Load your packages, e.g. library(drake).
+source("R/functions.R") # Define your custom code as a bunch of functions.
 # devtools::install(upgrade = FALSE)
 # drake::expose_imports("bookdown",
 #                       character_only = TRUE)
@@ -22,4 +22,5 @@ source("R/plan.R")      # Create your drake plan.
 # The arguments to drake_config() are basically the same as those to make().
 # drake::expose_imports("berlin.trees",
                       # character_only = TRUE)
-drake_config(plan, envir = getNamespace("berlin.trees"))
+# drake_config(plan, envir = getNamespace("berlin.trees"))
+drake_config(plan)

### 2020-10-22 analysis/scripts/07_LCZ_exploration.R

@@ -1,0 +1,71 @@
+loadd(bounding_box)
+loadd(berlin_polygons)
+loadd(full_data_set_clean)
+loadd(baumscheiben_in_lists)
+
+
+tifpath <- "./analysis/data/raw_data/WUDAPT_LCZ.geotiff"
+
+rgdal::GDALinfo(tifpath)
+
+
+lcz_europe <- raster::raster(tifpath)
+
+raster::extent(lcz_europe)
+lcz_proj <- raster::crs(lcz_europe)
+
+berlin_polygons <- sf::st_transform(berlin_polygons,
+                                    crs = lcz_proj)
+berlin_polygons %>% raster::extent()
+
+gdaluti::gdalbuildvrt(gdalfile = "yourraster.tif",
+             output.vrt = "tmp.vrt",
+             te = st_bbox(shp_layerbuffer1))
+
+tempraster <- raster("tmp.vrt")
+
+
+
+crop_raster <- raster::crop(lcz_europe,
+                            sf::st_union(berlin_polygons) %>% sf::st_centroid() %>%
+                                sf::st_buffer(35000)  %>% sf::st_as_sf())
+
+
+lcz_stars <- stars::st_as_stars(crop_raster)
+
+colourCount = length(unique(raster::values(crop_raster)))
+getPalette = colorRampPalette(RColorBrewer::brewer.pal(9, "Set1"))
+
+plot(lcz_stars, col = viridisLite::cividis(14))
+plot(lcz_stars, col =getPalette(colourCount))
+
+ggplot() +
+    stars::geom_stars(data = lcz_stars,
+                      aes(x = x,
+                          y = y,
+                          fill = as.factor(WUDAPT_LCZ))) +
+    geom_sf(data = berlin_polygons %>% dplyr::filter(NAMGEM == "Pankow"),
+            color = "black",
+            fill = NA,
+            size = 2)
+
+
+lcz_values_by_district <- raster::extract(crop_raster,
+                                          sf::st_make_valid(berlin_polygons)) %>%
+lapply(., function(x)prop.table(table(x))) %>%
+    setNames(berlin_polygons$NAMGEM) %>% dplyr::bind_rows(.id = "bezirk")
+
+lcz_values_by_district %>%
+    tidyr::pivot_longer(cols = tidyselect::matches("[0-9]"),
+                        names_to = "LCZ",
+                        values_to = "proportion_by_district") %>%
[…]

### 2020-10-22 make.R

@@ -19,4 +19,4 @@ source("R/plan.R")
 # Call make() to run your work.
 # Your targets will be stored in a hidden .drake/ cache,
 # and you can read them back into memory with loadd() and readd().
-make(plan, envir = getNamespace("berlin.trees"))
+make(plan)

### 2021-05-18 analysis/paper/paper.Rmd
added land use map and study area description

@@
 date: "`r format(Sys.time(), '%d %B, %Y')`"
+always_allow_html: true
 output: 
+    bookdown::html_document2:
+      fig_caption: yes
+      toc: true
     bookdown::word_document2:
@@
       - --lua-filter=../templates/pagebreak.lua
-    bookdown::html_document2:
-      fig_caption: yes
-      toc: true
     bookdown::pdf_document2:
@@
 - \usepackage[normalem]{ulem}
-bibliography: uhi_references.bib
+bibliography: 
+  - uhi_references.bib
+  - 04_gfz_uhi.bib
 csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
@@
 ```{r tab-tree-overview}
-loadd(overview_table)
+drake::loadd(overview_table)
 
@@
 
-# Proposed methods and data requirements {#sec:methods}
+# Methods {#sec:methods}
+
+## Study area
+
+Berlin is one of the largest metropolitan areas in Central Europe (892$~km^2$) with a population of approximately 3.6 million, and a maximum extent of 38$~km$ in North-South and 45$~km$ in East-West directions.
+It is located in North-Eastern Germany, and lies in the temperate zone with warm-humid climate (Dfb) according to the updated Köppen-Geiger classification [@beck2018], with mean annual temperature of approximately 10$^\circ C$ and precipitation of 575$~mm$ (Tempelhof weather station, DWD).
+Berlin features low relief (approximately 30$~m$ to 60$~m$ with 120$~m$ at solitary peaks), and is centered around a glacial outwash valley (sands, gravel), bordered by two plateaus consisting of glacial till and clay in the North-East and South, as well as sands in the South-West.
+The city provides extensive public green space covering around 30$~\%$ of its area [@suvkberlin2019], with an extensive urban forest of nearly 700000 publically-managed trees along streets, in parks and in riparian areas.
+It has a pronounced urban heat island, owing to its high degree of built-up and sealed surfaces, with air temperatures up to 12$^\circ C$ higher in urban compared to adjacent rural areas [@fenner2014].
+
+## Data sources
+
+An overview of all data used in this study, including sources, types, and application, is provided in Table X.
+Most notably, characteristics of street trees and their respective environments were collated from the city's tree inventory, as well as various elements of the Berlin Environment Atlas, which are openly available via daten.berlin.de and curated by the city's government.
+The data used from the inventory includes species, location, age, and diameter at breast height (dbh).
+The environment atlas holds information on several variables relevant to tree growth, such as soil type and nutrient exchange capacity, distance to groundwater, surface area of tree planting beds ("Baumscheibe"), as well as general characteristics of the urban environment like building height.
+It also includes spatially-explicit model outputs for local climate simulations representative of typical conditions, such as air temperature (2 m) on a 2015 summer's day at different hours (0400, 1400, 2200).
+These outputs are available at city block or planning unit level, and were used as a proxy for the general temperature distribution influenced by urbanization across the investigated growth period across the previous 80 years (see section X for details).
+This was deemed reasonable as Berlin's built-up area has not changed markedly over the past 50 years (i.e., from about 52 to 61 %) [@mohamed2017]. 
+The utility of using this high-resolution data set of air temperatures over a derived urban heat island product was assessed by… 
+Further insight into the role of the urban environment was assessed by including local climate zone classes derived by WUDAPT.
+These describe the proportion of urban area with a specific cover, such as dense high rises, sealed surfaces, etc. 
+For this study, the LCZ 2, representing dense mid-rise built up cover was used as a proxy for the degree of urbanization, and is the most frequent LCZ for Berlin [@fenner2017] (found ref in .
+Use percent sealed surface instead?
+
+```{r dsources-prep, include=FALSE}
+
+options(knitr.kable.NA = "")
+
+dsources_path <- "analysis/data/tables/data_sources.xlsx"
+dsources <- readxl::read_excel(dsources_path, na = "NA")
+dsources$Accessed <- format(dsources$Accessed, "%b '%y")
+
+```
+
+```{r dsources-tab}
+
+caption <- "Data description used for generating maps and route optimization.
+Resolution and radius are provided in $m$, the latter referring to the buffer size across which data was averaged to estimate trees' environmental conditions.
+A zero-radius refers to a point extraction from categorical and location specific data.
+**Planting Bed Area, and Soil type are not currently used.**"
+
+dsources <- dsources %>%
+  dplyr::select(Name = `Name-eng`, Accessed, Type, Unit = `Unit-latex`, Resolution, Radius, Source = `Source-short`, Reference = TextRef) %>%
+  dplyr::filter(Name != "UHI Berlin",
+                Name != "Park Trees",
+                Name != "Riparian Trees")
+
+if (grepl("docx", out_format) ||
+    knitr::is_latex_output() ) {
+  table_ds <- dsources %>% 
+  knitr::kable(booktabs = TRUE,
+
+             caption = caption,
+             caption.short = caption,
+             format = "latex",
+             # col.names = c("Subject/Relevance",
+             #              "Desc.",
+             #              "Type",
+             #              "Obsv. (n)",
+             #              "Resolution (m)",
+             #              "Source")
+             escape = FALSE)
+
+} else {
+table_ds <- dsources %>% 
+  knitr::kable(booktabs = TRUE,
+               format = "html",
+
+             caption = caption,
+             caption.short = caption_ov_short,
+             # col.names = c("Subject/Relevance",
+             #              "Desc.",
+             #              "Type",
+             #              "Obsv. (n)",
+             #              "Resolution (m)",
+             #              "Source")
+             escape = FALSE) 
+}
+
+table_ds %>% 
+  kableExtra::kable_styling(full_width = TRUE, 
+                            font_size = 11,
+                            latex_options = c("hold_position")) %>% 
+  kableExtra::row_spec(0, bold = TRUE, font_size = 12)
+
+# dsources %>%
+#   
+#   knitr::kable(caption = caption, escape = FALSE) %>%
+#   kableExtra::kable_styling(full_width = TRUE,
+#                             bootstrap_options = c("striped", "hover", "condensed", "responsive"))
+
+# flextable::set_flextable_defaults(padding.left = 0)
+# 
+# dsources %>%
+#   dplyr::select(`Name-eng`, Accessed, Type, Source, TextRef) %>%
+#   flextable::flextable() %>%
+#   flextable::autofit()
+
+```
 
@@
 
-loadd(age_tables)
+drake::loadd(age_tables)
 
@@
 
-age_tables <- bind_rows(age_tables[age_tables$Genera!="Other",],
+age_tables <- dplyr::bind_rows(age_tables[age_tables$Genera!="Other",],
                         age_tables[age_tables$Genera=="Other",])
@@
 
-age_tables <- bind_rows(age_tables,
+age_tables <- dplyr::bind_rows(age_tables,
                     add_row)

### 2021-05-18 analysis/paper/paper.tex
added land use map and study area description

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
  \usepackage{amssymb}
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Species-specific effects of the Urban Heat Island on tree growth across Berlin},
  pdfauthor={Alexander G. Hurley\^{}\{a *\}, Ingo Heinrich\^{}a, \ldots{}; \^{}a Geoforschungszentrum Potsdam, Section 4.3, Germany; * Corresponding author: hurley@gfz-potsdam.de},
  pdfkeywords={Uferbaeume; Anlagenbaeume; Strassenbaeume;},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage[normalem]{ulem}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Species-specific effects of the Urban Heat Island on tree growth across Berlin}
\author{Alexander G. Hurley\(^{a *}\), Ingo Heinrich\(^a\), \ldots{} \and \(^a\) \emph{Geoforschungszentrum Potsdam, Section 4.3, Germany} \and * Corresponding author: \href{mailto:hurley@gfz-potsdam.de}{\nolinkurl{hurley@gfz-potsdam.de}}}
\date{18 May, 2021}

\begin{document}
\maketitle
\begin{abstract}
\textbf{This document outlines the rationale for an analysis of tree growth (potential) and its relationship with the Urban Heat Island (UHI) effect in Berlin using an extensive, publicly available data set. It introduces preliminary results and provides an outlook for up-coming and potential work.}
\end{abstract}

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{sec:intro}{%
\section{Introduction}\label{sec:intro}}

Berlin features the most intense Urban Heat Island (UHI) in Germany due to its large extent and development intensity (Kuttler et al., 2015), with temperature increases of up to \(12~K\) during day-time and \(6~K\) on average for night-times (2001-2010, Fenner et al., 2014) in urban \(vs.\) rural areas.
Consequently, urban green (infrastructure) systems are subjected to increased heat more frequently, potentially affecting their process dynamics - either positively or adversely.
Their performance and health, however, is closely tied to local energy budgets (Grimmond et al., 1996 ; Hertel and Schlink, 2019), which in turn are decisive for controlling human wellbeing (e.g. Maras et al., 2016), amongst other factors.
Assessing the effect of increased temperatures on green infrastructure, as part of the urban landscape, is therefore instrumental for understanding, and ultimately mitigating, the potential impact of future warming on increasingly urban societies (Norton et al., 2015).

Trees, in particular, provide shading as well as transpirative cooling in their vicinity (Endlicher et al., 2016; Gillner et al., 2015; Oke, 1982), and therefore can reduce ambient temperatures, infrastructure power-consumption and (human) thermal discomfort (Akbari et al., 2001; e.g. Gulyás et al., 2006; Hoyano, 1988; Mayer and Höppe, 1987);
simultaneously, they provide numerous other environmental, cultural and psychological services and/or benefits (see Tzoulas et al., 2007 for review).
Further, recent tree growth dynamics as a proxy for on-going and future warming may provide an additional line of evidence to support the growing knowledge base on future climate-vegetation dynamics (Zhao et al., 2016) and may aid in mitigation and adaptation efforts (Brune, 2016; Pretzsch et al., 2017).

Trees and green infrastructure in urban areas show a tendency for enhanced growth rates and/or productivity compared to rural counterparts (Jia et al., 2018; Pretzsch et al., 2017), yet feature a broad range of effect size ranges and, in some cases, signs specific to species and locality.
Zhao et al. (2016) showed that growth rates increased within urban clusters as urbanization intensifies using remotely sensed vegetation indices.
Similarly, for Berlin, Dahlhausen et al. (2018), identified positive growth modulation in highly urbanized environments (using growth increments) for \emph{Tilia cordata} Mill, the most abundant tree of the city, which they attributed to the UHI effect, while intermediate development intensity showed indications of being least favorable for tree growth.
Further, Moser-Reischl et al. (2019) identified positive associations between air temperature and radial growth for two species commonly selected by urban planners (\emph{T. cordata}, \emph{Rubinia pseudoacacia}) in Munich.
By contrast, Gillner et al. (2014) highlight decreased growth for \emph{Acer} species (\emph{A. platanoides} and \emph{pseudoplatanus}), \emph{Platanus x hispanica} and \emph{Quercus rubra} with higher summer temperatures of the preceding year, especially when compounded with drought, in another German metropolis (Dresden).
Differences in growth trends may result from contrasting species-specific characteristics, but are indeed affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc. (Pauleit et al., 2002; Quigley, 2004; Randrup et al., 2001; Rhoades and Stipes, 1999).
Under climate change, atmospheric drought will likely be compounded with high temperatures - and intensified UHIs - more frequently, adding further stress to current urban disturbance regimes (Roloff et al., 2009).

Conditions affecting tree growth can vary greatly within urban areas or regions, and need to be accounted for when establishing relationships with pertinent drivers, such as the UHI effect.
This typically complicates the extrapolation from individual sampling sites toward predicting effect sizes across entire urban areas and tree stocks.
This is especially the case for studies reliant on labour-intensive methods which are limited logistically by sampling effort, reducing sample sizes, as well as species and spatial coverage.

To complement (existing) detailed dendroecological analyses of climate-growth relationships in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individuals covering 94 genera and at least 600 species and/or cultivars provided by the Berlin Senate Administration (Senatsverwaltung).
This data set contains information on location, species, trunk diameter (at breast height; \(DBH\); see Tab.\(~\)\ref{tab:tab-tree-overview}), and height, amongst other variables for the majority of street and park trees.

\begin{table}[!h]

\caption[Available records in Berlin tree database.]{\label{tab:tab-tree-overview}Available records by category in entire data set (n), and those with age and $DBH$ entries (n$_{full}$).}
\centering
\fontsize{11}{13}\selectfont
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedleft}X>{\raggedleft}X}
\toprule
\begingroup\fontsize{12}{14}\selectfont \textbf{Category}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{n}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{n$_{full}$}\endgroup\\
\midrule
Park & 257985 & 151527\\
Street & 363905 & 361381\\
Riparian & 46364 & 0\\
\bottomrule
\end{tabu}
\end{table}

In a space-for-time substitution, growth of individual species can be assessed across the entire city of Berlin, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.
Comparable applications are found, for example, in Quigley (2004) and Pretzsch et al. (2017).
The former inferred absolute growth potential for species across successional groups (early, mid, late stage), and between rural and urban conspecifics, yet lacked spatially-explicit effect size estimates across the urban-rural space and was limited to comparatively small sample sizes per group (\(n_{total}~=~230\) divided in 15 species, 3 groups and 2 locations).
Pretzsch et al. (2017) applied linear hierarchical models to infer growth modulation on annual basis for different time periods and urban \(vs.\) rural locations while accounting for stand-level variability;
however, for Berlin only 145 individuals of one species (\emph{T. cordata}) were assessed.
As mentioned previously, climate-growth relationships can vary substantially between species, and in fact, Quigley (2004) and Pretzsch et al. (2017) report contrasting results regarding average tree diameter, i.e.~smaller or larger for urban \(vs.\) rural trees of same age.
Consequently, this variability of effect sizes and directions calls for a more comprehensive assessment across species and with greater spatial coverage throughout Berlin.

\textbf{We therefore propose applying a statistical model that is fully spatially explicit, while also allowing to account for the nested nature of the data set (e.g.~streets and districts) as well as other pertinent factors using hierarchical, generalized additive models (see Section\(~\)\ref{sec:methods}).
As a result, the absolute growth potential of a species can be inferred given, for example, a specific location, age or UHI magnitude.
Further, the impact of UHI loading can be predicted for a single species across all of Berlin as a continuous surface.}
The inclusion of independent, tree-level growth data, however, is paramount as it allows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  validating the publicly-available Senate dataset (\(DBH\)) while providing a more reliable estimate of tree age,\\
\item
  applying the model with annual basal area as a response (enabling incorporating effects of varying climate/UHI intensities over time), and\\
\item
  comparing \(DBH\) (Senate data) and basal area-derived models for a subset of locations and species, increasing confidence in Berlin-wide predictions.
\end{enumerate}

\hypertarget{sec:methods}{%
\section{Methods}\label{sec:methods}}

\hypertarget{study-area}{%
\subsection{Study area}\label{study-area}}

Berlin is one of the largest metropolitan areas in Central Europe (892\(~km^2\)) with a population of approximately 3.6 million, and a maximum extent of 38\(~km\) in North-South and 45\(~km\) in East-West directions.
It is located in North-Eastern Germany, and lies in the temperate zone with warm-humid climate (Dfb) according to the updated Köppen-Geiger classification (Beck et al., 2018), with mean annual temperature of approximately 10\(^\circ C\) and precipitation of 575\(~mm\) (Tempelhof weather station, DWD).
Berlin features low relief (approximately 30\(~m\) to 60\(~m\) with 120\(~m\) at solitary peaks), and is centered around a glacial outwash valley (sands, gravel), bordered by two plateaus consisting of glacial till and clay in the North-East and South, as well as sands in the South-West.
The city provides extensive public green space covering around 30\(~\%\) of its area (SUVK, Berlin, 2019), with an extensive urban forest of nearly 700000 publically-managed trees along streets, in parks and in riparian areas.
It has a pronounced urban heat island, owing to its high degree of built-up and sealed surfaces, with air temperatures up to 12\(^\circ C\) higher in urban compared to adjacent rural areas (Fenner et al., 2014).

\hypertarget{data-sources}{%
\subsection{Data sources}\label{data-sources}}

An overview of all data used in this study, including sources, types, and application, is provided in Table X.
Most notably, characteristics of street trees and their respective environments were collated from the city's tree inventory, as well as various elements of the Berlin Environment Atlas, which are openly available via daten.berlin.de and curated by the city's government.
The data used from the inventory includes species, location, age, and diameter at breast height (dbh).
The environment atlas holds information on several variables relevant to tree growth, such as soil type and nutrient exchange capacity, distance to groundwater, surface area of tree planting beds (``Baumscheibe''), as well as general characteristics of the urban environment like building height.
It also includes spatially-explicit model outputs for local climate simulations representative of typical conditions, such as air temperature (2 m) on a 2015 summer's day at different hours (0400, 1400, 2200).
These outputs are available at city block or planning unit level, and were used as a proxy for the general temperature distribution influenced by urbanization across the investigated growth period across the previous 80 years (see section X for details).
This was deemed reasonable as Berlin's built-up area has not changed markedly over the past 50 years (i.e., from about 52 to 61 \%) (Mohamed, 2017).
The utility of using this high-resolution data set of air temperatures over a derived urban heat island product was assessed by\ldots{}
Further insight into the role of the urban environment was assessed by including local climate zone classes derived by WUDAPT.
These describe the proportion of urban area with a specific cover, such as dense high rises, sealed surfaces, etc.
For this study, the LCZ 2, representing dense mid-rise built up cover was used as a proxy for the degree of urbanization, and is the most frequent LCZ for Berlin (Fenner et al., 2017) (found ref in .
Use percent sealed surface instead?

\begin{table}[!h]

\caption[Data description used for generating maps and route optimization. Resolution and radius are provided in $m$, the latter referring to the buffer size across which data was averaged to estimate trees' environmental conditions. A zero-radius refers to a point extraction from categorical and location specific data. **Planting Bed Area, and Soil type are not currently used.**]{\label{tab:dsources-tab}Data description used for generating maps and route optimization. Resolution and radius are provided in $m$, the latter referring to the buffer size across which data was averaged to estimate trees' environmental conditions. A zero-radius refers to a point extraction from categorical and location specific data. **Planting Bed Area, and Soil type are not currently used.**}
\centering
\fontsize{11}{13}\selectfont
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedleft}X>{\raggedleft}X>{\raggedright}X>{\raggedright}X}
\toprule
\begingroup\fontsize{12}{14}\selectfont \textbf{Name}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{Accessed}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{Type}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{Unit}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{Resolution}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{Radius}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{Source}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{Reference}\endgroup\\
\midrule
Street Trees & Oct '20 & Point &  &  &  & https://daten.berlin.de/ & \\
Berlin Districts & Oct '20 & Polygon &  &  &  & https://daten.berlin.de/ & \\
Berlin Transport Network & Feb '21 & Polygon &  &  &  & OpenStreetMap Overpass API & \\
Berlin Water (Ways) & Feb '21 & Polygon &  &  &  & OpenStreetMap Overpass API & \\
Berlin Land-use & Apr '21 & Polygon &  &  &  & https://daten.berlin.de/ & \\
\addlinespace
WUDAPT LCZ & Oct '20 & Raster &  & 100 & 150 & https://www.wudapt.org/continental-lcz-maps/ & Demuzere et al. (2019)\\
Berlin Veg/Building Height & Oct '20 & Polygon & $m$ &  & 150 & https://daten.berlin.de/ & \\
Berlin Soil Nutrients,
 Bodenkundliche Kennwerte 2015 (Umweltatlas) & Nov '20 & Polygon & $mol~m^{-2}$ &  & 0 & https://daten.berlin.de/ & \\
Berlin Climate Model, Air temperature 2015 (Umweltatlas) & Feb '21 & Polygon & $^\circ C$ &  & 20 & https://daten.berlin.de/ & \\
Planting Bed Area & Oct '20 & Polygon & $m^2$ &  & 0 & https://daten.berlin.de/ & \\
\addlinespace
Berlin Soils & Oct '20 & Polygon &  &  & 0 & https://daten.berlin.de/ & \\
\bottomrule
\end{tabu}
\end{table}

\hypertarget{general-analyses}{%
\subsection{General analyses}\label{general-analyses}}

The proposed statistical method is from the class of hierarchical, generalized additive models (GAM, or GAMM for mixed models/hierarchical models).
In these models combinations of continuous and categorical predictor variables can be summed to estimate a response.
In particular, continuous variables that are linearly, as well as non-linearly related to the response can be represented by applying a transfer function, typically termed ``smoothing function'' (Wood, 2017); these are constructed using a number of base functions of varying complexity and form, which provides a high degree of flexibility, ideal for fitting ecosystem dynamics which are rarely linear (Pedersen et al., 2019), or correctly represented with deterministic functional forms (e.g.~quadratic equations).
In general, a GAM can be written as:

\begin{equation}
E (Y)~=~g^{-1}\left( \beta_0 + \sum_{i = 1}^{n} f_i (x_i) \right),
\label{eq:gam-general}
\end{equation}

and

\begin{equation}
y~=~E (Y) + \epsilon,
\label{eq:gam-expectation}
\end{equation}

where \(Y\) is taken from an appropriate distribution and corresponding link function \(g\), \(\beta_0\) is the intercept and \(f_i\) represents a smooth function of a predictor (Pedersen et al., 2019), and \(\epsilon \sim \mathcal{N}(0, \sigma ^2)\).
Note, that \(f_i\) consists of a smooth (e.g.~spline) constructed via basis functions of different form and complexity, multiplied by a coefficient:

\begin{equation}
f_i(x_i)~=~\sum_{k = 1}^{K} \beta_{i, k} b_{i,k}(x_i).
\label{eq:gam-smoothers}
\end{equation}

Nested data structures (e.g.~due to similar road {[}type{]}) can be accounted for by introducing random effects (Wood, 2017), while spatial dependence between observations can be included by constructing smoothing functions with e.g.~northings and eastings, as for example done in (Augustin et al., 2009).
Ultimately, the implementation of a such a GAMM will allow for establishing continuous prediction surfaces of growth potential (approximated via \(DBH\)) for individual species across urban areas (including parks) of Berlin.

Currently, \(DBH\) has been modeled using a hierarchical linear model (linear mixed effects model) with \texttt{lme4} (Bates et al., 2015) in R Core Team (2020) (see Section\(~\)\ref{sec:results}).
The general form of this model is:

\begin{equation}
Y_{i,j} = (\beta_0 + b_{0,i,j}) + (\beta_1 + b_{1,i,j}) \cdot x_i  + \epsilon_{i,j},
\label{eq:lme}
\end{equation}
where \(\beta_0\) is the intercept with its random component \(b_0\), and \(\beta_1\) the slope with its random component \(b_1\).
The random errors are assumed i.i.d. and distributed as \(b \sim \mathcal{N}(0, \tau ^2)\).
The model for which results are presented in Figure\(~\)\ref{fig:fig-lmestat} estimates \(DBH\) from tree age and the local UHI intensity as continuous covariates with random slopes and intercepts for each species;
note, that for computational efficiency each genera was modeled separately.
Further, models were established for the three most abundant species per genera with at least 1000 individuals.

\hypertarget{available-and-required-data}{%
\subsection{Available and required data}\label{available-and-required-data}}

Additional remote sensing data for local context, higher resolution UHI data, and most important, temporally-resolved tree growth data are still required and/or would greatly improve the confidence in effect size estimates (see Section\(~\)\ref{sec:intro}).
See Tab.\(~\)\ref{tab:tab-data-req} for details on data sets.

\begin{landscape}\begin{table}

\caption[Data requirements for analysis.]{\label{tab:tab-data-req}Data requirements for analysis, including currently available/accessed and required/desired data.}
\centering
\fontsize{11}{13}\selectfont
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedleft}X>{\raggedright}X>{\raggedright}X}
\toprule
\begingroup\fontsize{12}{14}\selectfont \textbf{Subject / Relevance}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{Desc.}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{Type}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{Obsv. (n)}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{Resolution (m)}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{Source}\endgroup\\
\midrule
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{available or accessed}}\\
\hspace{1em}urban trees & riparian, street, park; basic mensuration data (not for riparian) & tabular, spatial points & 668254 &  & Senatsverwaltung Berlin\\
\hspace{1em}UHI Effect & raster data set with summer day/night time (global coverage; Berlin included) & raster &  & 200 & UHI Explorer, [@chakraborty2019]\\
\hspace{1em}soil coverage (Baumscheibe) & available soil area and bounding infrastructure of trees & tabular, spatial polygons & 178576 &  & Senatsverwaltung Berlin\\
\hspace{1em}vegetation and building height & data specific for indidivual building (complexes) and vegetation & tabular, spatial polygons &  &  & Senatsverwaltung Berlin\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{required and/or desired}}\\
\hspace{1em}UHI Effect & raster data set with summer day/night time (global coverage; Berlin included) & raster &  & < 50 & ? Landsat / Sentinel\\
\hspace{1em}road / street characteristics & orientation and width of streets & tabular &  &  & ? Senatsverwaltung, urban planning\\
\hspace{1em}street tree density & planting density of trees as proxy for  potential density-dependent inhibition & tabular or raster &  & < 50 & ? Senatsverwaltung, urban planning\\
\hspace{1em}landcover data & providing local development intensity & tabular or raster &  & 100 - 200 & ? Landsat / Sentinel, Senatsverwaltung\\
\hspace{1em}\em{tree growth series} & \em{tree specific, incremental} & \em{tabular} & \em{} & \em{} & \em{? Field work, partners}\\
\bottomrule
\end{tabu}
\end{table}
\end{landscape}

\hypertarget{sec:results}{%
\section{Preliminary results}\label{sec:results}}

Tree locations are clustered and structured based on their category, i.e.~riparian, street and park trees (Fig.\(~\)\ref{fig:fig-tree-overview-map}).
Planting in space and time shows species-specific patterns (by districts), often related to major events, such as the start and end of armed and/or political conflict.
Table\(~\)\ref{tab:tab-age} shows the binned distribution of genera across age classes.

\begin{table}

\caption[Binned age-distribution for trees in Berlin data set.]{\label{tab:tab-age}Binned age-distribution for trees in Berlin data set, and entries missing age information.}
\centering
\fontsize{11}{13}\selectfont
\begin{tabu} to \linewidth {>{}l>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X>{\raggedleft}X}
\toprule
\begingroup\fontsize{12}{14}\selectfont \textbf{Genera}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{(0,30]}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{(30,60]}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{(60,90]}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{(90,120]}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{(120,150]}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{150+}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{Total (n)}\endgroup & \begingroup\fontsize{12}{14}\selectfont \textbf{Missing (n)}\endgroup\\
\midrule
\em{Tilia} & 43060 & 66929 & 37197 & 5403 & 205 & 113 & 163127 & 10220\\
\em{Acer} & 33027 & 52167 & 14941 & 3074 & 203 & 139 & 140815 & 37264\\
\em{Quercus} & 12085 & 20155 & 8808 & 4478 & 1012 & 1842 & 64464 & 16084\\
\em{Betula} & 6302 & 11026 & 2135 & 138 & 4 & 1 & 29013 & 9407\\
\em{Platanus} & 5242 & 12739 & 5105 & 1736 & 859 & 119 & 26714 & 914\\
\addlinespace
\em{Aesculus} & 5757 & 8651 & 6589 & 1808 & 201 & 45 & 25909 & 2858\\
\em{Robinia} & 4587 & 9351 & 2270 & 376 & 30 & 9 & 24238 & 7615\\
\em{Populus} & 1603 & 7819 & 3223 & 1108 & 207 & 100 & 19973 & 5913\\
\em{Carpinus} & 6411 & 5125 & 934 & 303 & 15 & 15 & 17951 & 5148\\
\em{Prunus} & 6837 & 3929 & 326 & 39 & 5 & 0 & 17904 & 6768\\
\addlinespace
\em{Fraxinus} & 6235 & 4862 & 1160 & 252 & 21 & 5 & 16835 & 4300\\
\em{Pinus} & 2118 & 3730 & 2039 & 345 & 18 & 4 & 14915 & 6661\\
\em{Other} & 33246 & 21960 & 5755 & 3832 & 285 & 863 & 106396 & 40455\\
\midrule
\em{Marg. Totals} & 166510 & 228443 & 90482 & 22892 & 3065 & 3255 & 668254 & 153607\\
\bottomrule
\end{tabu}
\end{table}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{C:/Users/ahurl/Documents/_work/p024_gfz_berlin-trees/berlin.trees/analysis/figures/map_01_overview}

}

\caption{Individual tree locations for three categories available in Berlin Senate urban tree data set. Note, that for each category 7000 observations were subsampled from the available pool to facilitate visualization.}\label{fig:fig-tree-overview-map}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{C:/Users/ahurl/Documents/_work/p024_gfz_berlin-trees/berlin.trees/analysis/figures/map_02_tree_sums_standardized}

}

\caption{Gridded counts for the 11 most frequent genera, as well as *Pinus* and remaining genera. Note, that counts are standardized to unity for individual genera.}\label{fig:fig-tree-count-map}
\end{figure}

The distribution of the UHI effect is highly irregular and clustered in space (Fig.\(~\)\ref{fig:fig-uhi-map}), and also shows variability through time (data not shown, refer to the \href{https://yceo.users.earthengine.app/view/uhimap}{urban heat island explorer}).

\begin{figure}

{\centering \includegraphics[width=0.5\linewidth]{C:/Users/ahurl/Documents/_work/p024_gfz_berlin-trees/berlin.trees/analysis/figures/map_03_uhi}

}

\caption{Estimate of UHI intensity based on the algorithm in (Chakraborty and Lee, 2019), comparing urban with rural pixels within the greater metropolitan cluster. Presented values are averaged over the summer of 2007.}\label{fig:fig-uhi-map}
\end{figure}

The exposure to increased heat-loading of individual genera (and consequently species) is highly uneven throughout the city (Fig.\(~\)\ref{fig:fig-density}).
Street and park trees of most genera are clustered in urban areas with intermediate to high UHI loading, while riparian trees, and some street and park trees of other genera tend to be spread more evenly across Berlin's UHI range.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{C:/Users/ahurl/Documents/_work/p024_gfz_berlin-trees/berlin.trees/analysis/figures/plot_02_genus_UHI_dens}

}

\caption{Empirical density distribution of all individuals within the presented genera along the UHI continuum. UHI intensities were extracted for each tree location, and the distribution hence represents the first detailed overview of the exposure of Berlin's trees to urban heat loading. The black line is the density across all three categories. Insets show corresponding tree totals.}\label{fig:fig-density}
\end{figure}

\textbf{Note, that results below are preliminary and should be considered as a template for future outputs, rather than used for inference.}
The effect of UHI loading on absolute growth potential varies between genera and species (Fig.\(~\)\ref{fig:fig-lmestat}).
Most notably, \emph{Quercus}, the 3rd-most frequent genera, shows decreased absolute growth with increasing UHI loading,
while the most frequent genera, \emph{Tilia}, features contrasting relationships between species.
The estimated effect sizes presented here are linear.
However, temperature may exert a non-linear control on absolute growth and, hence, applying a method able to capture such dynamics may result in somewhat different effect sizes / behavior.
Additionally, if temperatures increase in the future under climate warming, any non-linear effects may become more enhanced, stressing the need for a more flexible model fit and structure (i.e.~using GAMM over linear models-).

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{C:/Users/ahurl/Documents/_work/p024_gfz_berlin-trees/berlin.trees/analysis/figures/plot_03_ranef_species_dbh_uhi}

}

\caption{Impact of UHI loading on tree diameter (\(DBH\)), accounting for age and inter-specific differences from the linear mixed model (via random slopes and intercepts). Line-ranges are standard errors of predicted effect sizes (i.e. slopes). Differences between street and park trees are considerable for some species, and may be due to local clustering and/or spatial under-representation across the UHI continuum. Further investigations need to address the degree of spatial autocorrelation and account for it where required in linear mixed models, and with smoothing interactions in a GAMM implementation.}\label{fig:fig-lmestat}
\end{figure}

\hypertarget{outlook}{%
\section{Outlook}\label{outlook}}

We seek to build upon and improve the current analysis by:

\begin{itemize}
\tightlist
\item
  validating the database with independent observations
\item
  incorporating more pertinent covariates as dependent variables in the linear mixed model\\
\item
  testing multiple model structures with formal model selection procedures
\item
  checking model residuals for spatial auto-correlation and accounting for it where necessary to ensure unbiased estimates of effect sizes
\item
  repeating the above with a hierarchical GAM (i.e.~GAMM) to allow for:

  \begin{itemize}
  \tightlist
  \item
    estimating continuous prediction surfaces for UHI impacts on individual species' growth (similar to results in Figure\(~\)\ref{fig:fig-lmestat}) under recent conditions
  \item
    estimating absolute, species-specific growth potential under increased temperatures and UHI loading under climate change, ideally based on climate simulations (otherwise step-wise increases based on RCP scenarios) for the key species. Note, that complications presented by `out-of-sample' predictions will be addressed.
  \item
    assess potential age-dependent UHI impacts on individual species.\\
  \item
    repeating the above (GAMM) with total and incremental basal area as responses
  \end{itemize}
\end{itemize}

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-akbari2001}{}%
Akbari, H., Pomerantz, M., Taha, H., 2001. Cool surfaces and shade trees to reduce energy use and improve air quality in urban areas. Solar Energy, Urban {Environment} 70, 295--310. \url{https://doi.org/10.1016/S0038-092X(00)00089-X}

\leavevmode\hypertarget{ref-akbari2001}{}%
Akbari, H., Pomerantz, M., Taha, H., 2001. Cool surfaces and shade trees to reduce energy use and improve air quality in urban areas. Solar Energy, Urban {Environment} 70, 295--310. \url{https://doi.org/10.1016/S0038-092X(00)00089-X}

\leavevmode\hypertarget{ref-augustin2009}{}%
Augustin, N.H., Musio, M., von Wilpert, K., Kublin, E., Wood, S.N., Schumacher, M., 2009. Modeling {Spatiotemporal Forest Health Monitoring Data}. Journal of the American Statistical Association 104, 899--911. \url{https://doi.org/10.1198/jasa.2009.ap07058}

\leavevmode\hypertarget{ref-augustin2009}{}%
Augustin, N.H., Musio, M., von Wilpert, K., Kublin, E., Wood, S.N., Schumacher, M., 2009. Modeling {Spatiotemporal Forest Health Monitoring Data}. Journal of the American Statistical Association 104, 899--911. \url{https://doi.org/10.1198/jasa.2009.ap07058}

\leavevmode\hypertarget{ref-bates2015}{}%
Bates, D., Mächler, M., Bolker, B., Walker, S., 2015. Fitting linear mixed-effects models using {lme4}. Journal of Statistical Software 67, 1--48. \url{https://doi.org/10.18637/jss.v067.i01}

\leavevmode\hypertarget{ref-bates2015}{}%
Bates, D., Mächler, M., Bolker, B., Walker, S., 2015. Fitting linear mixed-effects models using {lme4}. Journal of Statistical Software 67, 1--48. \url{https://doi.org/10.18637/jss.v067.i01}

\leavevmode\hypertarget{ref-beck2018}{}%
Beck, H.E., Zimmermann, N.E., McVicar, T.R., Vergopolan, N., Berg, A., Wood, E.F., 2018. Present and future {Köppen}-{Geiger} climate classification maps at 1-km resolution. Scientific Data 5, 180214. \url{https://doi.org/10.1038/sdata.2018.214}

\leavevmode\hypertarget{ref-brune2016}{}%
Brune, M., 2016. Urban trees under climate change. {Potential} impacts of dry spells and heat waves in three {German} regions in the 2050s (No. Report 24). {Climate Service Center Germany}, {Hamburg}.

\leavevmode\hypertarget{ref-brune2016}{}%
Brune, M., 2016. Urban trees under climate change. {Potential} impacts of dry spells and heat waves in three {German} regions in the 2050s (No. Report 24). {Climate Service Center Germany}, {Hamburg}.

\leavevmode\hypertarget{ref-chakraborty2019}{}%
Chakraborty, T., Lee, X., 2019. A simplified urban-extent algorithm to characterize surface urban heat islands on a global scale and examine vegetation control on their spatiotemporal variability. International Journal of Applied Earth Observation and Geoinformation 74, 269--280. \url{https://doi.org/10.1016/j.jag.2018.09.015}

\leavevmode\hypertarget{ref-chakraborty2019}{}%
Chakraborty, T., Lee, X., 2019. A simplified urban-extent algorithm to characterize surface urban heat islands on a global scale and examine vegetation control on their spatiotemporal variability. International Journal of Applied Earth Observation and Geoinformation 74, 269--280. \url{https://doi.org/10.1016/j.jag.2018.09.015}

\leavevmode\hypertarget{ref-dahlhausen2018}{}%
Dahlhausen, J., Rötzer, T., Biber, P., Uhl, E., Pretzsch, H., 2018. Urban climate modifies tree growth in {Berlin}. Int J Biometeorol 62, 795--808. \url{https://doi.org/10.1007/s00484-017-1481-3}

\leavevmode\hypertarget{ref-dahlhausen2018}{}%
Dahlhausen, J., Rötzer, T., Biber, P., Uhl, E., Pretzsch, H., 2018. Urban climate modifies tree growth in {Berlin}. Int J Biometeorol 62, 795--808. \url{https://doi.org/10.1007/s00484-017-1481-3}

\leavevmode\hypertarget{ref-endlicher2016}{}%
Endlicher, W., Scherer, D., Büter, B., Kuttler, W., Mathey, J., Schneider, C., 2016. Stadtnatur fördert gutes {Stadtklima}, in: Ökosystemleistungen in Der {Stadt} -- {Gesundheit} Schützen Und {Lebensqualität} Erhöhen, 3.1. {TEEB DE. TU Berlin, UFZ Leipzig}, {Berlin, Leipzig}, pp. 51--63.

\leavevmode\hypertarget{ref-endlicher2016}{}%
Endlicher, W., Scherer, D., Büter, B., Kuttler, W., Mathey, J., Schneider, C., 2016. Stadtnatur fördert gutes {Stadtklima}, in: Ökosystemleistungen in Der {Stadt} -- {Gesundheit} Schützen Und {Lebensqualität} Erhöhen, 3.1. {TEEB DE. TU Berlin, UFZ Leipzig}, {Berlin, Leipzig}, pp. 51--63.

\leavevmode\hypertarget{ref-fenner2017}{}%
Fenner, D., Meier, F., Bechtel, B., Otto, M., Scherer, D., 2017. Intra and inter {`local climate zone'} variability of air temperature as observed by crowdsourced citizen weather stations in {Berlin}, {Germany}. metz 26, 525--547. \url{https://doi.org/10.1127/metz/2017/0861}

\leavevmode\hypertarget{ref-fenner2017}{}%
Fenner, D., Meier, F., Bechtel, B., Otto, M., Scherer, D., 2017. Intra and inter {`local climate zone'} variability of air temperature as observed by crowdsourced citizen weather stations in {Berlin}, {Germany}. metz 26, 525--547. \url{https://doi.org/10.1127/metz/2017/0861}

\leavevmode\hypertarget{ref-fenner2014}{}%
Fenner, D., Meier, F., Scherer, D., Polze, A., 2014. Spatial and temporal air temperature variability in {Berlin}, {Germany}, during the years 2001--2010. Urban Climate, {ICUC8}: {The} 8th {International Conference} on {Urban Climate} and the 10th {Symposium} on the {Urban Environment} 10, 308--331. \url{https://doi.org/10.1016/j.uclim.2014.02.004}

\leavevmode\hypertarget{ref-fenner2014}{}%
Fenner, D., Meier, F., Scherer, D., Polze, A., 2014. Spatial and temporal air temperature variability in {Berlin}, {Germany}, during the years 2001--2010. Urban Climate, {ICUC8}: {The} 8th {International Conference} on {Urban Climate} and the 10th {Symposium} on the {Urban Environment} 10, 308--331. \url{https://doi.org/10.1016/j.uclim.2014.02.004}

\leavevmode\hypertarget{ref-gillner2014}{}%
Gillner, S., Bräuning, A., Roloff, A., 2014. Dendrochronological analysis of urban trees: Climatic response and impact of drought on frequently used tree species. Trees 28, 1079--1093. \url{https://doi.org/10.1007/s00468-014-1019-9}

\leavevmode\hypertarget{ref-gillner2014}{}%
Gillner, S., Bräuning, A., Roloff, A., 2014. Dendrochronological analysis of urban trees: Climatic response and impact of drought on frequently used tree species. Trees 28, 1079--1093. \url{https://doi.org/10.1007/s00468-014-1019-9}

\leavevmode\hypertarget{ref-gillner2015}{}%
Gillner, S., Vogt, J., Tharang, A., Dettmann, S., Roloff, A., 2015. Role of street trees in mitigating effects of heat and drought at highly sealed urban sites. Landscape and Urban Planning 143, 33--42. \url{https://doi.org/10.1016/j.landurbplan.2015.06.005}

\leavevmode\hypertarget{ref-gillner2015}{}%
Gillner, S., Vogt, J., Tharang, A., Dettmann, S., Roloff, A., 2015. Role of street trees in mitigating effects of heat and drought at highly sealed urban sites. Landscape and Urban Planning 143, 33--42. \url{https://doi.org/10.1016/j.landurbplan.2015.06.005}

\leavevmode\hypertarget{ref-grimmond1996}{}%
Grimmond, C., Souch, C., Hubble, M., 1996. Influence of tree cover on summertime surface energy balance fluxes, {San Gabriel Valley}, {Los Angeles}. Clim. Res. 6, 45--57. \url{https://doi.org/10.3354/cr006045}

\leavevmode\hypertarget{ref-grimmond1996}{}%
Grimmond, C., Souch, C., Hubble, M., 1996. Influence of tree cover on summertime surface energy balance fluxes, {San Gabriel Valley}, {Los Angeles}. Clim. Res. 6, 45--57. \url{https://doi.org/10.3354/cr006045}

\leavevmode\hypertarget{ref-gulyas2006}{}%
Gulyás, Á., Unger, J., Matzarakis, A., 2006. Assessment of the microclimatic and human comfort conditions in a complex urban environment: {Modelling} and measurements. Building and Environment 41, 1713--1722. \url{https://doi.org/10.1016/j.buildenv.2005.07.001}

\leavevmode\hypertarget{ref-gulyas2006}{}%
Gulyás, Á., Unger, J., Matzarakis, A., 2006. Assessment of the microclimatic and human comfort conditions in a complex urban environment: {Modelling} and measurements. Building and Environment 41, 1713--1722. \url{https://doi.org/10.1016/j.buildenv.2005.07.001}

\leavevmode\hypertarget{ref-hertel2019}{}%
Hertel, D., Schlink, U., 2019. Decomposition of urban temperatures for targeted climate change adaptation. Environmental Modelling \& Software 113, 20--28. \url{https://doi.org/10.1016/j.envsoft.2018.11.015}

\leavevmode\hypertarget{ref-hertel2019}{}%
Hertel, D., Schlink, U., 2019. Decomposition of urban temperatures for targeted climate change adaptation. Environmental Modelling \& Software 113, 20--28. \url{https://doi.org/10.1016/j.envsoft.2018.11.015}

\leavevmode\hypertarget{ref-hoyano1988}{}%
Hoyano, A., 1988. Climatological uses of plants for solar control and the effects on the thermal environment of a building. Energy and Buildings 11, 181--199. \url{https://doi.org/10.1016/0378-7788(88)90035-7}

\leavevmode\hypertarget{ref-hoyano1988}{}%
Hoyano, A., 1988. Climatological uses of plants for solar control and the effects on the thermal environment of a building. Energy and Buildings 11, 181--199. \url{https://doi.org/10.1016/0378-7788(88)90035-7}

\leavevmode\hypertarget{ref-jia2018}{}%
Jia, W., Zhao, S., Liu, S., 2018. Vegetation growth enhancement in urban environments of the {Conterminous United States}. Global Change Biology 24, 4084--4094. \url{https://doi.org/10.1111/gcb.14317}

\leavevmode\hypertarget{ref-jia2018}{}%
Jia, W., Zhao, S., Liu, S., 2018. Vegetation growth enhancement in urban environments of the {Conterminous United States}. Global Change Biology 24, 4084--4094. \url{https://doi.org/10.1111/gcb.14317}

\leavevmode\hypertarget{ref-kuttler2015}{}%
Kuttler, W., Miethke, A., Dütemeyer, D., Barlag, A.-B. (Eds.), 2015. Das klima von essen = the climate of essen. {Westarp Wiss.}, {Hohenwarsleben}.

\leavevmode\hypertarget{ref-kuttler2015}{}%
Kuttler, W., Miethke, A., Dütemeyer, D., Barlag, A.-B. (Eds.), 2015. Das klima von essen = the climate of essen. {Westarp Wiss.}, {Hohenwarsleben}.

\leavevmode\hypertarget{ref-maras2016}{}%
Maras, I., Schmidt, T., Paas, B., Ziefle, M., Schneider, C., 2016. The impact of human-biometeorological factors on perceived thermal comfort in urban public places. https://doi.org/\url{http://dx.doi.org/10.18452/18162}

\leavevmode\hypertarget{ref-maras2016}{}%
Maras, I., Schmidt, T., Paas, B., Ziefle, M., Schneider, C., 2016. The impact of human-biometeorological factors on perceived thermal comfort in urban public places. https://doi.org/\url{http://dx.doi.org/10.18452/18162}

\leavevmode\hypertarget{ref-mayer1987}{}%
Mayer, H., Höppe, P., 1987. Thermal comfort of man in different urban environments. Theor Appl Climatol 38, 43--49. \url{https://doi.org/10.1007/BF00866252}

\leavevmode\hypertarget{ref-mayer1987}{}%
Mayer, H., Höppe, P., 1987. Thermal comfort of man in different urban environments. Theor Appl Climatol 38, 43--49. \url{https://doi.org/10.1007/BF00866252}

\leavevmode\hypertarget{ref-mohamed2017}{}%
Mohamed, M.A., 2017. Monitoring of {Temporal} and {Spatial Changes} of {Land Use} and {Land Cover} in {Metropolitan Regions} through {Remote Sensing} and {GIS}. NR 08, 353--369. \url{https://doi.org/10.4236/nr.2017.85022}

\leavevmode\hypertarget{ref-moser-reischl2019a}{}%
Moser-Reischl, A., Rahman, M.A., Pauleit, S., Pretzsch, H., Rötzer, T., 2019. Growth patterns and effects of urban micro-climate on two physiologically contrasting urban tree species. Landscape and Urban Planning 183, 88--99. \url{https://doi.org/10.1016/j.landurbplan.2018.11.004}

\leavevmode\hypertarget{ref-moser-reischl2019a}{}%
Moser-Reischl, A., Rahman, M.A., Pauleit, S., Pretzsch, H., Rötzer, T., 2019. Growth patterns and effects of urban micro-climate on two physiologically contrasting urban tree species. Landscape and Urban Planning 183, 88--99. \url{https://doi.org/10.1016/j.landurbplan.2018.11.004}

\leavevmode\hypertarget{ref-norton2015}{}%
Norton, B.A., Coutts, A.M., Livesley, S.J., Harris, R.J., Hunter, A.M., Williams, N.S.G., 2015. Planning for cooler cities: {A} framework to prioritise green infrastructure to mitigate high temperatures in urban landscapes. Landscape and Urban Planning 134, 127--138. \url{https://doi.org/10.1016/j.landurbplan.2014.10.018}

\leavevmode\hypertarget{ref-norton2015}{}%
Norton, B.A., Coutts, A.M., Livesley, S.J., Harris, R.J., Hunter, A.M., Williams, N.S.G., 2015. Planning for cooler cities: {A} framework to prioritise green infrastructure to mitigate high temperatures in urban landscapes. Landscape and Urban Planning 134, 127--138. \url{https://doi.org/10.1016/j.landurbplan.2014.10.018}

\leavevmode\hypertarget{ref-oke1982}{}%
Oke, T.R., 1982. The energetic basis of the urban heat island. Quarterly Journal of the Royal Meteorological Society 108, 1--24. \url{https://doi.org/10.1002/qj.49710845502}

\leavevmode\hypertarget{ref-oke1982}{}%
Oke, T.R., 1982. The energetic basis of the urban heat island. Quarterly Journal of the Royal Meteorological Society 108, 1--24. \url{https://doi.org/10.1002/qj.49710845502}

\leavevmode\hypertarget{ref-pauleit2002}{}%
Pauleit, S., Jones, N., Garcia-Martin, G., Garcia-Valdecantos, J.L., Rivière, L.M., Vidal-Beaudet, L., Bodson, M., Randrup, T.B., 2002. Tree establishment practice in towns and cities -- {Results} from a {European} survey. Urban Forestry \& Urban Greening 1, 83--96. \url{https://doi.org/10.1078/1618-8667-00009}

\leavevmode\hypertarget{ref-pauleit2002}{}%
Pauleit, S., Jones, N., Garcia-Martin, G., Garcia-Valdecantos, J.L., Rivière, L.M., Vidal-Beaudet, L., Bodson, M., Randrup, T.B., 2002. Tree establishment practice in towns and cities -- {Results} from a {European} survey. Urban Forestry \& Urban Greening 1, 83--96. \url{https://doi.org/10.1078/1618-8667-00009}

\leavevmode\hypertarget{ref-pedersen2019}{}%
Pedersen, E.J., Miller, D.L., Simpson, G.L., Ross, N., 2019. Hierarchical generalized additive models in ecology: An introduction with mgcv. PeerJ 7, e6876. \url{https://doi.org/10.7717/peerj.6876}

\leavevmode\hypertarget{ref-pedersen2019}{}%
Pedersen, E.J., Miller, D.L., Simpson, G.L., Ross, N., 2019. Hierarchical generalized additive models in ecology: An introduction with mgcv. PeerJ 7, e6876. \url{https://doi.org/10.7717/peerj.6876}

\leavevmode\hypertarget{ref-pretzsch2017}{}%
Pretzsch, H., Biber, P., Uhl, E., Dahlhausen, J., Schütze, G., Perkins, D., Rötzer, T., Caldentey, J., Koike, T., Con, T. van, Chavanne, A., Toit, B. du, Foster, K., Lefer, B., 2017. Climate change accelerates growth of urban trees in metropolises worldwide. Scientific Reports 7, 1--10. \url{https://doi.org/10.1038/s41598-017-14831-w}

\leavevmode\hypertarget{ref-pretzsch2017}{}%
Pretzsch, H., Biber, P., Uhl, E., Dahlhausen, J., Schütze, G., Perkins, D., Rötzer, T., Caldentey, J., Koike, T., Con, T. van, Chavanne, A., Toit, B. du, Foster, K., Lefer, B., 2017. Climate change accelerates growth of urban trees in metropolises worldwide. Scientific Reports 7, 1--10. \url{https://doi.org/10.1038/s41598-017-14831-w}

\leavevmode\hypertarget{ref-quigley2004}{}%
Quigley, M.F., 2004. Street trees and rural conspecifics: {Will} long-lived trees reach full size in urban conditions? Urban Ecosystems 7, 29--39. \url{https://doi.org/10.1023/B:UECO.0000020170.58404.e9}

\leavevmode\hypertarget{ref-quigley2004}{}%
Quigley, M.F., 2004. Street trees and rural conspecifics: {Will} long-lived trees reach full size in urban conditions? Urban Ecosystems 7, 29--39. \url{https://doi.org/10.1023/B:UECO.0000020170.58404.e9}

\leavevmode\hypertarget{ref-rcoreteam2020}{}%
R Core Team, 2020. R: {A} language and environment for statistical computing. {R Foundation for Statistical Computing}, {Vienna, Austria}.

\leavevmode\hypertarget{ref-rcoreteam2020}{}%
R Core Team, 2020. R: {A} language and environment for statistical computing. {R Foundation for Statistical Computing}, {Vienna, Austria}.

\leavevmode\hypertarget{ref-randrup2001}{}%
Randrup, T.B., McPherson, E.G., Costello, L.R., 2001. A review of tree root conflicts with sidewalks, curbs, and roads. Urban Ecosystems 5, 209--225. \url{https://doi.org/10.1023/A:1024046004731}

\leavevmode\hypertarget{ref-randrup2001}{}%
Randrup, T.B., McPherson, E.G., Costello, L.R., 2001. A review of tree root conflicts with sidewalks, curbs, and roads. Urban Ecosystems 5, 209--225. \url{https://doi.org/10.1023/A:1024046004731}

\leavevmode\hypertarget{ref-rhoades1999}{}%
Rhoades, R.W., Stipes, R.J., 1999. Growth of trees on the {Virgina Tech Campus} in response to various factors 7.

\leavevmode\hypertarget{ref-rhoades1999}{}%
Rhoades, R.W., Stipes, R.J., 1999. Growth of trees on the {Virgina Tech Campus} in response to various factors 7.

\leavevmode\hypertarget{ref-roloff2009}{}%
Roloff, A., Korn, S., Gillner, S., 2009. The {Climate}-{Species}-{Matrix} to select tree species for urban habitats considering climate change. Urban Forestry \& Urban Greening 8, 295--308. \url{https://doi.org/10.1016/j.ufug.2009.08.002}

\leavevmode\hypertarget{ref-roloff2009}{}%
Roloff, A., Korn, S., Gillner, S., 2009. The {Climate}-{Species}-{Matrix} to select tree species for urban habitats considering climate change. Urban Forestry \& Urban Greening 8, 295--308. \url{https://doi.org/10.1016/j.ufug.2009.08.002}

\leavevmode\hypertarget{ref-suvkberlin2019}{}%
SUVK, Berlin, 2019. Anteil öffentlicher {Grünflächen} in {Berlin}, Grünflächeninformationssystem ({GRIS}). {Senatsverwaltung für Umwelt, Verkehr und Klimaschutz Berlin, Referat Freiraumplanung und Stadtgrün}, {Berlin}.

\leavevmode\hypertarget{ref-tzoulas2007}{}%
Tzoulas, K., Korpela, K., Venn, S., Yli-Pelkonen, V., Kaźmierczak, A., Niemela, J., James, P., 2007. Promoting ecosystem and human health in urban areas using {Green Infrastructure}: {A} literature review. Landscape and Urban Planning 81, 167--178. \url{https://doi.org/10.1016/j.landurbplan.2007.02.001}

\leavevmode\hypertarget{ref-tzoulas2007}{}%
Tzoulas, K., Korpela, K., Venn, S., Yli-Pelkonen, V., Kaźmierczak, A., Niemela, J., James, P., 2007. Promoting ecosystem and human health in urban areas using {Green Infrastructure}: {A} literature review. Landscape and Urban Planning 81, 167--178. \url{https://doi.org/10.1016/j.landurbplan.2007.02.001}

\leavevmode\hypertarget{ref-wood2017}{}%
Wood, S.N., 2017. Generalized additive models: An introduction with {R}. {CRC press}.

\leavevmode\hypertarget{ref-wood2017}{}%
Wood, S.N., 2017. Generalized additive models: An introduction with {R}. {CRC press}.

\leavevmode\hypertarget{ref-zhao2016}{}%
Zhao, S., Liu, S., Zhou, D., 2016. Prevalent vegetation growth enhancement in urban environment. PNAS 113, 6313--6318. \url{https://doi.org/10.1073/pnas.1602312113}

\leavevmode\hypertarget{ref-zhao2016}{}%
Zhao, S., Liu, S., Zhou, D., 2016. Prevalent vegetation growth enhancement in urban environment. PNAS 113, 6313--6318. \url{https://doi.org/10.1073/pnas.1602312113}

\end{CSLReferences}

\newpage

\hypertarget{colophon}{%
\subsubsection{Colophon}\label{colophon}}

This report was generated on 2021-05-18 14:55:27 using the following computational environment and dependencies:

The current Git commit details are:

\begin{verbatim}
#> Local:    master C:/Users/ahurl/Documents/_work/p024_gfz_berlin-trees/berlin.trees
#> Remote:   master @ origin (https://github.com/the-Hull/berlin.trees.git)
#> Head:     [43beb0c] 2021-05-18: add refs
\end{verbatim}

\end{document}

### 2021-05-18 R/functions.R

@@ -565,6 +565,44 @@ download_berlin_polygons_as_sf <- function(path = "./analysis/data/raw_data/spat
 
 
 
+
+#' Download  Berlin Land use to green, water and urban
+#'
+#' @param path character, path to read file from
+#'
+#' @return sf-tibble with polygons of Berlin landuse
+#' @import httr
+#' dplyr
+#'
+download_berlin_lu <- function(path = "./analysis/data/raw_data/spatial_ancillary/berlin_landuse.geojson"){
+
+    # blu <- sf::st_read(path)
+
+    wfs_lu <- "https://fbinter.stadt-berlin.de/fb/wfs/data/senstadt/sach_nutz2015_nutzsa"
+
+
+    query <- list(service = "WFS",
+                  request = "GetFeature",
+                  version = "2.0.0",
+                  TypeNames = "fis:sach_nutz2015_nutzsa",
+                  # count = 10,
+                  outputFormat = 'application/geo+json')
+
+    # request data via GET (REST API and save to disk)
+
+    httr::GET(wfs_lu,
+              query = query,
+              httr::write_disk(path = path, overwrite = TRUE)
+
+    )
+
+
+
+    return(path)
+}
+
+
+
 #' Load spatial-features data sets of all Berlin trees
 #'
 #' @param download_data Character, file paths for downloaded .geojson files from WFS service
@@ -1090,6 +1128,9 @@ prep_model_df <- function(dset,
     return(df_nest)
 
 }
+
+
+
 
 # Spatial -----------------------------------------------------------------
 
@@ -2518,6 +2559,138 @@ make_ranef_plot <- function(model_out,
 
 
 }
+
+
+#' Generate study site area
+#'
+#' @param blu character, path to berlin land use
+#' @param berlin_poly character, path to berlin poly
+#'
+#' @return ggplot object
+#' @export
+#'
+#' @examples
+make_map_study_area <- function(blu, berlin_poly, path_out, height, width, dpi){
+
+
+
+    blu <-  sf::st_read(blu, quiet = TRUE) %>%
+        sf::st_make_valid()
+
+    blu <- blu %>%
+        dplyr::mutate(lu_adjusted = dplyr::case_when(
+            GRZ_NAME %in% c("Baumschule / Gartenbau",
+                            "Park / Grünfläche",
+                            "Friedhof",
+                            "Brachfläche, wiesenartiger Vegetationsbestand",
[…]

### 2021-05-19 analysis/paper/paper.Rmd
paper update

@@
 It also includes spatially-explicit model outputs for local climate simulations representative of typical conditions, such as air temperature (2 m) on a 2015 summer's day at different hours (0400, 1400, 2200).
+These model outputs were preferred over *in situ* measurements as **more representative based on averaged conditions, physiologically more relevant than surface urban heat island products**.
 These outputs are available at city block or planning unit level, and were used as a proxy for the general temperature distribution influenced by urbanization across the investigated growth period across the previous 80 years (see section X for details).
-This was deemed reasonable as Berlin's built-up area has not changed markedly over the past 50 years (i.e., from about 52 to 61 %) [@mohamed2017]. 
-The utility of using this high-resolution data set of air temperatures over a derived urban heat island product was assessed by… 
+This was deemed reasonable as Berlin's built-up area has not changed markedly over the past 50 years, i.e., about 52 to 61$\%$ [@mohamed2017]. 
 Further insight into the role of the urban environment was assessed by including local climate zone classes derived by WUDAPT.
 These describe the proportion of urban area with a specific cover, such as dense high rises, sealed surfaces, etc. 
-For this study, the LCZ 2, representing dense mid-rise built up cover was used as a proxy for the degree of urbanization, and is the most frequent LCZ for Berlin [@fenner2017] (found ref in .
+For this study, the LCZ 6, representing dense mid-rise built up cover was used as a proxy for the degree of urbanization, and is the most frequent LCZ for Berlin [@fenner2017] (found ref in .
 Use percent sealed surface instead?
@@
 
+# Discussion
+
+- Temperature, environmental and urban controls on tree growth  
+	+ Intra specific differences
+	+ Sensitive trees in hot spots
+	+ Temporal dynamics  
+	+ Validity and limitations  
+- Implications:  
+  + Policy / planning:  
+    - tree species
+	  - Increase temporal coverage and accuracy  
+		- which trees to prioritize for measurement campaigns  
+		- Supplement species matrix etc. with these analyses  
+  + Open data important --> drives innovation
+
 # Outlook

### 2021-09-15 R/functions.R
moran's checks

@@ -4044,6 +4044,50 @@ summarize_age_groups <- function(dset,model_df, tempvar, age_break_expr){
     return(pred_groups)
 }
 
+
+
+
+#' Calculate Moran's I for spatial subset
+#'
+#' @param dframe sf data frame
+#' @param grid sf polygon to subset dframe
+#' @param min_obs numeric, fail-safe to ensure enough data is assessed
+#' @param var character, variable name in dframe to be assessed
+#'
+#' @return list, output from `ape::Moran.I()`
+check_moran <- function(dframe, grid, min_obs = 1000, var){
+
+    if(!{var %in% colnames(dframe)}){
+        stop(sprintf("%s is not a column in dframe.", var))
+    }
+
+    subdf <- dframe[grid, ]
+    subdf <- tidyr::drop_na(subdf, tidyselect::all_of(var))
+
+
+    if(nrow(subdf) < min_obs){
+        message("not enough observations, inrease min_obs")
+        return(NULL)
+    }
+    if(length(unique(diff(as.numeric(subdf[ , var, drop = TRUE])))) > 1 &&
+       unique(diff(as.numeric(subdf[ , var, drop = TRUE]))) == 0){
+        message("all obs are equal")
+        return(NULL)
+    }
+
+    # calc inverse distance
+    dists <- 1/as.matrix(dist(cbind(subdf$X, subdf$Y)))
+    diag(dists) <- 0
+    dists[is.infinite(dists)] <- 0
+
+
+    calc_weights <- ape::Moran.I(x = subdf[ , var, drop = TRUE], weight = dists, scaled = TRUE)
+
+    return(calc_weights)
+}
+
+
+
 # BIWI analyses -----------------------------------------------------------
 
 #' Moving or sliding windows for rwl time series

### 2021-09-15 R/temp_gam.R

@@ -938,7 +938,7 @@ drake::loadd(model_df_stat_filtered)
 
 
 
-bam_dbh_fulldf[18:21, "model_file_path"]
+# bam_dbh_fulldf[18:21, "model_file_path"]
 
 
 
@@ -1157,7 +1157,8 @@ mod_summary_list <-
     setNames(levels(mod_groups))
 future::plan(future::sequential())
 
-saveRDS(mod_summary_list, "TEMP_mod_summary_list.Rds")
+# saveRDS(mod_summary_list, "TEMP_mod_summary_list.Rds")
+# readRDS("TEMP_mod_summary_list.Rds")
 
 # grab model paths --------------------------------------------------------
 
@@ -1186,8 +1187,8 @@ mod_group_list <-
 
 
 # Predict for each model --------------------------------------------------
-mod_test_list <- mod_group_list[1]
-mod_test_list[[1]] <- mod_group_list[[1]][1:3]
+# mod_test_list <- mod_group_list[1]
+# mod_test_list[[1]] <- mod_group_list[[1]][1:3]
 
 
 mod_prediction_list <- purrr::map_depth(
@@ -1412,6 +1413,8 @@ mod_n <- mod_summary_list %>%
     rename(n_sample = 2)
 
 
+
+
 mod_dev <- mod_summary_list %>%
     purrr::map_depth(2, "summary") %>%
     purrr::map_depth(2, "dev.expl") %>%
@@ -1445,3 +1448,96 @@ ggplot(mod_dev,
     theme_minimal() +
     scale_fill_brewer(type = "qual", palette = "Set3")
 
+
+
+
+
+
+ggplot(mod_dev %>%
+           mutate(is_spatial = grepl("spatial", mod_group)),
+       aes(x = expvar,
+           y = deviance_explained)) +
+    # geom_line(data = mod_means, aes(x = y),  color = "black", group = 1, alpha = 0.3) +
+    geom_boxplot() +
+    geom_jitter( shape = 21,width = .1, aes(size = n_sample,
+                                             fill = mod_group), alpha = 0.8) +
+    # geom_linerange(data = mod_means, aes(x = y, xmin = ymin, xmax = ymax), color = "black") +
+    # geom_point(data = mod_means, aes(y = y), size = 3, shape = 21, color = "white", fill = "black") +
+    guides(fill = guide_legend(override.aes = list(size = 3),
+                               ncol = 3)) +
+    theme_minimal() +
[…]

### 2021-09-15 analysis/paper/paper.Rmd

@@
   - uhi_references.bib
-  - 04_gfz_uhi.bib
+  - 04_gfz_UHI.bib
+  - 07_phenoTram.bib
 csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
@@
 
-```{r dsources-tab}
+```{r dsources-tab, eval = TRUE}
 
@@
              #              "Source")
-             escape = FALSE)
+             escape = FALSE) %>% 
+      kableExtra::landscape()
 
@@
 
-- Temperature, environmental and urban controls on tree growth  
+1. Temperature, environmental and urban controls on tree growth  
 	+ Intra specific differences
@@
 	+ Validity and limitations
-- Implications:  
+
+2. Implications:  
   + Policy / planning:  

### 2021-09-15 analysis/paper/manuscript.Rmd
updating ms

--- 2021-09-15 analysis/paper/paper.Rmd
@@
 Berlin is one of the largest metropolitan areas in Central Europe (892$~km^2$) with a population of approximately 3.6 million, and a maximum extent of 38$~km$ in North-South and 45$~km$ in East-West directions.
-It is located in North-Eastern Germany, and lies in the temperate zone with warm-humid climate (Dfb) according to the updated Köppen-Geiger classification [@beck2018], with mean annual temperature of approximately 10$^\circ C$ and precipitation of 575$~mm$ (Tempelhof weather station, DWD).
-Berlin features low relief (approximately 30$~m$ to 60$~m$ with 120$~m$ at solitary peaks), and is centered around a glacial outwash valley (sands, gravel), bordered by two plateaus consisting of glacial till and clay in the North-East and South, as well as sands in the South-West.
+It is located in North-Eastern Germany, and lies in the temperate zone with warm-humid climate (Dfb) according to the updated Köppen-Geiger classification [@beck2018], with mean annual temperature of approximately 10$^\circ C$ and precipitation of 575$~mm$ (Tempelhof weather station, German Weather Service DWD).
+Berlin has low relief (approximately 30$~m$ to 60$~m$ with 120$~m$ at solitary peaks), and is centered around a glacial outwash valley (sands, gravel), bordered by two plateaus consisting of glacial till and clay in the North-East and South, as well as sands in the South-West.
 The city provides extensive public green space covering around 30$~\%$ of its area [@suvkberlin2019], with an extensive urban forest of nearly 700000 publically-managed trees along streets, in parks and in riparian areas.
-It has a pronounced urban heat island, owing to its high degree of built-up and sealed surfaces, with air temperatures up to 12$^\circ C$ higher in urban compared to adjacent rural areas [@fenner2014].
+It has a pronounced urban heat island, owing to its high degree of built-up and sealed surfaces, with air temperatures up to 12$^\circ C$ higher in urban compared to adjacent rural areas at night times [@fenner2014].
 
-## Data sources
+# General approach
+
+- Assess differences in mean growth across age for individual species produced by environmental factors, including air temperature (or measure of urban heat island loading), soil characteristics and rooting area, adjacent building height) in a space-for-time approach
+- Focus on street trees where species have a minimum of X individuals
+- Account for potential differences in tree management by authorities
+- Use flexible models that can produce curves of varying modality 
+- Understand / confirm urban tree growth (assessed for single species)
+
+## Data sources and processing
 
@@
 Use percent sealed surface instead?
+
+- datacleanr
+- adjusting bad records
+- final original, final data set available online, with cleaning scripts
 
@@
 
-# Preliminary results {#sec:results}
+# Results {#sec:results}
 
@@
 1. Temperature, environmental and urban controls on tree growth  
-	+ Intra specific differences
+	+ Urban tree growth accelerated (BIWI data)
+	+ Intra specific differences in UHI response (note temp. products)
 	+ Sensitive trees in hot spots
-	+ Temporal dynamics  
-	+ Validity and limitations
+	+ Other controls: soil, urban fabric, 
+	+ Validity and limitations (spatial auto-correlation, data quality, coverage)
 
@@
     - tree species
-	  - Increase temporal coverage and accuracy  
-		- which trees to prioritize for measurement campaigns  
-		- Supplement species matrix etc. with these analyses  
+	  - Increase temporal coverage and accuracy of surveys
+		- which trees to prioritize for measurement campaigns (low-abundance species, highly sensitive)
+		- Compare and Supplement species matrix etc. with these analyses  
+		- Generate co-located environmental observations (e.g., micro-climate)
   + Open data important --> drives innovation
 
-# Outlook
-
-We seek to build upon and improve the current analysis by:
-
-- validating the database with independent observations
-- incorporating more pertinent covariates as dependent variables in the linear mixed model  
-- testing multiple model structures with formal model selection procedures
-- checking model residuals for spatial auto-correlation and accounting for it where necessary to ensure unbiased estimates of effect sizes
-- repeating the above with a hierarchical GAM (i.e. GAMM) to allow for:  
-  * estimating continuous prediction surfaces for UHI impacts on individual species' growth (similar to results in Figure$~$\@ref(fig:fig-lmestat)) under recent conditions
-  * estimating absolute, species-specific growth potential under increased temperatures and UHI loading under climate change, ideally based on climate simulations (otherwise step-wise increases based on RCP scenarios) for the key species.
-Note, that  complications presented by 'out-of-sample' predictions will be addressed.
-  * assess potential age-dependent UHI impacts on individual species.  
-  * repeating the above (GAMM) with total and incremental basal area as responses
+# Conclusion
 

### 2021-10-14 analysis/paper/paper.Rmd
lcz 300

@@
 # Introduction {#sec:intro}
+
+The manifold ecological and societal benefits urban trees provide [e.g., @roy2012] depend critically on their health and performance. 
+For instance, trees alter local energy budgets [@grimmond1996 ; @hertel2019] through shading and transpiration [@gillner2015;@endlicher2016], and therefore can reduce ambient temperatures, infrastructure power-consumption and (human) thermal discomfort  [e.g.
+@gulyas2006; @akbari2001; @mayer1987; @hoyano1988].
+However, excess heat common for cities [i.e.
+Urban Heat Island, UHI; @oke1982], combined with other urban conditions, affects tree physiological functioning with outcomes ranging from enhanced growth to early senescence, branch die-back, and even mortality [e.g., @hilbert2019;@au2018;@gillner2014].
+Thus, assessing the effect of increased temperatures on trees, as part of urban green infrastructure, is instrumental for understanding as well as adapting to current and expected conditions in this century [@ward2007], especially considering ever more urbanized societies  and the potential for UHI effects to compound with more frequent atmospheric drought [@roloff2009;@brune2016;@norton2015].
+
+The UHI effect, i.e., the difference between urban and adjacent rural (air) temperatures, has been intensively studied for several decades [cf. @oke1982;@stewart2011].
+It is typically related to the structure and density of urban land-use [@kuttler2015], which can be characterized through local climate zones, and modulated by physiographic and urban characteristics, such as vicinity to water bodies, predominant wind and street direction, etc.
+[@stewart2012]; yet, the physical basis for the excess heat in cities is to a large extent found in the altered surface energy balance as the proportional cover of vegetation decreases compared to rural (or reference) systems [@hertel2019; @oke1992].
+In temperate climates, this results in strongest UHI magnitudes at night [cf. @fenner2014].
+For example, Berlin features the most intense UHI in Germany due to its large extent and development intensity  with an average air temperature increase of around 5 K at night-times [2001-2010] with maxima of up to 11 K [@fenner2014] in urban $vs.$ rural areas.
+
+**Add refs**
+Increased air temperatures due to UHIs can affect growth through altering several physiological processes.
+Higher temperatures lead to more rapid reaction times at cellular level in the cambium (incremental growth) and during photosynthesis, can extend leaf-on phases, and  may be associated with less cloudy conditions, indirectly increasing the availability of energy for photosynthesis.
+However, if temperatures exceed photosynthetic optima, productivity may decline rapidly followed by (permanent) tissue damage in leaves.
+As warmer air is typically drier, high temperatures can also lead to closure of stomata (major leaf gas exchange pathway) to prevent excessive loss of water to a "demanding" atmosphere. 
+Additionally, high temperatures at night can increase respiration rates, leading to increased loss of carbohydrates.
+Where high temperatures and dry air are compounded by atmospheric or soil drought, trees thus not only experience decreased growth, but also may face extensive tissue damage and ultimately starvation.
+	- Compounded by decreased Precip in cities?! Moser reischl?
+
+**Urban tree growth is highly location specific**
+Urban trees show a tendency for enhanced growth rates and/or productivity compared to rural conspecifics, which is typically attributed to increased temperatures [@jia2018; @pretzsch2017], yet feature a broad range of effect size ranges and signs (i.e., reduced growth) specific to species and location.
+@zhao2016 showed that productivity rates, as a proxy for growth, increased within urban clusters as urbanization intensifies using remotely sensed vegetation indices.
+Further, @moser-reischl2019a identified positive associations between air temperature and radial growth for two species (total of 20 individuals) commonly selected by urban planners (*T.
+cordata*, *Rubinia pseudoacacia*) in Munich.
+
+[@briber2015; @obrien2012] … generally increased growth
+Contrastingly,  @gillner2014 highlight decreased growth for *Acer* species  (*A.
+platanoides* and *pseudoplatanus*), *Platanus x hispanica* and *Quercus rubra* with higher summer temperatures of the preceding year, especially when compounded with drought, in another German metropolis (Dresden).
+@quigley2004 identified absolute growth potential decreased for species between rural and urban conspecifics, yet 
+<!-- lacked spatially-explicit effect size estimates across the urban-rural space and  -->
+was limited to comparatively small sample sizes per group ($n_{total}~=~230$ divided in 15 species, 3 groups and 2 locations) - also lacked multi-modal responses
+@pretzsch2017 inferred enhanced growth in recent decades and across urban locations spanning several latitudes, including Berlin - however, only 145 individuals of one species (*T.
+cordata*) were assessed there.
+As mentioned previously, climate-growth relationships can vary substantially between species, and in fact, @quigley2004 and @pretzsch2017 report contrasting results regarding average tree diameter, i.e.
+smaller or larger for urban $vs.$ rural trees of same age, **consider other studies mentioned above**
+Similarly, for Berlin, @dahlhausen2018, identified enhanced growth in highly urbanized environments (using basal area increments of 252 trees) for *Tilia cordata* Mill, the most abundant tree of the city, which they attributed to the UHI effect, while intermediate development intensity was adverse for tree growth.
+
+These differences in growth trends may result from contrasting species-specific responses to increased temperatures, but are indeed affected by other (time-varying) factors and stochastic processes, such as water availability, pollution and road-salt loading, structural impedance by infrastructure, or management, etc.
+[@rhoades1999; @quigley2004; @randrup2001; @pauleit2002].
+**expand on location-specific factors that can affect growth**
+This can hinder the extrapolation from individual sampling sites toward predicting effect across entire urban areas and tree stocks, especially where studies rely on labor-intensive methods which are limited logistically by sampling effort, reducing sample sizes, and species as well as spatial coverage.
+This is exacerbated by a lack of co-located environmental variables (i.e.
+measured in situ) at pertinent spatial scales, for instance, as noted by @wohlfart2019 for air temperature and tree leaf phenology, which may lead to incorrect inferences and interpretations for the role of climate change on growth/productivity when applying space-for-time substitutions.
+
+Space-for-time substitutions are a common approach (cf.
+studies above) to generate inferences in observational (rather than treatment-control) studies, where manipulations are costly or logistically infeasible due to time and/or financial constraints.
+However, they require  accounting for confounding factors specific to trees' environments, such as street characteristics, development intensity, available soil volume, etc.
+While several of the aforementioned studies applied the space-for-time approach to quantify temperature and excess heat on growth, they typically compare trees grouped using qualitative or summary descriptors of sampling sites, disregarding the spatio-temporal variability in location-specific factors noted above.
+This can hinder the extrapolation from individual sampling sites toward predicting effects across entire urban areas and tree stocks, especially when studies rely on labor-intensive methods which are limited logistically by sampling effort, reducing sample sizes, and species as well as spatial coverage.
+This can be  exacerbated by a lack of co-located environmental variables (i.e.
+measured in situ) at pertinent spatial scales, for instance, as noted by @wohlfart2019 for air temperature and tree leaf phenology, which may lead to incorrect inferences and interpretations for the role of climate change on growth/productivity when applying space-for-time substitutions.
+It is thus likely that the varying and even contrasting growth responses observed for urban trees across and within studies is at least modulated by several confounding factors, making the attribution to a single driver, such as excess heat, more difficult and possibly less accurate.
+
+These limitations could be overcome by relying on large data sets with extensive ancillary data coverage. 
+Berlin, as one of the greenest cities in Europe, provides an openly accessible tree inventory, with spatio-temporal environmental data sets pertinent to tree growth, through its Environmental Agency. 
+It features a total of 650000 individuals covering 94 genera and at least 600 species and/or cultivars, listing information on location, trunk diameter (at breast height; $DBH$; see Tab.$~$\@ref(tab:tab-tree-overview)), and stem height, amongst other variables, for the majority of street and park trees.
+For this study, our object was to assess the impact of excess urban heat, i.e.
+the UHI effect, on tree growth using this openly available inventory data set, complemented by additional open data sources as well as incremental growth data from tree cores.
+The assessment relied on flexible statistical models that could capture species and location-specific responses to heat and other urban factors.
+Specifically, we aimed to (1) assess heat exposure of most abundant species; (2) determine the impact of (excess) heat on stem growth with space-for-time substitution to increase coverage; (3) h ighlight the role of location-specific environmental factors in mediating temperature responses.
+While this work constitutes a case study for Berlin, we believe the results are a valuable contribution toward Berlin's current and future management of its tree stock. 
+This is because the high temporal and spatial variability of controls on growth, and thus variability in responses, require that assessments are developed for a specific region, because well-understood tree characteristics [e.g., see @brune2016;@roloff2009], could be strongly modulated  in predicable ways at a given location or time due to management, planting practices, or other environmental controls; for example, if drought hardiness is related to extensive root networks, small soil volumes available to street trees could render a species vulnerable to water stress.
+As such, the approach can also serve as a framework for base-line assessments for other cities with available or growing inventory and ancillary data [cite something on tree diversity in tree stocks, management, etc.], and also inform species-climate matrices regarding temperature sensitivity…..
+
+---
+
+**OLD**
 

### 2021-10-14 R/functions.R

@@ -2243,14 +2243,14 @@ make_model_grid <- function(){
                  "dbh_cm ~  s(X,Y, k = k_spatial, bs = 'gp', m = 3) + te(STANDALTER, %s, by = species_corrected, m = 1, k = k_te) + species_corrected + s(BEZIRK, bs = 're')"),
         list("mI_spatial_age_x_temp_by_species_building_height_reBEZIRK" =
                  "dbh_cm ~  s(X,Y, k = k_spatial, bs = 'gp', m = 3) + te(STANDALTER, %s, by = species_corrected, m = 1, k = k_te) + species_corrected + +s(building_height_m, k = k_uni) + s(BEZIRK, bs = 're')"),
-        list("mI_spatial_age_x_temp_by_species_building_height30_reBEZIRK" =
-                 "dbh_cm ~  s(X,Y, k = k_spatial, bs = 'gp', m = 3) + te(STANDALTER, %s, by = species_corrected, m = 1, k = k_te) + species_corrected + +s(building_height_m30, k = k_uni) + s(BEZIRK, bs = 're')"),
+        list("mI_spatial_age_x_temp_by_species_building_height300_reBEZIRK" =
+                 "dbh_cm ~  s(X,Y, k = k_spatial, bs = 'gp', m = 3) + te(STANDALTER, %s, by = species_corrected, m = 1, k = k_te) + species_corrected + +s(building_height_m300, k = k_uni) + s(BEZIRK, bs = 're')"),
         list("mI_spatial_age_x_temp_by_species_lcz6_reBEZIRK" =
                  "dbh_cm ~  s(X,Y, k = k_spatial, bs = 'gp', m = 3) + te(STANDALTER, %s, by = species_corrected, m = 1, k = k_te) + species_corrected + +s(lcz_prop_6, k = k_uni) + s(BEZIRK, bs = 're')"),
         # list("mI_spatial_age_x_temp_by_species_lcz6_30_reBEZIRK" =
         #          "dbh_cm ~  s(X,Y, k = k_spatial, bs = 'gp', m = 3) + te(STANDALTER, %s, by = species_corrected, m = 1, k = k_te) + species_corrected + +s(lcz_prop30_6, k = k_uni) + s(BEZIRK, bs = 're')"),
-        list("mI_spatial_age_x_temp_by_species_lcz6_75_reBEZIRK" =
[…]

### 2021-10-14 R/plan.R

@@ -155,9 +155,9 @@ plan <- drake::drake_plan(
     #                                                   5000),
     #                                        ~assess_relative_lcz_cover(.x, wudapt_lcz, 30),
     #                                        .progress = FALSE),
-    lcz_cover_prop75 = furrr::future_map_dfr(split_by_n(full_data_set_clean,
+    lcz_cover_prop300 = furrr::future_map_dfr(split_by_n(full_data_set_clean,
                                                       5000),
-                                           ~assess_relative_lcz_cover(.x, wudapt_lcz, 75),
+                                           ~assess_relative_lcz_cover(.x, wudapt_lcz, 300),
                                            .progress = FALSE),
 
     building_height_mean_m = furrr::future_map(split_by_n(full_data_set_clean,
@@ -171,11 +171,11 @@ plan <- drake::drake_plan(
 
 
 
-    building_height_mean_m_30 = furrr::future_map(split_by_n(full_data_set_clean,
+    building_height_mean_m_300 = furrr::future_map(split_by_n(full_data_set_clean,
                                                           75000),
                                                ~assess_relative_building_height(sf_data =  .x,
                                                                                 bh_raster = berlin_building_height,
-                                                                                buff_dist = 30),
+                                                                                buff_dist = 300),
                                                .progress = FALSE) %>%
         unlist() %>%
         unname(),
@@ -205,10 +205,10 @@ plan <- drake::drake_plan(
                             soil_type = soil_type_data,
                             soil_nutrients = soil_nutrient_data,
                             building_height_mean_m = building_height_mean_m,
-                            building_height_mean_m_30 = building_height_mean_m_30,
+                            building_height_mean_m_300 = building_height_mean_m_300,
                             lcz_cover_prop = lcz_cover_prop,
                             # lcz_cover_prop30 = lcz_cover_prop30,
-                            lcz_cover_prop75 = lcz_cover_prop75,
+                            lcz_cover_prop300 = lcz_cover_prop300,
[…]

### 2021-10-14 R/temp_gam.R

@@ -1558,16 +1558,20 @@ short_mod[[1]] <- short_mod[[1]][1:2]
 #' Compare Variable and residuals
 #'
 #' @param mod_list list, mod_group_list
+#' @param bam_df data.frame, for dependency management only
 #' @param var_response character, column name of response variable, e.g. dbh_cm, must be used in model from `mod_list`
 #' @param var_resid character, column name of residuals, typically `.resid` from `broom::augment`
 #' @param gridp sf df, 2x2 grid squares to subset data - currently split by column `grid_region`
 #'
 #' @return list with nesting from mod_list, with ouput from `ape::Moran.I()`
 assess_morans_spatialmod <- function(mod_list,
+                                     bam_df,
                                      gridp,
                                      var_response,
                                      var_resid){
 
+
+    invisible(is.null(bam_df))
 
     # cycle through
     moran_mods <- purrr::map2(mod_list,
@@ -1718,3 +1722,194 @@ plot_obs_predicted_model <- function(path_model,
 
 
 }
+
+
+
+# BIWI --------------------------------------------------------------------
+
+series_long <- prep_rwl_data(path_meta_cores = "C:/Users/ahurl/Documents/_work/p024_gfz_berlin-trees/ext_data/BIWI/dendro_da/Dendro/BIWi_INV_20190123.xlsx",
+                             path_meta_trees = "C:/Users/ahurl/Documents/_work/p024_gfz_berlin-trees/ext_data/BIWI/dendro_da/Dendro/BIWi_INV_20190123.xlsx",
+                             path_meta_sites = "C:/Users/ahurl/Documents/_work/p024_gfz_berlin-trees/ext_data/BIWI/dendro_da/Dendro/BIWi_INV_20190123.xlsx",
+                             path_dir_fh = "C:/Users/ahurl/Documents/_work/p024_gfz_berlin-trees/ext_data/BIWI/dendro_da/Dendro/RAW/")
+
+
+slide_rwl <- make_rwl_windows(df = series_long,
+                              window_n = 5,
+                              type = "slide")
+
+move_rwl <- make_rwl_windows(df = series_long,
+                             window_n = 5,
+                             type = "move")
+
+
+
+#' Estimates time and age trend in BIWI data
+#'
+#' @param df data.frame, either all series raw or moving data
+#' @return mgcv::gamm output
+apply_gam_biwi <- function(df){
+
+    # define inner fncs
[…]

### 2021-10-18 analysis/paper/paper.Rmd
intro writing added lcz6_300

@@
 
-**Add refs**
-Increased air temperatures due to UHIs can affect growth through altering several physiological processes.
-Higher temperatures lead to more rapid reaction times at cellular level in the cambium (incremental growth) and during photosynthesis, can extend leaf-on phases, and  may be associated with less cloudy conditions, indirectly increasing the availability of energy for photosynthesis.
-However, if temperatures exceed photosynthetic optima, productivity may decline rapidly followed by (permanent) tissue damage in leaves.
-As warmer air is typically drier, high temperatures can also lead to closure of stomata (major leaf gas exchange pathway) to prevent excessive loss of water to a "demanding" atmosphere. 
-Additionally, high temperatures at night can increase respiration rates, leading to increased loss of carbohydrates.
-Where high temperatures and dry air are compounded by atmospheric or soil drought, trees thus not only experience decreased growth, but also may face extensive tissue damage and ultimately starvation.
-	- Compounded by decreased Precip in cities?! Moser reischl?
+Increased air temperatures due to UHIs can affect growth through altering several physiological processes across plant organs directly or indirectly [@dusenge2019].
+Generally, reaction times at cellular level increase with temperature up to a maximum, after which a drop in enzymatic activity results in a species-dependent optimum curve  [@arcus2016; @parent2010].
+In leaves this optimum response is reflected in the net assimilation rate of carbohydrates, as a balance of photosynthesis and respiration, with losses exceeding gains more rapidly with increasing temperatures [@long1991].
+These responses vary between species [@tjoelker2001] as well as intra-specifically due to local acclimation, i.e., a shift of optimum temperature responses after prolonged exposure [@yamori2014], and threshold temperatures before tissue damage occurs [for review see @geange2021].
+High temperatures in temperate areas are often coincident with low relative air humidity (i.e., large vapor pressure deficit), which in turn can decrease stomatal conductance governing the majority of gas exchange in leaves [@grossiord2020], and thus the capacity for photosynthesis.
+Under prolonged stomatal closure (or decreased conductance) with high temperatures, trees may thus face decreased growth (in subsequent years) or even starvation as their carbohydrate reserves are depleted yet not replenished at sufficient rates [@mcdowell2008].
+Furthermore, air (and soil temperatures) affect the initiation, speed and cessation of cambial activity, and thus radial growth throughout a growing season [e.g., see @rathgeber2016; @begum2013].
+Radial growth is increasingly considered to be limited by wood formation dynamics and their relation with environmental drivers, rather than solely by photosynthetic activity [@korner2015].
+In particular, the availability of soil water is critical for cell expansion [e.g., @peters2021a] and limits  radial growth before photosynthesis [@fatichi2014]; however, this water availability is again linked to local climate as higher temperatures drive evaporation and thus may contribute to the depletion of soil water storage.
+
+<!-- Higher temperatures lead to more rapid reaction times at cellular level in the cambium , i.e., controlling incremental growth [@begum2013],  and during photosynthesis [@yamori2014], can extend leaf-on phases, and  may be associated with less cloudy conditions, indirectly increasing the availability of energy for photosynthesis.
+-->
+<!-- However, if temperatures exceed photosynthetic optima, productivity may decline rapidly followed by (permanent) tissue damage in leaves. -->
+<!-- As warmer air is typically drier, high temperatures can also lead to closure of stomata (major leaf gas exchange pathway) to prevent excessive loss of water to a "demanding" atmosphere.  -->
+<!-- Additionally, high temperatures at night can increase respiration rates, leading to increased loss of carbohydrates. -->
+<!-- Where high temperatures and dry air are compounded by atmospheric or soil drought, trees thus not only experience decreased growth, but also may face extensive tissue damage and ultimately starvation.
+-->
+<!-- 	- Compounded by decreased Precip in cities?! Moser reischl? -->
 
@@
 	+ Intra specific differences
+	  + management
+	  + root vs. canopy (huber value?)
+	  + comparison young vs. old -> increased CO2, perhaps limitation of space for time,
+	     as lower conductance under high CO2 may result in more water savings (check WUE
+	     in old vs. young trees? future direction).[@dusenge2019] [@mccarthy2011]
 	+ Sensitive trees in hot spots

### 2021-10-18 R/functions.R

@@ -2492,6 +2492,7 @@ make_model_prediction_df <- function(
         lcz_name <- 'lcz_prop_6'
 
     }
+
 #
 #     lcz <- {if(!is.null(fixed_vars$lcz_prop_6)){
 #         fixed_vars$lcz_prop_6
@@ -2604,7 +2605,7 @@ make_model_prediction_df <- function(
                 bheight_name,
                 "soil_nutrients_swert",
                 "baumsch_flaeche_m2",
-                "lcz_prop_6",
+                lcz_name,
                 "BEZIRK"
             )
 
@@ -2637,6 +2638,7 @@ make_model_prediction_df <- function(
 
         }) %>%
         setNames(mod_names)
+
 
 
 
@@ -3923,6 +3925,7 @@ make_deviance_plot <- function(deviance_list,
         deviance_list$deviances %>%
             dplyr::mutate(is_spatial = grepl("spatial", mod_group),
                           is_tensor = grepl("_x_", fixed = TRUE, ignore.case = FALSE, mod_group),
+                          mod_class = sub("(.*[_]species[_])(.*)", replacement = "\\2", mod_group),
                           mod_group = forcats::fct_relevel(mod_group, stringr::str_sort(levels(mod_group))),
                           expvar = forcats::fct_relevel(
                               expvar,
@@ -3930,8 +3933,9 @@ make_deviance_plot <- function(deviance_list,
                               "soil_nutrients",
                               "baumsch_flaeche",
                               "building_height",
-                              "building_height30",
+                              "building_height300",
                               "lcz6",
+                              "lcz6_300",
                               "urbclim_mod_morning_3_5",
                               "urbclim_mod_afternoon_13_15",
                               "urbclim_mod_night_21_23",
@@ -3945,20 +3949,21 @@ make_deviance_plot <- function(deviance_list,
         # geom_line(data = mod_means, aes(x = y),  color = "black", group = 1, alpha = 0.3) +
         # annotate("text", x = c(2.5, 5.5, 8, 10), y = rep(0.82, 4), label = c("no temp", "Urbclim", "Landsat", "Berlin UA")) +
         # annotate("text", x = c(1:4), y = rep(0.82, 4), label = c("bla", "blu", "bli", "kli")) +
[…]

### 2021-10-18 R/plan.R

@@ -378,11 +378,12 @@ plan <- drake::drake_plan(
 
 
 
-   mod_summaries_filtered = model_summarize(mgroups = mod_groups_filtered,
+   mod_summaries_filtered = drake::target(model_summarize(mgroups = mod_groups_filtered,
                                             bam_df = bam_dbh_filtered,
                                     path_model_dir = drake::file_in("analysis/data/models/stat/filtered/"),
                                     path_model_files = path_model_files_filtered,
                                     path_out = drake::file_out("./analysis/data/models/stat/summary/mod_filtered_summary.Rds")),
+                                    format = "file"),
 
 
    ## predictions
@@ -507,7 +508,7 @@ plan <- drake::drake_plan(
            X = 388141,
            Y = 5818534,
            STANDALTER = c(30:35, 45:50, 60:65, 75:80, 90:95),
-           lcz_prop_6 = seq(0, 1, by = 0.025)),
+           lcz_prop300_6 = seq(0, 1, by = 0.025)),
        age_expression = age_expr,
        group_vars = "lcz_prop300_6"),
 
@@ -661,6 +662,17 @@ plan <- drake::drake_plan(
        width = 12,
        dpi = 300),
 
+
+   ### stat: GAMM AIC ------------------
+   gam_aic = extract_mod_AIC(readRDS(mod_summaries_filtered)),
+
+   plot_gam_aic = make_aic_plot(
+       aic_list = gam_aic,
+       base_size = 18,
+       file = drake::file_out("./analysis/figures/fig_model_aic.png"),
+       height = 10,
+       width = 12,
+       dpi = 300),
    ### stat: GAMM day_2007 prediction ----------------
 
    plot_gam_temp_prediction = plot_dbh_temp_single_var(

### 2021-10-18 R/temp_gam.R

@@ -1701,17 +1701,30 @@ plot_obs_predicted_model <- function(path_model,
 
     mod <- readRDS(path_model)
 
-    mod_broomed <- broom::augment(mod)
+    mod_broomed <- broom::augment(mod) %>%
+        dplyr::mutate(.fitted_response = mod$family$linkinv(.fitted))
+
+    mod_preds <- lm(dbh_cm ~ .fitted_response, data = mod_broomed)
+
+    adjr <- summary(mod_preds)$adj.r.squared
+
+    mod_rmse <- mod_preds %>%
+        broom::augment() %>%
+        dplyr::summarise(sqrt(sum(.fitted - dbh_cm)^2/dplyr::n()))
+
+
+
 
     ifun <- Gamma(link = "log")$linkinv
 
     gplot <- mod_broomed %>%
-        ggplot(aes(x = mod$family$linkinv(.fitted), y = dbh_cm)) +
+        ggplot(aes(x = .fitted_response, y = dbh_cm)) +
         geom_hex(bins = 75, color = "transparent") +
         geom_smooth(method = "lm", color = "red") +
         scale_fill_viridis_c() +
         theme_minimal(base_size = 18) +
-        labs(x = "Predicted", y = "Observed", fill = "N")
+        labs(x = "Predicted", y = "Observed", fill = "N") +
+        annotate("text", x = 20, y = )
 
 
     ggplot2::ggsave(filename = file,

### 2021-10-18 R/functions.R
added details on obs-preds plot

@@ -4601,17 +4601,40 @@ plot_obs_predicted_model <- function(path_model,
 
     mod <- readRDS(path_model)
 
-    mod_broomed <- broom::augment(mod)
+
+    mod_broomed <- broom::augment(mod) %>%
+        dplyr::mutate(predicted = mod$family$linkinv(.fitted))
+
+    mod_preds <- lm(dbh_cm ~ predicted, data = mod_broomed)
+
+    adjr <- format(summary(mod_preds)$adj.r.squared, digits = 3)
+
+    mod_rmse <- mod_preds %>%
+        broom::augment() %>%
+        dplyr::summarise(sqrt(sum(.fitted - dbh_cm)^2/dplyr::n())) %>%
+        as.numeric() %>%
+        format(digits = 2)
+
+    mod_mae <- mod_preds %>%
+        broom::augment() %>%
+        dplyr::summarise(mean(abs(.fitted - dbh_cm))) %>%
+        as.numeric() %>%
+        round(2)
 
     ifun <- Gamma(link = "log")$linkinv
 
     gplot <- mod_broomed %>%
-        ggplot(aes(x = mod$family$linkinv(.fitted), y = dbh_cm)) +
+        ggplot(aes(x = predicted, y = dbh_cm)) +
         geom_hex(bins = 75, color = "transparent") +
-        geom_smooth(method = "lm", formula = "y ~ x - 1", color = "red") +
+        geom_smooth(method = "lm", formula = "y ~ x", color = "red") +
+        geom_abline(slope = 1, intercept = 0, linetype = 2, color = "gray40") +
         scale_fill_viridis_c() +
         theme_minimal(base_size = 18) +
-        labs(x = "Predicted", y = "Observed", fill = "N")
+        labs(x = "Predicted", y = "Observed", fill = "N") +
+        annotate("text",
+                 x = 50,
+                 y = 200,
+                 label = as.expression(bquote(R[adj]^2~'='~.(adjr)~~~~RMSE~'='~.(mod_rmse)~~~~MAE~'='~.(mod_mae))))
 
 
     ggplot2::ggsave(filename = file,

### 2021-10-18 R/plan.R

@@ -801,8 +801,13 @@ plan <- drake::drake_plan(
    ### stat: Obs vs. Pred --------------------------------------------------
 
 
-   plot_obs_pred = plot_obs_predicted_model(path_model = bam_dbh_filtered[bam_dbh_filtered$model=='mI_spatial_age_x_temp_by_species_reBEZIRK_var-day_2007', 'model_file_path'],
+   plot_obs_pred_single_temp = plot_obs_predicted_model(path_model = bam_dbh_filtered[bam_dbh_filtered$model=='mI_spatial_age_x_temp_by_species_reBEZIRK_var-day_2007', 'model_file_path'],
                                             file = drake::file_out("./analysis/figures/fig-gam-dbh_temp-day2007_obs_pred.png"),
+                                            height = 7,
+                                            width = 8,
+                                            dpi = 300),
+   plot_obs_pred_single_temp_lcz6 = plot_obs_predicted_model(path_model = bam_dbh_filtered[bam_dbh_filtered$model=='mI_spatial_age_x_temp_by_species_lcz6_reBEZIRK_var-day_2007', 'model_file_path'],
+                                            file = drake::file_out("./analysis/figures/fig-gam-dbh_temp-day2007_lcz6_obs_pred.png"),
                                             height = 7,
                                             width = 8,
                                             dpi = 300),

### 2021-10-22 analysis/paper/paper.Rmd
added SI temp comparison

@@
 @gulyas2006; @akbari2001; @mayer1987; @hoyano1988].
-However, excess heat common for cities [i.e.
-Urban Heat Island, UHI; @oke1982], combined with other urban conditions, affects tree physiological functioning with outcomes ranging from enhanced growth to early senescence, branch die-back, and even mortality [e.g., @hilbert2019;@au2018;@gillner2014].
+However, excess heat common for cities [i.e., Urban Heat Island, UHI, @oke1982], combined with other urban conditions, affects tree physiological functioning with outcomes ranging from enhanced growth to early senescence, branch die-back, and even mortality [e.g., @hilbert2019;@au2018;@gillner2014].
 Thus, assessing the effect of increased temperatures on trees, as part of urban green infrastructure, is instrumental for understanding as well as adapting to current and expected conditions in this century [@ward2007], especially considering ever more urbanized societies  and the potential for UHI effects to compound with more frequent atmospheric drought [@roloff2009;@brune2016;@norton2015].
@@
 In temperate climates, this results in strongest UHI magnitudes at night [cf. @fenner2014].
-For example, Berlin features the most intense UHI in Germany due to its large extent and development intensity  with an average air temperature increase of around 5 K at night-times [2001-2010] with maxima of up to 11 K [@fenner2014] in urban $vs.$ rural areas.
+For example, Berlin features the most intense UHI in Germany due to its large extent and development intensity  with an average air temperature increase of around 5 K at night-times (2001-2010) with maxima of up to 11 K [@fenner2014] in urban $vs.$ rural areas.
 
-Increased air temperatures due to UHIs can affect growth through altering several physiological processes across plant organs directly or indirectly [@dusenge2019].
+Increased air temperatures due to UHIs can affect tree growth through altering several physiological processes across plant organs directly or indirectly [@dusenge2019].
 Generally, reaction times at cellular level increase with temperature up to a maximum, after which a drop in enzymatic activity results in a species-dependent optimum curve  [@arcus2016; @parent2010].
@@
 Under prolonged stomatal closure (or decreased conductance) with high temperatures, trees may thus face decreased growth (in subsequent years) or even starvation as their carbohydrate reserves are depleted yet not replenished at sufficient rates [@mcdowell2008].
-Furthermore, air (and soil temperatures) affect the initiation, speed and cessation of cambial activity, and thus radial growth throughout a growing season [e.g., see @rathgeber2016; @begum2013].
+Furthermore, air (and soil temperatures) affect the initiation, speed and cessation of cambial activity, and thus radial growth throughout a growing season [e.g., see @begum2013; @rathgeber2016].
 Radial growth is increasingly considered to be limited by wood formation dynamics and their relation with environmental drivers, rather than solely by photosynthetic activity [@korner2015].
-In particular, the availability of soil water is critical for cell expansion [e.g., @peters2021a] and limits  radial growth before photosynthesis [@fatichi2014]; however, this water availability is again linked to local climate as higher temperatures drive evaporation and thus may contribute to the depletion of soil water storage.
+In particular, the availability of soil water is critical for cell expansion [e.g., @peters2021a] and most likely limits radial growth before photosynthesis [@fatichi2014]; however, this water availability is again linked to local climate as higher temperatures drive evaporation and thus may contribute to the depletion of soil water storage, impeding growth.
 
@@
 
-**Urban tree growth is highly location specific**
 Urban trees show a tendency for enhanced growth rates and/or productivity compared to rural conspecifics, which is typically attributed to increased temperatures [@jia2018; @pretzsch2017], yet feature a broad range of effect size ranges and signs (i.e., reduced growth) specific to species and location.
@@
 This is exacerbated by a lack of co-located environmental variables (i.e.
-measured in situ) at pertinent spatial scales, for instance, as noted by @wohlfart2019 for air temperature and tree leaf phenology, which may lead to incorrect inferences and interpretations for the role of climate change on growth/productivity when applying space-for-time substitutions.
+measured in situ) at pertinent spatial scales, for instance, as noted by @wohlfahrt2019 for air temperature and tree leaf phenology, which may lead to incorrect inferences and interpretations for the role of climate change on growth/productivity when applying space-for-time substitutions.
 
@@
 This can be  exacerbated by a lack of co-located environmental variables (i.e.
-measured in situ) at pertinent spatial scales, for instance, as noted by @wohlfart2019 for air temperature and tree leaf phenology, which may lead to incorrect inferences and interpretations for the role of climate change on growth/productivity when applying space-for-time substitutions.
+measured in situ) at pertinent spatial scales, for instance, as noted by @wohlfahrt2019 for air temperature and tree leaf phenology, which may lead to incorrect inferences and interpretations for the role of climate change on growth/productivity when applying space-for-time substitutions.
 It is thus likely that the varying and even contrasting growth responses observed for urban trees across and within studies is at least modulated by several confounding factors, making the attribution to a single driver, such as excess heat, more difficult and possibly less accurate.
@@
 
-## Data sources
+## Dendrochronological sampling and analyses
 
-An overview of all data used in this study, including sources, types, and application, is provided in Table X.
-Most notably, characteristics of street trees and their respective environments were collated from the city's tree inventory, as well as various elements of the Berlin Environment Atlas, which are openly available via daten.berlin.de and curated by the city's government.
+## Statistical Analyses
+
+### Model development and selection
+#### General approach: space-for-time analyses
+
+#### Data sources
+
+An overview of data used for models, including sources, types, and application, is provided in Table X.
+
+**Trees**  
+Characteristics of street trees and their respective environments were collated from the city's tree inventory, as well as various elements of the Berlin Environment Atlas, which are openly available via https://daten.berlin.de and curated by the city's government.
 The data used from the inventory includes species, location, age, and diameter at breast height (dbh).
-The environment atlas holds information on several variables relevant to tree growth, such as soil type and nutrient exchange capacity, distance to groundwater, surface area of tree planting beds ("Baumscheibe"), as well as general characteristics of the urban environment like building height.
+
+**Temperature/UHI data sets**
+
+**Ancillary data sets**  
+Following the general approach described above, four ancillary dependent variables next to a temperature measure were employed in models; these were chosen due to their availability at high spatial resolution and coverage, and because their influence on growth was previously identified in literature or their likely impact could be deduced using ecophysiological principles.
+We included planting bed area, the sum of exchangeable basic cation as a proxy for soil nutrient availability, adjacent building height,  and the proportional coverage of local climate zone 6 (LCZ6; open mid-rise, see @demuzere2019 and @stewart2012 for details).
+The latter was chosen as an increase reflects a transition away from densely urbanized areas.
+Note that only street trees in urban, not rural areas or within greenspaces, were considered here, but individual trees may grow along streets adjacent to greenspaces and parks of varying sizes.
+
+were considered to have 
+The environment atlas also provided soil nutrient exchange capacity as a measure of growth-favoring soil conditions,
+
+surface area of tree planting beds ("Baumscheibe"), as well as general characteristics of the urban environment like building height.
 It also includes spatially-explicit model outputs for local climate simulations representative of typical conditions, such as air temperature (2 m) on a 2015 summer's day at different hours (0400, 1400, 2200).
@@
 
+dsource <- dsources[-which(dsources$File == 'Wintertime_gridded_UHI_data.zip'), ]
+
 dsources <- dsources %>%
+    dplyr::mutate(Category = forcats::fct_relevel(Category, 'trees', 'temperature', 'landuse', 'ancillary', 'gov')) %>% 
+    arrange(Category) %>%
   dplyr::select(Name = `Name-eng`, Accessed, Type, Unit = `Unit-latex`, Resolution, Radius, Source = `Source-short`, Reference = TextRef) %>%
-  dplyr::filter(Name != "UHI Berlin",
-                Name != "Park Trees",
+  # dplyr::filter(Name != "UHI Berlin",
+  dplyr::filter(Name != "Park Trees",
                 Name != "Riparian Trees")
@@
 
-## Available and required data
+<!-- ## Available and required data -->
 
-Additional remote sensing data for local context, higher resolution UHI data, and most important, temporally-resolved tree growth data are still required and/or would greatly improve the confidence in effect size estimates (see Section$~$\@ref(sec:intro)).
-See Tab.$~$\@ref(tab:tab-data-req) for details on data sets.
+<!-- Additional remote sensing data for local context, higher resolution UHI data, and most important, temporally-resolved tree growth data are still required and/or would greatly improve the confidence in effect size estimates (see Section$~$\@ref(sec:intro)).
+-->
+<!-- See Tab.$~$\@ref(tab:tab-data-req) for details on data sets. -->
 
-```{r tab-data-req,echo=FALSE}
+```{r tab-data-req,echo=FALSE, eval = FALSE}
 
@@
 
+	[@bussotti2014]
+
 2. Implications:  

### 2021-10-22 R/functions.R

@@ -4230,7 +4230,8 @@ plot_dbh_temp_single_var <- function(pred_list,
                      alpha = 0.4) +
 
 
-        geom_ribbon( alpha = 0.5, color = "transparent") +
+        geom_ribbon(data = plot_data %>% filter(prediction_range == "within"),
+                    alpha = 0.5, color = "transparent") +
         # geom_line(color = "black")  +
 
         geom_smooth(aes(group = 1),
@@ -4242,7 +4243,7 @@ plot_dbh_temp_single_var <- function(pred_list,
 
 
         # geom_line(linetype = 1)  +
-        geom_line(linetype = 3, color = "black", size = 0.75)  +
+        geom_line(linetype = 3, color = "black", size = 0.35)  +
         geom_line(data = plot_data %>% filter(prediction_range == "within")) +
 
 
@@ -4312,18 +4313,20 @@ plot_dbh_temp_single_var_single_species <- function(pred_list,
 ){
 
     if(is.null(species_filter)){
-        species_filter <- unique(model_df$species_corrected)
+        species_filter <- shorten_species(unique(model_df$species_corrected))
     }
 
 
     `%nin%` <- Negate(`%in%`)
 
     model_df <- model_df %>%
-        dplyr::filter(age_group %nin% age_filter,
-                      species_corrected %in% species_filter) %>%
         mutate(age_group = eval(age_expression),
                species_corrected = shorten_species(species_corrected)) %>%
+        dplyr::filter(
+            age_group %nin% age_filter,
+            species_corrected %in% species_filter) %>%
         tidyr::drop_na(age_group)
+
 
 
     plot_data <- pred_list$pred_groups %>%
@@ -4331,15 +4334,20 @@ plot_dbh_temp_single_var_single_species <- function(pred_list,
         tidyr::pivot_longer(cols = dplyr::all_of(names(pred_list$pred_var)),
                             names_to = "uhi_tempvar",
                             values_to = "temp_degc") %>%
+        dplyr::mutate(temp_degc_scaled = scales::rescale(temp_degc, to = c(0,1)),
+                      species_corrected = shorten_species(species_corrected)) %>%
         dplyr::filter(prediction_range == prediction_range) %>%
         dplyr::filter(age_group %nin% age_filter) %>%
-        dplyr::filter(species_corrected %in% species_filter) %>%
+        dplyr::filter(species_corrected %in% shorten_species(species_filter)) %>%
[…]

### 2021-10-22 R/plan.R

@@ -410,6 +410,17 @@ plan <- drake::drake_plan(
        group_vars = NULL),
 
 
+   pred_data_single_tempvar_fixed_lcz6 =  pred_dbh_temp_single_var(
+       path_model =    bam_dbh_filtered[bam_dbh_filtered$model=='mI_spatial_age_x_temp_by_species_lcz6_reBEZIRK_var-day_2007', 'model_file_path'],
+       model_df = model_df_stat_filtered,
+       fixed_vars = list(X = 388141,
+                         Y = 5818534,
+                         STANDALTER = c(30:35, 45:50, 60:65, 75:80, 90:95),
+                         lcz_prop_6 = 0.5),
+       age_expression = age_expr,
+       group_vars = NULL),
+
+
    pred_data_single_tempvar_fullmodel =  pred_dbh_temp_single_var(
        path_model =    bam_dbh_filtered[bam_dbh_filtered$model=='mI_spatial_age_x_temp_by_species_reBEZIRK_full_var-day_2007', 'model_file_path'],
        model_df = model_df_stat_filtered,
@@ -487,7 +498,7 @@ plan <- drake::drake_plan(
 
 
 
-   pred_data_single_tempvar_fixed_lcz6 =  pred_dbh_temp_single_var(
+   pred_data_single_tempvar_multi_lcz6 =  pred_dbh_temp_single_var(
        path_model =    bam_dbh_filtered[bam_dbh_filtered$model=='mI_spatial_age_x_temp_by_species_lcz6_reBEZIRK_var-day_2007', 'model_file_path'],
        model_df = model_df_stat_filtered,
        fixed_vars = list(
@@ -687,6 +698,17 @@ plan <- drake::drake_plan(
        width = 18,
        dpi = 300),
 
+   plot_gam_temp_lcz6_prediction = plot_dbh_temp_single_var(
+       pred_list = pred_data_single_tempvar_fixed_lcz6,
+       model_df = model_df_stat_filtered,
+       age_filter = NULL,
+       age_expression = age_expr,
+       prediction_range = "within",
+       base_size = 18,
+       file = drake::file_out("./analysis/figures/fig-gam-dbh_temp-day2007_lcz6.png"),
+       height = 11,
+       width = 18,
+       dpi = 300),
 
    plot_gam_temp_prediction_fullmodel = plot_dbh_temp_single_var(
        pred_list = pred_data_single_tempvar_fullmodel,
@@ -704,18 +726,35 @@ plan <- drake::drake_plan(
 
 
 
[…]

### 2021-10-26 analysis/paper/paper.Rmd
added biwi

@@
 
-An overview of data used for models, including sources, types, and application, is provided in Table X.
+An overview of data used for models, including sources, types, and application, is provided in Table$~$\@ref(tab:dsources).
 
@@
 **Temperature/UHI data sets**  
+Temperature and UHI data were summarized temporally either by the provider or manually to provide a characteristic representation of heat loading during summer at different times (morning, afternoon/day, night), from which tree averages (radius of 150$~m$) were calculated.
+Two data sets of urban air and one surface temperature were tested as explanatory variables in GAMM models.
+The air temperatures from the Berlin environmental atlas (EnvAt) were model outputs that are representations of typical summer conditions at 0400, 1400 and 2200 hours; these data are provided at city block basis (spatial polygons), from which weighted averages were extracted.
+UrbClim air temperatures are hourly model outputs (100$~m$ resolution, **ref**) based on ERA5 re-analyses data (ECMWF) for which data of the hottest month in the record (June, 2011) were averaged to hours equivalent to EnvAt data by using a window of $\pm~1~$hour (i.e., 0300 to 0500, etc.).
+Subsequently, a land-use and landcover mask (2018,      Copernicus CORINE **ref**) was used to define urban and rural/forested areas, by which the urban heat loading was calculated as $UHI_{x,y} = T_{Air_{2m}~x,y} - \overline{T_{Air_{2m}~Rural}}$, where $x$ and $y$ define an urban grid cell.
+The surface UHI data set by @chakraborty2019 derives its measure in a similar fashion and the reader is referred to the detailed description therein; note this data set provides day and night-time averaged UHI estimates at 500$~m$ resolution, which were extracted for the hottest summer in this record (2007).
 

### 2021-10-26 R/functions.R

@@ -5340,7 +5340,7 @@ summarize_age_groups <- function(dset,model_df, groupvars, age_break_expr){
         # filter(STANDALTER < 100) %>%
         group_by(age_group, !!! groupvars_sym, species_corrected) %>%
         summarise(mean_dbh = mean(pred.fit, na.rm = TRUE),
-                  mean_se = sqrt(sum(pred.se.fit))/n()) %>%
+                  mean_se = sqrt(sum(pred.se.fit^2)/n())) %>%
         mutate(response.fit.mean = ifun(mean_dbh),
                response.low.mean = ifun(mean_dbh - 1.96 * mean_se),
                response.high.mean = ifun(mean_dbh + 1.96 * mean_se)) %>%
@@ -5668,6 +5668,17 @@ make_rwl_windows <- function(df, window_n = 5, type = "slide"){
 
 
 
+#' Makes df of rwl time series, custom for BIWI data structure
+#'
+#' @param path_meta_cores  character, inventory file
+#' @param path_meta_trees character, inventory file
+#' @param path_meta_sites character, inventory file
+#' @param path_dir_fh character, path to raw rwl folders
+#'
+#' @return
+#' @export
+#'
+#' @examples
 prep_rwl_data <- function(path_meta_cores,
                           path_meta_trees,
                           path_meta_sites,
@@ -5881,6 +5892,86 @@ prep_rwl_data <- function(path_meta_cores,
 
 
     return(all_series_rwl_long)
+
+
+}
+
+
+
+#' Estimates time and age trend in BIWI data
+#'
+#' @param df data.frame, either all series raw or moving data
+#' @return mgcv::gamm output
+apply_gam_biwi <- function(df){
+
+    # define inner fncs
+    make_urban_df <- function(df, year_min, year_max,year_break, cambial_age_max){
+        urban_rwl <- df %>%
+            mutate(age_group = cut(cambial_age, seq(from = 0, to = 120, by = 5)),
+                   site_global = case_when(
+                       site_type %in% c("natural", "rural") ~ "forest",
+                       TRUE ~ "urban"),
+                   year_break = as.factor(
+                       ifelse(
+                           year <= year_break,
+                           sprintf("<=%s", year_break),
+                           sprintf(">%s", year_break))),
+                   location_short = as.factor(location_short)) %>%
+            filter(site_global == "urban",
+                   year >= year_min,
+                   year <= year_max,
[…]

### 2021-10-26 R/plan.R

@@ -814,7 +814,7 @@ plan <- drake::drake_plan(
                                  prediction_range = "within",
                                  base_size = 18,
                                  x_label = expression('Proportional Cover - LCZ6'[bar(150~m)]),
-                                 file = drake::file_out("./analysis/figures/fig-gam-dbh_temp-day2007_lcz6_tilia.png"),
+                                 file = drake::file_out("./analysis/figures/fig-gam-dbh_temp-day2007_xvar-lcz6_tilia.png"),
                                  height = 7,
                                  width = 8,
                                  dpi = 300),

### 2021-10-26 R/temp_gam.R

@@ -1740,102 +1740,23 @@ plot_obs_predicted_model <- function(path_model,
 
 # BIWI --------------------------------------------------------------------
 
-series_long <- prep_rwl_data(path_meta_cores = "C:/Users/ahurl/Documents/_work/p024_gfz_berlin-trees/ext_data/BIWI/dendro_da/Dendro/BIWi_INV_20190123.xlsx",
-                             path_meta_trees = "C:/Users/ahurl/Documents/_work/p024_gfz_berlin-trees/ext_data/BIWI/dendro_da/Dendro/BIWi_INV_20190123.xlsx",
-                             path_meta_sites = "C:/Users/ahurl/Documents/_work/p024_gfz_berlin-trees/ext_data/BIWI/dendro_da/Dendro/BIWi_INV_20190123.xlsx",
-                             path_dir_fh = "C:/Users/ahurl/Documents/_work/p024_gfz_berlin-trees/ext_data/BIWI/dendro_da/Dendro/RAW/")
-
-
-slide_rwl <- make_rwl_windows(df = series_long,
-                              window_n = 5,
-                              type = "slide")
-
-move_rwl <- make_rwl_windows(df = series_long,
-                             window_n = 5,
-                             type = "move")
+series_long <- prep_rwl_data(path_meta_cores = "./analysis/data/raw_data/biwi/BIWi_INV_20190123.xlsx",
+                             path_meta_trees = "./analysis/data/raw_data/biwi/BIWi_INV_20190123.xlsx",
+                             path_meta_sites = "./analysis/data/raw_data/biwi/BIWi_INV_20190123.xlsx",
+                             path_dir_fh = "./analysis/data/raw_data/biwi/rwl/")
 
 
 
-#' Estimates time and age trend in BIWI data
-#'
-#' @param df data.frame, either all series raw or moving data
-#' @return mgcv::gamm output
-apply_gam_biwi <- function(df){
-
-    # define inner fncs
-    make_urban_df <- function(df, year_min, year_max,year_break, cambial_age_max){
-        urban_rwl <- df %>%
-            mutate(age_group = cut(cambial_age, seq(from = 0, to = 120, by = 5)),
-                   site_global = case_when(
-                       site_type %in% c("natural", "rural") ~ "forest",
-                       TRUE ~ "urban"),
-                   year_break = as.factor(
-                       ifelse(
-                           year <= year_break,
[…]

### 2021-11-17 analysis/paper/paper.Rmd
added hicam

@@
 
----
+<!-- Berlin features the most intense Urban Heat Island (UHI) in Germany due to its large extent and development intensity [@kuttler2015], with temperature increases of up to $12~K$ during day-time and $6~K$ on average for night-times [2001-2010, @fenner2014] in urban $vs.$ rural areas.
+-->
+<!-- Consequently, urban green (infrastructure) systems are subjected to increased heat more frequently, potentially affecting their process dynamics - either positively or adversely.  -->
+<!-- Their performance and health, however, is closely tied to local energy budgets [@grimmond1996 ; @hertel2019], which in turn are decisive for controlling human wellbeing [e.g.
+@maras2016], amongst other factors.
+-->
+<!-- Assessing the effect of increased temperatures on green infrastructure, as part of the urban landscape, is therefore instrumental for understanding, and ultimately mitigating, the potential impact of future warming on increasingly urban societies [@norton2015].
+-->
 
-**OLD**
+<!-- Trees, in particular, provide shading as well as transpirative cooling in their vicinity [@gillner2015;@oke1982;@endlicher2016], and therefore can reduce ambient temperatures, infrastructure power-consumption and (human) thermal discomfort  [e.g.
+@gulyas2006; @akbari2001; @mayer1987; @hoyano1988]; -->
+<!-- simultaneously, they provide numerous other environmental, cultural and psychological services and/or benefits [see @tzoulas2007 for review]. -->
+<!-- Further, recent tree growth dynamics as a proxy for on-going and future warming may provide an additional line of evidence to support the growing knowledge base on future climate-vegetation dynamics [@zhao2016] and may aid in mitigation and adaptation efforts [@pretzsch2017;@brune2016].
+-->
 
-Berlin features the most intense Urban Heat Island (UHI) in Germany due to its large extent and development intensity [@kuttler2015], with temperature increases of up to $12~K$ during day-time and $6~K$ on average for night-times [2001-2010, @fenner2014] in urban $vs.$ rural areas.
-Consequently, urban green (infrastructure) systems are subjected to increased heat more frequently, potentially affecting their process dynamics - either positively or adversely. 
-Their performance and health, however, is closely tied to local energy budgets [@grimmond1996 ; @hertel2019], which in turn are decisive for controlling human wellbeing [e.g.
-@maras2016], amongst other factors.
-Assessing the effect of increased temperatures on green infrastructure, as part of the urban landscape, is therefore instrumental for understanding, and ultimately mitigating, the potential impact of future warming on increasingly urban societies [@norton2015].
-
-Trees, in particular, provide shading as well as transpirative cooling in their vicinity [@gillner2015;@oke1982;@endlicher2016], and therefore can reduce ambient temperatures, infrastructure power-consumption and (human) thermal discomfort  [e.g.
-@gulyas2006; @akbari2001; @mayer1987; @hoyano1988];
-simultaneously, they provide numerous other environmental, cultural and psychological services and/or benefits [see @tzoulas2007 for review].
-Further, recent tree growth dynamics as a proxy for on-going and future warming may provide an additional line of evidence to support the growing knowledge base on future climate-vegetation dynamics [@zhao2016] and may aid in mitigation and adaptation efforts [@pretzsch2017;@brune2016].
-
-Trees and green infrastructure in urban areas show a tendency for enhanced growth rates and/or productivity compared to rural counterparts [@jia2018; @pretzsch2017], yet feature a broad range of effect size ranges and, in some cases, signs specific to species and locality.
-@zhao2016 showed that growth rates increased within urban clusters as urbanization intensifies using remotely sensed vegetation indices.
-Similarly, for Berlin, @dahlhausen2018, identified positive growth modulation in highly urbanized environments (using growth increments) for *Tilia cordata* Mill, the most abundant tree of the city, which they attributed to the UHI effect, while intermediate development intensity showed indications of being least favorable for tree growth.
-<!-- For Berlin, @dahlhausen2018, showed divergent patterns for young and old trees dependent on urban development intensity for *Tilia cordata* Mill, the most abundant tree of the city, but overall reported an increased growth rate which they attributed to the UHI effect.
+<!-- Trees and green infrastructure in urban areas show a tendency for enhanced growth rates and/or productivity compared to rural counterparts [@jia2018; @pretzsch2017], yet feature a broad range of effect size ranges and, in some cases, signs specific to species and locality.
 -->
-Further, @moser-reischl2019a identified positive associations between air temperature and radial growth for two species commonly selected by urban planners (*T.
+<!-- @zhao2016 showed that growth rates increased within urban clusters as urbanization intensifies using remotely sensed vegetation indices. -->
+<!-- Similarly, for Berlin, @dahlhausen2018, identified positive growth modulation in highly urbanized environments (using growth increments) for *Tilia cordata* Mill, the most abundant tree of the city, which they attributed to the UHI effect, while intermediate development intensity showed indications of being least favorable for tree growth.
+-->
+<!-- <!-- For Berlin, @dahlhausen2018, showed divergent patterns for young and old trees dependent on urban development intensity for *Tilia cordata* Mill, the most abundant tree of the city, but overall reported an increased growth rate which they attributed to the UHI effect.
+--> -->
+<!-- Further, @moser-reischl2019a identified positive associations between air temperature and radial growth for two species commonly selected by urban planners (*T.
 cordata*, *Rubinia pseudoacacia*) in Munich.
-By contrast, @gillner2014 highlight decreased growth for *Acer* species  (*A.
+-->
+<!-- By contrast, @gillner2014 highlight decreased growth for *Acer* species  (*A.
 platanoides* and *pseudoplatanus*), *Platanus x hispanica* and *Quercus rubra* with higher summer temperatures of the preceding year, especially when compounded with drought, in another German metropolis (Dresden).
-Differences in growth trends may result from contrasting species-specific characteristics, but are indeed affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc.
-[@rhoades1999; @quigley2004; @randrup2001; @pauleit2002].
-Under climate change, atmospheric drought will likely be compounded with high temperatures - and intensified UHIs - more frequently, adding further stress to current urban disturbance regimes [@roloff2009].
-
-Conditions affecting tree growth can vary greatly within urban areas or regions, and need to be accounted for when establishing relationships with pertinent drivers, such as the UHI effect.
-This typically complicates the extrapolation from individual sampling sites toward predicting effect sizes across entire urban areas and tree stocks.
-This is especially the case for studies reliant on labour-intensive methods which are limited logistically by sampling effort, reducing sample sizes, as well as species and spatial coverage.
-<!-- While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e.
-genera and/or species, and space).
 -->
-<!-- However, urban tree growth in urban settings is affected by various processes, which  -->
-<!-- However, growth in urban settings is affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc.
+<!-- Differences in growth trends may result from contrasting species-specific characteristics, but are indeed affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc.
 [@rhoades1999; @quigley2004; @randrup2001; @pauleit2002].
 -->
-<!-- While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e.
-genera and/or species, and space).
--->
-<!-- This typically hinders the extrapolation from individual sampling sites toward prediction of effect sizes across entire urban areas and tree stocks.  -->
-<!-- Therefore, studies reliant on labor-intensive sampling, must carefully address confounding factors to ensure effect sizes are estimated adequately. -->
-
-<!-- Detailed, but spatially-limited dendroecological analyses of climate-growth relationships in Berlin for key species can be enhanced by inferring growth modulation from a large data set in excess of 650000 individuals provided by the Berlin Senate Administration (Senatsverwaltung).
+<!-- Under climate change, atmospheric drought will likely be compounded with high temperatures - and intensified UHIs - more frequently, adding further stress to current urban disturbance regimes [@roloff2009].
 -->
 
-To complement (existing) detailed dendroecological analyses of climate-growth relationships in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individuals covering 94 genera and at least 600 species and/or cultivars provided by the Berlin Senate Administration (Senatsverwaltung).
-This data set contains information on location, species, trunk diameter (at breast height; $DBH$; see Tab.$~$\@ref(tab:tab-tree-overview)), and height, amongst other variables for the majority of street and park trees.
+<!-- Conditions affecting tree growth can vary greatly within urban areas or regions, and need to be accounted for when establishing relationships with pertinent drivers, such as the UHI effect. -->
+<!-- This typically complicates the extrapolation from individual sampling sites toward predicting effect sizes across entire urban areas and tree stocks. -->
+<!-- This is especially the case for studies reliant on labour-intensive methods which are limited logistically by sampling effort, reducing sample sizes, as well as species and spatial coverage. -->
+<!-- <!-- While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e.
+genera and/or species, and space).
+--> -->
+<!-- <!-- However, urban tree growth in urban settings is affected by various processes, which  --> -->
+<!-- <!-- However, growth in urban settings is affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc.
+[@rhoades1999; @quigley2004; @randrup2001; @pauleit2002].
+--> -->
+<!-- <!-- While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e.
+genera and/or species, and space).
+--> -->
+<!-- <!-- This typically hinders the extrapolation from individual sampling sites toward prediction of effect sizes across entire urban areas and tree stocks.  --> -->
+<!-- <!-- Therefore, studies reliant on labor-intensive sampling, must carefully address confounding factors to ensure effect sizes are estimated adequately. --> -->
 
-```{r tab-tree-overview}
+<!-- <!-- Detailed, but spatially-limited dendroecological analyses of climate-growth relationships in Berlin for key species can be enhanced by inferring growth modulation from a large data set in excess of 650000 individuals provided by the Berlin Senate Administration (Senatsverwaltung).
+--> -->
+
+<!-- To complement (existing) detailed dendroecological analyses of climate-growth relationships in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individuals covering 94 genera and at least 600 species and/or cultivars provided by the Berlin Senate Administration (Senatsverwaltung).
+-->
+<!-- This data set contains information on location, species, trunk diameter (at breast height; $DBH$; see Tab.$~$\@ref(tab:tab-tree-overview)), and height, amongst other variables for the majority of street and park trees.
+-->
+
+```{r tab-tree-overview, eval = FALSE, include = FALSE}
 drake::loadd(overview_table)
@@
 Berlin features low relief (approximately 30$~m$ to 60$~m$ with 120$~m$ at solitary peaks), and is centered around a glacial outwash valley (sands, gravel), bordered by two plateaus consisting of glacial till and clay in the North-East and South, as well as sands in the South-West.
-The city provides extensive public green space covering around 30$~\%$ of its area [@suvkberlin2019], with an extensive urban forest of nearly 700000 publically-managed trees along streets, in parks and in riparian areas.
-It has a pronounced urban heat island, owing to its high degree of built-up and sealed surfaces, with air temperatures up to 12$^\circ C$ higher in urban compared to adjacent rural areas [@fenner2014].
+The city provides extensive public green space covering around 30$~\%$ of its area [@suvkberlin2019], with an extensive urban forest of nearly 700000 publicly-managed trees along streets, in parks and in riparian areas.
+<!-- It has a pronounced urban heat island, owing to its high degree of built-up and sealed surfaces, with air temperatures up to 12$^\circ C$ higher in urban compared to adjacent rural areas [@fenner2014].
+-->
 
-## Dendrochronological sampling and analyses
+## Data sources
 
-## Statistical Analyses
+An overview of data used for models, including sources, types, and application, is provided in Table$~$\@ref(tab:tab-dsources).
 
-### Model development and selection
-#### General approach: space-for-time analyses
-
-#### Data sources
-
-An overview of data used for models, including sources, types, and application, is provided in Table$~$\@ref(tab:dsources).
-
-**Trees**  
+### Street trees  
 Characteristics of street trees and their respective environments were collated from the city's tree inventory, as well as various elements of the Berlin Environment Atlas, which are openly available via https://daten.berlin.de and curated by the city's government.
 The data used from the inventory includes species, location, age, and diameter at breast height (dbh).
+**quality control** - filtered dataset to remove drastic outliers --> fit poly model, used Median Absolute deviance with multiplier of X to exclude points. 
+Focused on 9 most abundant species to and Cultivars and hybrids were combined to increase sample sizes. **check which ones**
 
-**Temperature/UHI data sets**  
+Table$~$\@ref(tab:tab-age) shows the binned distribution of genera across age classes.
+
+```{r tab-age}
+
+drake::loadd(age_tables)
+
+age_tables <- age_tables %>% 
+  dplyr::arrange(desc(n_total))
+colnames(age_tables)[c(1,7,8, 9)] <- c("Genera", "150+", "Total (n)", "Missing (n)")
+
+age_tables <- dplyr::bind_rows(age_tables[age_tables$Genera!="Other",],
+                        age_tables[age_tables$Genera=="Other",])
+
+add_row <- as.data.frame(matrix(c(NA, colSums(age_tables[,-1])), nrow = 1))
+add_row[1,1] <- "Marg. Totals"
+colnames(add_row) <- colnames(age_tables)
+
+age_tables <- dplyr::bind_rows(age_tables,
+                    add_row)
+
+caption_trees <- "Binned age-distribution for trees in Berlin data set, and entries missing age information. "
+caption_trees_short <- "Binned age-distribution for trees in Berlin data set."
+
+if (grepl("docx", out_format) ||
+    knitr::is_latex_output() ) {
+  table_dat_req_kable <- age_tables %>% 
+  knitr::kable(booktabs = TRUE,
+               format = "latex",
+             caption = caption_trees,
+             caption.short = caption_trees_short
+             # col.names = c("Subject/Relevance",
+             #              "Desc.",
+             #              "Type",
+             #              "Obsv. (n)",
+             #              "Resolution (m)",
+             #              "Source")
+             ) %>% 
+  kableExtra::row_spec(13, hline_after = TRUE)
+
+} else {
+table_dat_req_kable <- age_tables %>% 
+  knitr::kable(booktabs = TRUE,
+               format = "html",
+              caption = caption_trees,
+             caption.short = caption_trees_short
+             # col.names = c("Subject/Relevance",
+             #              "Desc.",
+             #              "Type",
+             #              "Obsv. (n)",
+             #              "Resolution (m)",
+             #              "Source")
+             ) %>% 
+  kableExtra::row_spec(13, extra_css = "border-bottom: 1px solid")
+
+# %>% 
+#   kableExtra::kable_styling(full_width = TRUE, 
+#                             font_size = 10) %>% 
+#   kableExtra::column_spec(1, italic = TRUE) %>%
+#   kableExtra::row_spec(0, bold = TRUE) %>% 
+#     kableExtra::row_spec(13, hline_after = TRUE)
+
+}
+
+   table_dat_req_kable %>% 
+  kableExtra::kable_styling(full_width = TRUE, 
+                            font_size = 11) %>% 
+  kableExtra::column_spec(1, italic = TRUE) %>%  
+  kableExtra::row_spec(0, italic = FALSE, bold = TRUE, font_size = 12)
+
+```
+
+### Temperature/UHI data sets
 Temperature and UHI data were summarized temporally either by the provider or manually to provide a characteristic representation of heat loading during summer at different times (morning, afternoon/day, night), from which tree averages (radius of 150$~m$) were calculated.
@@
 
-**Ancillary data sets**  
+### Ancillary data sets
 Following the general approach described above, four ancillary dependent variables next to a temperature measure were employed in models; these were chosen due to their availability at high spatial resolution and coverage, and because their influence on growth was previously identified in literature or their likely impact could be deduced using ecophysiological principles.
@@
 
-```{r dsources-tab, eval = TRUE}
+```{r tab-dsources, eval = TRUE}
 
-caption <- "Data description used for generating maps and route optimization.
-Resolution and radius are provided in $m$, the latter referring to the buffer size across which data was averaged to estimate trees' environmental conditions.
-A zero-radius refers to a point extraction from categorical and location specific data.
-**Planting Bed Area, and Soil type are not currently used.**"
+caption <- "Data description used for maps/visualizations and analyses.
+Resolution and radius are provided in $m$, the latter is the buffer in which data was averaged around each trees.
+A zero-radius refers to a point extraction from categorical and location specific data.**"
+caption_ov_short <- "Data descriptions"
 
@@
 ```
+
+## Dendrochronological sampling and analyses
+
+## Statistical Analyses
+
+### General approach: space-for-time analyses
+
+#### Model development and selection
 
@@
 
-Tree locations are clustered and structured based on their category, i.e. riparian, street and park trees (Fig.$~$\@ref(fig:fig-tree-overview-map)).
-Planting in space and time shows species-specific patterns (by districts), often related to major events, such as the start and end of armed and/or political conflict.
-Table$~$\@ref(tab:tab-age) shows the binned distribution of genera across age classes.
-
-```{r tab-age}
-
-drake::loadd(age_tables)
-
-age_tables <- age_tables %>% 
-  dplyr::arrange(desc(n_total))
-colnames(age_tables)[c(1,7,8, 9)] <- c("Genera", "150+", "Total (n)", "Missing (n)")
-
-age_tables <- dplyr::bind_rows(age_tables[age_tables$Genera!="Other",],
-                        age_tables[age_tables$Genera=="Other",])
-
-add_row <- as.data.frame(matrix(c(NA, colSums(age_tables[,-1])), nrow = 1))
-add_row[1,1] <- "Marg. Totals"
-colnames(add_row) <- colnames(age_tables)
-
-age_tables <- dplyr::bind_rows(age_tables,
-                    add_row)
-
-caption_trees <- "Binned age-distribution for trees in Berlin data set, and entries missing age information."
-caption_trees_short <- "Binned age-distribution for trees in Berlin data set."
-
-if (grepl("docx", out_format) ||
-    knitr::is_latex_output() ) {
-  table_dat_req_kable <- age_tables %>% 
-  knitr::kable(booktabs = TRUE,
-               format = "latex",
-             caption = caption_trees,
-             caption.short = caption_trees_short
-             # col.names = c("Subject/Relevance",
-             #              "Desc.",
-             #              "Type",
-             #              "Obsv. (n)",
-             #              "Resolution (m)",
-             #              "Source")
-             ) %>% 
-  kableExtra::row_spec(13, hline_after = TRUE)
-
-} else {
-table_dat_req_kable <- age_tables %>% 
-  knitr::kable(booktabs = TRUE,
-               format = "html",
-              caption = caption_trees,
-             caption.short = caption_trees_short
-             # col.names = c("Subject/Relevance",
-             #              "Desc.",
-             #              "Type",
-             #              "Obsv. (n)",
-             #              "Resolution (m)",
-             #              "Source")
-             ) %>% 
-  kableExtra::row_spec(13, extra_css = "border-bottom: 1px solid")
-
-# %>% 
-#   kableExtra::kable_styling(full_width = TRUE, 
-#                             font_size = 10) %>% 
-#   kableExtra::column_spec(1, italic = TRUE) %>%
-#   kableExtra::row_spec(0, bold = TRUE) %>% 
-#     kableExtra::row_spec(13, hline_after = TRUE)
-
-}
-
-   table_dat_req_kable %>% 
-  kableExtra::kable_styling(full_width = TRUE, 
-                            font_size = 11) %>% 
-  kableExtra::column_spec(1, italic = TRUE) %>%  
-  kableExtra::row_spec(0, italic = FALSE, bold = TRUE, font_size = 12)
-
-```
+<!-- Tree locations are clustered and structured based on their category, i.e. riparian, street and park trees (Fig.$~$\@ref(fig:fig-tree-overview-map)). -->
+<!-- Planting in space and time shows species-specific patterns (by districts), often related to major events, such as the start and end of armed and/or political conflict. -->
 
@@
 1. Temperature, environmental and urban controls on tree growth
+
 	+ Intra specific differences
@@
 	  + root vs. canopy (huber value?)
+	  +  accelerated recent growth
 	  + comparison young vs. old -> increased CO2, perhaps limitation of space for time,

### 2021-11-17 R/functions.R

@@ -2010,8 +2010,9 @@ prep_model_df <- function(dset,
                           path = "./analysis/figures/diagnostic_01_model_data.png"){
 
 
-    dset <- sf::st_drop_geometry(dset) %>%
-        mutate(species_corrected = as.factor(gsub("Tilia intermedia.*", "Tilia intermedia", x = species_corrected)))
+    dset <- sf::st_drop_geometry(dset)
+    # %>%
+        # mutate(species_corrected = as.factor(gsub("Tilia intermedia.*", "Tilia intermedia", x = species_corrected)))
 
     #
     top_species <- dset %>%
@@ -2205,10 +2206,11 @@ make_model_grid <- function(){
 
     model_params <- new.env()
 
-    model_params$k_uni <- 25
+    model_params$k_uni <- 7
     model_params$k_age <- 5
     # model_params$k_te <- c(7, 20)
-    model_params$k_te <- c(5, 15)
+    # model_params$k_te <- c(5, 15)
+    model_params$k_te <- c(5, 7)
     model_params$k_spatial <- 400
     model_params$k_spatial_soilnutmodel <- 250
     model_params$k_soilnut <- 25
@@ -4762,6 +4764,98 @@ make_plot_pairs <- function(dframe,
 
 }
 
+
+
+
+#' BIWI Generate growth ~ age plot
+#'
+#' @param biwi_preds predictions from biwi gam
+#' @param base_size
+#' @param file
+#' @param height
+#' @param width
+#' @param dpi
+#'
+#' @return
+#' @export
+#'
+#' @examples
+make_biwi_plot <- function(biwi_preds,
+                           base_size = 18,
+                           file,
+                           height,
+                           width,
+                           dpi){
+
+
+
+
+    # helper for error propagation
+    serror <- function(x){
+
+        return(sqrt(sum(x^2)/length(x)))
+
+
+        # sqrt(sum(pred.se.fit^2))/n())
+
+    }
+
+    means <- biwi_preds %>%
+        mutate(year_break = as.factor(ifelse(year <= 1960, "1920 - 1960", "1961 - 2001"))) %>%
+        group_by(cambial_age, year_break) %>%
+        summarize(mean_link = mean(fit),
+                  mean_fit_response = Gamma(link="log")$linkinv(mean_link),
+                  mean_se_response_low = Gamma(link="log")$linkinv(mean_link - 1.96 * serror(se.fit)),
[…]

### 2021-11-17 R/plan.R

@@ -184,7 +184,7 @@ plan <- drake::drake_plan(
     # heat polygons based on block level - 20 m captures canopy
     berlin_heat_model = assess_mean_temps(full_data_set_clean,
                                           berlin_heat_model_2015,
-                                          20),
+                                          buff_dist = 150),
 
     # raster has 100 m resolution
     berlin_urbclim_heat_model = assess_mean_temps_urbclim(full_data_set_clean,
@@ -867,16 +867,21 @@ plan <- drake::drake_plan(
                          dpi = 300),
 
 
-   ### BIWI: cambial age vs. annual growth
+   ### BIWI: cambial age vs. annual growth -------------------
 
-   series_long = prep_rwl_data(path_meta_cores = "./analysis/data/raw_data/biwi/BIWi_INV_20190123.xlsx",
-                                path_meta_trees = "./analysis/data/raw_data/biwi/BIWi_INV_20190123.xlsx",
-                                path_meta_sites = "./analysis/data/raw_data/biwi/BIWi_INV_20190123.xlsx",
-                                path_dir_fh = "./analysis/data/raw_data/biwi/rwl/"),
+   series_long = prep_rwl_data(path_meta_cores = drake::file_in("./analysis/data/raw_data/biwi/BIWi_INV_20190123.xlsx"),
+                                path_meta_trees = drake::file_in("./analysis/data/raw_data/biwi/BIWi_INV_20190123.xlsx"),
+                                path_meta_sites = drake::file_in("./analysis/data/raw_data/biwi/BIWi_INV_20190123.xlsx"),
+                                path_dir_fh = drake::file_in("./analysis/data/raw_data/biwi/rwl/")),
 
    biwi_preds = biwi_mod_predict(series_long),
 
-   biwi_plot =
+   biwi_plot = make_biwi_plot(biwi_preds,
+                              base_size = 18,
+                              file = drake::file_out("./analysis/figures/fig-biwi-growth.png"),
+                              height = 8,
+                              width = 8,
+                              dpi = 300),
 
 
    ### SI: Heat comparison ------------------------------
@@ -918,7 +923,7 @@ plan <- drake::drake_plan(
         output_file = file_out("paper_knit.html"),
         output_format = bookdown::html_document2(),
         quiet = TRUE
-    ),
+    )
     # paper_word = rmarkdown::render(
     #     knitr_in("./analysis/paper/paper.Rmd"),
     #     output_dir = "./analysis/paper/",

### 2021-11-23 R/functions.R
updated paper

@@ -4266,7 +4266,7 @@ plot_dbh_temp_single_var <- function(pred_list,
 
         labs(color = "Age Group",
              fill = "Age Group",
-             x = expression(UHI~Magnitude~(degree~C)),
+             x = expression(UHI~Magnitude~(degree*C)),
              y = expression(bar(DBH)~(cm)))
 
 
@@ -4410,7 +4410,7 @@ plot_dbh_temp_single_var_single_species <- function(pred_list,
 
         labs(color = "Species",
              fill = "Species",
-             x = expression(UHI~Magnitude~(degree~C)),
+             x = expression(UHI~Magnitude~(degree*C)),
              y = expression(bar(DBH)~(cm)))
 
 
@@ -4661,6 +4661,8 @@ plot_obs_predicted_model <- function(path_model,
                     height = height,
                     width = width)
 
+    return(gplot)
+
 
 }
 
@@ -4855,6 +4857,540 @@ make_biwi_plot <- function(biwi_preds,
     return(file)
 
 }
+
+
+
+
+## HICAM report ---------------------------------------------------------
+
+
+
+plot_dbh_temp_single_var_GG <- function(pred_list,
+                                     model_df,
+                                     age_filter,
+                                     age_expression,
+                                     base_size = 18,
+                                     prediction_range = "full"){
+
+    `%nin%` <- Negate(`%in%`)
+
+    model_df <- model_df %>%
+        mutate(age_group = eval(age_expression),
+               species_corrected = shorten_species(species_corrected) %>%
+                   as.character()) %>%
+        tidyr::drop_na(age_group) %>%
+        dplyr::mutate(species_corrected =
+                          dplyr::case_when(
+                              species_corrected == "A. hippocastanum" ~ "A. hippo- castanum",
+                              species_corrected == "A. platanoides" ~ "A. plata- noides",
+                              species_corrected == "A. pseudoplatanus" ~ "A. pseudopla- tanus",
+                              species_corrected == "T. platyphyllos" ~ "T. platy- phyllos",
+                              TRUE ~ species_corrected))
+
+
+
+
+    plot_data <- pred_list$pred_groups %>%
+        # dplyr::select(-tempvar) %>%
+        tidyr::pivot_longer(cols = dplyr::all_of(names(pred_list$pred_var)),
+                            names_to = "uhi_tempvar",
+                            values_to = "temp_degc") %>%
[…]

### 2021-11-23 R/plan.R

@@ -896,6 +896,77 @@ plan <- drake::drake_plan(
                    path_out = "./analysis/figures/"),
 
 
+
+
+
+    ### HICAM Report --------------------------------------------------------
+
+   # plot_modperformance_predictions <- make_plot_mod_res(
+   #     path_model = bam_dbh_filtered[bam_dbh_filtered$model=='mI_spatial_age_x_temp_by_species_lcz6_reBEZIRK_var-day_2007', 'model_file_path'],
+   #     file = drake::file_out("./analysis/figures/fig-gam-dbh_temp-day2007_lcz6_obs_pred.png"),
+   #     height = 7,
+   #     width = 8,
+   #     dpi = 300
+   # ),
+
+
+   plot_gam_temp_prediction_genus_comparison_hicam = plot_dbh_temp_single_var_single_species_GG(
+       pred_list = pred_data_single_tempvar_fixed_lcz6,
+       model_df = model_df_stat_filtered,
+       age_filter = c('[45 - 50]', '[75 - 80]', '[90 - 95]'),
+       species_filter = NULL,
+       # species_filter = c("Tilia cordata", "Tilia platyphyllos"),
+       # species_filter = c("Tilia cordata","Platanus acerifolia"),
+       age_expression = age_expr,
+       prediction_range = "within",
+       base_size = 18),
+
+
+   plot_gam_temp_lcz6_prediction_hicam = plot_dbh_temp_single_var_GG(
+       pred_list = pred_data_single_tempvar_fixed_lcz6,
+       model_df = model_df_stat_filtered,
+       age_filter = NULL,
+       age_expression = age_expr,
+       prediction_range = "within",
+       base_size = 18),
+
+
+
+   plot_hicam_temp_effects = plot_hicam_species_overview(plot_ages = plot_gam_temp_lcz6_prediction_hicam,
+                          plot_comparison = plot_gam_temp_prediction_genus_comparison_hicam,
+                          path_out = drake::file_out("analysis/figures/hicam_report/temp_effects_species_overview.png"),
+                          width = 12,
+                          height = 15,
+                          dpi = 300),
+
+
+
+   plot_uhi_map_hicam = make_uhi_plot_hicam(uhi_stacks = uhi_stacks,
+                                berlin_poly = berlin_polygons,
+                                dsf = full_data_set_clean,
+                                base_size = 18),
+
+
+
[…]

### 2021-11-23 analysis/paper/paper.Rmd
updated paper
intro done

@@
 ---
-title: "Species-specific effects of the Urban Heat Island on tree growth across Berlin"
+title: "Assessing the impact of urban heating on tree growth in Berlin with open inventory and environmental data"
 author:
@@
 
-Urban trees show a tendency for enhanced growth rates and/or productivity compared to rural conspecifics, which is typically attributed to increased temperatures [@jia2018; @pretzsch2017], yet feature a broad range of effect size ranges and signs (i.e., reduced growth) specific to species and location.
+Urban trees show a tendency for enhanced growth rates and/or productivity compared to rural conspecifics [e.g., @briber2015; @obrien2012], which is typically attributed to increased temperatures [@jia2018; @pretzsch2017], yet feature a broad range of effect sizes and signs (i.e., reduced growth) specific to species and location.
 @zhao2016 showed that productivity rates, as a proxy for growth, increased within urban clusters as urbanization intensifies using remotely sensed vegetation indices.
-Further, @moser-reischl2019a identified positive associations between air temperature and radial growth for two species (total of 20 individuals) commonly selected by urban planners (*T.
-cordata*, *Rubinia pseudoacacia*) in Munich.
-
-[@briber2015; @obrien2012] … generally increased growth
+Further, @moser-reischl2019a identified positive associations between air temperature and radial growth for two species (total of 20 individuals) commonly selected by urban planners (*Tilia  cordata* MilL., *Rubinia pseudoacacia*) in Munich.
+<!-- [@briber2015; @obrien2012] … generally increased growth -->
 Contrastingly,  @gillner2014 highlight decreased growth for *Acer* species  (*A.
 platanoides* and *pseudoplatanus*), *Platanus x hispanica* and *Quercus rubra* with higher summer temperatures of the preceding year, especially when compounded with drought, in another German metropolis (Dresden).
-@quigley2004 identified absolute growth potential decreased for species between rural and urban conspecifics, yet 
+@quigley2004 identified absolute growth potential decreased for species between rural and urban conspecifics, yet assessments were
 <!-- lacked spatially-explicit effect size estimates across the urban-rural space and  -->
-was limited to comparatively small sample sizes per group ($n_{total}~=~230$ divided in 15 species, 3 groups and 2 locations) - also lacked multi-modal responses
+limited to comparatively small sample sizes per group ($n_{total}~=~230$ divided in 15 species, 3 groups and 2 locations).
 @pretzsch2017 inferred enhanced growth in recent decades and across urban locations spanning several latitudes, including Berlin - however, only 145 individuals of one species (*T.
@@
 As mentioned previously, climate-growth relationships can vary substantially between species, and in fact, @quigley2004 and @pretzsch2017 report contrasting results regarding average tree diameter, i.e.
-smaller or larger for urban $vs.$ rural trees of same age, **consider other studies mentioned above**
-Similarly, for Berlin, @dahlhausen2018, identified enhanced growth in highly urbanized environments (using basal area increments of 252 trees) for *Tilia cordata* Mill, the most abundant tree of the city, which they attributed to the UHI effect, while intermediate development intensity was adverse for tree growth.
-
+smaller or larger for urban $vs.$ rural trees of same age.
+Similarly, for Berlin, @dahlhausen2018, identified enhanced growth in highly urbanized environments (using basal area increments of a large sample of 252 trees) for *T.
+cordata*, the most abundant tree of the city, which they attributed to the UHI effect, while intermediate development intensity was adverse for tree growth.
 These differences in growth trends may result from contrasting species-specific responses to increased temperatures, but are indeed affected by other (time-varying) factors and stochastic processes, such as water availability, pollution and road-salt loading, structural impedance by infrastructure, or management, etc.
 [@rhoades1999; @quigley2004; @randrup2001; @pauleit2002].
-**expand on location-specific factors that can affect growth**
-This can hinder the extrapolation from individual sampling sites toward predicting effect across entire urban areas and tree stocks, especially where studies rely on labor-intensive methods which are limited logistically by sampling effort, reducing sample sizes, and species as well as spatial coverage.
-This is exacerbated by a lack of co-located environmental variables (i.e.
+Further, the variability in responses may require that assessments are developed for a specific region, because well-understood tree characteristics [e.g., see @brune2016;@roloff2009], could be strongly modulated predictably due to management, planting practices, or other environmental controls; for example, if drought hardiness is related to extensive root networks, restricted soil volumes available to street trees will render a species more vulnerable to water stress.
+<!-- **expand on location-specific factors that can affect growth** -->
+
+<!-- This can hinder the extrapolation from individual sampling sites toward predicting effect across entire urban areas and tree stocks, especially where studies rely on labor-intensive methods which are limited logistically by sampling effort, reducing sample sizes, and species as well as spatial coverage.
+-->
+<!-- This is exacerbated by a lack of co-located environmental variables (i.e.
 measured in situ) at pertinent spatial scales, for instance, as noted by @wohlfahrt2019 for air temperature and tree leaf phenology, which may lead to incorrect inferences and interpretations for the role of climate change on growth/productivity when applying space-for-time substitutions.
-
-Space-for-time substitutions are a common approach (cf.
-studies above) to generate inferences in observational (rather than treatment-control) studies, where manipulations are costly or logistically infeasible due to time and/or financial constraints.
+-->
+Space-for-time substitutions and time series comparisons between and within locations are a common approach (cf.
+studies above) to generate inferences in observational (rather than treatment-control) studies, where manipulations are costly or logistically unfeasible due to time and/or financial constraints.
 However, they require  accounting for confounding factors specific to trees' environments, such as street characteristics, development intensity, available soil volume, etc.
-While several of the aforementioned studies applied the space-for-time approach to quantify temperature and excess heat on growth, they typically compare trees grouped using qualitative or summary descriptors of sampling sites, disregarding the spatio-temporal variability in location-specific factors noted above.
-This can hinder the extrapolation from individual sampling sites toward predicting effects across entire urban areas and tree stocks, especially when studies rely on labor-intensive methods which are limited logistically by sampling effort, reducing sample sizes, and species as well as spatial coverage.
+While several of the aforementioned studies applied these approaches to quantify temperature and excess heat on growth, they typically compare trees grouped using qualitative or summary descriptors of sampling sites, disregarding the spatio-temporal variability in location-specific factors noted above.
+This can hinder the extrapolation from individual sampling sites toward predicting effects across entire urban areas and tree stocks, especially when studies rely on labor-intensive methods, which are limited logistically by sampling effort, reducing sample sizes and coverage of species and space.
 This can be  exacerbated by a lack of co-located environmental variables (i.e.
 measured in situ) at pertinent spatial scales, for instance, as noted by @wohlfahrt2019 for air temperature and tree leaf phenology, which may lead to incorrect inferences and interpretations for the role of climate change on growth/productivity when applying space-for-time substitutions.
-It is thus likely that the varying and even contrasting growth responses observed for urban trees across and within studies is at least modulated by several confounding factors, making the attribution to a single driver, such as excess heat, more difficult and possibly less accurate.
+It is thus likely that the varying and even contrasting growth responses observed for urban trees across and within studies are at least modulated by some confounding factors, making the attribution to a single driver, such as excess heat, more difficult and possibly less accurate.
 
-These limitations could be overcome by relying on large data sets with extensive ancillary data coverage. 
-Berlin, as one of the greenest cities in Europe, provides an openly accessible tree inventory, with spatio-temporal environmental data sets pertinent to tree growth, through its Environmental Agency. 
-It features a total of 650000 individuals covering 94 genera and at least 600 species and/or cultivars, listing information on location, trunk diameter (at breast height; $DBH$; see Tab.$~$\@ref(tab:tab-tree-overview)), and stem height, amongst other variables, for the majority of street and park trees.
-For this study, our object was to assess the impact of excess urban heat, i.e.
+These limitations could be overcome by relying on dendroecological surveys (i.e., incremental growth) or inventories (single or repeat) combined with pertinent environmental data with adequate spatio-temporal coverage and resolution.
+Such data are increasingly more available due to open data policies and because their value is recognized across several domains [e.g. @ossola2020].
+Berlin, as one of the greenest cities in Europe, provides an openly accessible tree inventory, with spatio-temporal environmental data sets relevant to tree growth. 
+It features a total of 650000 individuals covering 94 genera and at least 600 species and/or cultivars, listing information on location, stem diameter (at breast height; $DBH$), and stem height, amongst other variables, for the majority of street and park trees.
+For this study, our objective was to assess the impact of excess urban heat, i.e.
 the UHI effect, on tree growth using this openly available inventory data set, complemented by additional open data sources as well as incremental growth data from tree cores.
 The assessment relied on flexible statistical models that could capture species and location-specific responses to heat and other urban factors.
-Specifically, we aimed to (1) assess heat exposure of most abundant species; (2) determine the impact of (excess) heat on stem growth with space-for-time substitution to increase coverage; (3) h ighlight the role of location-specific environmental factors in mediating temperature responses.
-While this work constitutes a case study for Berlin, we believe the results are a valuable contribution toward Berlin's current and future management of its tree stock. 
-This is because the high temporal and spatial variability of controls on growth, and thus variability in responses, require that assessments are developed for a specific region, because well-understood tree characteristics [e.g., see @brune2016;@roloff2009], could be strongly modulated  in predicable ways at a given location or time due to management, planting practices, or other environmental controls; for example, if drought hardiness is related to extensive root networks, small soil volumes available to street trees could render a species vulnerable to water stress.
-As such, the approach can also serve as a framework for base-line assessments for other cities with available or growing inventory and ancillary data [cite something on tree diversity in tree stocks, management, etc.], and also inform species-climate matrices regarding temperature sensitivity…..
-
+Specifically, we aimed to (1) assess heat exposure of most abundant species; (2) determine the impact of (excess) heat on stem growth with space-for-time substitution to increase coverage; (3) highlight the role of location-specific environmental factors in mediating temperature responses.
+<!-- This work constitutes a case study for Berlin, the results are a valuable contribution toward Berlin's current and future management of its tree stock.  -->
+Our results are a contribution toward Berlin's current and future management of its tree stock and may help drive adaptation to climate change.
+Despite being a case study for a single city, we believe our work may provide a flexible approach for other cities with available or growing inventories, as well as ancillary environmental data, and may also inform the use of other planning tools, such as species-climate matrices [@roloff2009] regarding temperature sensitivity.
+<!-- This is because the high temporal and spatial variability of controls on growth, and thus variability in responses, require that assessments are developed for a specific region, because well-understood tree characteristics [e.g., see @brune2016;@roloff2009], could be strongly modulated  in predicable ways at a given location or time due to management, planting practices, or other environmental controls; for example, if drought hardiness is related to extensive root networks, small soil volumes available to street trees could render a species vulnerable to water stress.
+-->
+<!-- As such, the approach can also serve as a framework for base-line assessments for other cities with available or growing inventory and ancillary data [cite something on tree diversity in tree stocks, management, etc.], and also inform species-climate matrices regarding temperature sensitivity -->
 <!-- Berlin features the most intense Urban Heat Island (UHI) in Germany due to its large extent and development intensity [@kuttler2015], with temperature increases of up to $12~K$ during day-time and $6~K$ on average for night-times [2001-2010, @fenner2014] in urban $vs.$ rural areas.
@@
 -->
-
 <!-- Trees, in particular, provide shading as well as transpirative cooling in their vicinity [@gillner2015;@oke1982;@endlicher2016], and therefore can reduce ambient temperatures, infrastructure power-consumption and (human) thermal discomfort  [e.g.
@@
 -->
-
 <!-- Trees and green infrastructure in urban areas show a tendency for enhanced growth rates and/or productivity compared to rural counterparts [@jia2018; @pretzsch2017], yet feature a broad range of effect size ranges and, in some cases, signs specific to species and locality.
@@
 -->
-<!-- <!-- For Berlin, @dahlhausen2018, showed divergent patterns for young and old trees dependent on urban development intensity for *Tilia cordata* Mill, the most abundant tree of the city, but overall reported an increased growth rate which they attributed to the UHI effect.
---> -->
+<!--  For Berlin, @dahlhausen2018, showed divergent patterns for young and old trees dependent on urban development intensity for *Tilia cordata* Mill, the most abundant tree of the city, but overall reported an increased growth rate which they attributed to the UHI effect.
+-->
 <!-- Further, @moser-reischl2019a identified positive associations between air temperature and radial growth for two species commonly selected by urban planners (*T.
@@
 -->
-
 <!-- Conditions affecting tree growth can vary greatly within urban areas or regions, and need to be accounted for when establishing relationships with pertinent drivers, such as the UHI effect. -->
@@
 <!-- This is especially the case for studies reliant on labour-intensive methods which are limited logistically by sampling effort, reducing sample sizes, as well as species and spatial coverage. -->
-<!-- <!-- While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e.
+<!-- While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e.
 genera and/or species, and space).
---> -->
-<!-- <!-- However, urban tree growth in urban settings is affected by various processes, which  --> -->
-<!-- <!-- However, growth in urban settings is affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc.
+-->
+<!--  However, urban tree growth in urban settings is affected by various processes, which  --> 
+<!--  However, growth in urban settings is affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc.
 [@rhoades1999; @quigley2004; @randrup2001; @pauleit2002].
---> -->
-<!-- <!-- While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e.
+-->
+<!-- While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e.
 genera and/or species, and space).
---> -->
-<!-- <!-- This typically hinders the extrapolation from individual sampling sites toward prediction of effect sizes across entire urban areas and tree stocks.  --> -->
-<!-- <!-- Therefore, studies reliant on labor-intensive sampling, must carefully address confounding factors to ensure effect sizes are estimated adequately. --> -->
+-->
+<!-- This typically hinders the extrapolation from individual sampling sites toward prediction of effect sizes across entire urban areas and tree stocks.  --> 
+<!--  Therefore, studies reliant on labor-intensive sampling, must carefully address confounding factors to ensure effect sizes are estimated adequately. -->
 
-<!-- <!-- Detailed, but spatially-limited dendroecological analyses of climate-growth relationships in Berlin for key species can be enhanced by inferring growth modulation from a large data set in excess of 650000 individuals provided by the Berlin Senate Administration (Senatsverwaltung).
---> -->
+<!--  Detailed, but spatially-limited dendroecological analyses of climate-growth relationships in Berlin for key species can be enhanced by inferring growth modulation from a large data set in excess of 650000 individuals provided by the Berlin Senate Administration (Senatsverwaltung).
+-->
 

### 2021-11-23 analysis/paper/paper.Rmd
methods - street trees

@@
 For this study, our objective was to assess the impact of excess urban heat, i.e.
-the UHI effect, on tree growth using this openly available inventory data set, complemented by additional open data sources as well as incremental growth data from tree cores.
+the UHI effect, on tree growth ($DBH$) using this openly available inventory data set, complemented by additional open data sources as well as incremental growth data from tree cores.
 The assessment relied on flexible statistical models that could capture species and location-specific responses to heat and other urban factors.
@@
 
-In a space-for-time substitution, growth of individual species can be assessed across the entire city of Berlin, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.
-Comparable applications are found, for example, in @quigley2004 and @pretzsch2017.
-The former  inferred absolute growth potential for species across successional groups (early, mid, late stage), and between rural and urban conspecifics, yet lacked spatially-explicit effect size estimates across the urban-rural space and was limited to comparatively small sample sizes per group ($n_{total}~=~230$ divided in 15 species, 3 groups and 2 locations).
-@pretzsch2017 applied linear hierarchical models to infer growth modulation on annual basis for different time periods and urban $vs.$ rural locations while accounting for stand-level variability;
-however, for Berlin only 145 individuals of one species (*T. cordata*) were assessed.
-As mentioned previously, climate-growth relationships can vary substantially between species, and in fact, @quigley2004 and @pretzsch2017 report contrasting results regarding average tree diameter, i.e.
+<!-- In a space-for-time substitution, growth of individual species can be assessed across the entire city of Berlin, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.
+-->
+<!-- Comparable applications are found, for example, in @quigley2004 and @pretzsch2017. -->
+<!-- The former  inferred absolute growth potential for species across successional groups (early, mid, late stage), and between rural and urban conspecifics, yet lacked spatially-explicit effect size estimates across the urban-rural space and was limited to comparatively small sample sizes per group ($n_{total}~=~230$ divided in 15 species, 3 groups and 2 locations).
+-->
+<!-- @pretzsch2017 applied linear hierarchical models to infer growth modulation on annual basis for different time periods and urban $vs.$ rural locations while accounting for stand-level variability; -->
+<!-- however, for Berlin only 145 individuals of one species (*T. cordata*) were assessed. -->
+<!-- As mentioned previously, climate-growth relationships can vary substantially between species, and in fact, @quigley2004 and @pretzsch2017 report contrasting results regarding average tree diameter, i.e.
 smaller or larger for urban $vs.$ rural trees of same age.
-Consequently, this variability of effect sizes and directions calls for a more comprehensive assessment across species and with greater spatial coverage throughout Berlin.
+-->
+<!-- Consequently, this variability of effect sizes and directions calls for a more comprehensive assessment across species and with greater spatial coverage throughout Berlin. -->
 
-**We therefore propose applying a statistical model that is fully spatially explicit, while also allowing to account for the nested nature of the data set (e.g.
+<!-- **We therefore propose applying a statistical model that is fully spatially explicit, while also allowing to account for the nested nature of the data set (e.g.
 streets and districts) as well as other pertinent factors using hierarchical, generalized additive models (see Section$~$\@ref(sec:methods)).
-As a result, the absolute growth potential of a species can be inferred given, for example, a specific location, age or UHI magnitude. 
-Further, the impact of UHI loading can be predicted for a single species across all of Berlin as a continuous surface.**
-The inclusion of independent, tree-level growth data, however, is paramount as it allows:
-
-1) validating the publicly-available Senate dataset ($DBH$) while providing a more reliable estimate of tree age,  
-2) applying the model with annual basal area as a response (enabling incorporating effects of varying climate/UHI intensities over time), and  
-3) comparing $DBH$ (Senate data) and basal area-derived models for a subset of locations and species, increasing confidence in Berlin-wide predictions.
-
+-->
+<!-- As a result, the absolute growth potential of a species can be inferred given, for example, a specific location, age or UHI magnitude.  -->
+<!-- Further, the impact of UHI loading can be predicted for a single species across all of Berlin as a continuous surface.** -->
+<!-- The inclusion of independent, tree-level growth data, however, is paramount as it allows:   -->
+<!-- 1) validating the publicly-available Senate dataset ($DBH$) while providing a more reliable estimate of tree age,   -->
+<!-- 2) applying the model with annual basal area as a response (enabling incorporating effects of varying climate/UHI intensities over time), and   -->
+<!-- 3) comparing $DBH$ (Senate data) and basal area-derived models for a subset of locations and species, increasing confidence in Berlin-wide predictions.  -->
 <!-- We acknolwedge results from [@gregg2003], which indicated impeded growth adjacent to intense urban clusters, rather than enhancement within them. -->
@@
 
-## Data sources
+## General approach: space-for-time analyses
 
-An overview of data used for models, including sources, types, and application, is provided in Table$~$\@ref(tab:tab-dsources).
+Species-specifc growth responses in relation across Berlin were assessed in relat
+
+We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of excess heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; LandSat-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section$~$\@ref(stat-gam) for details).
+Covariates were extracted at 150 and 300$~m$ to infer the impact of reference scale of the urban fabric on tree growth. 
+From all tested models the most suitable (i.e., parsimonious with highest explanatory) was employed for further analyses.
+An overview of data used for models, including sources, types, and application, is provided in Section$~$\@ref(dsources) and in Table$~$\@ref(tab:tab-dsources).
+
+## Data sources {#dsources}
+
+<!-- Berlin’s open data provided tree inventories (species, age, location, diameter which were processed with a bespoke software datacleanr by Hurley et al., submitted), planting bed area, estimates of soil nutrient availability, and adjacent building height.
+With WUDAPT local climate zones (Demuzere et al., 2019), these were used to determine trees’ growing conditions (locally, or within a radius of 150 m / 300 m).
+-->
 
 ### Street trees
-Characteristics of street trees and their respective environments were collated from the city's tree inventory, as well as various elements of the Berlin Environment Atlas, which are openly available via https://daten.berlin.de and curated by the city's government.
-The data used from the inventory includes species, location, age, and diameter at breast height (dbh).
-**quality control** - filtered dataset to remove drastic outliers --> fit poly model, used Median Absolute deviance with multiplier of X to exclude points. 
-Focused on 9 most abundant species to and Cultivars and hybrids were combined to increase sample sizes. **check which ones**
 
+Berlin’s open data provided tree inventories including species, age, location, and circumference.
+Unplausible observations, likely from erroneous data entry, were removed.
+Additional manual data processing for quality control was done with a bespoke software  `datacleanr` by Hurley *et al*.
+(submitted), where obvious outliers or clearly interpolated data were removed; the latter was deemed necessary, as several observations in multiple city districts were derived by a linear relationship, which does not captures the ontogenic growth dynamics of trees, while leaving no unexplained variation related to variables other than age.
+All of these operations were recorded, and can be viewed and reproduced via the supplementary code.
+Lastly, observations with unlikely diameter-age combinations were identified via the residuals of a generalized linear model between diameter and age with a Gamma log-link distribution: if individual residuals exceeded seven times the median absolute deviation of all residuals, they were removed.
+The median absolute deviation (MAD) is comparable to the inter quartile range, yet more robust to outliers:
+
+$$MAD = median(|X_i - median(X)|) $$
+This approach is considered conservative (see supplementary information), yet all analyses were carried forward with the filtered and unfiltered data set.
+No considerable differences were found for the unfiltered and filtered data, yet all subsequent sections use the latter.
+<!-- which were processed with the bespoke software datacleanr by Hurley et al., submitted), -->
+<!-- Characteristics of street trees and their respective environments were collated from the city's tree inventory, as well as various elements of the Berlin Environment Atlas, which are openly available via https://daten.berlin.de and curated by the city's government.
+-->
+<!-- The data used from the inventory includes species, location, age, and diameter at breast height (dbh). -->
+<!-- **quality control** - filtered dataset to remove drastic outliers --> fit poly model, used Median Absolute deviance with multiplier of X to exclude points.  -->
 Table$~$\@ref(tab:tab-age) shows the binned distribution of genera across age classes.
+Final sample sizes
 
@@
 
-## Statistical Analyses
+## Statistical Analyses {#stat-gam}
 

### 2021-11-23 analysis/paper/paper.Rmd
cont methods

@@
 $$MAD = median(|X_i - median(X)|) $$
-This approach is considered conservative (see supplementary information), yet all analyses were carried forward with the filtered and unfiltered data set.
-No considerable differences were found for the unfiltered and filtered data, yet all subsequent sections use the latter.
+This approach is considered conservative (see supplementary information), yet all analyses were carried forward with the unfiltered and filtered data - no considerable differences were found, thus subsequent sections are based on the filtered data.
 <!-- which were processed with the bespoke software datacleanr by Hurley et al., submitted), -->
@@
 <!-- The data used from the inventory includes species, location, age, and diameter at breast height (dbh). -->
-<!-- **quality control** - filtered dataset to remove drastic outliers --> fit poly model, used Median Absolute deviance with multiplier of X to exclude points.  -->
 Table$~$\@ref(tab:tab-age) shows the binned distribution of genera across age classes.
-Final sample sizes
+Final samples applied in models were smaller, following the availability of ancillary data for a given observation, and limited to a maximum age of 125 years to increase confidence in model estimates.
 

### 2021-11-23 analysis/paper/paper.Rmd
methods-data

@@
 
-Berlin’s open data provided tree inventories including species, age, location, and circumference.
-Unplausible observations, likely from erroneous data entry, were removed.
+Berlin’s open data provided tree inventories including species, age, location, and circumference which was transformed into diameter.
+Implausible observations, likely from erroneous data entry, were removed.
 Additional manual data processing for quality control was done with a bespoke software  `datacleanr` by Hurley *et al*.
-(submitted), where obvious outliers or clearly interpolated data were removed; the latter was deemed necessary, as several observations in multiple city districts were derived by a linear relationship, which does not captures the ontogenic growth dynamics of trees, while leaving no unexplained variation related to variables other than age.
+(submitted), where obvious outliers or clearly interpolated data were removed; the latter was deemed necessary, as several observations in multiple city districts were derived by linear relationships, which does not capture the ontogenic growth dynamics of trees, and leaves no variation related to variables other than age.
 All of these operations were recorded, and can be viewed and reproduced via the supplementary code.
@@
 Temperature and UHI data were summarized temporally either by the provider or manually to provide a characteristic representation of heat loading during summer at different times (morning, afternoon/day, night), from which tree averages (radius of 150$~m$) were calculated.
-Two data sets of urban air and one surface temperature were tested as explanatory variables in GAMM models.
-The air temperatures from the Berlin environmental atlas (EnvAt) were model outputs that are representations of typical summer conditions at 0400, 1400 and 2200 hours; these data are provided at city block basis (spatial polygons), from which weighted averages were extracted.
-UrbClim air temperatures are hourly model outputs (100$~m$ resolution, **ref**) based on ERA5 re-analyses data (ECMWF) for which data of the hottest month in the record (June, 2011) were averaged to hours equivalent to EnvAt data by using a window of $\pm~1~$hour (i.e., 0300 to 0500, etc.).
-Subsequently, a land-use and landcover mask (2018,      Copernicus CORINE **ref**) was used to define urban and rural/forested areas, by which the urban heat loading was calculated as $UHI_{x,y} = T_{Air_{2m}~x,y} - \overline{T_{Air_{2m}~Rural}}$, where $x$ and $y$ define an urban grid cell.
+Two data sets of urban air and one surface temperature were tested as explanatory variables in GAM models.
+The air temperatures from the Berlin environmental atlas (EnvAt) are model outputs that are representations of typical summer conditions at 0400, 1400 and 2200 hours; these data are provided at city block basis (spatial polygons), from which weighted averages were extracted.
+UrbClim air temperatures are hourly model outputs [100$~m$ resolution; @deridder2015]) based on ERA5 re-analyses data (ECMWF) for which observations from the hottest month  available (June, 2011) were averaged to hours equivalent to EnvAt data by using a window of $\pm~1~$hour (i.e., 0300 to 0500, etc.).
+Subsequently, a land-use and land-cover mask (CORINE; European Union, Copernicus Land Monitoring Service 2018, European Environment Agency) was used to define urban and rural/forested areas.
+This was deemed reasonable as Berlin's built-up area has not changed markedly over the past 50 years, i.e., about 52 to 61$\%$ [@mohamed2017]. 
+The urban heat loading was then calculated as $UHI_{x,y} = T_{Air_{2m}~x,y} - \overline{T_{Air_{2m}~Rural}}$, where $x$ and $y$ define an urban grid cell.
 The surface UHI data set by @chakraborty2019 derives its measure in a similar fashion and the reader is referred to the detailed description therein; note this data set provides day and night-time averaged UHI estimates at 500$~m$ resolution, which were extracted for the hottest summer in this record (2007).
@@
 ### Ancillary data sets
-Following the general approach described above, four ancillary dependent variables next to a temperature measure were employed in models; these were chosen due to their availability at high spatial resolution and coverage, and because their influence on growth was previously identified in literature or their likely impact could be deduced using ecophysiological principles.
-We included planting bed area, the sum of exchangeable basic cation as a proxy for soil nutrient availability, adjacent building height,  and the proportional coverage of local climate zone 6 (LCZ6; open mid-rise, see @demuzere2019 and @stewart2012 for details).
-The latter was chosen as an increase reflects a transition away from densely urbanized areas.
-Note that only street trees in urban, not rural areas or within greenspaces, were considered here, but individual trees may grow along streets adjacent to greenspaces and parks of varying sizes.
+Following the general approach described above, four ancillary covariates next to a temperature measure were employed in models; these were chosen due to their availability at high spatial resolution and coverage, and because their influence on growth was previously identified in literature or their likely impact could be deduced using ecophysiological principles.
+We included planting bed area and the sum of exchangeable basic cation as a proxy for soil nutrient availability (point extractions), as well as the proportional coverage of local climate zone 6 (LCZ6; open mid-rise, see @demuzere2019 and @stewart2012 for details) and adjacent building height (spatial averages).
+The latter was chosen as an increase reflects a transition away from densely urbanized areas and had the highest coverage for the processed tree inventory. 
+Note that only street trees in urban, not rural areas or within green spaces, were considered here, but individual trees may grow along streets adjacent to green spaces and parks of varying sizes.
 
-were considered to have 
-The environment atlas also provided soil nutrient exchange capacity as a measure of growth-favoring soil conditions,
-
-surface area of tree planting beds ("Baumscheibe"), as well as general characteristics of the urban environment like building height.
-It also includes spatially-explicit model outputs for local climate simulations representative of typical conditions, such as air temperature (2 m) on a 2015 summer's day at different hours (0400, 1400, 2200).
-These model outputs were preferred over *in situ* measurements as **more representative based on averaged conditions, physiologically more relevant than surface urban heat island products**.
-These outputs are available at city block or planning unit level, and were used as a proxy for the general temperature distribution influenced by urbanization across the investigated growth period across the previous 80 years (see section X for details).
-This was deemed reasonable as Berlin's built-up area has not changed markedly over the past 50 years, i.e., about 52 to 61$\%$ [@mohamed2017]. 
-Further insight into the role of the urban environment was assessed by including local climate zone classes derived by WUDAPT.
-These describe the proportion of urban area with a specific cover, such as dense high rises, sealed surfaces, etc. 
-For this study, the LCZ 6, representing dense mid-rise built up cover was used as a proxy for the degree of urbanization, and is the most frequent LCZ for Berlin [@fenner2017] (found ref in .
-Use percent sealed surface instead?
+<!-- surface area of tree planting beds ("Baumscheibe"), as well as general characteristics of the urban environment like building height. -->
+<!-- It also includes spatially-explicit model outputs for local climate simulations representative of typical conditions, such as air temperature (2 m) on a 2015 summer's day at different hours (0400, 1400, 2200).
+-->
+<!-- These model outputs were preferred over *in situ* measurements as **more representative based on averaged conditions, physiologically more relevant than surface urban heat island products**. -->
+<!-- These outputs are available at city block or planning unit level, and were used as a proxy for the general temperature distribution influenced by urbanization across the investigated growth period across the previous 80 years (see section X for details).
+-->
+<!-- This was deemed reasonable as Berlin's built-up area has not changed markedly over the past 50 years, i.e., about 52 to 61$\%$ [@mohamed2017].  -->
+<!-- Further insight into the role of the urban environment was assessed by including local climate zone classes derived by WUDAPT. -->
+<!-- These describe the proportion of urban area with a specific cover, such as dense high rises, sealed surfaces, etc.  -->
+<!-- For this study, the LCZ 6, representing dense mid-rise built up cover was used as a proxy for the degree of urbanization, and is the most frequent LCZ for Berlin [@fenner2017] (found ref in . -->
+<!-- Use percent sealed surface instead? -->
 

### 2021-11-24 analysis/paper/paper.Rmd
methods - dendro

@@
 header-includes:
+- \usepackage{amsmath}
 - \usepackage{booktabs}
@@
 
+(ref:study-area) Berlin's generalized land-use derived from @suvkberlin2019 and location within the European context (inset).
+
+```{r fig-study-area, fig.cap="(ref:study-area)",fig.width = 12, fig.height = 14, out.width="50%"}
+
+knitr::include_graphics(here::here("analysis/figures/map_00_studyarea.png"))
+
+# plot_count_map
+
+# plot(1:10)
+```
+
 ## General approach: space-for-time analyses
 
-Species-specifc growth responses in relation across Berlin were assessed in relat
-
+<!-- Species-specifc growth responses in relation across Berlin were assessed in relat -->
 We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of excess heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; LandSat-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section$~$\@ref(stat-gam) for details).
@@
 From all tested models the most suitable (i.e., parsimonious with highest explanatory) was employed for further analyses.
-An overview of data used for models, including sources, types, and application, is provided in Section$~$\@ref(dsources) and in Table$~$\@ref(tab:tab-dsources).
 
@@
 
-### Street trees
-
-Berlin’s open data provided tree inventories including species, age, location, and circumference which was transformed into diameter.
-Implausible observations, likely from erroneous data entry, were removed.
-Additional manual data processing for quality control was done with a bespoke software  `datacleanr` by Hurley *et al*.
-(submitted), where obvious outliers or clearly interpolated data were removed; the latter was deemed necessary, as several observations in multiple city districts were derived by linear relationships, which does not capture the ontogenic growth dynamics of trees, and leaves no variation related to variables other than age.
-All of these operations were recorded, and can be viewed and reproduced via the supplementary code.
-Lastly, observations with unlikely diameter-age combinations were identified via the residuals of a generalized linear model between diameter and age with a Gamma log-link distribution: if individual residuals exceeded seven times the median absolute deviation of all residuals, they were removed.
-The median absolute deviation (MAD) is comparable to the inter quartile range, yet more robust to outliers:
-
-$$MAD = median(|X_i - median(X)|) $$
-This approach is considered conservative (see supplementary information), yet all analyses were carried forward with the unfiltered and filtered data - no considerable differences were found, thus subsequent sections are based on the filtered data.
-<!-- which were processed with the bespoke software datacleanr by Hurley et al., submitted), -->
-<!-- Characteristics of street trees and their respective environments were collated from the city's tree inventory, as well as various elements of the Berlin Environment Atlas, which are openly available via https://daten.berlin.de and curated by the city's government.
--->
-<!-- The data used from the inventory includes species, location, age, and diameter at breast height (dbh). -->
-Table$~$\@ref(tab:tab-age) shows the binned distribution of genera across age classes.
-Final samples applied in models were smaller, following the availability of ancillary data for a given observation, and limited to a maximum age of 125 years to increase confidence in model estimates.
-
-```{r tab-age}
-
-drake::loadd(age_tables)
-
-age_tables <- age_tables %>% 
-  dplyr::arrange(desc(n_total))
-colnames(age_tables)[c(1,7,8, 9)] <- c("Genera", "150+", "Total (n)", "Missing (n)")
-
-age_tables <- dplyr::bind_rows(age_tables[age_tables$Genera!="Other",],
-                        age_tables[age_tables$Genera=="Other",])
-
-add_row <- as.data.frame(matrix(c(NA, colSums(age_tables[,-1])), nrow = 1))
-add_row[1,1] <- "Marg. Totals"
-colnames(add_row) <- colnames(age_tables)
-
-age_tables <- dplyr::bind_rows(age_tables,
-                    add_row)
-
-caption_trees <- "Binned age-distribution for trees in Berlin data set, and entries missing age information. "
-caption_trees_short <- "Binned age-distribution for trees in Berlin data set."
-
-if (grepl("docx", out_format) ||
-    knitr::is_latex_output() ) {
-  table_dat_req_kable <- age_tables %>% 
-  knitr::kable(booktabs = TRUE,
-               format = "latex",
-             caption = caption_trees,
-             caption.short = caption_trees_short
-             # col.names = c("Subject/Relevance",
-             #              "Desc.",
-             #              "Type",
-             #              "Obsv. (n)",
-             #              "Resolution (m)",
-             #              "Source")
-             ) %>% 
-  kableExtra::row_spec(13, hline_after = TRUE)
-
-} else {
-table_dat_req_kable <- age_tables %>% 
-  knitr::kable(booktabs = TRUE,
-               format = "html",
-              caption = caption_trees,
-             caption.short = caption_trees_short
-             # col.names = c("Subject/Relevance",
-             #              "Desc.",
-             #              "Type",
-             #              "Obsv. (n)",
-             #              "Resolution (m)",
-             #              "Source")
-             ) %>% 
-  kableExtra::row_spec(13, extra_css = "border-bottom: 1px solid")
-
-# %>% 
-#   kableExtra::kable_styling(full_width = TRUE, 
-#                             font_size = 10) %>% 
-#   kableExtra::column_spec(1, italic = TRUE) %>%
-#   kableExtra::row_spec(0, bold = TRUE) %>% 
-#     kableExtra::row_spec(13, hline_after = TRUE)
-
-}
-
-   table_dat_req_kable %>% 
-  kableExtra::kable_styling(full_width = TRUE, 
-                            font_size = 11) %>% 
-  kableExtra::column_spec(1, italic = TRUE) %>%  
-  kableExtra::row_spec(0, italic = FALSE, bold = TRUE, font_size = 12)
-
-```
-
-### Temperature/UHI data sets
-Temperature and UHI data were summarized temporally either by the provider or manually to provide a characteristic representation of heat loading during summer at different times (morning, afternoon/day, night), from which tree averages (radius of 150$~m$) were calculated.
-Two data sets of urban air and one surface temperature were tested as explanatory variables in GAM models.
-The air temperatures from the Berlin environmental atlas (EnvAt) are model outputs that are representations of typical summer conditions at 0400, 1400 and 2200 hours; these data are provided at city block basis (spatial polygons), from which weighted averages were extracted.
-UrbClim air temperatures are hourly model outputs [100$~m$ resolution; @deridder2015]) based on ERA5 re-analyses data (ECMWF) for which observations from the hottest month  available (June, 2011) were averaged to hours equivalent to EnvAt data by using a window of $\pm~1~$hour (i.e., 0300 to 0500, etc.).
-Subsequently, a land-use and land-cover mask (CORINE; European Union, Copernicus Land Monitoring Service 2018, European Environment Agency) was used to define urban and rural/forested areas.
-This was deemed reasonable as Berlin's built-up area has not changed markedly over the past 50 years, i.e., about 52 to 61$\%$ [@mohamed2017]. 
-The urban heat loading was then calculated as $UHI_{x,y} = T_{Air_{2m}~x,y} - \overline{T_{Air_{2m}~Rural}}$, where $x$ and $y$ define an urban grid cell.
-The surface UHI data set by @chakraborty2019 derives its measure in a similar fashion and the reader is referred to the detailed description therein; note this data set provides day and night-time averaged UHI estimates at 500$~m$ resolution, which were extracted for the hottest summer in this record (2007).
-
-### Ancillary data sets
-Following the general approach described above, four ancillary covariates next to a temperature measure were employed in models; these were chosen due to their availability at high spatial resolution and coverage, and because their influence on growth was previously identified in literature or their likely impact could be deduced using ecophysiological principles.
-We included planting bed area and the sum of exchangeable basic cation as a proxy for soil nutrient availability (point extractions), as well as the proportional coverage of local climate zone 6 (LCZ6; open mid-rise, see @demuzere2019 and @stewart2012 for details) and adjacent building height (spatial averages).
-The latter was chosen as an increase reflects a transition away from densely urbanized areas and had the highest coverage for the processed tree inventory. 
-Note that only street trees in urban, not rural areas or within green spaces, were considered here, but individual trees may grow along streets adjacent to green spaces and parks of varying sizes.
-
-<!-- surface area of tree planting beds ("Baumscheibe"), as well as general characteristics of the urban environment like building height. -->
-<!-- It also includes spatially-explicit model outputs for local climate simulations representative of typical conditions, such as air temperature (2 m) on a 2015 summer's day at different hours (0400, 1400, 2200).
--->
-<!-- These model outputs were preferred over *in situ* measurements as **more representative based on averaged conditions, physiologically more relevant than surface urban heat island products**. -->
-<!-- These outputs are available at city block or planning unit level, and were used as a proxy for the general temperature distribution influenced by urbanization across the investigated growth period across the previous 80 years (see section X for details).
--->
-<!-- This was deemed reasonable as Berlin's built-up area has not changed markedly over the past 50 years, i.e., about 52 to 61$\%$ [@mohamed2017].  -->
-<!-- Further insight into the role of the urban environment was assessed by including local climate zone classes derived by WUDAPT. -->
-<!-- These describe the proportion of urban area with a specific cover, such as dense high rises, sealed surfaces, etc.  -->
-<!-- For this study, the LCZ 6, representing dense mid-rise built up cover was used as a proxy for the degree of urbanization, and is the most frequent LCZ for Berlin [@fenner2017] (found ref in . -->
-<!-- Use percent sealed surface instead? -->
+An overview of data used for models, including sources, types, and application, is provided in Table$~$\@ref(tab:tab-dsources), with detailed descriptions in the following subsections.
 
@@
 
-## Dendrochronological sampling and analyses
+### Street trees
+
+Berlin’s open data provided tree inventories including species, age, location, and circumference which was transformed into diameter.
+Note that only street trees in urban, not rural areas or within green spaces, were considered here, but individual trees may grow along streets adjacent to green spaces and parks of varying sizes.
+Implausible observations, likely from erroneous data entry, were removed.
+Additional manual data processing for quality control was done with a bespoke software  `datacleanr` by Hurley *et al*.
+(submitted), where obvious outliers or clearly interpolated data were removed; the latter was deemed necessary, as several observations in multiple city districts were derived by linear relationships (i.e., straight-line), which do not capture the ontogenic growth dynamics of trees, and leave no variation related to variables other than age.
+All of these operations were recorded, and can be viewed and reproduced via the supplementary code.
+Lastly, observations with unlikely diameter-age combinations were identified via the residuals of a generalized linear model between diameter and age with a Gamma log-link distribution: if individual residuals exceeded seven times the median absolute deviation of all residuals, they were removed.
+The median absolute deviation (MAD) is comparable to the inter-quartile range, yet more robust to outliers:
+
+$$MAD = median(|X_i - median(X)|) $$
+This approach is considered conservative (see supplementary information), yet all analyses were carried forward with the unfiltered and filtered data - no considerable differences were found, thus subsequent sections are based on the filtered data.
+<!-- which were processed with the bespoke software datacleanr by Hurley et al., submitted), -->
+<!-- Characteristics of street trees and their respective environments were collated from the city's tree inventory, as well as various elements of the Berlin Environment Atlas, which are openly available via https://daten.berlin.de and curated by the city's government.
+-->
+<!-- The data used from the inventory includes species, location, age, and diameter at breast height (dbh). -->
+Table$~$\@ref(tab:tab-age) shows the binned distribution of genera across age classes.
+Final samples applied in models were smaller, following the availability of ancillary data for a given observation, and limited to a maximum age of 125 years to increase confidence in reported values, and ultimately model estimates.
+
+```{r tab-age}
+
+drake::loadd(age_tables)
+
+age_tables <- age_tables %>% 
+  dplyr::arrange(desc(n_total))
+colnames(age_tables)[c(1,7,8, 9)] <- c("Genera", "150+", "Total (n)", "Missing (n)")
+
+age_tables <- dplyr::bind_rows(age_tables[age_tables$Genera!="Other",],
+                        age_tables[age_tables$Genera=="Other",])
+
+add_row <- as.data.frame(matrix(c(NA, colSums(age_tables[,-1])), nrow = 1))
+add_row[1,1] <- "Marg. Totals"
+colnames(add_row) <- colnames(age_tables)
+
+age_tables <- dplyr::bind_rows(age_tables,
+                    add_row)
+
+caption_trees <- "Binned age-distribution for trees in Berlin data set, and entries missing age information. "
+caption_trees_short <- "Binned age-distribution for trees in Berlin data set."
+
+if (grepl("docx", out_format) ||
+    knitr::is_latex_output() ) {
+  table_dat_req_kable <- age_tables %>% 
+  knitr::kable(booktabs = TRUE,
+               format = "latex",
+             caption = caption_trees,
+             caption.short = caption_trees_short
+             # col.names = c("Subject/Relevance",
+             #              "Desc.",
+             #              "Type",
+             #              "Obsv. (n)",
+             #              "Resolution (m)",
+             #              "Source")
+             ) %>% 
+  kableExtra::row_spec(13, hline_after = TRUE)
+
+} else {
+table_dat_req_kable <- age_tables %>% 
+  knitr::kable(booktabs = TRUE,
+               format = "html",
+              caption = caption_trees,
+             caption.short = caption_trees_short
+             # col.names = c("Subject/Relevance",
+             #              "Desc.",
+             #              "Type",
+             #              "Obsv. (n)",
+             #              "Resolution (m)",
+             #              "Source")
+             ) %>% 
+  kableExtra::row_spec(13, extra_css = "border-bottom: 1px solid")
+
+# %>% 
+#   kableExtra::kable_styling(full_width = TRUE, 
+#                             font_size = 10) %>% 
+#   kableExtra::column_spec(1, italic = TRUE) %>%
+#   kableExtra::row_spec(0, bold = TRUE) %>% 
+#     kableExtra::row_spec(13, hline_after = TRUE)
+
+}
+
+   table_dat_req_kable %>% 
+  kableExtra::kable_styling(full_width = TRUE, 
+                            font_size = 11) %>% 
+  kableExtra::column_spec(1, italic = TRUE) %>%  
+  kableExtra::row_spec(0, italic = FALSE, bold = TRUE, font_size = 12)
+
+```
+
+### Temperature/UHI data sets
+Temperature and UHI data were summarized temporally either by the provider or manually to provide a characteristic representation of heat loading during summer at different times (morning, afternoon/day, night), from which tree averages (radius of 150$~m$) were calculated.
+Two data sets of urban air and one surface temperature were tested as explanatory variables in GAM models.
+The air temperatures from the Berlin environmental atlas (EnvAt) are model outputs that are representations of typical summer conditions at 0400, 1400 and 2200 hours; these data are provided at city block basis (spatial polygons), from which weighted averages were extracted.
+UrbClim air temperatures are hourly model outputs [100$~m$ resolution, @deridder2015] based on ERA5 re-analyses data (ECMWF) for which observations from the hottest month  available (June, 2011) were averaged to hours equivalent to EnvAt data by using a window of $\pm~1~$hour (i.e., 0300 to 0500, etc.).
+Subsequently, a land-use and land-cover mask (CORINE; European Union, Copernicus Land Monitoring Service 2018, European Environment Agency) was used to define urban and rural/forested areas.
+Using this mask was deemed reasonable as Berlin's built-up area has not changed markedly over the past 50 years, i.e., about 52 to 61$\%$ [@mohamed2017]. 
+The urban heat loading was then calculated as $UHI_{x,y} = T_{Air_{2m}~x,y} - \overline{T_{Air_{2m}~Rural}}$, where $x$ and $y$ define an urban grid cell.
+The surface UHI data set by @chakraborty2019 derives its measure in a similar fashion and the reader is referred to the detailed description therein; note this data set provides day and night-time averaged UHI estimates at 500$~m$ resolution, which were extracted for the hottest summer in this record (2007).
+
+### Ancillary environmental data sets
+
+Following the general approach described above, four ancillary covariates next to a temperature measure were employed in models; these were chosen due to their availability at high spatial resolution and coverage, and/or because their influence on growth was previously identified in literature or their likely impact could be deduced using ecophysiological principles.
+We included planting bed area and the sum of exchangeable basic cation as a proxy for soil nutrient availability (point extractions), as well as the proportional coverage of local climate zone 6 (LCZ6; open mid-rise, see @demuzere2019 and @stewart2012 for details) and adjacent building height (spatial averages).
+The latter was chosen as an increase reflects a transition away from densely urbanized areas and had the highest coverage for the processed tree inventory.
+
+<!-- surface area of tree planting beds ("Baumscheibe"), as well as general characteristics of the urban environment like building height. -->
+<!-- It also includes spatially-explicit model outputs for local climate simulations representative of typical conditions, such as air temperature (2 m) on a 2015 summer's day at different hours (0400, 1400, 2200).
+-->
+<!-- These model outputs were preferred over *in situ* measurements as **more representative based on averaged conditions, physiologically more relevant than surface urban heat island products**. -->
+<!-- These outputs are available at city block or planning unit level, and were used as a proxy for the general temperature distribution influenced by urbanization across the investigated growth period across the previous 80 years (see section X for details).
+-->
+<!-- This was deemed reasonable as Berlin's built-up area has not changed markedly over the past 50 years, i.e., about 52 to 61$\%$ [@mohamed2017].  -->
+<!-- Further insight into the role of the urban environment was assessed by including local climate zone classes derived by WUDAPT. -->
+<!-- These describe the proportion of urban area with a specific cover, such as dense high rises, sealed surfaces, etc.  -->
+<!-- For this study, the LCZ 6, representing dense mid-rise built up cover was used as a proxy for the degree of urbanization, and is the most frequent LCZ for Berlin [@fenner2017] (found ref in . -->
+<!-- Use percent sealed surface instead? -->
+
+## Dendrochronological sampling
+
+- external data from Schneider et al., in review
+
+Kolonie Heimaterde Alpenrose, Gutschmidtstraße britz sued, Hasenheide park,  "urban-smallpark_weigandufer (weigandufer, werra strase)
+
+```{r tab-dendro-sites}
+
+drake::loadd(series_long)
+drake::loadd(species_lookup)
+
+sampling_site_overview <-     series_long %>% 
+        dplyr::filter(grepl("urban", x = site_type)) %>% 
+        dplyr::select(sample_location, species, tree_id) %>% 
+        dplyr::group_by(sample_location, species) %>%
+        dplyr::summarize(tree_id = unique(tree_id)) %>% 
+        # mutate(site_type = gsub("[-].*$", "", site_type)) %>% 
+        dplyr::tally() %>% 
+    dplyr::mutate(sample_location = stringr::str_replace(sample_location, "[-][a-zA-Zä]+$", "") %>% 
+               stringr::str_replace(pattern = "-StEi", ""))
+
+spec_matched <- species_lookup$Species[match(sampling_site_overview$species, species_lookup$Code)]
+
+sampling_site_overview$species_long <- spec_matched
+
+sampling_site_overview <- sampling_site_overview %>% 
+    dplyr::select(Location = sample_location, Species = species_long, n)
+
+caption <- "Tree sampling locations"
+caption_ov_short <- caption
+
+if (grepl("docx", out_format) ||
+    knitr::is_latex_output() ) {
+  table_ds <- sampling_site_overview %>% 
+  knitr::kable(booktabs = TRUE,
+
+             caption = caption,
+             caption.short = caption,
+             format = "latex",
+             # col.names = c("Subject/Relevance",
+             #              "Desc.",
+             #              "Type",
+             #              "Obsv. (n)",
+             #              "Resolution (m)",
+             #              "Source")
+             escape = FALSE)
+
+} else {
+table_ds <- sampling_site_overview %>% 
+  knitr::kable(booktabs = TRUE,
+               format = "html",
+
+             caption = caption,
+             caption.short = caption_ov_short,
+             # col.names = c("Subject/Relevance",
+             #              "Desc.",
+             #              "Type",
+             #              "Obsv. (n)",
+             #              "Resolution (m)",
+             #              "Source")
+             escape = FALSE) 
+}
+
+table_ds %>% 
+  kableExtra::kable_styling(full_width = TRUE, 
+                            font_size = 11,
+                            latex_options = c("hold_position")) %>% 
+  kableExtra::row_spec(0, bold = TRUE, font_size = 12)
+
+```
+
+- Sampling locations
+- standard dendroecological processing (sample preparation, cross dating, estimation of missing years to pith, mean tree series)
 
@@
 
-### General approach: space-for-time analyses
+### GAMs
 
-#### Model development and selection
+### Dendrochronological analyses
+- GAM for cambial age of urban trees, split like @pretzsch2017 (<=1960, >1960)
+-
+
+### Model development and selection
+
+**Table with model x species sample sizes or infer from figure?**
+
+Model forms:
+
+\begin{align}
+
+g(E(Y)) &= f(utmx_i,utmy_i) + f_j(age_i) + f_j(temp_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i \\
+
+g(E(Y)) &= f_j(age_i) + f_j(temp_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
+
+g(E(Y)) &= f_j(age_i) + f_j(temp_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
+
+g(E(Y)) &= f_j(age_i) +  f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
+
+g(E(Y)) &= f_j(age_i) +  c_{i,j} + \tau_{i,k} + \epsilon_i  \\
+
+\end{align}
+
+As the spatial extent and coverage varies between temperature and ancillary data, more complex models typically also had fewer total observations.
 

### 2021-11-24 analysis/paper/supplemental_information.Rmd
methods - dendro

--- 2021-09-15 analysis/paper/manuscript.Rmd
@@
 ---
-title: "Species-specific effects of the Urban Heat Island on tree growth across Berlin"
+title: "Supplement to: Assessing the impact of urban heating on tree growth in Berlin with open inventory and environmental data"
 author:
@@
 
-# Introduction {#sec:intro}
+# Overview
 
-Berlin features the most intense Urban Heat Island (UHI) in Germany due to its large extent and development intensity [@kuttler2015], with temperature increases of up to $12~K$ during day-time and $6~K$ on average for night-times [2001-2010, @fenner2014] in urban $vs.$ rural areas.
-Consequently, urban green (infrastructure) systems are subjected to increased heat more frequently, potentially affecting their process dynamics - either positively or adversely. 
-Their performance and health, however, is closely tied to local energy budgets [@grimmond1996 ; @hertel2019], which in turn are decisive for controlling human wellbeing [e.g.
-@maras2016], amongst other factors.
-Assessing the effect of increased temperatures on green infrastructure, as part of the urban landscape, is therefore instrumental for understanding, and ultimately mitigating, the potential impact of future warming on increasingly urban societies [@norton2015].
-
-Trees, in particular, provide shading as well as transpirative cooling in their vicinity [@gillner2015;@oke1982;@endlicher2016], and therefore can reduce ambient temperatures, infrastructure power-consumption and (human) thermal discomfort  [e.g.
-@gulyas2006; @akbari2001; @mayer1987; @hoyano1988];
-simultaneously, they provide numerous other environmental, cultural and psychological services and/or benefits [see @tzoulas2007 for review].
-Further, recent tree growth dynamics as a proxy for on-going and future warming may provide an additional line of evidence to support the growing knowledge base on future climate-vegetation dynamics [@zhao2016] and may aid in mitigation and adaptation efforts [@pretzsch2017;@brune2016].
-
-Trees and green infrastructure in urban areas show a tendency for enhanced growth rates and/or productivity compared to rural counterparts [@jia2018; @pretzsch2017], yet feature a broad range of effect size ranges and, in some cases, signs specific to species and locality.
-@zhao2016 showed that growth rates increased within urban clusters as urbanization intensifies using remotely sensed vegetation indices.
-Similarly, for Berlin, @dahlhausen2018, identified positive growth modulation in highly urbanized environments (using growth increments) for *Tilia cordata* Mill, the most abundant tree of the city, which they attributed to the UHI effect, while intermediate development intensity showed indications of being least favorable for tree growth.
-<!-- For Berlin, @dahlhausen2018, showed divergent patterns for young and old trees dependent on urban development intensity for *Tilia cordata* Mill, the most abundant tree of the city, but overall reported an increased growth rate which they attributed to the UHI effect.
--->
-Further, @moser-reischl2019a identified positive associations between air temperature and radial growth for two species commonly selected by urban planners (*T.
-cordata*, *Rubinia pseudoacacia*) in Munich.
-By contrast, @gillner2014 highlight decreased growth for *Acer* species  (*A.
-platanoides* and *pseudoplatanus*), *Platanus x hispanica* and *Quercus rubra* with higher summer temperatures of the preceding year, especially when compounded with drought, in another German metropolis (Dresden).
-Differences in growth trends may result from contrasting species-specific characteristics, but are indeed affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc.
-[@rhoades1999; @quigley2004; @randrup2001; @pauleit2002].
-Under climate change, atmospheric drought will likely be compounded with high temperatures - and intensified UHIs - more frequently, adding further stress to current urban disturbance regimes [@roloff2009].
-
-Conditions affecting tree growth can vary greatly within urban areas or regions, and need to be accounted for when establishing relationships with pertinent drivers, such as the UHI effect.
-This typically complicates the extrapolation from individual sampling sites toward predicting effect sizes across entire urban areas and tree stocks.
-This is especially the case for studies reliant on labour-intensive methods which are limited logistically by sampling effort, reducing sample sizes, as well as species and spatial coverage.
-<!-- While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e.
-genera and/or species, and space).
--->
-<!-- However, urban tree growth in urban settings is affected by various processes, which  -->
-<!-- However, growth in urban settings is affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc.
-[@rhoades1999; @quigley2004; @randrup2001; @pauleit2002].
--->
-<!-- While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e.
-genera and/or species, and space).
--->
-<!-- This typically hinders the extrapolation from individual sampling sites toward prediction of effect sizes across entire urban areas and tree stocks.  -->
-<!-- Therefore, studies reliant on labor-intensive sampling, must carefully address confounding factors to ensure effect sizes are estimated adequately. -->
-
-<!-- Detailed, but spatially-limited dendroecological analyses of climate-growth relationships in Berlin for key species can be enhanced by inferring growth modulation from a large data set in excess of 650000 individuals provided by the Berlin Senate Administration (Senatsverwaltung).
--->
-
-To complement (existing) detailed dendroecological analyses of climate-growth relationships in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individuals covering 94 genera and at least 600 species and/or cultivars provided by the Berlin Senate Administration (Senatsverwaltung).
-This data set contains information on location, species, trunk diameter (at breast height; $DBH$; see Tab.$~$\@ref(tab:tab-tree-overview)), and height, amongst other variables for the majority of street and park trees.
-
-```{r tab-tree-overview}
-drake::loadd(overview_table)
-
-caption_ov <- "Available records by category in entire data set (n), and those with age and $DBH$ entries (n$_{full}$)."
-caption_ov_short <- "Available records in Berlin tree database."
-
-if (grepl("docx", out_format) ||
-    knitr::is_latex_output() ) {
-  table_ov <- overview_table %>% 
-  knitr::kable(booktabs = TRUE,
-
-             caption = caption_ov,
-             caption.short = caption_ov_short,
-             format = "latex",
-             # col.names = c("Subject/Relevance",
-             #              "Desc.",
-             #              "Type",
-             #              "Obsv. (n)",
-             #              "Resolution (m)",
-             #              "Source")
-             escape = FALSE)
-
-} else {
-table_ov <- overview_table %>% 
-  knitr::kable(booktabs = TRUE,
-               format = "html",
-
-             caption = caption_ov,
-             caption.short = caption_ov_short,
-             # col.names = c("Subject/Relevance",
-             #              "Desc.",
-             #              "Type",
-             #              "Obsv. (n)",
-             #              "Resolution (m)",
-             #              "Source")
-             escape = FALSE) 
-}
-
- table_ov %>% 
-  kableExtra::kable_styling(full_width = TRUE, 
-                            font_size = 11,
-                            latex_options = c("hold_position")) %>% 
-  kableExtra::row_spec(0, bold = TRUE, font_size = 12)
-
-```
-
-In a space-for-time substitution, growth of individual species can be assessed across the entire city of Berlin, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.
-Comparable applications are found, for example, in @quigley2004 and @pretzsch2017.
-The former  inferred absolute growth potential for species across successional groups (early, mid, late stage), and between rural and urban conspecifics, yet lacked spatially-explicit effect size estimates across the urban-rural space and was limited to comparatively small sample sizes per group ($n_{total}~=~230$ divided in 15 species, 3 groups and 2 locations).
-@pretzsch2017 applied linear hierarchical models to infer growth modulation on annual basis for different time periods and urban $vs.$ rural locations while accounting for stand-level variability;
-however, for Berlin only 145 individuals of one species (*T. cordata*) were assessed.
-As mentioned previously, climate-growth relationships can vary substantially between species, and in fact, @quigley2004 and @pretzsch2017 report contrasting results regarding average tree diameter, i.e.
-smaller or larger for urban $vs.$ rural trees of same age.
-Consequently, this variability of effect sizes and directions calls for a more comprehensive assessment across species and with greater spatial coverage throughout Berlin.
-
-**We therefore propose applying a statistical model that is fully spatially explicit, while also allowing to account for the nested nature of the data set (e.g.
-streets and districts) as well as other pertinent factors using hierarchical, generalized additive models (see Section$~$\@ref(sec:methods)).
-As a result, the absolute growth potential of a species can be inferred given, for example, a specific location, age or UHI magnitude. 
-Further, the impact of UHI loading can be predicted for a single species across all of Berlin as a continuous surface.**
-The inclusion of independent, tree-level growth data, however, is paramount as it allows:
-
-1) validating the publicly-available Senate dataset ($DBH$) while providing a more reliable estimate of tree age,  
-2) applying the model with annual basal area as a response (enabling incorporating effects of varying climate/UHI intensities over time), and  
-3) comparing $DBH$ (Senate data) and basal area-derived models for a subset of locations and species, increasing confidence in Berlin-wide predictions.
-
-<!-- We acknolwedge results from [@gregg2003], which indicated impeded growth adjacent to intense urban clusters, rather than enhancement within them. -->
-
-# Methods {#sec:methods}
-
-## Study area
-
-Berlin is one of the largest metropolitan areas in Central Europe (892$~km^2$) with a population of approximately 3.6 million, and a maximum extent of 38$~km$ in North-South and 45$~km$ in East-West directions.
-It is located in North-Eastern Germany, and lies in the temperate zone with warm-humid climate (Dfb) according to the updated Köppen-Geiger classification [@beck2018], with mean annual temperature of approximately 10$^\circ C$ and precipitation of 575$~mm$ (Tempelhof weather station, German Weather Service DWD).
-Berlin has low relief (approximately 30$~m$ to 60$~m$ with 120$~m$ at solitary peaks), and is centered around a glacial outwash valley (sands, gravel), bordered by two plateaus consisting of glacial till and clay in the North-East and South, as well as sands in the South-West.
-The city provides extensive public green space covering around 30$~\%$ of its area [@suvkberlin2019], with an extensive urban forest of nearly 700000 publically-managed trees along streets, in parks and in riparian areas.
-It has a pronounced urban heat island, owing to its high degree of built-up and sealed surfaces, with air temperatures up to 12$^\circ C$ higher in urban compared to adjacent rural areas at night times [@fenner2014].
-
-# General approach
-
-- Assess differences in mean growth across age for individual species produced by environmental factors, including air temperature (or measure of urban heat island loading), soil characteristics and rooting area, adjacent building height) in a space-for-time approach
-- Focus on street trees where species have a minimum of X individuals
-- Account for potential differences in tree management by authorities
-- Use flexible models that can produce curves of varying modality 
-- Understand / confirm urban tree growth (assessed for single species)
-
-## Data sources and processing
-
-An overview of all data used in this study, including sources, types, and application, is provided in Table X.
-Most notably, characteristics of street trees and their respective environments were collated from the city's tree inventory, as well as various elements of the Berlin Environment Atlas, which are openly available via daten.berlin.de and curated by the city's government.
-The data used from the inventory includes species, location, age, and diameter at breast height (dbh).
-The environment atlas holds information on several variables relevant to tree growth, such as soil type and nutrient exchange capacity, distance to groundwater, surface area of tree planting beds ("Baumscheibe"), as well as general characteristics of the urban environment like building height.
-It also includes spatially-explicit model outputs for local climate simulations representative of typical conditions, such as air temperature (2 m) on a 2015 summer's day at different hours (0400, 1400, 2200).
-These model outputs were preferred over *in situ* measurements as **more representative based on averaged conditions, physiologically more relevant than surface urban heat island products**.
-These outputs are available at city block or planning unit level, and were used as a proxy for the general temperature distribution influenced by urbanization across the investigated growth period across the previous 80 years (see section X for details).
-This was deemed reasonable as Berlin's built-up area has not changed markedly over the past 50 years, i.e., about 52 to 61$\%$ [@mohamed2017]. 
-Further insight into the role of the urban environment was assessed by including local climate zone classes derived by WUDAPT.
-These describe the proportion of urban area with a specific cover, such as dense high rises, sealed surfaces, etc. 
-For this study, the LCZ 6, representing dense mid-rise built up cover was used as a proxy for the degree of urbanization, and is the most frequent LCZ for Berlin [@fenner2017] (found ref in .
-Use percent sealed surface instead?
-
-- datacleanr
-- adjusting bad records
-- final original, final data set available online, with cleaning scripts
-
-```{r dsources-prep, include=FALSE}
-
-options(knitr.kable.NA = "")
-
-dsources_path <- "analysis/data/tables/data_sources.xlsx"
-dsources <- readxl::read_excel(dsources_path, na = "NA")
-dsources$Accessed <- format(dsources$Accessed, "%b '%y")
-
-```
-
-```{r dsources-tab, eval = TRUE}
-
-caption <- "Data description used for generating maps and route optimization.
-Resolution and radius are provided in $m$, the latter referring to the buffer size across which data was averaged to estimate trees' environmental conditions.
-A zero-radius refers to a point extraction from categorical and location specific data.
-**Planting Bed Area, and Soil type are not currently used.**"
-
-dsources <- dsources %>%
-  dplyr::select(Name = `Name-eng`, Accessed, Type, Unit = `Unit-latex`, Resolution, Radius, Source = `Source-short`, Reference = TextRef) %>%
-  dplyr::filter(Name != "UHI Berlin",
-                Name != "Park Trees",
-                Name != "Riparian Trees")
-
-if (grepl("docx", out_format) ||
-    knitr::is_latex_output() ) {
-  table_ds <- dsources %>% 
-  knitr::kable(booktabs = TRUE,
-
-             caption = caption,
-             caption.short = caption,
-             format = "latex",
-             # col.names = c("Subject/Relevance",
-             #              "Desc.",
-             #              "Type",
-             #              "Obsv. (n)",
-             #              "Resolution (m)",
-             #              "Source")
-             escape = FALSE) %>% 
-      kableExtra::landscape()
-
-} else {
-table_ds <- dsources %>% 
-  knitr::kable(booktabs = TRUE,
-               format = "html",
-
-             caption = caption,
-             caption.short = caption_ov_short,
-             # col.names = c("Subject/Relevance",
-             #              "Desc.",
-             #              "Type",
-             #              "Obsv. (n)",
-             #              "Resolution (m)",
-             #              "Source")
-             escape = FALSE) 
-}
-
-table_ds %>% 
-  kableExtra::kable_styling(full_width = TRUE, 
-                            font_size = 11,
-                            latex_options = c("hold_position")) %>% 
-  kableExtra::row_spec(0, bold = TRUE, font_size = 12)
-
-# dsources %>%
-#   
-#   knitr::kable(caption = caption, escape = FALSE) %>%
-#   kableExtra::kable_styling(full_width = TRUE,
-#                             bootstrap_options = c("striped", "hover", "condensed", "responsive"))
-
-# flextable::set_flextable_defaults(padding.left = 0)
-# 
-# dsources %>%
-#   dplyr::select(`Name-eng`, Accessed, Type, Source, TextRef) %>%
-#   flextable::flextable() %>%
-#   flextable::autofit()
-
-```
-
-## General analyses
-
-The proposed statistical method is from the class of hierarchical, generalized additive models (GAM, or GAMM for mixed models/hierarchical models).
-In these models combinations of continuous and categorical predictor variables can be summed to estimate a response. 
-In particular, continuous variables that are linearly, as well as non-linearly related to the response can be represented by applying a transfer function, typically termed "smoothing function" [@wood2017]; these are constructed using a number of base functions of varying complexity and form, which provides a high degree of flexibility, ideal for fitting ecosystem dynamics which are rarely linear [@pedersen2019], or correctly represented with deterministic functional forms (e.g.
-quadratic equations).
-In general, a GAM can be written as:
-
-\begin{equation}
-E (Y)~=~g^{-1}\left( \beta_0 + \sum_{i = 1}^{n} f_i (x_i) \right),
-(\#eq:gam-general)
-\end{equation}
-
-and
-
-\begin{equation}
-y~=~E (Y) + \epsilon,
-(\#eq:gam-expectation)
-\end{equation}
-
-where $Y$ is taken from an appropriate distribution and corresponding link function $g$, $\beta_0$ is the intercept and $f_i$ represents a smooth function of a predictor [@pedersen2019], and $\epsilon \sim \mathcal{N}(0, \sigma ^2)$.
-Note, that $f_i$ consists of a smooth (e.g. spline) constructed via basis functions of different form and complexity, multiplied by a coefficient:
-
-\begin{equation}
-f_i(x_i)~=~\sum_{k = 1}^{K} \beta_{i, k} b_{i,k}(x_i).
-(\#eq:gam-smoothers)
-\end{equation}
-
-Nested data structures (e.g.
-due to similar road [type]) can be accounted for by introducing random effects [@wood2017], while spatial dependence between observations can be included by constructing smoothing functions with e.g.
-northings and eastings, as for example done in [@augustin2009].
-Ultimately, the implementation of a such a GAMM will allow for establishing continuous prediction surfaces of growth potential (approximated via $DBH$) for individual species across urban areas (including parks) of Berlin.
-
-Currently, $DBH$ has been modeled using a hierarchical linear model (linear mixed effects model) with `lme4` [@bates2015] in @rcoreteam2020 (see Section$~$\@ref(sec:results)).
-The general form of this model is:
-
-\begin{equation}
-Y_{i,j} = (\beta_0 + b_{0,i,j}) + (\beta_1 + b_{1,i,j}) \cdot x_i  + \epsilon_{i,j},
-(\#eq:lme)
-\end{equation}
-where $\beta_0$ is the intercept with its random component $b_0$, and $\beta_1$ the slope with its random component $b_1$.
-The random errors are assumed i.i.d. and distributed as $b \sim \mathcal{N}(0, \tau ^2)$. 
-The model for which results are presented in Figure$~$\@ref(fig:fig-lmestat) estimates $DBH$ from tree age and the local UHI intensity as continuous covariates with random slopes and intercepts for each species;
-note, that for computational efficiency each genera was modeled separately.
-Further, models were established for the three most abundant species per genera with at least 1000 individuals.
-
-<!-- We caution, that 1) the presented model results are preliminary, 2) initial diagnostics indicate that modification of the current model structure (e.g.
-by accounting for individual-level variability) may be required and 3) that further, time-intensive data cleaning is necessary.
--->
-<!-- First tests, however, showed that while the magnitude of effects changes with different model structures, their direction appears stable. -->
-
-## Available and required data
-
-Additional remote sensing data for local context, higher resolution UHI data, and most important, temporally-resolved tree growth data are still required and/or would greatly improve the confidence in effect size estimates (see Section$~$\@ref(sec:intro)).
-See Tab.$~$\@ref(tab:tab-data-req) for details on data sets.
-
-```{r tab-data-req,echo=FALSE}
-
-dat_req <- read.csv(here::here("analysis", "data", "tables", "01_data_requirements.csv"), header = TRUE)
-
-caption_data_req <- "Data requirements for analysis, including currently available/accessed and required/desired data."
-caption_data_req_short <- "Data requirements for analysis."
-
-if (grepl("docx", out_format) ||
-    knitr::is_latex_output() ) {
-  table_dat_req <- dat_req %>% 
-  dplyr::select(-owned) %>% 
-  knitr::kable(booktabs = TRUE,
-
-             caption = caption_data_req,
-             caption.short = caption_data_req_short,
-             format = "latex",
-             col.names = c("Subject / Relevance",
-                          "Desc.",
-                          "Type",
-                          "Obsv. (n)",
-                          "Resolution (m)",
-                          "Source")) %>% 
-  kableExtra::kable_styling(full_width = TRUE, 
-                            font_size = 11) %>% 
-  kableExtra::row_spec(0, bold = TRUE, font_size = 12) %>% 
-  kableExtra::row_spec(9,italic =  TRUE) %>% 
-  kableExtra::pack_rows("available or accessed", 1, 4) %>%
-  kableExtra::pack_rows("required and/or desired", 5, 9) %>% kableExtra::landscape()
-
-} else {
-  table_dat_req <- dat_req %>% 
-  dplyr::select(-owned) %>% 
-  knitr::kable(booktabs = TRUE,
-               format = "html",
-               caption = "Data requirements for analysis, including currently available/accessed and required/desired data.",
-             caption.short = "Data requirements for analysis",
-             col.names = c("Subject/Relevance",
-                          "Desc.",
-                          "Type",
-                          "Obsv. (n)",
-                          "Resolution (m)",
-                          "Source")) %>% 
-  kableExtra::kable_styling(full_width = TRUE, 
-                            font_size = 10) %>% 
-  kableExtra::row_spec(0, bold = TRUE) %>% 
-  kableExtra::pack_rows("available or accessed", 1, 4) %>%
-  kableExtra::pack_rows("required and/or desired", 5, 9)
-
-}
-
- table_dat_req
-
-```
-
-# Results {#sec:results}
-
-Tree locations are clustered and structured based on their category, i.e. riparian, street and park trees (Fig.$~$\@ref(fig:fig-tree-overview-map)).
-Planting in space and time shows species-specific patterns (by districts), often related to major events, such as the start and end of armed and/or political conflict.
-Table$~$\@ref(tab:tab-age) shows the binned distribution of genera across age classes.
-
-```{r tab-age}
-
-drake::loadd(age_tables)
-
-age_tables <- age_tables %>% 
-  dplyr::arrange(desc(n_total))
-colnames(age_tables)[c(1,7,8, 9)] <- c("Genera", "150+", "Total (n)", "Missing (n)")
-
-age_tables <- dplyr::bind_rows(age_tables[age_tables$Genera!="Other",],
-                        age_tables[age_tables$Genera=="Other",])
-
-add_row <- as.data.frame(matrix(c(NA, colSums(age_tables[,-1])), nrow = 1))
-add_row[1,1] <- "Marg. Totals"
-colnames(add_row) <- colnames(age_tables)
-
-age_tables <- dplyr::bind_rows(age_tables,
-                    add_row)
-
-caption_trees <- "Binned age-distribution for trees in Berlin data set, and entries missing age information."
-caption_trees_short <- "Binned age-distribution for trees in Berlin data set."
-
-if (grepl("docx", out_format) ||
-    knitr::is_latex_output() ) {
-  table_dat_req_kable <- age_tables %>% 
-  knitr::kable(booktabs = TRUE,
-               format = "latex",
-             caption = caption_trees,
-             caption.short = caption_trees_short
-             # col.names = c("Subject/Relevance",
-             #              "Desc.",
-             #              "Type",
-             #              "Obsv. (n)",
-             #              "Resolution (m)",
-             #              "Source")
-             ) %>% 
-  kableExtra::row_spec(13, hline_after = TRUE)
-
-} else {
-table_dat_req_kable <- age_tables %>% 
-  knitr::kable(booktabs = TRUE,
-               format = "html",
-              caption = caption_trees,
-             caption.short = caption_trees_short
-             # col.names = c("Subject/Relevance",
-             #              "Desc.",
-             #              "Type",
-             #              "Obsv. (n)",
-             #              "Resolution (m)",
-             #              "Source")
-             ) %>% 
-  kableExtra::row_spec(13, extra_css = "border-bottom: 1px solid")
-
-# %>% 
-#   kableExtra::kable_styling(full_width = TRUE, 
-#                             font_size = 10) %>% 
-#   kableExtra::column_spec(1, italic = TRUE) %>%
-#   kableExtra::row_spec(0, bold = TRUE) %>% 
-#     kableExtra::row_spec(13, hline_after = TRUE)
-
-}
-
-   table_dat_req_kable %>% 
-  kableExtra::kable_styling(full_width = TRUE, 
-                            font_size = 11) %>% 
-  kableExtra::column_spec(1, italic = TRUE) %>%  
-  kableExtra::row_spec(0, italic = FALSE, bold = TRUE, font_size = 12)
-
-```
-
-<!-- ```{r fig-bar-tree, fig.cap="A plot of random numbers",fig.width = 15, fig.height = 8, out.width="100%", include=FALSE, eval=FALSE} -->
-
-<!-- readd(plot_tree_sums_bar) -->
-
-<!-- ``` -->
-
-```{r fig-tree-overview-map, fig.cap="Individual tree locations for three categories available in Berlin Senate urban tree data set.
-Note, that for each category 7000 observations were subsampled from the available pool to facilitate visualization.",fig.width = 15, fig.height = 8, out.width="100%", include=TRUE, eval=TRUE}
-
-# loadd(full_data_set_clean)
-# loadd(plot_overview_map)
-# print(class(full_data_set_clean))
-
-knitr::include_graphics(here::here("analysis/figures/map_01_overview.png"))
-
-# readd(plot_overview_map) + 
-#      +
-#     labs(caption = NULL)
-
-# plot_overview_map
-
-```
-
-```{r fig-tree-count-map, fig.cap="Gridded counts for the 11 most frequent genera, as well as *Pinus* and remaining genera.
-Note, that counts are standardized to unity for individual genera.",fig.width = 12, fig.height = 14, out.width="100%"}
-
-knitr::include_graphics(here::here("analysis/figures/map_02_tree_sums_standardized.png"))
-
-# plot_count_map
-
-# plot(1:10)
-```
-
-The distribution of the UHI effect is highly irregular and clustered in space (Fig.$~$\@ref(fig:fig-uhi-map)), and also shows variability through time (data not shown, refer to the [urban heat island explorer](https://yceo.users.earthengine.app/view/uhimap)).
-
-(ref:chakraborty2019) [@chakraborty2019]
-
-```{r fig-uhi-map, fig.cap="Estimate of UHI intensity based on the algorithm in (ref:chakraborty2019), comparing urban with rural pixels within the greater metropolitan cluster.
-Presented values are averaged over the summer of 2007.",fig.width = 12, fig.height = 14, out.width="50%"}
-
-knitr::include_graphics(here::here("analysis/figures/map_03_uhi.png"))
-
-# plot_count_map
-
-# plot(1:10)
-```
-
-The exposure to increased heat-loading of individual genera (and consequently species) is highly uneven throughout the city (Fig.$~$\@ref(fig:fig-density)).
-Street and park trees of most genera are clustered in urban areas with intermediate to high UHI loading, while riparian trees, and some street and park trees of other genera tend to be spread more evenly across Berlin's UHI range.
-
-```{r fig-density, fig.cap="Empirical density distribution of all individuals within the presented genera along the UHI continuum.
-UHI intensities were extracted for each tree location, and the distribution hence represents the first detailed overview of the exposure of Berlin's trees to urban heat loading.
-The black line is the density across all three categories.
-Insets show corresponding tree totals.",fig.width = 12, fig.height = 12, out.width="100%"}
-knitr::include_graphics(here::here("analysis/figures/plot_02_genus_UHI_dens.png"))
-
-# plot_count_map
-
-# plot(1:10)
-```
-
-**Note, that results below are preliminary and should be considered as a template for future outputs, rather than used for inference.**
-The effect of UHI loading on absolute growth potential varies between genera and species (Fig.$~$\@ref(fig:fig-lmestat)).
-Most notably, *Quercus*, the 3rd-most frequent genera, shows decreased absolute growth with increasing UHI loading,
-while the most frequent genera, *Tilia*, features contrasting relationships between species.
-The estimated effect sizes presented here are linear.
-However, temperature may exert a non-linear control on absolute growth and, hence, applying a  method able to capture such dynamics may result in somewhat different effect sizes / behavior.
-Additionally, if temperatures increase in the future under climate warming, any non-linear effects may become more enhanced, stressing the need for a more flexible model fit and structure (i.e.
-using GAMM over linear models-).
-
-```{r fig-lmestat, fig.cap="Impact of UHI loading on tree diameter (\\(DBH\\)), accounting for age and inter-specific differences from the linear mixed model (via random slopes and intercepts).
-Line-ranges are standard errors of predicted effect sizes (i.e.
-slopes).
-Differences between street and park trees are considerable for some species, and may be due to local clustering and/or spatial under-representation across the UHI continuum.
-Further investigations need to address the degree of spatial autocorrelation and account for it where required in linear mixed models, and with smoothing interactions in a GAMM implementation.",fig.width = 20, fig.height = 12, out.width="100%"}
-knitr::include_graphics(here::here("analysis/figures/plot_03_ranef_species_dbh_uhi.png"))
-# plot_count_map
-
-# plot(1:10)
-```
-
-# Discussion
-
-1. Temperature, environmental and urban controls on tree growth  
-	+ Urban tree growth accelerated (BIWI data)
-	+ Intra specific differences in UHI response (note temp. products)
-	+ Sensitive trees in hot spots
-	+ Other controls: soil, urban fabric, 
-	+ Validity and limitations (spatial auto-correlation, data quality, coverage)
-
-2. Implications:  
-  + Policy / planning:  
-    - tree species
-	  - Increase temporal coverage and accuracy of surveys
-		- which trees to prioritize for measurement campaigns (low-abundance species, highly sensitive)
-		- Compare and Supplement species matrix etc. with these analyses  
-		- Generate co-located environmental observations (e.g., micro-climate)
-  + Open data important --> drives innovation
-
-# Conclusion
+- Upload full code with DOI
+- datacleanr imgs
+- diagnostic lm imgs
+- spatial comparison (Moran's I)
 

### 2021-11-24 R/plan.R

@@ -869,6 +869,10 @@ plan <- drake::drake_plan(
 
    ### BIWI: cambial age vs. annual growth -------------------
 
+   species_lookup = read.csv(drake::file_in("analysis/data/tables/species_lookup.csv"), header = TRUE),
+
+
+
    series_long = prep_rwl_data(path_meta_cores = drake::file_in("./analysis/data/raw_data/biwi/BIWi_INV_20190123.xlsx"),
                                 path_meta_trees = drake::file_in("./analysis/data/raw_data/biwi/BIWi_INV_20190123.xlsx"),
                                 path_meta_sites = drake::file_in("./analysis/data/raw_data/biwi/BIWi_INV_20190123.xlsx"),
@@ -994,7 +998,7 @@ plan <- drake::drake_plan(
         output_file = file_out("paper_knit.html"),
         output_format = bookdown::html_document2(),
         quiet = TRUE
-    )
+    ),
     # paper_word = rmarkdown::render(
     #     knitr_in("./analysis/paper/paper.Rmd"),
     #     output_dir = "./analysis/paper/",
@@ -1010,6 +1014,15 @@ plan <- drake::drake_plan(
     #     output_format = bookdown::pdf_document2(),
     #     quiet = TRUE
     # )
+
+
+   si_html = rmarkdown::render(
+       knitr_in("./analysis/paper/supplemental_information.Rmd"),
+       output_dir = "./analysis/paper/",
+       output_file = file_out("supplemental_information.html"),
+       output_format = bookdown::html_document2(),
+       quiet = TRUE
+   )
 
 )
 

### 2021-11-24 analysis/paper/paper.Rmd
methods - gam

@@
 The assessment relied on flexible statistical models that could capture species and location-specific responses to heat and other urban factors.
-Specifically, we aimed to (1) assess heat exposure of most abundant species; (2) determine the impact of (excess) heat on stem growth with space-for-time substitution to increase coverage; (3) highlight the role of location-specific environmental factors in mediating temperature responses.
+Specifically, we aimed to (1) assess heat exposure of the most abundant species; (2) determine the impact of (excess) heat on stem growth across tree age classes with a space-for-time substitution; (3) highlight the role of location-specific environmental factors in mediating temperature responses.
 <!-- This work constitutes a case study for Berlin, the results are a valuable contribution toward Berlin's current and future management of its tree stock.  -->
@@
 
-### Temperature/UHI data sets
+### Temperature/UHI data
+
 Temperature and UHI data were summarized temporally either by the provider or manually to provide a characteristic representation of heat loading during summer at different times (morning, afternoon/day, night), from which tree averages (radius of 150$~m$) were calculated.
@@
 Using this mask was deemed reasonable as Berlin's built-up area has not changed markedly over the past 50 years, i.e., about 52 to 61$\%$ [@mohamed2017]. 
-The urban heat loading was then calculated as $UHI_{x,y} = T_{Air_{2m}~x,y} - \overline{T_{Air_{2m}~Rural}}$, where $x$ and $y$ define an urban grid cell.
+The urban heat loading was then calculated as 
+$$UHI_{x,y} = T_{Air_{2m}~x,y} - \overline{T_{Air_{2m}~Rural}},$$
+where $T$ is temperature ($^\circ C$) $x$ and $y$ define an urban grid cell.
 The surface UHI data set by @chakraborty2019 derives its measure in a similar fashion and the reader is referred to the detailed description therein; note this data set provides day and night-time averaged UHI estimates at 500$~m$ resolution, which were extracted for the hottest summer in this record (2007).
 
-### Ancillary environmental data sets
+### Ancillary environmental
 
@@
 
-- external data from Schneider et al., in review
-
-Kolonie Heimaterde Alpenrose, Gutschmidtstraße britz sued, Hasenheide park,  "urban-smallpark_weigandufer (weigandufer, werra strase)
+To contextualize tree growth patterns between age groups derived from Berlin's inventory data, we drew upon a recently established data set from Schneider *et al.* (in review), who sampled several common tree species across a rural-urban gradient.
+For our purposes, we grouped trees sampled in parks, green spaces and along streets into a single urban category, and focused analyses on these.
+Two to three cores were extracted at breast height from each tree.
+These were then prepared using standard dendro-ecological methods (i.e., mounting, sanding, measuring), and cross-dated with TSAP-Win\textsuperscript{TM} and COFECHA [@holmes1986], producing mean tree series of incremental growth.
+Additionally, cambial age of each increment was established by counting years from the inner most ($a = 0$) ring at the pith outward; on tangentially bored cores, missing rings to the pith were estimated.
 
@@
 
-caption <- "Tree sampling locations"
+caption <- "Overview of urban sampling locations and respective tree species coverage. Individual trees were sampled up to to three times to obtain a mean-tree ring width series."
 caption_ov_short <- caption
@@
 
-- Sampling locations
-- standard dendroecological processing (sample preparation, cross dating, estimation of missing years to pith, mean tree series)
-
 ## Statistical Analyses {#stat-gam}
@@
 ### GAMs
+
+ (GAM, or GAMM for mixed models/hierarchical models).
+We applied hierarchical generalized additive models (GAM) to estimate the relationship of several covariates with stem diamater growth ($DBH$).
+GAMs, as an extension of generalized linear models [@wood2017], allow modeling response variables as parametric and non-parametric combinations of smoothed explanatory covariates, and can assume non-normal errors and response distributions.
+These smooths are constructed by summation of base functions of varying complexity and form, analogous to scatterplot smoothing [@hastie2017], which provides a high degree of flexibility, ideal for fitting ecosystem dynamics which are rarely linear [@pedersen2019], or correctly represented with deterministic functional forms (e.g.
+quadratic equations).
+In general, a GAM can be written as:
+
+\begin{equation}
+E (Y)~=~g^{-1}\left( \beta_0 + \sum_{i = 1}^{n} f_i (x_i) \right),
+(\#eq:gam-general)
+\end{equation}
+
+and
+
+\begin{equation}
+y~=~E (Y) + \epsilon,
+(\#eq:gam-expectation)
+\end{equation}
+
+where $Y$ is taken from an appropriate distribution and corresponding link function $g$, $\beta_0$ is the intercept and $f_i$ represents a smooth function of a predictor [@pedersen2019], and $\epsilon \sim \mathcal{N}(0, \sigma ^2)$.
+Note, that $f_i$ consists of a smooth (e.g. spline) constructed via basis functions of different form and complexity, multiplied by a coefficient:
+
+\begin{equation}
+f_i(x_i)~=~\sum_{k = 1}^{K} \beta_{i, k} b_{i,k}(x_i).
+(\#eq:gam-smoothers)
+\end{equation}
+
+Nested data structures (e.g., city districts) can be accounted for by introducing random effects [@wood2017], while spatial dependence between observations can be accounted for constructing smoothing functions with, for instance, northings and eastings [cf.
+@wood2017], which is  applied here.
 
@@
 As the spatial extent and coverage varies between temperature and ancillary data, more complex models typically also had fewer total observations.
-
-## General analyses
-
-The proposed statistical method is from the class of hierarchical, generalized additive models (GAM, or GAMM for mixed models/hierarchical models).
-In these models combinations of continuous and categorical predictor variables can be summed to estimate a response. 
-In particular, continuous variables that are linearly, as well as non-linearly related to the response can be represented by applying a transfer function, typically termed "smoothing function" [@wood2017]; these are constructed using a number of base functions of varying complexity and form, which provides a high degree of flexibility, ideal for fitting ecosystem dynamics which are rarely linear [@pedersen2019], or correctly represented with deterministic functional forms (e.g.
-quadratic equations).
-In general, a GAM can be written as:
-
-\begin{equation}
-E (Y)~=~g^{-1}\left( \beta_0 + \sum_{i = 1}^{n} f_i (x_i) \right),
-(\#eq:gam-general)
-\end{equation}
-
-and
-
-\begin{equation}
-y~=~E (Y) + \epsilon,
-(\#eq:gam-expectation)
-\end{equation}
-
-where $Y$ is taken from an appropriate distribution and corresponding link function $g$, $\beta_0$ is the intercept and $f_i$ represents a smooth function of a predictor [@pedersen2019], and $\epsilon \sim \mathcal{N}(0, \sigma ^2)$.
-Note, that $f_i$ consists of a smooth (e.g. spline) constructed via basis functions of different form and complexity, multiplied by a coefficient:
-
-\begin{equation}
-f_i(x_i)~=~\sum_{k = 1}^{K} \beta_{i, k} b_{i,k}(x_i).
-(\#eq:gam-smoothers)
-\end{equation}
-
-Nested data structures (e.g.
-due to similar road [type]) can be accounted for by introducing random effects [@wood2017], while spatial dependence between observations can be included by constructing smoothing functions with e.g.
-northings and eastings, as for example done in [@augustin2009].
-Ultimately, the implementation of a such a GAMM will allow for establishing continuous prediction surfaces of growth potential (approximated via $DBH$) for individual species across urban areas (including parks) of Berlin.
-
-Currently, $DBH$ has been modeled using a hierarchical linear model (linear mixed effects model) with `lme4` [@bates2015] in @rcoreteam2020 (see Section$~$\@ref(sec:results)).
-The general form of this model is:
-
-\begin{equation}
-Y_{i,j} = (\beta_0 + b_{0,i,j}) + (\beta_1 + b_{1,i,j}) \cdot x_i  + \epsilon_{i,j},
-(\#eq:lme)
-\end{equation}
-where $\beta_0$ is the intercept with its random component $b_0$, and $\beta_1$ the slope with its random component $b_1$.
-The random errors are assumed i.i.d. and distributed as $b \sim \mathcal{N}(0, \tau ^2)$. 
-The model for which results are presented in Figure$~$\@ref(fig:fig-lmestat) estimates $DBH$ from tree age and the local UHI intensity as continuous covariates with random slopes and intercepts for each species;
-note, that for computational efficiency each genera was modeled separately.
-Further, models were established for the three most abundant species per genera with at least 1000 individuals.
 

### 2021-11-24 analysis/paper/paper.Rmd
methods gam cont

@@
 These were then prepared using standard dendro-ecological methods (i.e., mounting, sanding, measuring), and cross-dated with TSAP-Win\textsuperscript{TM} and COFECHA [@holmes1986], producing mean tree series of incremental growth.
-Additionally, cambial age of each increment was established by counting years from the inner most ($a = 0$) ring at the pith outward; on tangentially bored cores, missing rings to the pith were estimated.
+Additionally, cambial age of each increment was established by counting years from the inner most ring at the pith ($a = 0$) outward; on tangentially bored cores, missing rings to the pith were estimated.
 
@@
 
-caption <- "Overview of urban sampling locations and respective tree species coverage. Individual trees were sampled up to to three times to obtain a mean-tree ring width series."
+caption <- "Overview of urban sampling locations and respective tree species coverage. Individual trees were sampled two or tree times to obtain a mean-tree ring width series."
 caption_ov_short <- caption
@@
 
- (GAM, or GAMM for mixed models/hierarchical models).
 We applied hierarchical generalized additive models (GAM) to estimate the relationship of several covariates with stem diamater growth ($DBH$).
-GAMs, as an extension of generalized linear models [@wood2017], allow modeling response variables as parametric and non-parametric combinations of smoothed explanatory covariates, and can assume non-normal errors and response distributions.
+GAMs, as an extension of generalized linear models [@wood2017], allow modeling response variables as parametric and non-parametric combinations of smoothed explanatory covariates, and can assume non-normal response distributions.
 These smooths are constructed by summation of base functions of varying complexity and form, analogous to scatterplot smoothing [@hastie2017], which provides a high degree of flexibility, ideal for fitting ecosystem dynamics which are rarely linear [@pedersen2019], or correctly represented with deterministic functional forms (e.g.
@@
 where $Y$ is taken from an appropriate distribution and corresponding link function $g$, $\beta_0$ is the intercept and $f_i$ represents a smooth function of a predictor [@pedersen2019], and $\epsilon \sim \mathcal{N}(0, \sigma ^2)$.
-Note, that $f_i$ consists of a smooth (e.g. spline) constructed via basis functions of different form and complexity, multiplied by a coefficient:
-
-\begin{equation}
-f_i(x_i)~=~\sum_{k = 1}^{K} \beta_{i, k} b_{i,k}(x_i).
-(\#eq:gam-smoothers)
-\end{equation}
-
-Nested data structures (e.g., city districts) can be accounted for by introducing random effects [@wood2017], while spatial dependence between observations can be accounted for constructing smoothing functions with, for instance, northings and eastings [cf.
-@wood2017], which is  applied here.
+Nested data structures (e.g., city districts) can be accounted for by introducing random effects, while spatial dependence between observations can be accounted for by constructing smoothing functions with, for instance, northings and eastings [cf.
+@wood2017].
 
@@
 
-### Model development and selection
+### Stem diameter model development and selection
 
@@
 
-g(E(Y)) &= f(utmx_i,utmy_i) + f_j(age_i) + f_j(temp_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i \\
+g(E(DBH_i)) &=  f_j(age_i) + f_j(temp_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i \\
 
-g(E(Y)) &= f_j(age_i) + f_j(temp_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
+g(E(DBH_i)) &= f_j(age_i) + f_j(temp_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
 
-g(E(Y)) &= f_j(age_i) + f_j(temp_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
+g(E(DBH_i)) &= f_j(age_i) + f_j(temp_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
 
-g(E(Y)) &= f_j(age_i) +  f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
+g(E(DBH_i)) &= f_j(age_i) +  f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
 
-g(E(Y)) &= f_j(age_i) +  c_{i,j} + \tau_{i,k} + \epsilon_i  \\
+g(E(DBH_i)) &= f_j(age_i) +  c_{i,j} + \tau_{i,k} + \epsilon_i  \\
 
@@
 
-As the spatial extent and coverage varies between temperature and ancillary data, more complex models typically also had fewer total observations.
+where $g()$ is a log-link for $DBH \sim gamma$, and $i$, $j$ and $k$ are indices for observations, species and districts, respectively. 
+Above models were tested in combination with a global spatial smooth $f(utmx_i,utmy_i)$ to account for the spatial structure of observations, which reduced auto-correlation of model residuals to considerably (see supplemental information).
+**what's the impact of remaining auto correlation? inflated errors, larger coefficients?**
+As the spatial extent and coverage varies between temperature and ancillary data, more complex models typically also had fewer total observations, which prevents a direct comparison with information-based model selection criteria, such as Aikake's Information Criterion.
+We thus selected the model with the highest explanatory power, and comparatively larger sample size for final analyses.
+**further analyses with equal sample sizes (requires data collection / cutting down); but this is first look at potential of approach and initial insights**
 

### 2021-11-25 analysis/paper/paper.Rmd
method gam detal

@@
 
+- models fit using mgcv **citation** in R **citation**
+
 ### Dendrochronological analyses
+
 - GAM for cambial age of urban trees, split like @pretzsch2017 (<=1960, >1960)
--
+- lme4, used AR process to account for autocorrelation between observations
+- calculataed model averages for all years within year groups
 
@@
 
-g(E(DBH_i)) &=  f_j(age_i) + f_j(temp_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i \\
+g(E(DBH_i)) &=  f_j(age_i) + f_j(temp_i, age_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i \\
 
-g(E(DBH_i)) &= f_j(age_i) + f_j(temp_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
+g(E(DBH_i)) &= f_j(age_i) + f_j(temp_i, age_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
 
-g(E(DBH_i)) &= f_j(age_i) + f_j(temp_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
+g(E(DBH_i)) &= f_j(age_i) + f_j(temp_i, age_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
 
@@
 where $g()$ is a log-link for $DBH \sim gamma$, and $i$, $j$ and $k$ are indices for observations, species and districts, respectively. 
-Above models were tested in combination with a global spatial smooth $f(utmx_i,utmy_i)$ to account for the spatial structure of observations, which reduced auto-correlation of model residuals to considerably (see supplemental information).
+Models were implemented in `R` [@rcoreteam2021] with `mgcv::bam()` [@wood2017a;@li2020a] for a dramatic increase computational efficiency, allowing for rapid model development/comparison
+All models above were also tested with a global spatial smooth $f(utmx_i,utmy_i)$ using Gaussian process smooth [cf.
+p.
+242 in @wood2017] to account for the spatial structure of observations, which reduced auto-correlation of model residuals to considerably (see supplemental information).
 **what's the impact of remaining auto correlation? inflated errors, larger coefficients?**
+
+- $f_j$ for temperature interaction with age, implemented as tensor smooth [@wood2006], assuming that temperature impacts vary across trees' lives.
+- Reader is referred to detailed implementation in the supplemental material code
+
 As the spatial extent and coverage varies between temperature and ancillary data, more complex models typically also had fewer total observations, which prevents a direct comparison with information-based model selection criteria, such as Aikake's Information Criterion.

### 2021-11-25 analysis/paper/paper.Rmd
stat gam cont

--- 2021-11-24 analysis/paper/paper.Rmd
@@
 <!-- Species-specifc growth responses in relation across Berlin were assessed in relat -->
-We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of excess heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; LandSat-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section$~$\@ref(stat-gam) for details).
+We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of excess heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; LandSat-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section$~$\@ref(sec:stat-gam) for details).
 Covariates were extracted at 150 and 300$~m$ to infer the impact of reference scale of the urban fabric on tree growth. 
@@
 
-## Data sources {#dsources}
+## Data sources {#sec:dsources}
 
@@
 Resolution and radius are provided in $m$, the latter is the buffer in which data was averaged around each trees.
-A zero-radius refers to a point extraction from categorical and location specific data.**"
+A zero-radius refers to a point extraction from categorical and location specific data."
 caption_ov_short <- "Data descriptions"
@@
 
-caption_trees <- "Binned age-distribution for trees in Berlin data set, and entries missing age information. "
-caption_trees_short <- "Binned age-distribution for trees in Berlin data set."
+caption_trees <- "Binned age-distribution for genera in Berlin data set, and entries missing age information. "
+caption_trees_short <- "Binned age-distribution for genera in Berlin data set."
 
@@
 The air temperatures from the Berlin environmental atlas (EnvAt) are model outputs that are representations of typical summer conditions at 0400, 1400 and 2200 hours; these data are provided at city block basis (spatial polygons), from which weighted averages were extracted.
-UrbClim air temperatures are hourly model outputs [100$~m$ resolution, @deridder2015] based on ERA5 re-analyses data (ECMWF) for which observations from the hottest month  available (June, 2011) were averaged to hours equivalent to EnvAt data by using a window of $\pm~1~$hour (i.e., 0300 to 0500, etc.).
+UrbClim air temperatures are hourly model outputs [100$~m$ resolution, @deridder2015] based on ERA5 re-analyses data (ECMWF) for which observations from the hottest month  available (June, 2011) were averaged to hours equivalent to Berlin Environmental Atlas (referred to as Berlin EnvAt) data by using a window of $\pm~1~$hour (i.e., 0300 to 0500, etc.).
 Subsequently, a land-use and land-cover mask (CORINE; European Union, Copernicus Land Monitoring Service 2018, European Environment Agency) was used to define urban and rural/forested areas.
@@
 where $T$ is temperature ($^\circ C$) $x$ and $y$ define an urban grid cell.
-The surface UHI data set by @chakraborty2019 derives its measure in a similar fashion and the reader is referred to the detailed description therein; note this data set provides day and night-time averaged UHI estimates at 500$~m$ resolution, which were extracted for the hottest summer in this record (2007).
+The LandSat-derived surface UHI data set by @chakraborty2019 (referred to as LandSat) estimates its measure in a similar fashion and the reader is referred to the detailed description therein; note this data set provides day and night-time averaged UHI estimates at 500$~m$ resolution, which were extracted for the hottest summer in this record (2007).
 
-### Ancillary environmental
+### Ancillary environmental data
 
@@
 
-## Statistical Analyses {#stat-gam}
+## Statistical Analyses {#sec:stat-gam}
 
@@
 @wood2017].
+All models were implemented in `R` [@rcoreteam2021] using functions available in the package `mgcv` [@wood2017].
 
 ### Dendrochronological analyses
+
 - GAM for cambial age of urban trees, split like @pretzsch2017 (<=1960, >1960)
--
+- Models were implemented with `mgcv::gamm()` to leverage correlation structures made available through package `nlme` [@pinheiro2021].
+A $3^{rd}$-order autocorrelation-moving average (ARMA) correlation structure was assumed applied to account for the dependency of observations across years for each tree, as is frequently the case for tree growth dynamics [e.g., see @Fritts1989].
+- calculated model averages for all years within year groups
+
+$e_i = \phi e_{i−1} + \epsilon_i$
 
@@
 
-g(E(DBH_i)) &=  f_j(age_i) + f_j(temp_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i \\
+g(DBH_i) &=  f_j(age_i) + f_j(temp_{t,i}, age_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i \\
 
-g(E(DBH_i)) &= f_j(age_i) + f_j(temp_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
+g(DBH_i) &= f_j(age_i) + f_j(temp_{t,i}, age_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
 
-g(E(DBH_i)) &= f_j(age_i) + f_j(temp_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
+g(DBH_i) &= f_j(age_i) + f_j(temp_{t,i}, age_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
 
-g(E(DBH_i)) &= f_j(age_i) +  f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
+g(DBH_i) &= f_j(age_i) +  f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
 
-g(E(DBH_i)) &= f_j(age_i) +  c_{i,j} + \tau_{i,k} + \epsilon_i  \\
+g(DBH_i) &= f_j(age_i) +  c_{i,j} + \tau_{i,k} + \epsilon_i \\
 
@@
 
-where $g()$ is a log-link for $DBH \sim gamma$, and $i$, $j$ and $k$ are indices for observations, species and districts, respectively. 
-Above models were tested in combination with a global spatial smooth $f(utmx_i,utmy_i)$ to account for the spatial structure of observations, which reduced auto-correlation of model residuals to considerably (see supplemental information).
+where $g()$ is a log-link for $DBH \sim gamma$, and $i$, $j$ and $k$ are indices for observations, species and districts, respectively, and $t$ refers to an (excess) heat measure from UrbClim, Berlin EnvAt or LandSat at different times (morning, afternoon/day, night; cf.
+Section$~$\@ref(sec:dsources)).
+Models were implemented with `mgcv::bam()` [@wood2017a;@li2020a] for a dramatic increase computational efficiency, allowing for rapid model development/comparison.
+All models above were also tested with a global spatial smooth $f(utmx_i,utmy_i)$ using Gaussian process smooth [cf.
+p.
+242 in @wood2017] to account for the spatial structure of observations, which reduced auto-correlation of model residuals considerably (see supplemental information).
 **what's the impact of remaining auto correlation? inflated errors, larger coefficients?**
-As the spatial extent and coverage varies between temperature and ancillary data, more complex models typically also had fewer total observations, which prevents a direct comparison with information-based model selection criteria, such as Aikake's Information Criterion.
-We thus selected the model with the highest explanatory power, and comparatively larger sample size for final analyses.
-**further analyses with equal sample sizes (requires data collection / cutting down); but this is first look at potential of approach and initial insights**
+The functions $f_j$ are for species-specific smooths [i.e., with individual smoothness penalties and functional shapes as detailed by @pedersen2019]; the interaction between temperature interaction and age were implemented as tensor smooths [@wood2006] to account for the different variable scales (i.e., units).
+Readers are referred to the detailed implementation in the supplemental material code.
+The covariates for planting bed area and soil nutrient availability were log transformed to account for their skewed distribution, improving the estimation of coefficients for their respective basis functions.
 
-<!-- We caution, that 1) the presented model results are preliminary, 2) initial diagnostics indicate that modification of the current model structure (e.g.
-by accounting for individual-level variability) may be required and 3) that further, time-intensive data cleaning is necessary.
--->
-<!-- First tests, however, showed that while the magnitude of effects changes with different model structures, their direction appears stable. -->
+We selected the model with the highest explanatory power, based on residual deviance and observed vs. predicted fit, with comparatively largest sample size for final analyses.
+As the spatial extent and coverage varied between temperature and ancillary data, more complex models (and specifically those including planting bed area and LandSat temperatures) typically also had fewer total observations, which prevents a direct comparison with information-based model selection criteria, such as Aikake's.
+For this reason, model structures were generally kept relatively simple (i.e., few terms, interactions and restricted number of basis functions) reducing the potential for choosing overly complex or over-fitted models.
+Future research may focus on collecting additional data (e.g., increasing the coverage on planting bed area).
+We chose to carry out the analysis in its current form rather than on considerably smaller, but comparable sample sizes to identify the strongest relationships in the existing data and thus to highlight the utility of the approach *per se* and for Berlin in particular.
 
-<!-- ## Available and required data -->
-
-<!-- Additional remote sensing data for local context, higher resolution UHI data, and most important, temporally-resolved tree growth data are still required and/or would greatly improve the confidence in effect size estimates (see Section$~$\@ref(sec:intro)).
--->
-<!-- See Tab.$~$\@ref(tab:tab-data-req) for details on data sets. -->
-
-```{r tab-data-req,echo=FALSE, eval = FALSE}
-
-dat_req <- read.csv(here::here("analysis", "data", "tables", "01_data_requirements.csv"), header = TRUE)
-
-caption_data_req <- "Data requirements for analysis, including currently available/accessed and required/desired data."
-caption_data_req_short <- "Data requirements for analysis."
-
-if (grepl("docx", out_format) ||
-    knitr::is_latex_output() ) {
-  table_dat_req <- dat_req %>% 
-  dplyr::select(-owned) %>% 
-  knitr::kable(booktabs = TRUE,
-
-             caption = caption_data_req,
-             caption.short = caption_data_req_short,
-             format = "latex",
-             col.names = c("Subject / Relevance",
-                          "Desc.",
-                          "Type",
-                          "Obsv. (n)",
-                          "Resolution (m)",
-                          "Source")) %>% 
-  kableExtra::kable_styling(full_width = TRUE, 
-                            font_size = 11) %>% 
-  kableExtra::row_spec(0, bold = TRUE, font_size = 12) %>% 
-  kableExtra::row_spec(9,italic =  TRUE) %>% 
-  kableExtra::pack_rows("available or accessed", 1, 4) %>%
-  kableExtra::pack_rows("required and/or desired", 5, 9) %>% kableExtra::landscape()
-
-} else {
-  table_dat_req <- dat_req %>% 
-  dplyr::select(-owned) %>% 
-  knitr::kable(booktabs = TRUE,
-               format = "html",
-               caption = "Data requirements for analysis, including currently available/accessed and required/desired data.",
-             caption.short = "Data requirements for analysis",
-             col.names = c("Subject/Relevance",
-                          "Desc.",
-                          "Type",
-                          "Obsv. (n)",
-                          "Resolution (m)",
-                          "Source")) %>% 
-  kableExtra::kable_styling(full_width = TRUE, 
-                            font_size = 10) %>% 
-  kableExtra::row_spec(0, bold = TRUE) %>% 
-  kableExtra::pack_rows("available or accessed", 1, 4) %>%
-  kableExtra::pack_rows("required and/or desired", 5, 9)
-
-}
-
- table_dat_req
-
-```
-
-# Preliminary results {#sec:results}
+# Results {#sec:results}
 

### 2021-11-25 analysis/paper/paper.Rmd
methods done?

@@
 
-These limitations could be overcome by relying on dendroecological surveys (i.e., incremental growth) or inventories (single or repeat) combined with pertinent environmental data with adequate spatio-temporal coverage and resolution.
-Such data are increasingly more available due to open data policies and because their value is recognized across several domains [e.g. @ossola2020].
+These limitations could be overcome by developing extensive dendroecological surveys (i.e., incremental growth) and/or inventories (single or repeat) combined with pertinent environmental data with adequate spatio-temporal coverage and resolution.
+Inventories are logistically and financially more feasible, and - together with environmental data - are increasingly more available [e.g.
+@ossola2020] due to open data policies and their value being recognized across domains for urban greenspace planning and adaptation [@hansen2019; @monteiro2020].
 Berlin, as one of the greenest cities in Europe, provides an openly accessible tree inventory, with spatio-temporal environmental data sets relevant to tree growth. 
@@
 
-- GAM for cambial age of urban trees, split like @pretzsch2017 (<=1960, >1960)
-- Models were implemented with `mgcv::gamm()` to leverage correlation structures made available through package `nlme` [@pinheiro2021].
-A $3^{rd}$-order autocorrelation-moving average (ARMA) correlation structure was assumed applied to account for the dependency of observations across years for each tree, as is frequently the case for tree growth dynamics [e.g., see @Fritts1989].
-- calculated model averages for all years within year groups
+We assessed trends in annual growth dynamics of urban trees across the 20$^{th}$ century for 1920-1960 and 1961-2001 [similar grouping as  @pretzsch2017 and @dahlhausen2018] with a hierarchical GAM implemented in `mgcv::gamm()` to leverage auto-correlation structures made available through the package `nlme` [@pinheiro2021].
+Annual growth was modeled as:
 
-$e_i = \phi e_{i−1} + \epsilon_i$
+$$ g(\Delta r_i) =  f(year_i) + f_j(cambial~age_i)  + c_{i,j} + \tau_{i,k} + e_i $$
+where $g()$ is a log-link for $\Delta r \sim Gamma$, $\Delta r$ is the annual radial increment for observation $i$. 
+A global temporal (by year) and time-dependent ($j$, $\leq 1960$ or $>1960$) trend in cambial age were implemented with thin plate regression splines (default smoothing function in `mgcv`); $c_j$ is a time-group dependent intercept, while $tau$ represents a matrix of random effect coefficients for $k$ species to account for differences in growth patterns, and $e_i = \phi e_{i−1} + \epsilon_i$.
+A $3^{rd}$-order autocorrelation-moving average (ARMA) correlation structure was applied (i.e., $\phi(3,1)$) to account for the dependency of $\Delta r$ across years for each tree, as is frequently the case for tree growth [e.g., see @Fritts1989]; the detailed implementation for this model is given in the supplemental material code.
+$\Delta r$ was then derived for a range of cambial ages, and averaged for both time groups, allowing a comparison of recent to earlier growth.
+We acknowledge that tree cores obtained at breast height do not represent absolute tree age.
+However, here they serve as a proxy for growth between young ($>1960$) and older individuals to contextualize growth patterns inferred from the larger-scale tree inventory.
 
@@
 
-**Table with model x species sample sizes or infer from figure?**
-
-Model forms:
+The diameter ($DBH$) of the ten most abundant species were modeled using GAMs as follows:
 
@@
 
-where $g()$ is a log-link for $DBH \sim gamma$, and $i$, $j$ and $k$ are indices for observations, species and districts, respectively, and $t$ refers to an (excess) heat measure from UrbClim, Berlin EnvAt or LandSat at different times (morning, afternoon/day, night; cf.
-Section$~$\@ref(sec:dsources)).
+where $g()$ is a log-link for $DBH \sim Gamma$, and $i$, $j$ are indices for observations and species, respectively, and $t$ refers to an (excess) heat measure from UrbClim, Berlin EnvAt or LandSat at different times (morning, afternoon/day, night; cf.
+Section$~$\@ref(sec:dsources)); $c$ is a species-dependent intercept, while $tau$ represents a matrix of random effect coefficients for $k$ districts to account for differing management regimes across the city.
 Models were implemented with `mgcv::bam()` [@wood2017a;@li2020a] for a dramatic increase computational efficiency, allowing for rapid model development/comparison.
@@
 The covariates for planting bed area and soil nutrient availability were log transformed to account for their skewed distribution, improving the estimation of coefficients for their respective basis functions.
+We selected the model with the highest explanatory power, based on residual deviance and observed vs. predicted fit, with comparatively largest sample size for final analyses.
+From this model we derived age and species dependent $DBH$ averages across a temperature measure from predicted values in 5-year age groups starting at 30, 45, 60, 75, 90.
 
-We selected the model with the highest explanatory power, based on residual deviance and observed vs. predicted fit, with comparatively largest sample size for final analyses.
-As the spatial extent and coverage varied between temperature and ancillary data, more complex models (and specifically those including planting bed area and LandSat temperatures) typically also had fewer total observations, which prevents a direct comparison with information-based model selection criteria, such as Aikake's.
-For this reason, model structures were generally kept relatively simple (i.e., few terms, interactions and restricted number of basis functions) reducing the potential for choosing overly complex or over-fitted models.
-Future research may focus on collecting additional data (e.g., increasing the coverage on planting bed area).
-We chose to carry out the analysis in its current form rather than on considerably smaller, but comparable sample sizes to identify the strongest relationships in the existing data and thus to highlight the utility of the approach *per se* and for Berlin in particular.
+As the spatial extent and coverage varied between temperature and ancillary data, more complex models (and specifically those including planting bed area and LandSat temperatures) typically also had fewer total observations, which prevented a direct comparison with information-based model selection criteria, such as Aikake's.
+For this reason, simpler model structures were preferred (i.e., few terms, interactions and restricted number of basis functions), reducing the potential for choosing over-fitted models.
+Future research may focus on collecting additional data (e.g., increasing the coverage on planting bed area) and subsequently deriving species-specific smooths for ancillary environmental covariates.
+We chose to carry out the analysis in its current form rather than on considerably smaller but comparable sample sizes across models, to identify the strongest relationships in the existing data.
+This allowed us to highlight the utility of the approach *per se* and for Berlin in particular.
 

### 2021-11-29 analysis/paper/paper.Rmd
Adding figures

@@
 We acknowledge that tree cores obtained at breast height do not represent absolute tree age.
-However, here they serve as a proxy for growth between young ($>1960$) and older individuals to contextualize growth patterns inferred from the larger-scale tree inventory.
+However, here they serve as a proxy for growth between young ($\gtrapprox 1960$) and older individuals to contextualize growth patterns inferred from the larger-scale tree inventory.
 
@@
 **what's the impact of remaining auto correlation? inflated errors, larger coefficients?**
-The functions $f_j$ are for species-specific smooths [i.e., with individual smoothness penalties and functional shapes as detailed by @pedersen2019]; the interaction between temperature interaction and age were implemented as tensor smooths [@wood2006] to account for the different variable scales (i.e., units).
+Further, we implemented the interaction between temperature and age (i.e., $f_j(temp, age)$ as tensor smooths [@wood2006] to account for the different variable scales (i.e., units); all models were also tested without this interaction using a thin plate regression spline smooth for temperature (not shown in equations above).
+The functions $f_j$ are for species-specific smooths [i.e., with individual smoothness penalties and functional shapes as detailed by @pedersen2019].
 Readers are referred to the detailed implementation in the supplemental material code.
 The covariates for planting bed area and soil nutrient availability were log transformed to account for their skewed distribution, improving the estimation of coefficients for their respective basis functions.
+
+Considering all combinations of (excess) heat measures and covariates (with point, as well as 150/300$~m$ extractions), a total of `r nrow(drake::readd(bam_dbh_filtered))` models were applied.
 We selected the model with the highest explanatory power, based on residual deviance and observed vs. predicted fit, with comparatively largest sample size for final analyses.
@@
 # Results {#sec:results}
+
+(ref:cambial-effect) Contrasting growth trends for recent and earlier radial increments (i.e. stem growth)
+
+```{r fig-cambial-effect, fig.cap="(ref:cambial-effect)", out.width="50%", include=TRUE, eval=TRUE}
+
+knitr::include_graphics(here::here("analysis/figures/fig-biwi-growth.png"))
+
+```
+
+(ref:model-deviance) Fig caption
+
+```{r fig-model-deviance, fig.cap="(ref:model-dev)", out.width="100%", include=TRUE, eval=TRUE}
+
+knitr::include_graphics(here::here("analysis/figures/fig_model_deviance.png"))
+
+```
+
+(ref:model-performance) Predicted vs. observed best model day2007 and LCZ6
+
+```{r fig-model-performance, fig.cap="(ref:model-performance)", out.width="50%", include=TRUE, eval=TRUE}
+
+knitr::include_graphics(here::here("analysis/figures/fig-gam-dbh_temp-day2007_lcz6_obs_pred.png"))
+
+```
+
+(ref:heat-effect) Age group-dependent temperature effect
+
+```{r fig-heat-effect, fig.cap="(ref:heat-effect)", out.width="100%", include=TRUE, eval=TRUE}
+
+knitr::include_graphics(here::here("analysis/figures/fig-gam-dbh_temp-day2007_lcz6.png"))
+
+```
+
+(ref:lcz6-effect) Global effect of LCZ6 on growth
+
+```{r fig-lcz6-effect, fig.cap="(ref:lcz6-effect)", out.width="50%", include=TRUE, eval=TRUE}
+
+knitr::include_graphics(here::here("analysis/figures/fig-gam-dbh_temp-day2007_xvar-lcz6_tilia.png"))
+
+```
 

### 2021-11-29 R/functions.R
adjusted raster resolution of building height to 5 to match other rasterize/fasterize calls

@@ -345,7 +345,7 @@ download_building_height <- function(fpath = "./analysis/data/raw_data/spatial_a
 
     rm(building_height1, building_height2)
 
-    building_full_raster <- raster::raster(building_full, resolution=10)
+    building_full_raster <- raster::raster(building_full, resolution=5)
 
     building_rasterized <- raster::rasterize(building_full,
                                              building_full_raster,

### 2021-11-29 R/functions.R
replaced rasterize with fasterize

@@ -347,9 +347,14 @@ download_building_height <- function(fpath = "./analysis/data/raw_data/spatial_a
 
     building_full_raster <- raster::raster(building_full, resolution=5)
 
-    building_rasterized <- raster::rasterize(building_full,
-                                             building_full_raster,
-                                             field = "mean_ndom")
+    # building_rasterized <- raster::rasterize(building_full,
+    #                                          building_full_raster,
+    #                                          field = "mean_ndom")
+
+    building_rasterized <- fasterize::fasterize(building_full,
+                         building_full_raster,
+                         field = "mean_ndom",
+                         fun = "max")
 
 
 

### 2021-11-29 analysis/paper/paper.Rmd
added temp maps

@@
 Resolution and radius are provided in $m$, the latter is the buffer in which data was averaged around each trees.
-A zero-radius refers to a point extraction from categorical and location specific data."
+A zero-radius refers to a point extraction from categorical and location specific data.
+Polygons with radius data were rasterized to a resolution of $5~m$"
 caption_ov_short <- "Data descriptions"

### 2021-11-29 R/functions.R

@@ -3301,6 +3301,560 @@ make_uhi_urbclim_plot <- function(uhi_rast,
 
 }
 
+#' Make temp overview plot
+#'
+#' @param landsat list, uhi stacks
+#' @param envat rasterstack, berlin heat model
+#' @param urbclim list, urbclim rasters
+#' @param berlin_poly sf, outlines
+#' @param ylim numeric, UTM coordinates
+#' @param xlim numeric, UTM coordinates
+#' @param base_size
+#' @param file
+#' @param height
+#' @param width
+#' @param dpi
+plot_temp_maps <- function(landsat,
+                           envat,
+                           urbclim,
+                           berlin_poly,
+                           ylim = c(5789783, 5853683),
+                           xlim = c(360062, 424844),
+                           # ylim = c(52.275, 52.85),
+                           base_size = 18,
+                           file,
+                           height,
+                           width,
+                           dpi){
+
+    # extrafont::loadfonts(device = "win",quiet = TRUE)
+    extrafont::loadfonts(device = "pdf",quiet = TRUE)
+
+    mid_rescaler <- function(mid = 0) {
+        function(x, to = c(0, 1), from = range(x, na.rm = TRUE)) {
+            scales::rescale_mid(x, to, from, mid)
+        }
+    }
+
+    # main_crs <-  raster::crs(landsat$Summertime_gridded_UHI_data$day)
+    main_crs <- berlin_poly %>% sf::st_crs()
+
+
+    landsat$Summertime_gridded_UHI_data$day <-
+        raster::projectRaster(
+            landsat$Summertime_gridded_UHI_data$day,
+            crs = main_crs$wkt
+        )
+    landsat$Summertime_gridded_UHI_data$night <-
+        raster::projectRaster(
+            landsat$Summertime_gridded_UHI_data$night,
+            crs = main_crs$wkt
+        )
+
+
+
+    # ensure all assets have same proj
+    berlin_poly <- sf::st_transform(berlin_poly,
+                                    crs = main_crs)
+    urbclim <- raster::projectRaster(from = urbclim,
+                                     crs = main_crs$wkt)  %>%
+        stars::st_as_stars(proxy = FALSE)
+    urbclim <- urbclim[berlin_poly]
+    envat <- raster::projectRaster(from = envat,
+                                   crs = main_crs$wkt) %>%
+        stars::st_as_stars(proxy = FALSE)
+    #
+
+
[…]

### 2021-11-29 R/plan.R

@@ -600,6 +600,19 @@ plan <- drake::drake_plan(
                                                                                UHI~(degree*C)))),
 
 
+   ### map: temp/UHI overview
+
+   plot_uhi_overview_maps = plot_temp_maps(landsat = uhi_stacks,
+                          envat = berlin_heat_model_2015,
+                          urbclim = raster::stack(uhi_urbclim),
+                          berlin_poly = berlin_polygons,
+                          base_size = 18,
+                          file = "./analysis/figures/fig_temp_maps.png",
+                          dpi = 300,
+                          height = 12,
+                          width = 12),
+
+
    ### bar: Tree counts ------------
 
     plot_tree_sums_bar = tree_sums_bar_plot(full_data_set_clean,

### 2021-11-29 analysis/paper/paper.Rmd
added temp maps

@@
 knitr::include_graphics(here::here("analysis/figures/map_00_studyarea.png"))
-
-# plot_count_map
-
-# plot(1:10)
 ```
@@
 
-Temperature and UHI data were summarized temporally either by the provider or manually to provide a characteristic representation of heat loading during summer at different times (morning, afternoon/day, night), from which tree averages (radius of 150$~m$) were calculated.
+Temperature and UHI data were summarized temporally either by the provider or manually to provide a characteristic representation of heat loading during summer at different times (morning, afternoon/day, night), from which tree averages (radius of 150$~m$) were calculated (Figure$~$\@ref(fig:fig-temp-overview).
 Two data sets of urban air and one surface temperature were tested as explanatory variables in GAM models.
@@
 The LandSat-derived surface UHI data set by @chakraborty2019 (referred to as LandSat) estimates its measure in a similar fashion and the reader is referred to the detailed description therein; note this data set provides day and night-time averaged UHI estimates at 500$~m$ resolution, which were extracted for the hottest summer in this record (2007).
+
+(ref:temp-overview) Temp overviews; rows are landsat, berlin environmental atlas, and UrbClim, columns are morning, day/afternoon, night.
+
+```{r fig-temp-overview, fig.cap="(ref:study-temp)",fig.width = 12, fig.height = 12, out.width="100%"}
+
+knitr::include_graphics(here::here("./analysis/figures/fig_temp_maps.png"))
+```
 

### 2021-11-30 analysis/paper/paper.Rmd
adjusting eqs

@@
 
-$$MAD = median(|X_i - median(X)|) $$
+\begin{equation}
+  MAD = median(\|X_i - median(X)\|) 
+  (\#eq:mad)
+\end{equation}
+
 This approach is considered conservative (see supplementary information), yet all analyses were carried forward with the unfiltered and filtered data - no considerable differences were found, thus subsequent sections are based on the filtered data.
@@
 
-Temperature and UHI data were summarized temporally either by the provider or manually to provide a characteristic representation of heat loading during summer at different times (morning, afternoon/day, night), from which tree averages (radius of 150$~m$) were calculated (Figure$~$\@ref(fig:fig-temp-overview).
+Temperature and UHI data were summarized temporally either by the provider or manually to provide a characteristic representation of heat loading during summer at different times (morning, afternoon/day, night), from which tree averages (radius of 150$~m$) were calculated (Figure$~$\@ref(fig:fig-temp-overview)).
 Two data sets of urban air and one surface temperature were tested as explanatory variables in GAM models.
@@
 The urban heat loading was then calculated as
-$$UHI_{x,y} = T_{Air_{2m}~x,y} - \overline{T_{Air_{2m}~Rural}},$$
+
+\begin{equation}
+ UHI_{x,y} = T_{Air_{2m}~x,y} - \overline{T_{Air_{2m}~Rural}},
+  (\#eq:uhi)
+\end{equation}
+
 where $T$ is temperature ($^\circ C$) $x$ and $y$ define an urban grid cell.
@@
 (ref:temp-overview) Temp overviews; rows are landsat, berlin environmental atlas, and UrbClim, columns are morning, day/afternoon, night.
+**ADD Labels for Data source and time of day at left and bottom**
 
-```{r fig-temp-overview, fig.cap="(ref:study-temp)",fig.width = 12, fig.height = 12, out.width="100%"}
+```{r fig-temp-overview, fig.cap="(ref:temp-overview)",fig.width = 12, fig.height = 12, out.width="100%"}
 
@@
 
-## Dendrochronological sampling
+## Dendrochronological sampling {#sec:methods-dendro}
 
@@
 
-$$ g(\Delta r_i) =  f(year_i) + f_j(cambial~age_i)  + c_{i,j} + \tau_{i,k} + e_i $$
+\begin{equation}
+g(\Delta r_i) =  f(year_i) + f_j(cambial~age_i)  + c_{i,j} + \tau_{i,k} + e_i 
+(\#eq:gam-cambial)
+\end{equation}
+
 where $g()$ is a log-link for $\Delta r \sim Gamma$, $\Delta r$ is the annual radial increment for observation $i$. 
@@
 \begin{align}
+ UHI_{x,y} = T_{Air_{2m}~x,y} - \overline{T_{Air_{2m}~Rural}},
+  (\#eq:uhi)
+\end{align}
 
-g(DBH_i) &=  f_j(age_i) + f_j(temp_{t,i}, age_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i \\
+test
 
-g(DBH_i) &= f_j(age_i) + f_j(temp_{t,i}, age_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
+$$ g(DBH_i) = f(x_i,y_i) + f_j(age_i) + f_j(temp_{t,i}, age_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i$$
 
-g(DBH_i) &= f_j(age_i) + f_j(temp_{t,i}, age_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
+$$g(DBH_i) = f(x_i,y_i) + f_j(age_i) + f_j(temp_{t,i}, age_i) + c_{i,j} + \tau_{i,k} + \epsilon_i$$
 
-g(DBH_i) &= f_j(age_i) +  f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i  \\
+$$g(DBH_i) = f(x_i,y_i) + f_j(age_i) +  f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i$$
 
-g(DBH_i) &= f_j(age_i) +  c_{i,j} + \tau_{i,k} + \epsilon_i \\
-
-\end{align}
+$$g(DBH_i) = f(x_i,y_i) + f_j(age_i) +  c_{i,j} + \tau_{i,k} + \epsilon_i $$
 
@@
 Section$~$\@ref(sec:dsources)); $c$ is a species-dependent intercept, while $tau$ represents a matrix of random effect coefficients for $k$ districts to account for differing management regimes across the city.
-Models were implemented with `mgcv::bam()` [@wood2017a;@li2020a] for a dramatic increase computational efficiency, allowing for rapid model development/comparison.
-All models above were also tested with a global spatial smooth $f(utmx_i,utmy_i)$ using Gaussian process smooth [cf.
+A global spatial smooth $f(x_i,y_i)$ (representing projected coordinates in UTM) via a Gaussian process [cf.
 p.
-242 in @wood2017] to account for the spatial structure of observations, which reduced auto-correlation of model residuals considerably (see supplemental information).
+242 in @wood2017] was included to account for the spatial structure of observations, which reduced auto-correlation of model residuals considerably (see supplemental information).
+These were compared to a sub-set of models without a spatial component (cf. Equation$~$\@ref(eq:gams-nocovariate), and Figure$~$\@ref(fig:fig-model-deviance) but not further discussed there).
+We also tested a suite of models without the spatial smooth for comparison
 **what's the impact of remaining auto correlation? inflated errors, larger coefficients?**
+
 Further, we implemented the interaction between temperature and age (i.e., $f_j(temp, age)$ as tensor smooths [@wood2006] to account for the different variable scales (i.e., units); all models were also tested without this interaction using a thin plate regression spline smooth for temperature (not shown in equations above).
 The functions $f_j$ are for species-specific smooths [i.e., with individual smoothness penalties and functional shapes as detailed by @pedersen2019].
-Readers are referred to the detailed implementation in the supplemental material code.
 The covariates for planting bed area and soil nutrient availability were log transformed to account for their skewed distribution, improving the estimation of coefficients for their respective basis functions.
+Note, that Equation$~$\@ref(eq:gams-age) was considered as the appropriate null model for interpretations.
+Models were implemented with `mgcv::bam()` [@wood2017a;@li2020a] and readers are referred to the detailed implementation in the supplemental material code.
 
@@
 
-As the spatial extent and coverage varied between temperature and ancillary data, more complex models (and specifically those including planting bed area and LandSat temperatures) typically also had fewer total observations, which prevented a direct comparison with information-based model selection criteria, such as Aikake's.
-For this reason, simpler model structures were preferred (i.e., few terms, interactions and restricted number of basis functions), reducing the potential for choosing over-fitted models.
+As the spatial extent and coverage varied between temperature and ancillary data, more complex models (and specifically those including planting bed area and LandSat temperatures) typically also had fewer total observations.
+While this prevented a full comparison with information-based model selection criteria, such as Aikake's (AIC), the appropriateness of models that differed only in their implementation of the temperature-age interaction (i.e., $f_j(temp_i)$ $vs.$ $f_j(temp_i, age_i)$) could be assessed.
+For this reason, the suite of developed models presented above were limited to comparatively simple structures (i.e., few terms, interactions and restricted number of basis functions), reducing the potential for choosing over-fitted models without formal comparison.
+AIC is calculated as:
+
+$$AIC= −2log(L) \cdot 2k$$
+
+where $L$ is the maximum likelihood estimate, and $k$ is the number of parameters.
+Where any two comparable models had a $| \Delta AIC| > 10$, we considered the model with the lower score more suitable. 
 Future research may focus on collecting additional data (e.g., increasing the coverage on planting bed area) and subsequently deriving species-specific smooths for ancillary environmental covariates.
@@
 This allowed us to highlight the utility of the approach *per se* and for Berlin in particular.
+**Measures used to asses goodness-of-fit were residual deviance, RSME and MAE (add formuals)**
 
@@
 
-(ref:cambial-effect) Contrasting growth trends for recent and earlier radial increments (i.e. stem growth)
+## Growth trend dynamics
+
+Recently established annual, incremental growth is on average greater as compared to earlier times (i.e., prior 1960; Figure$~$\@ref(fig:fig-cambial-effect)).
+The contrast is strongest in the first 30 years of cambial development, with clear indication that averages are statistically different up to approximately 22 years (cf.
+overlap of 95$~\%$confidence intervals).
+The predicted trajectories for both periods follow typical ontogenetic patterns, yet differ in shape.
+This may be due to inaccurately estimated cambial ages, explaining the largely monotonic decrease for recent growth due to missing the typical (but not always present) initial rise and fall in pith-near stages.
+This would inadvertently create a left-shift in Figure$~$\@ref(fig:fig-cambial-effect) for recent growth.
+Assuming that to be the case, the actual difference in average rates would be greater than presented here.
+
+(ref:cambial-effect) Rates for annual incremental growth differ between recent and earlier times, as predicted by a hierarchical GAM (see Section$~$\@ref(sec:methods-dendro)).
+Thick lines and bands are for mean and 95$~\%$ confidence interval of all annual predictions across a time group (fine lines).
+On average, early stages of recent growth (up to approximately 22 years) exceed that of earlier periods discernibly.
 
@@
 
-(ref:model-deviance) Fig caption
+## Model selection
 
-```{r fig-model-deviance, fig.cap="(ref:model-dev)", out.width="100%", include=TRUE, eval=TRUE}
+* Null-model high, but for predictive purposes more is better
+* Temperature-based models on average better than without, but planting bed area and lcz6 individually outperform - indication that these variables are indeed important; choice of best performing model, which includes lcz6, coroborated.
+*Landsat temp best on average - easily accessible/maintainable as compared to model outputs
+* $lcz6_{150}$ vs. $lcz6_{300}$ - scale of impact (further investigation, note in discussion) 
+* building height, as a proxy for urban fabric and incoming radiation had minimal effect (regardless of 150/300)
+
+(ref:model-deviance) Overview of model fit, expressed as explained deviance, for all tested models.
+The Panels distinguish between models without (top) and with (bottom) spatial smooths to account for autocorrelation of residuals.
+Colored sub-sections define models without and with the applied (excess) heat measures.
+Symbols identify models with an interaction between temperature and age (diamond), and without (circle), while colors highlight included covariates and their reference radius (i.e.
+150/300$~m$, but also note x-axis labels), and size indicates the model's observations.
+Generally, including temperature (interactions) improve model fits above the null-model (age only), and excluding planting bed area and LCZ6 cover, also above other non-temperature models.
+The LandSat-derived UHI measures provide the best fit on average, with the model including a temperature-age interaction and LCZ6 cover scoring highest.
+
+```{r fig-model-deviance, fig.cap="(ref:model-deviance)", out.width="100%", include=TRUE, eval=TRUE}
 
@@
 ```
+
+**Add table with ANOVA for interaction, non-interaction and MAE/RSME**
 
@@
 
-(ref:heat-effect) Age group-dependent temperature effect
+## Urban heat and urban fabric effects on stem growth
+
+(ref:heat-effect) Age group-dependent temperature effect; only within range
 
@@
 ```
+
+## Exposed trees
+
+**Add figure with tree locations where growth is below the average - color = UHI?**
 

### 2021-11-30 R/functions.R

@@ -2711,7 +2711,7 @@ model_summarize <- function(mgroups, bam_df, path_model_dir, path_model_files,pa
 
 
 
-    future::plan(future::multisession(workers = 12))
+    future::plan(future::multisession(workers = 4))
     mod_summary_list <-
         furrr::future_map(
             levels(mgroups),
@@ -3468,7 +3468,7 @@ plot_temp_maps <- function(landsat,
     # ENVAT Morning -----------------------------------------------------------
 
 
-    rn <- envat %>% slice(band, 1) %>% as.data.frame() %>% select(3) %>% range(na.rm = TRUE)
+    rn <- envat %>% dplyr::slice(band, 1) %>% as.data.frame() %>% dplyr::select(3) %>% range(na.rm = TRUE)
 
     rn[1] <- floor(rn[1])
     rn[2] <- ceiling(rn[2])
@@ -3526,7 +3526,7 @@ plot_temp_maps <- function(landsat,
 
 
     # ENVAT AFTERNOON -----------------------------------------------------------
-    rn <- envat %>% slice(band, 2) %>% as.data.frame() %>% select(3) %>% range(na.rm = TRUE)
+    rn <- envat %>% dplyr::slice(band, 2) %>% as.data.frame() %>% dplyr::select(3) %>% range(na.rm = TRUE)
     rn[1] <- floor(rn[1])
     rn[2] <- ceiling(rn[2])
 
@@ -3580,7 +3580,7 @@ plot_temp_maps <- function(landsat,
 
 
     # ENVAT NIGHT -----------------------------------------------------------
-    rn <- envat %>% slice(band, 3) %>% as.data.frame() %>% select(3) %>% range(na.rm = TRUE)
+    rn <- envat %>% dplyr::slice(band, 3) %>% as.data.frame() %>% dplyr::select(3) %>% range(na.rm = TRUE)
     rn[1] <- floor(rn[1])
     rn[2] <- ceiling(rn[2])
 
@@ -3640,7 +3640,7 @@ plot_temp_maps <- function(landsat,
     # urbclim Morning -----------------------------------------------------------
 
 
-    rn <- urbclim %>% slice(band, 1) %>% as.data.frame() %>% select(3) %>% range(na.rm = TRUE)
+    rn <- urbclim %>% dplyr::slice(band, 1) %>% as.data.frame() %>% dplyr::select(3) %>% range(na.rm = TRUE)
 
[…]

### 2021-11-30 R/plan.R

@@ -600,7 +600,7 @@ plan <- drake::drake_plan(
                                                                                UHI~(degree*C)))),
 
 
-   ### map: temp/UHI overview
+   ### map: temp/UHI overview ----------------
 
    plot_uhi_overview_maps = plot_temp_maps(landsat = uhi_stacks,
                           envat = berlin_heat_model_2015,
@@ -1012,14 +1012,14 @@ plan <- drake::drake_plan(
         output_format = bookdown::html_document2(),
         quiet = TRUE
     ),
-    # paper_word = rmarkdown::render(
-    #     knitr_in("./analysis/paper/paper.Rmd"),
-    #     output_dir = "./analysis/paper/",
-    #     output_file = file_out("paper_knit.docx"),
-    #     # output_file = "./paper_knit.html",
-    #     output_format = bookdown::word_document2(),
-    #     quiet = TRUE
-    # ),
+    paper_word = rmarkdown::render(
+        knitr_in("./analysis/paper/paper.Rmd"),
+        output_dir = "./analysis/paper/",
+        output_file = file_out("paper_knit.docx"),
+        # output_file = "./paper_knit.html",
+        output_format = bookdown::word_document2(),
+        quiet = TRUE
+    ),
     # paper_pdf = rmarkdown::render(
     #     knitr_in("./analysis/paper/paper.Rmd"),
     #     output_dir = "./analysis/paper/",

### 2021-12-01 analysis/paper/paper.Rmd
results sections heat temp map

@@
 
-if (grepl("docx", out_format) ||
-    knitr::is_latex_output() ) {
+if (grepl("docx", out_format)) {
+    table_ds <- dsources %>% 
+        knitr::kable(booktabs = TRUE,
+
+                     caption = caption,
+                     caption.short = caption,
+                     format = "pipe",
+                     # col.names = c("Subject/Relevance",
+                     #              "Desc.",
+                     #              "Type",
+                     #              "Obsv. (n)",
+                     #              "Resolution (m)",
+                     #              "Source")
+                     escape = FALSE) %>% 
+        kableExtra::landscape()
+} else if(knitr::is_latex_output() ){
     table_ds <- dsources %>% 
@@
 
-caption_trees <- "Binned age-distribution for genera in Berlin data set, and entries missing age information. "
+caption_trees <- "Binned age-distribution for street tree genera in Berlin data set, and entries missing age information. "
 caption_trees_short <- "Binned age-distribution for genera in Berlin data set."
 
-if (grepl("docx", out_format) ||
-    knitr::is_latex_output() ) {
+if (grepl("docx", out_format)) {
+    table_dat_req_kable <- age_tables %>% 
+        knitr::kable(booktabs = TRUE,
+                     format = "pipe",
+                     caption = caption_trees,
+                     caption.short = caption_trees_short
+        ) %>% 
+        kableExtra::row_spec(13, hline_after = TRUE) 
+} else if(knitr::is_latex_output() ){
     table_dat_req_kable <- age_tables %>% 
@@
                      caption.short = caption_trees_short
-             # col.names = c("Subject/Relevance",
-             #              "Desc.",
-             #              "Type",
-             #              "Obsv. (n)",
-             #              "Resolution (m)",
-             #              "Source")
         ) %>% 
@@
 
-if (grepl("docx", out_format) ||
-    knitr::is_latex_output() ) {
+if (grepl("docx", out_format) ){
+  table_ds <- sampling_site_overview %>% 
+  knitr::kable(booktabs = TRUE,
+
+             caption = caption,
+             caption.short = caption,
+             format = "pipe",
+             escape = FALSE) 
+} else if (knitr::is_latex_output() ) {
   table_ds <- sampling_site_overview %>% 
@@
 
-\begin{align}
- UHI_{x,y} = T_{Air_{2m}~x,y} - \overline{T_{Air_{2m}~Rural}},
-  (\#eq:uhi)
-\end{align}
+\begin{equation}
+g(DBH_i) = f(x_i,y_i) + f_j(age_i) + f_j(temp_{t,i}, age_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i 
+(\#eq:gam-full)
+\end{equation}
 
-test
+\begin{equation}
+g(DBH_i) = f(x_i,y_i) + f_j(age_i) + f_j(temp_{t,i}, age_i) + c_{i,j} + \tau_{i,k} + \epsilon_i
+(\#eq:gam-nocovariate)
+\end{equation}
 
-$$ g(DBH_i) = f(x_i,y_i) + f_j(age_i) + f_j(temp_{t,i}, age_i) + f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i$$
+\begin{equation}
+g(DBH_i) = f(x_i,y_i) + f_j(age_i) +  f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i 
+(\#eq:gam-notemp)
+\end{equation}
 
-$$g(DBH_i) = f(x_i,y_i) + f_j(age_i) + f_j(temp_{t,i}, age_i) + c_{i,j} + \tau_{i,k} + \epsilon_i$$
-
-$$g(DBH_i) = f(x_i,y_i) + f_j(age_i) +  f(covariate_i) + c_{i,j} + \tau_{i,k} + \epsilon_i$$
-
-$$g(DBH_i) = f(x_i,y_i) + f_j(age_i) +  c_{i,j} + \tau_{i,k} + \epsilon_i $$
+\begin{equation}
+g(DBH_i) = f(x_i,y_i) + f_j(age_i) +  c_{i,j} + \tau_{i,k} + \epsilon_i
+(\#eq:gam-age)
+\end{equation}
 
@@
 242 in @wood2017] was included to account for the spatial structure of observations, which reduced auto-correlation of model residuals considerably (see supplemental information).
-These were compared to a sub-set of models without a spatial component (cf. Equation$~$\@ref(eq:gams-nocovariate), and Figure$~$\@ref(fig:fig-model-deviance) but not further discussed there).
+These were compared to a sub-set of models without a spatial component (cf. Equation$~$\@ref(eq:gam-nocovariate), and Figure$~$\@ref(fig:fig-model-deviance) but not further discussed there).
 We also tested a suite of models without the spatial smooth for comparison
@@
 The covariates for planting bed area and soil nutrient availability were log transformed to account for their skewed distribution, improving the estimation of coefficients for their respective basis functions.
-Note, that Equation$~$\@ref(eq:gams-age) was considered as the appropriate null model for interpretations.
+Note, that Equation$~$\@ref(eq:gam-age) was considered as the appropriate null model for interpretations, and a model including all covariates and the temperature-age interaction was tested as a saturated model.
 Models were implemented with `mgcv::bam()` [@wood2017a;@li2020a] and readers are referred to the detailed implementation in the supplemental material code.
@@
 
-(ref:model-performance) Predicted vs. observed best model day2007 and LCZ6
+(ref:model-performance) Predicted $vs.$ observed data for the best model according to explained deviance and 'local' AIC comparison, based on the LandSat-derived UHI measure [@chakraborty2019] for average summer conditions (2007) and LCZ6 [open to mid-rise; @stewart2012].
+Hexagons and colors represent x-y bins (i.e., in the observed-predicted space) and their counts, and the red line is a least-squares fit.
+The model captures the mean response well, as indicated by lighters colors along dashed 1:1 line, and its close approximation by the least squares fit, as well as model metrics (see text label in plot).
+However, the generally large scatter around this mean response indicates that predictions on individual-level would be inappropriate.
 
@@
 
-(ref:heat-effect) Age group-dependent temperature effect; only within range
+Relationships derived via the GAM in Equation$~$\@ref(eq:gam-full) with LCZ6 as covariate show fairly consistent responses across age groups considering trajectories, yet they differ somewhat in shape (Figure$~$\@ref(fig:fig-heat-effect)).
+Generally, younger age groups (30-35) showed increased growth, or little to no impact (*Acer platanoides*, *Acer pseudoplatanus*), with greater excess heat.
+At older ages, this trajectory gradually shifted toward decreasing stem diameter at higher temperatures.
+Notably, *Platanus acerifolia* showed an earlier reversal of it at age group 45-50 as compared to other species. 
+This may be due to the stronger upward curvature towards the estimates' lower temperature limit (see respective panel in Figure$~$\@ref(fig:fig-heat-effect)).
+Although the uncertainty around this edge is comparatively well-constrained, the low number of observations at the end of the continuum does call for caution for further interpretation.
+The distribution of individual street trees along the excess heat continuum, evident in the density estimates in Figure$~$\@ref(fig:fig-heat-effect), show that the majority of species investigated here are subjected to intermediate to high levels of excess heat.
+Trees in younger age groups are typically further toward the upper end of this continuum.
+Several species - including *A.
+platanoides*, *A.
+pseudoplatanus*, *Acer hippocastanum*, *Quercus robur*,  and *Tilia euchlora* - have good coverage across the entire excess heat continuum (although skewed toward positive values), which increases the confidence in the observed trajectory shift from young to old.
+Contrastingly, the cultivar *Tilia intermedia* 'Palllida' is predominantly distributed along the upper end of the continuum, resulting in greater uncertainty in the estimated relationship across age groups and excess heat.
+
+The assessed tree species showed fairly consistent responses to exchess urban heating
+
+(ref:heat-effect) Age-group averaged relationships between the LandSat-derived excess heat measure and stem diameter for the ten most abundant species from the best model (i.e., Equation$~$\@ref(eq:gam-full) with LCZ6 as covariate).
+Note the differing y-axis scales between panel rows.
+The colored lines and bands are the averaged GAM fit, and gray lines are least-squares estimates fitted to these averages.
+Estimates were only generated within temperature ranges that a given species experienced.
+Density plots (scaled to match 100$~cm$ on the y-axis) show the distribution of individual trees along the urban heat continuum, aiding both with interpreting uncertainties (band width, "wiggliness") and assessing species' exposure to urban heat.
+A general shift from increased or stable growth with temperature at younger ages to a decrease for older trees is apparent.
 
@@
 **Add figure with tree locations where growth is below the average - color = UHI?**
-
-<!-- Tree locations are clustered and structured based on their category, i.e. riparian, street and park trees (Fig.$~$\@ref(fig:fig-tree-overview-map)). -->
-<!-- Planting in space and time shows species-specific patterns (by districts), often related to major events, such as the start and end of armed and/or political conflict. -->
-
-<!-- ```{r fig-bar-tree, fig.cap="A plot of random numbers",fig.width = 15, fig.height = 8, out.width="100%", include=FALSE, eval=FALSE} -->
-
-<!-- readd(plot_tree_sums_bar) -->
-
-<!-- ``` -->
-
-```{r fig-tree-overview-map, fig.cap="Individual tree locations for three categories available in Berlin Senate urban tree data set.
-Note, that for each category 7000 observations were subsampled from the available pool to facilitate visualization.",fig.width = 15, fig.height = 8, out.width="100%", include=TRUE, eval=TRUE}
-
-# loadd(full_data_set_clean)
-# loadd(plot_overview_map)
-# print(class(full_data_set_clean))
-
-knitr::include_graphics(here::here("analysis/figures/map_01_overview.png"))
-
-# readd(plot_overview_map) + 
-#      +
-#     labs(caption = NULL)
-
-# plot_overview_map
-
-```
-
-```{r fig-tree-count-map, fig.cap="Gridded counts for the 11 most frequent genera, as well as *Pinus* and remaining genera.
-Note, that counts are standardized to unity for individual genera.",fig.width = 12, fig.height = 14, out.width="100%"}
-
-knitr::include_graphics(here::here("analysis/figures/map_02_tree_sums_standardized.png"))
-
-# plot_count_map
-
-# plot(1:10)
-```
-
-The distribution of the UHI effect is highly irregular and clustered in space (Fig.$~$\@ref(fig:fig-uhi-map)), and also shows variability through time (data not shown, refer to the [urban heat island explorer](https://yceo.users.earthengine.app/view/uhimap)).
-
-(ref:chakraborty2019) [@chakraborty2019]
-
-```{r fig-uhi-map, fig.cap="Estimate of UHI intensity based on the algorithm in (ref:chakraborty2019), comparing urban with rural pixels within the greater metropolitan cluster.
-Presented values are averaged over the summer of 2007.",fig.width = 12, fig.height = 14, out.width="50%"}
-
-knitr::include_graphics(here::here("analysis/figures/map_03_uhi.png"))
-
-# plot_count_map
-
-# plot(1:10)
-```
-
-The exposure to increased heat-loading of individual genera (and consequently species) is highly uneven throughout the city (Fig.$~$\@ref(fig:fig-density)).
-Street and park trees of most genera are clustered in urban areas with intermediate to high UHI loading, while riparian trees, and some street and park trees of other genera tend to be spread more evenly across Berlin's UHI range.
-
-```{r fig-density, fig.cap="Empirical density distribution of all individuals within the presented genera along the UHI continuum.
-UHI intensities were extracted for each tree location, and the distribution hence represents the first detailed overview of the exposure of Berlin's trees to urban heat loading.
-The black line is the density across all three categories.
-Insets show corresponding tree totals.",fig.width = 12, fig.height = 12, out.width="100%"}
-knitr::include_graphics(here::here("analysis/figures/plot_02_genus_UHI_dens.png"))
-
-# plot_count_map
-
-# plot(1:10)
-```
-
-**Note, that results below are preliminary and should be considered as a template for future outputs, rather than used for inference.**
-The effect of UHI loading on absolute growth potential varies between genera and species (Fig.$~$\@ref(fig:fig-lmestat)).
-Most notably, *Quercus*, the 3rd-most frequent genera, shows decreased absolute growth with increasing UHI loading,
-while the most frequent genera, *Tilia*, features contrasting relationships between species.
-The estimated effect sizes presented here are linear.
-However, temperature may exert a non-linear control on absolute growth and, hence, applying a  method able to capture such dynamics may result in somewhat different effect sizes / behavior.
-Additionally, if temperatures increase in the future under climate warming, any non-linear effects may become more enhanced, stressing the need for a more flexible model fit and structure (i.e.
-using GAMM over linear models-).
-
-```{r fig-lmestat, fig.cap="Impact of UHI loading on tree diameter (\\(DBH\\)), accounting for age and inter-specific differences from the linear mixed model (via random slopes and intercepts).
-Line-ranges are standard errors of predicted effect sizes (i.e.
-slopes).
-Differences between street and park trees are considerable for some species, and may be due to local clustering and/or spatial under-representation across the UHI continuum.
-Further investigations need to address the degree of spatial autocorrelation and account for it where required in linear mixed models, and with smoothing interactions in a GAMM implementation.",fig.width = 20, fig.height = 12, out.width="100%"}
-knitr::include_graphics(here::here("analysis/figures/plot_03_ranef_species_dbh_uhi.png"))
-# plot_count_map
-
-# plot(1:10)
-```
 
@@
 
-# Outlook
+## Conclusions
 
-We seek to build upon and improve the current analysis by:
-
-- validating the database with independent observations
-- incorporating more pertinent covariates as dependent variables in the linear mixed model  
-- testing multiple model structures with formal model selection procedures
-- checking model residuals for spatial auto-correlation and accounting for it where necessary to ensure unbiased estimates of effect sizes
-- repeating the above with a hierarchical GAM (i.e. GAMM) to allow for:  
-  * estimating continuous prediction surfaces for UHI impacts on individual species' growth (similar to results in Figure$~$\@ref(fig:fig-lmestat)) under recent conditions
-  * estimating absolute, species-specific growth potential under increased temperatures and UHI loading under climate change, ideally based on climate simulations (otherwise step-wise increases based on RCP scenarios) for the key species.
-Note, that  complications presented by 'out-of-sample' predictions will be addressed.
-  * assess potential age-dependent UHI impacts on individual species.  
-  * repeating the above (GAMM) with total and incremental basal area as responses
-
-<!-- # Acknowledgements -->
+# Acknowledgements
 

### 2021-12-01 R/functions.R

@@ -3380,7 +3380,7 @@ plot_temp_maps <- function(landsat,
         ggplot2::geom_sf(data = berlin_poly,
                          color = "gray60",
                          fill = "gray80",
-                         size = 1,
+                         size = 0.75,
                          show.legend = FALSE) +
 
 
@@ -3392,7 +3392,7 @@ plot_temp_maps <- function(landsat,
         # ggplot2::geom_sf(data = berlin_poly,
         #                  color = "gray60",
         #                  fill = "transparent",
-        #                  size = 1,
+        #                  size = 0.75,
         #                  show.legend = FALSE) +
 
         ggplot2::coord_sf(xlim = xlim, ylim = ylim) +
@@ -3425,7 +3425,7 @@ plot_temp_maps <- function(landsat,
         ggplot2::geom_sf(data = berlin_poly,
                          color = "gray60",
                          fill = "gray80",
-                         size = 1,
+                         size = 0.75,
                          show.legend = FALSE) +
 
 
@@ -3437,7 +3437,7 @@ plot_temp_maps <- function(landsat,
         # ggplot2::geom_sf(data = berlin_poly,
         #                  color = "gray60",
         #                  fill = "transparent",
-        #                  size = 1,
+        #                  size = 0.75,
         #                  show.legend = FALSE) +
 
         ggplot2::coord_sf(xlim = xlim, ylim = ylim) +
@@ -3479,7 +3479,7 @@ plot_temp_maps <- function(landsat,
         # ggplot2::geom_sf(data = berlin_poly,
         #                  color = "gray60",
         #                  fill = "gray80",
-        #                  size = 1,
+        #                  size = 0.75,
         #                  show.legend = FALSE) +
 
 
@@ -3491,7 +3491,7 @@ plot_temp_maps <- function(landsat,
         # ggplot2::geom_sf(data = berlin_poly,
         #                  color = "gray60",
         #                  fill = "transparent",
-        #                  size = 1,
+        #                  size = 0.75,
         #                  show.legend = FALSE) +
 
         ggplot2::coord_sf(xlim = xlim, ylim = ylim) +
@@ -3536,7 +3536,7 @@ plot_temp_maps <- function(landsat,
         # ggplot2::geom_sf(data = berlin_poly,
         #                  color = "gray60",
         #                  fill = "gray80",
[…]

### 2021-12-01 R/packages.R

@@ -35,3 +35,4 @@ library(readxl)
 library(ncdf4)
 library(patchwork)
 library(rnaturalearth)
+library(grid)

### 2021-12-01 R/plan.R

@@ -1012,14 +1012,14 @@ plan <- drake::drake_plan(
         output_format = bookdown::html_document2(),
         quiet = TRUE
     ),
-    paper_word = rmarkdown::render(
-        knitr_in("./analysis/paper/paper.Rmd"),
-        output_dir = "./analysis/paper/",
-        output_file = file_out("paper_knit.docx"),
-        # output_file = "./paper_knit.html",
-        output_format = bookdown::word_document2(),
-        quiet = TRUE
-    ),
+    # paper_word = rmarkdown::render(
+    #     knitr_in("./analysis/paper/paper.Rmd"),
+    #     output_dir = "./analysis/paper/",
+    #     output_file = file_out("paper_knit.docx"),
+    #     # output_file = "./paper_knit.html",
+    #     output_format = bookdown::word_document2(),
+    #     quiet = TRUE
+    # ),
     # paper_pdf = rmarkdown::render(
     #     knitr_in("./analysis/paper/paper.Rmd"),
     #     output_dir = "./analysis/paper/",

### 2021-12-01 analysis/paper/paper.Rmd
cont results

@@
 <!-- Species-specifc growth responses in relation across Berlin were assessed in relat -->
-We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of excess heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; LandSat-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section$~$\@ref(sec:stat-gam) for details).
+We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of (excess) heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; LandSat-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section$~$\@ref(sec:stat-gam) for details).
 Covariates were extracted at 150 and 300$~m$ to infer the impact of reference scale of the urban fabric on tree growth. 
@@
 Following the general approach described above, four ancillary covariates next to a temperature measure were employed in models; these were chosen due to their availability at high spatial resolution and coverage, and/or because their influence on growth was previously identified in literature or their likely impact could be deduced using ecophysiological principles.
-We included planting bed area and the sum of exchangeable basic cation as a proxy for soil nutrient availability (point extractions), as well as the proportional coverage of local climate zone 6 (LCZ6; open mid-rise, see @demuzere2019 and @stewart2012 for details) and adjacent building height (spatial averages).
+We included planting bed area and the sum of exchangeable basic cation as a proxy for soil nutrient availability (point extractions), as well as the proportional coverage of local climate zone 6 (LCZ6; open mid-rise urban land cover, see @demuzere2019 and @stewart2012 for details) and adjacent building height (spatial averages).
 The latter was chosen as an increase reflects a transition away from densely urbanized areas and had the highest coverage for the processed tree inventory.
@@
 
-### Stem diameter model development and selection
+### Stem diameter model development and selection {#sec:methods-gam-heat}
 
@@
 
-* Null-model high, but for predictive purposes more is better
-* Temperature-based models on average better than without, but planting bed area and lcz6 individually outperform - indication that these variables are indeed important; choice of best performing model, which includes lcz6, coroborated.
-*Landsat temp best on average - easily accessible/maintainable as compared to model outputs
-* $lcz6_{150}$ vs. $lcz6_{300}$ - scale of impact (further investigation, note in discussion) 
-* building height, as a proxy for urban fabric and incoming radiation had minimal effect (regardless of 150/300)
+* null model high, but for predictive purposes more is better
+Models including (excess) heat measures were on average better than those without.
+
+However, this was not the case for planting bed area and $LCZ6_{\over{150~m}}$, which outperformed several temperature-based models.
+This indicated the importance of both covariates, with $LCZ6_{\over{150~m}}$ also represented in the best-performing model, corroborating its impact on stem diameter development.
+As $LCZ6_{\over{300~m}}$ only marginally increased performance compared to the null model (age only), the reference scale for the impact of the urban fabric (i.e., LCZ6) has a limit between $150$ to $300~m$.
+Nutrient availablity, similar to $LCZ6_{\over{300~m}}$, also only improved model performance marginally.
+Building height, regardless of its reference scale (i.e., $150$ $vs$ $300~m$), as another proxy for urban fabric, showed minimal to no improvement compared to the age-only null model.
 
@@
 150/300$~m$, but also note x-axis labels), and size indicates the model's observations.
-Generally, including temperature (interactions) improve model fits above the null-model (age only), and excluding planting bed area and LCZ6 cover, also above other non-temperature models.
+Generally, including temperature (interactions) improve model fits above the null model (age only), and excluding planting bed area and LCZ6 cover, also above other non-temperature models.
 The LandSat-derived UHI measures provide the best fit on average, with the model including a temperature-age interaction and LCZ6 cover scoring highest.
@@
 
-(ref:model-performance) Predicted $vs.$ observed data for the best model according to explained deviance and 'local' AIC comparison, based on the LandSat-derived UHI measure [@chakraborty2019] for average summer conditions (2007) and LCZ6 [open to mid-rise; @stewart2012].
+(ref:model-performance) Predicted $vs.$ observed data for the best model according to explained deviance and 'local' AIC comparison, based on the LandSat-derived UHI measure [@chakraborty2019] for average summer conditions (2007) and LCZ6 [open to mid-rise,  @stewart2012].
 Hexagons and colors represent x-y bins (i.e., in the observed-predicted space) and their counts, and the red line is a least-squares fit.
@@
 Relationships derived via the GAM in Equation$~$\@ref(eq:gam-full) with LCZ6 as covariate show fairly consistent responses across age groups considering trajectories, yet they differ somewhat in shape (Figure$~$\@ref(fig:fig-heat-effect)).
-Generally, younger age groups (30-35) showed increased growth, or little to no impact (*Acer platanoides*, *Acer pseudoplatanus*), with greater excess heat.
+Generally, younger age groups (30-35) showed increased growth, or little to no impact (*Acer platanoides*, *Acer pseudoplatanus*), with stronger UHI magnitude.
 At older ages, this trajectory gradually shifted toward decreasing stem diameter at higher temperatures.
@@
 Although the uncertainty around this edge is comparatively well-constrained, the low number of observations at the end of the continuum does call for caution for further interpretation.
-The distribution of individual street trees along the excess heat continuum, evident in the density estimates in Figure$~$\@ref(fig:fig-heat-effect), show that the majority of species investigated here are subjected to intermediate to high levels of excess heat.
+The distribution of individual street trees along the UHI continuum, evident in the density estimates in Figure$~$\@ref(fig:fig-heat-effect), show that the majority of species investigated here are subjected to intermediate to high UHI levels.
 Trees in younger age groups are typically further toward the upper end of this continuum.
@@
 platanoides*, *A.
-pseudoplatanus*, *Acer hippocastanum*, *Quercus robur*,  and *Tilia euchlora* - have good coverage across the entire excess heat continuum (although skewed toward positive values), which increases the confidence in the observed trajectory shift from young to old.
-Contrastingly, the cultivar *Tilia intermedia* 'Palllida' is predominantly distributed along the upper end of the continuum, resulting in greater uncertainty in the estimated relationship across age groups and excess heat.
+pseudoplatanus*, *Acer hippocastanum*, *Quercus robur*,  and *Tilia euchlora* - have good coverage across the entire UHI continuum (although skewed toward positive values), which increases the confidence in the observed trajectory shift from young to old.
+Contrastingly, the cultivar *Tilia intermedia* 'Palllida' is predominantly distributed along the upper end of the continuum, resulting in greater uncertainty in the estimated relationship across age groups and UHI magnitudes.
 
-The assessed tree species showed fairly consistent responses to exchess urban heating
-
-(ref:heat-effect) Age-group averaged relationships between the LandSat-derived excess heat measure and stem diameter for the ten most abundant species from the best model (i.e., Equation$~$\@ref(eq:gam-full) with LCZ6 as covariate).
-Note the differing y-axis scales between panel rows.
+(ref:heat-effect) Age-group averaged relationships between the LandSat-derived UHI and stem diameter for the ten most abundant species from the best model (i.e., Equation$~$\@ref(eq:gam-full) with LCZ6 as covariate).
+Note that estimates were derived with $LCZ6_{\over{150~m}} = 0.5$, and that y-axis scales differ between panel rows.
 The colored lines and bands are the averaged GAM fit, and gray lines are least-squares estimates fitted to these averages.
@@
 
-(ref:lcz6-effect) Global effect of LCZ6 on growth
+The chosen model also highlighted that, on average, trees growing in less urbanized environments (i.e., more open to mid-rise urban land cover; $LCZ6_{\over{150~m}}$) achieve somewhat greater diameters.
+Note, that planting bed area, similar to $LCZ6_{\over{150~m}}$, was identified as a covariate increasing explained deviance, both in models with and without a temperature measure, and represents an opportunity for further investigation, especially following additional data collection.
+
+(ref:lcz6-effect) Age-group averaged relationships between the LCZ6 and stem diameter for two $Tilia$ genera from the best model (i.e., Equation$~$\@ref(eq:gam-full) with LandSat-derived UHI as excess heat measure).
+LCZ6 cover was derived as a weighted average within a radius of 150$~m$ around each tree.
+Note the identical shape, but difference in intercept as a result of the model specification with a global smooth the non-temperature covariate for all species (see Section$~$\@ref(sec:methods-gam-heat)); the UHI magnitude was fixed at $3^\circ C$ where most trees are present across age groups.
+The model identified a steady increase of average stem diameter with open to mid-rise urban land cover.
 
@@
 		- Supplement species matrix etc. with these analyses  
+		- *Landsat temp best on average - easily accessible/maintainable as compared to model outputs
+* $lcz6_{150}$ vs. $lcz6_{300}$ - scale of impact (further investigation, note in discussion) 
+        - planting bed area .. more data!
   + Open data important --> drives innovation

### 2021-12-01 R/functions.R

@@ -5229,7 +5229,7 @@ plot_obs_predicted_model <- function(path_model,
         geom_hex(bins = 75, color = "transparent") +
         geom_smooth(method = "lm", formula = "y ~ x", color = "red") +
         geom_abline(slope = 1, intercept = 0, linetype = 2, color = "gray40") +
-        scale_fill_viridis_c() +
+        scale_fill_viridis_c(option = "plasma") +
         theme_minimal(base_size = 18) +
         labs(x = "Predicted", y = "Observed", fill = "N") +
         annotate("text",

### 2021-12-06 analysis/paper/paper.Rmd
updated k_age = 3
cont writing

@@
 Temperature and UHI data were summarized temporally either by the provider or manually to provide a characteristic representation of heat loading during summer at different times (morning, afternoon/day, night), from which tree averages (radius of 150$~m$) were calculated (Figure$~$\@ref(fig:fig-temp-overview)).
-Two data sets of urban air and one surface temperature were tested as explanatory variables in GAM models.
-The air temperatures from the Berlin environmental atlas (EnvAt) are model outputs that are representations of typical summer conditions at 0400, 1400 and 2200 hours; these data are provided at city block basis (spatial polygons), from which weighted averages were extracted.
+Two urban air temperature and one surface UHI data set were tested as explanatory variables in GAM models.
+The air temperatures from the Berlin environmental atlas (EnvAt) are processed model outputs [FITNAH-3D; cf.
+@gross1994] that are representations of typical summer conditions at 0400, 1400 and 2200 hours; these data are provided at city block basis (spatial polygons), from which weighted averages were extracted after rasterizing ($5~m$ resolution).
 UrbClim air temperatures are hourly model outputs [100$~m$ resolution, @deridder2015] based on ERA5 re-analyses data (ECMWF) for which observations from the hottest month  available (June, 2011) were averaged to hours equivalent to Berlin Environmental Atlas (referred to as Berlin EnvAt) data by using a window of $\pm~1~$hour (i.e., 0300 to 0500, etc.).
@@
 Following the general approach described above, four ancillary covariates next to a temperature measure were employed in models; these were chosen due to their availability at high spatial resolution and coverage, and/or because their influence on growth was previously identified in literature or their likely impact could be deduced using ecophysiological principles.
-We included planting bed area and the sum of exchangeable basic cation as a proxy for soil nutrient availability (point extractions), as well as the proportional coverage of local climate zone 6 (LCZ6; open mid-rise urban land cover, see @demuzere2019 and @stewart2012 for details) and adjacent building height (spatial averages).
-The latter was chosen as an increase reflects a transition away from densely urbanized areas and had the highest coverage for the processed tree inventory.
+We included planting bed area and the sum of exchangeable basic cation as a proxy for soil nutrient availability (point extractions), as well as the proportional coverage of local climate zone 6 (LCZ6; open mid-rise urban land cover, see @demuzere2019 and @stewart2012 for details) and adjacent building height (rasterized to $5~m$ resolution).
+The latter was chosen as an increase reflects a transition away from densely urbanized areas and had the highest coverage (i.e., within [0-1]) for the processed tree inventory.
 
@@
 
+## Implications
+
+### Green infrastructure management and planning
+
+The availability and quality of green infrastructure data, such as street tree inventories, and environmental covariates, determines the confidence in any inference of infrastructure performance in relation to its given context (e.g., temperature, planting bed area).
+This is fundamental to inform management and planning to ensure that current and future targets are met. 
+Indeed, the execution of this study was possible due to the readily accessible and rich data sources provided by Berlin’s senate.
+However, few additions or alterations to data collection and provision schemes would have greatly  enhanced the utility and applicability of the data sets applied here.
+Note that we acknowledge the financial and logistic constraints that authorities are subject to, yet the suggestions below are likely valuable for this and other urban green infrastucture initiatives.
+Urban tree databases typically represent a single time point (e.g., see https://opentrees.org), despite repeated inventories at regular and/or irregular intervals within cities or their districts.
+Storing and providing multi-temporal inventories, rather than over-writing existing records, would give planners and researchers a wealth of additional information on individual tree performance, which – even at low temporal coverage – would allow verifying space-for-time substitutions, or make them superfluous.
+We also urge authorities within and across cities to enhance and ideally standardize their (meta) data collection and provisioning to facilitate data quality control, increase confidence, and enable synthesis studies across districts and regions [@ossola2020].
+Key information includes (1) location, absolute age and planting date, (2) management regimes (e.g., planting regime, crown pruning, irrigation), (3) data collection protocol and body (e.g., actual measurement $vs.$ estimate or interpolation), as well as (4) tree provenance to account for potential differences in ecotype and/or acclimation [@watkins2020].
+However, there are financial and logistic trade-offs for data collection of the above, with contrasting needs across stakeholder groups [e.g., academics and authorities, @ostberg2013] where data that supports meeting long-term objectives should be prioritized.
+Lastly, less abundant species, which represent canditates for future tree selection, could be assessed with higher-quality standards regarding points measured above, to build a more reliable data source even at potentially lower coverage demographically and spatially.
+
+### Tree selection
+
+Tree species selection for urban forests depends on several often competing factors -  such as aesthetic appeal and ecosystem service provision [e.g., @roman2020] - and has been focused on identifying trees suitable for current and future conditions for the last decades with increasingly strong evidence basis [e.g., @hirons2019].
+Here, biogeographical approaches aim to match natural (i.e., non-urban) species distributions within current bioclimatic envelopes with future conditions at a planting site at larger scales [e.g., @broadmeadow2005].
+@watkins2020 used such "climate matching" to identify the space along environmental dimensions (heat, precipitation) which urban regions inhabit, falling within or extending the natural envelope. 
+Our work extends this approach by identifying the heat space (i.e., UHI range) occupied by several key species.
+This not only allows identifying species-specific and intra-specific (age-dependent) sensitivities to urban conditions, but can also inform how age cohorts may respond as temperatures at a given location shift along the present continuum in the future, while accounting for the impacts additional urban conditions (e.g., management, disturbance, urban fabric).
+The insights derived from this study, and potentially others following this approach in the future, can thus add another dimension to tree selection approaches implemented down to individual trees and locations.
+It thus complements quantitative and qualititaive tools, such as the climate matching tool, ecological site classification or species-climate matrix [see @hirons2019; @roloff2009; @brune 2019; and references therein].
+
 ## Conclusions

### 2021-12-06 R/functions.R

@@ -2212,10 +2212,10 @@ make_model_grid <- function(){
     model_params <- new.env()
 
     model_params$k_uni <- 7
-    model_params$k_age <- 5
+    model_params$k_age <- 3
     # model_params$k_te <- c(7, 20)
     # model_params$k_te <- c(5, 15)
-    model_params$k_te <- c(5, 7)
+    model_params$k_te <- c(3, 7)
     model_params$k_spatial <- 400
     model_params$k_spatial_soilnutmodel <- 250
     model_params$k_soilnut <- 25
@@ -2785,7 +2785,7 @@ pred_dbh_temp_single_var <- function(path_model,
 
 
 
-    `%nin%` <- Negate(`%in%`)
+    # `%nin%` <- Negate(`%in%`)
 
 
 

### 2021-12-07 analysis/paper/paper.Rmd
cont discussion

@@
 
+## Heat, environmental and urban controls on tree growth
+
+Intra specific differences
+
+Quercus robur, Tilia intermedia and Aesculum hippocastanum performed best, matching with climate-species matrix in roloff and brune
+
+## Utility, transferrability and limitations of approach
+
+The major strength of this work is the estimation of growth sensitivity of urban trees to heat and other urban conditions at city-scale with little data and processing power requirements. 
+In particular,  differences within and between species across several dimensions (depending on model specification) are readily explored where only tree locations, age, diameter (or any other target) and selected variables are needed to implement this approach.
+Indeed, due to the high temporal and spatial variability of controls on growth, and thus variability in responses, such assessments should be developed specifically for a given setting, as well-understood tree characteristics [e.g., @roloff2009], could be strongly mediated at a given location or time.
+For instance, if drought hardiness is related to extensive root networks, the limited soil volumes available to street trees could render a species vulnerable to water stress.
+In addition, studies on climatic suitability and physiological responses are difficult to compare due to local acclimation of species to controls on growth and performance, as well as different methodological protocols [e.g., see @brune2016].
+Thus, the approach followed here, with its comparatively simple data requirements, can be implemented in a variety of settings to gain initial and/or guiding insights on tree performance in a specific urban region, and also serve as a benchmark to compare other empirical estimates or mechanistic modelswith.
+It may even allow estimating tree performance under future conditions to some extent (e.g., considering increase in temperature alone for this study), but is of course contingent on data coverage (e.g., demographics, environmental gradients).
+Furthermore, the implementation of GAM models here with `mgcv::bam()` is computationally efficient [@wood2017a], making model testing, selection and final application a rapid process even for abundant observations (here $n >150000$ and processing times of $t < 60~s$ on a desktop computer for the selected model).
+
+Two noteworthy caveats of the approach followed here lie in its empirical, data-driven nature, and its dependence on data quality.
+The presented results here are correlative. 
+While their interpretation follows first principles and may also allow anticipating the performance of a given tree under a range of increased temperatures, insights derived from mechanistic understanding are indispensable.
+This is because a change in one environmental condition, e.g., temperature, may also entail alterations to other drivers, and ultimately, affect physiological processes in ways not yet captured in existing data.
+The data provided in Berlin's tree inventory required extensive quality controlling, and it is likely that inaccurate (diameter, age) or interpolated (diameter) observations were including in the analyses.
+However, the minimum requirement of 10000 observations per species, and the focus on average diameter responses across age groups decreased potential impacts such observations may have had to a degree that was considered satisfactory, while uncertainty is communicated effectively (i.e., larger confidence bands for *T.
+intermedia 'Pallida'*).
+Further, the utility and transferability of the approach  *per se* were considered of value to the community and practitioners.
+
 ## Implications
 
-### Green infrastructure management and planning
+### Tree selection
+
+Tree species selection for urban forests depends on several often competing factors -  such as aesthetic appeal and ecosystem service provision [e.g., @roman2020] - and has been focused on identifying trees suitable for current and future conditions for the last decades with increasingly strong evidence basis [e.g., @hirons2019].
+Here, biogeographical approaches aim to match natural (i.e., non-urban) species distributions within current bioclimatic envelopes with future conditions at a planting site [e.g., @broadmeadow2005].
+@watkins2020 used such "climate matching" to identify the space along environmental dimensions (heat, precipitation) which urban regions inhabit, falling within or extending the natural envelope. 
+Our work extends this approach by identifying the heat space (i.e., UHI range) occupied by several key species within a city.
+This not only allows identifying species-specific and intra-specific (age-dependent) sensitivities to urban conditions, but can also inform how age cohorts may respond as temperatures at a given location shift along the present continuum in the future, while accounting for the impacts of additional urban conditions (e.g., management, disturbance, urban fabric).
+The insights derived from this study, and potentially others following this approach in the future, can thus add another dimension to tree selection approaches implemented down to individual trees and locations.
+It thus complements quantitative and qualitative tools, such as the climate matching tool, ecological site classification or species-climate matrix [see @hirons2019; @roloff2009; @brune 2019; and references therein].
+
+### Green infrastructure management, planning and open data
 
@@
 However, few additions or alterations to data collection and provision schemes would have greatly  enhanced the utility and applicability of the data sets applied here. 
-Note that we acknowledge the financial and logistic constraints that authorities are subject to, yet the suggestions below are likely valuable for this and other urban green infrastucture initiatives.
+We acknowledge the financial and logistic constraints that authorities are subject to, yet the suggestions below are likely valuable for this and other urban green infrastructure initiatives, as several of them are easily implemented.
+
 Urban tree databases typically represent a single time point (e.g., see https://opentrees.org), despite repeated inventories at regular and/or irregular intervals within cities or their districts.
@@
 However, there are financial and logistic trade-offs for data collection of the above, with contrasting needs across stakeholder groups [e.g., academics and authorities, @ostberg2013] where data that supports meeting long-term objectives should be prioritized.
-Lastly, less abundant species, which represent canditates for future tree selection, could be assessed with higher-quality standards regarding points measured above, to build a more reliable data source even at potentially lower coverage demographically and spatially.
+Less abundant species which are candidates for future plantings could be assessed with higher-quality standards regarding points measured above, to build a more reliable data source even at potentially lower coverage demographically and spatially.
+Managers and planners also should be aware of potential scale-limit of the urban fabric (e.g., $LCZ6_{\over{150~m}} vs.
+$LCZ6_{\over{300~m}}) on tree performance when altering surroundings or planting in a given area; efforts should be directed toward identifying the most pertinent environmental covariates and how their influence scales in space.
 
-### Tree selection
-
-Tree species selection for urban forests depends on several often competing factors -  such as aesthetic appeal and ecosystem service provision [e.g., @roman2020] - and has been focused on identifying trees suitable for current and future conditions for the last decades with increasingly strong evidence basis [e.g., @hirons2019].
-Here, biogeographical approaches aim to match natural (i.e., non-urban) species distributions within current bioclimatic envelopes with future conditions at a planting site at larger scales [e.g., @broadmeadow2005].
-@watkins2020 used such "climate matching" to identify the space along environmental dimensions (heat, precipitation) which urban regions inhabit, falling within or extending the natural envelope. 
-Our work extends this approach by identifying the heat space (i.e., UHI range) occupied by several key species.
-This not only allows identifying species-specific and intra-specific (age-dependent) sensitivities to urban conditions, but can also inform how age cohorts may respond as temperatures at a given location shift along the present continuum in the future, while accounting for the impacts additional urban conditions (e.g., management, disturbance, urban fabric).
-The insights derived from this study, and potentially others following this approach in the future, can thus add another dimension to tree selection approaches implemented down to individual trees and locations.
-It thus complements quantitative and qualititaive tools, such as the climate matching tool, ecological site classification or species-climate matrix [see @hirons2019; @roloff2009; @brune 2019; and references therein].
+Finally, open data was key for this study.
+We expressedly urge authorities to implement FAIR (findable, accessible, inter-operable and reusable) data principles to enable work similar to ours, as well as the creation of larger-scale databases and/or execution of synoptic studies [e.g., @ossola2020].
 
 ## Conclusions
+
+- novel application of open data to infer sensitivity to urban heat and other pertinent environmental drivers
+- identified age-shift and most tolerant species
+- provide policy-relevant suggestions for authorities and practicioners
+- open several avenues for further inquiry:
+  + multi-temporal
+  + increase covariate coverage to see if predictable performance alteration (planting bed!)
+  + less abundant species --> potential to identify suitable trees in a specific urban context (e.g., bespoke for Berlin)
 

### 2021-12-08 R/functions.R
added temp sens figure

@@ -5260,6 +5260,10 @@ plot_obs_predicted_model <- function(path_model,
 #' @param age_expression expression, used to define age groups
 #' @param prediction_range character, within or full?
 #' @param base_size numeric
+#' @param file
+#' @param height
+#' @param width
+#' @param dpi
 #'
 #' @return list with plot and lms of slopes for temp sensitivities
 #' @export
@@ -5272,7 +5276,11 @@ plot_dbh_temp_comparison <- function(
     species_filter,
     age_expression,
     base_size = 18,
-    prediction_range = "full"
+    prediction_range = "full",
+    file,
+    height,
+    width,
+    dpi
 ){
 
     if(is.null(species_filter)){
@@ -5393,8 +5401,7 @@ plot_dbh_temp_comparison <- function(
         theme_minimal(base_size = base_size) +
         theme(legend.position = 'top',
               legend.direction = "horizontal",
-              panel.spacing = unit(2, "lines"),
-              strip.text = element_text(size = 12)) +
+              panel.spacing = unit(1.5, "lines")) +
 
         # scale_color_brewer(palette = 2, type = "qual") +
         # scale_fill_brewer(palette = 2, type = "qual") +
@@ -5404,8 +5411,8 @@ plot_dbh_temp_comparison <- function(
             aesthetics = c("fill", "color")) +
 
 
-        labs(color = "Species",
-             fill = "Species",
+        labs(color = NULL,
+             fill = NULL,
              x = expression(UHI~Magnitude~(degree*C)),
              y = expression(bar(DBH)[centered]~(cm)))
 
@@ -5441,7 +5448,21 @@ plot_dbh_temp_comparison <- function(
                                               ticks.colour = "gray30"))
 
 
-    # plot_dbh_sens_lms$plot + plot_dbh_sens_lms$est_plot + plot_layout(guides = "collect", widths = c(0.7, 0.3)) & theme(legend.position='bottom', legend.box = 'vertical')
+    p_plot <- gplot +
+        estimate_plot +
+        plot_annotation(tag_levels = "A") +
+        plot_layout(guides = "collect",
+                    widths = c(0.7, 0.3)) &
+        theme(legend.position='bottom',
+              legend.box = 'vertical',
+              legend.text = element_text(face = "italic"))
+
+    ggsave(filename = file,
+           plot = p_plot,
+           width = width,
+           height = height,
+           dpi = dpi,
+           units = "in")
 
[…]

### 2021-12-08 R/plan.R

@@ -877,7 +877,11 @@ plan <- drake::drake_plan(
                                                 # species_filter = c("Tilia cordata","Platanus acerifolia"),
                                                 age_expression = age_expr,
                                                 prediction_range = "within",
-                                                base_size = 18),
+                                                base_size = 18,
+                                                file = "./analysis/figures/fig_model_day2007_lcz6_growthsens.png",
+                                                height = 6,
+                                                width = 12,
+                                                dpi = 300),
 
 
    ### stat: moran comparison ------------------

### 2022-01-10 analysis/paper/paper.Rmd
added temp sens figure
update writing

@@
 abstract: |
-  **This document outlines the rationale for an analysis of tree growth (potential) and its relationship with the Urban Heat Island (UHI) effect in Berlin using an extensive, publicly available data set.
-It introduces preliminary results and provides an outlook for up-coming and potential work.**
+  **Populate..We analzed tree growth (potential) and its relationship with the Urban Heat Island (UHI) effect, as well as other characteristics of the urban fabric, for the ten most abundant street tree species/cultivars in Berlin using open data.
+We identify the most heat-sensitive species, give recommendations for future tree selections and provide guidelines to for local authorities for the prioitiziation of data acquisition (i.e., inventories) as well as data provision.**
 keywords: |
@@
 
+(ref:temp-sens) Temperature sensitivity ....
+
+```{r fig-temp-sens, fig.cap="(ref:temp-sens)", out.width="100%", include=TRUE, eval=TRUE}
+
+knitr::include_graphics(here::here("analysis/figures/fig_model_day2007_lcz6_growthsens.png"))
+
+```
+
 The chosen model also highlighted that, on average, trees growing in less urbanized environments (i.e., more open to mid-rise urban land cover; $LCZ6_{\over{150~m}}$) achieve somewhat greater diameters.
@@
 ```
-
-## Exposed trees
-
-**Add figure with tree locations where growth is below the average - color = UHI?**
 

### 2022-01-10 analysis/paper/paper.Rmd
cont disc

@@
 
-Intra specific differences
+- Young to old shift
+increase in growth rates, temperature
+
+### Demographic patterns
+
+We showed that initial growth rates ($\lessapprox 30~a$) are greater in  recently established trees, as compared to older ones. 
+This pattern was found in annually-resolved incremental growth observations, and corresponds with prior work on urban-rural [e.g., @pretzsch2017] and intra-urban comparisons [@dahlhausen2019], where it was attributed to increased temperatures - namely, the UHI effect.
+Furthermore, the space-for-time substitution highlighted that recently established, i.e.
+younger, trees have greater (positive) sensitivity to heat than intermediate age groups, while the oldest trees showed lower average absolute growth (diameters) with increasing heat.
+We are not aware of other studies reporting such an age-dependent shift in sensitivity, and by contrast, for example, @pretzsch2017 showed ever increasing basal area increments within the age range investigated here for recent as compared to earlier growth.
+However, the consistency of this pattern across species allows for a satisfactory level of confidence in it.
+Indeed, @dahlhausen2019 report an age-dependent shift in growth rates depending on urban development intensity, ..where urban fabric may be indicative can be assumed indicative of temperatures..)
+Explanation found in individual optimum responses, and potentially their interaction, in relation to different drivers.
+For instance, higher temperatures (phys optimum), CO2/WUE.. but larger trees need to support greater leaf area at same soil volume (considering restriction).
+Thus it can be assumed that as trees grow and their requirements for resources (water, nutrients) increase, higher temperatures may be increasingly limiting. 
+Counter-intuitive perhaps? Young trees struggle to establish... management effort (irrigation) may also be crucial, but no data available.
+
+- live fast die young
+- water use efficiency
+
+- Intra specific differences
 
 Quercus robur, Tilia intermedia and Aesculum hippocastanum performed best, matching with climate-species matrix in roloff and brune
+
+- Urban fabric: LCZ density and sensitive trees in hot spots
 
@@
 The insights derived from this study, and potentially others following this approach in the future, can thus add another dimension to tree selection approaches implemented down to individual trees and locations.
-It thus complements quantitative and qualitative tools, such as the climate matching tool, ecological site classification or species-climate matrix [see @hirons2019; @roloff2009; @brune 2019; and references therein].
+It thus complements quantitative and qualitative tools, such as the climate matching tool, ecological site classification or species-climate matrix [see @hirons2019; @roloff2009; @brune2016; and references therein].
 
@@
 
-The availability and quality of green infrastructure data, such as street tree inventories, and environmental covariates, determines the confidence in any inference of infrastructure performance in relation to its given context (e.g., temperature, planting bed area).
+The availability and quality of green infrastructure data, such as street tree inventories, and environmental covariates, determine the confidence in any inference of infrastructure performance in relation to its context (e.g., temperature, planting bed area).
 This is fundamental to inform management and planning to ensure that current and future targets are met. 
@@
 Less abundant species which are candidates for future plantings could be assessed with higher-quality standards regarding points measured above, to build a more reliable data source even at potentially lower coverage demographically and spatially.
-Managers and planners also should be aware of potential scale-limit of the urban fabric (e.g., $LCZ6_{\over{150~m}} vs.
-$LCZ6_{\over{300~m}}) on tree performance when altering surroundings or planting in a given area; efforts should be directed toward identifying the most pertinent environmental covariates and how their influence scales in space.
+Managers and planners also should be aware of potential scale-limit of the urban fabric (e.g., $LCZ6_{\over{150~m}}~vs.~LCZ6_{\over{300~m}}$) on tree performance when altering surroundings or planting in a given area; efforts should be directed toward identifying the most pertinent environmental covariates and how their influence scales in space.
 

### 2022-01-11 analysis/paper/paper.Rmd
cont discussion

@@
 We showed that initial growth rates ($\lessapprox 30~a$) are greater in  recently established trees, as compared to older ones. 
-This pattern was found in annually-resolved incremental growth observations, and corresponds with prior work on urban-rural [e.g., @pretzsch2017] and intra-urban comparisons [@dahlhausen2019], where it was attributed to increased temperatures - namely, the UHI effect.
+This pattern was found in annually-resolved incremental growth observations, and corresponds with prior work on urban-rural [e.g., @pretzsch2017;@zhao2016] and intra-urban comparisons [e.g., @dahlhausen2018], where it was attributed to increased temperatures - namely, the UHI effect.
 Furthermore, the space-for-time substitution highlighted that recently established, i.e.
 younger, trees have greater (positive) sensitivity to heat than intermediate age groups, while the oldest trees showed lower average absolute growth (diameters) with increasing heat.
-We are not aware of other studies reporting such an age-dependent shift in sensitivity, and by contrast, for example, @pretzsch2017 showed ever increasing basal area increments within the age range investigated here for recent as compared to earlier growth.
-However, the consistency of this pattern across species allows for a satisfactory level of confidence in it.
-Indeed, @dahlhausen2019 report an age-dependent shift in growth rates depending on urban development intensity, ..where urban fabric may be indicative can be assumed indicative of temperatures..)
-Explanation found in individual optimum responses, and potentially their interaction, in relation to different drivers.
-For instance, higher temperatures (phys optimum), CO2/WUE.. but larger trees need to support greater leaf area at same soil volume (considering restriction).
-Thus it can be assumed that as trees grow and their requirements for resources (water, nutrients) increase, higher temperatures may be increasingly limiting. 
-Counter-intuitive perhaps? Young trees struggle to establish... management effort (irrigation) may also be crucial, but no data available.
+We are not aware of other studies reporting such an age-dependent shift in sensitivity.
+<!-- , and by contrast, for example, @pretzsch2017 showed ever increasing basal area increments within the age range investigated here for recent as compared to earlier growth. -->
+However, the consistency of this pattern across species provides a satisfactory level of confidence in it, and indeed, @dahlhausen2018 also report a time - and by implication an age-dependent - shift of temperature sensitivity on growth depending on urban development intensity, but do not further distinguish between age groups.
+The age-dependent shift may be explained by a combination of physiological and environmental changes over time, and/or may be related to management practices, but we acknowledge that additional analyses is required to disentangle these processes.
+For instance, higher temperatures may have benefited physiological processes close to their respective optima in both young and old individuals. 
+But, as older trees must support greater leaf area through rooting in similarly small (restricted) soil volumes, water availability and nutrients may become increasingly limiting.
+Higher rates of transpiration and photosynthesis, driven through temperature increases, could thus turn detrimental as the ratios between soil volume, root and leaf area change throughout a trees life span and in time.
+These e
+In addition to this, younger trees are typically more rigorously managed and irrigated, which may (further) alter age-dependent temperature sensitivity.
+We lack data on demographic differences in management to confirm this, and acknowledge that younger trees here also surpassed their establishment phase, which entails high mortality for recently planted individuals [@harris2017]  during which the above would not apply
 

### 2022-01-11 analysis/paper/paper.Rmd
cont disc

@@
 author:
-  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^a$, ...
-  - $^a$ *Geoforschungszentrum Potsdam, Section 4.3, Germany*
+  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^{a,b}$, ...
+  - $^a$ *Geoforschungszentrum Potsdam, Section 4.3, Potsdam, Germany*
+  - $^b$ *Deutsches Archäologisches Institut, Dendroclimatology, Berlin, Germany*
   - "* Corresponding author: hurley@gfz-potsdam.de"
@@
 
-- Young to old shift
-increase in growth rates, temperature
-
 ### Demographic patterns
@@
 Higher rates of transpiration and photosynthesis, driven through temperature increases, could thus turn detrimental as the ratios between soil volume, root and leaf area change throughout a trees life span and in time.
-These e
-In addition to this, younger trees are typically more rigorously managed and irrigated, which may (further) alter age-dependent temperature sensitivity.
-We lack data on demographic differences in management to confirm this, and acknowledge that younger trees here also surpassed their establishment phase, which entails high mortality for recently planted individuals [@harris2017]  during which the above would not apply
+In addition to this, younger trees are typically more rigorously managed and irrigated [e.g., [@koeser2014]], which may (further) alter age-dependent temperature sensitivity.
+We lack data on demographic differences in management to confirm this, and acknowledge that younger trees here also surpassed their establishment phase, which entails high mortality during the first five years after planting [@harris2017;@hilbert2019] during which the above would not apply
 
-- live fast die young
-- water use efficiency
+### Intra-specific differences and the urban fabric
 
-- Intra specific differences
-
-Quercus robur, Tilia intermedia and Aesculum hippocastanum performed best, matching with climate-species matrix in roloff and brune
-
-- Urban fabric: LCZ density and sensitive trees in hot spots
+- Quercus robur, Tilia intermedia and Aesculum hippocastanum performed best, matching with climate-species matrix in roloff and brune		
+- least: Platanus acerifolia, Acer pseudoplatanus.. data issues or location specific factors where trees were planted. Avenue for further work to confirm and identify drivers.
+- Urban fabric: LCZ density 
+- and sensitive trees in hot spots
 
@@
 We acknowledge the financial and logistic constraints that authorities are subject to, yet the suggestions below are likely valuable for this and other urban green infrastructure initiatives, as several of them are easily implemented.
+
+**Increased growth rates.. live fast, die young - high turnover! requires focused planting and/or maintenance to counter [@smith2019].**
 

### 2022-01-11 analysis/paper/paper.Rmd
refs

@@
 Higher rates of transpiration and photosynthesis, driven through temperature increases, could thus turn detrimental as the ratios between soil volume, root and leaf area change throughout a trees life span and in time.
-In addition to this, younger trees are typically more rigorously managed and irrigated [e.g., [@koeser2014]], which may (further) alter age-dependent temperature sensitivity.
+In addition to this, younger trees are typically more rigorously managed and irrigated [e.g., @koeser2014], which may (further) alter age-dependent temperature sensitivity.
 We lack data on demographic differences in management to confirm this, and acknowledge that younger trees here also surpassed their establishment phase, which entails high mortality during the first five years after planting [@harris2017;@hilbert2019] during which the above would not apply

### 2022-01-12 analysis/paper/paper.Rmd
disc

@@
 
-- Quercus robur, Tilia intermedia and Aesculum hippocastanum performed best, matching with climate-species matrix in roloff and brune		
-- least: Platanus acerifolia, Acer pseudoplatanus.. data issues or location specific factors where trees were planted. Avenue for further work to confirm and identify drivers.
+Generally, the ten investigated species and cultivars showed consistent responses in sign, but differed substantially in magnitude, modulated by age.
+*Aesculum hippocastanum* , *Tilia intermedia*, *T. intermedia 'pallida'* and showed greatest absolute growth at high temperatures at younger (30-35) as well as older ages (60-65).
+The drought tolerance and performance under high temperatures is well documented for the latter two and features in tree selection frameworks, such as species climate matrices in @roloff2009 and in @brune2016, where several studies investigating (hydro)climatic sensitivity were summarized.
+Contrastingly, *Acer pseudoplatanus* showed decreased absolute growth with higher temperatures; this corresponds with their classification as *not very suitable* under [@roloff2009].
+*Platanus acerifolia* and *Acer platanoides* are considered *suitable* and *very suitable* under @roloff2009, yet show the greatest decline with temperatures here.
+Indeed, several several studies reviewed in @brune2016 highlight *A. platanoides* as drought-tolerant.
+Such contrasting indications are also reported for *A. pseudoplatanus*, and even report *A. hippocastanum* as only moderately tolerant, despite performing best here. 
+These differences may be due to idiosyncrasies of planting sites, such as low soil volume and increased potential for water stress, (long-term) interactions with pathogens and herbivores [e.g., @meineke2018], or perhaps related to data quality.
+Most important, these differences highlight the need for assessments of performance (e.g., as growth) under location-specific conditions, and this study provides a first attempt and insights into temperature sensitivity at the scale of an entire metropolis.
+
 - Urban fabric: LCZ density 
-- and sensitive trees in hot spots
+- and sensitive trees in hot spots ..**Increased growth rates.. live fast, die young - high turnover! requires focused planting and/or maintenance to counter [@smith2019].**
 
@@
 We acknowledge the financial and logistic constraints that authorities are subject to, yet the suggestions below are likely valuable for this and other urban green infrastructure initiatives, as several of them are easily implemented.
-
-**Increased growth rates.. live fast, die young - high turnover! requires focused planting and/or maintenance to counter [@smith2019].**
 

### 2022-01-12 analysis/paper/paper.Rmd
cont disc

@@
 
-However, this was not the case for planting bed area and $LCZ6_{\over{150~m}}$, which outperformed several temperature-based models.
+The null models containing only age to explain diameter captured a large portion of deviance.
+
+Including additional covariates thus resulted only in comparatively small buincreases of predictive skill
+The inclusion of temperature and temperature-age interactions improved the predictive skill, with models including the LandSat-derived UHI measures performing best on average.
+Note that non-temperature models containing planting bed area and $LCZ6_{\over{150~m}}$ alone outperformed several of those including other temperature measures and covariate combinations.
 This indicated the importance of both covariates, with $LCZ6_{\over{150~m}}$ also represented in the best-performing model, corroborating its impact on stem diameter development.
-As $LCZ6_{\over{300~m}}$ only marginally increased performance compared to the null model (age only), the reference scale for the impact of the urban fabric (i.e., LCZ6) has a limit between $150$ to $300~m$.
-Nutrient availablity, similar to $LCZ6_{\over{300~m}}$, also only improved model performance marginally.
+As $LCZ6_{\over{300~m}}$ only marginally increased performance compared to the null model (age only), the reference scale for the impact of the urban fabric (i.e., LCZ6) has a limit at $<300~m$.
+Nutrient availability, similar to $LCZ6_{\over{300~m}}$, also only improved model performance marginally.
 Building height, regardless of its reference scale (i.e., $150$ $vs$ $300~m$), as another proxy for urban fabric, showed minimal to no improvement compared to the age-only null model.
@@
 
-(ref:temp-sens) Temperature sensitivity ....
+(ref:temp-sens) Temperature sensitivity of diameter growth centered on species-specific means for comparability and to aid interpretation of inter-specific differences.
+Lines and bands are from ordinary least squares fits on the GAM estimates in Fig.$~$\@ref(fig:fig-heat-effect) (A) and the corresponding coefficients (slopes, B).
 
@@
 
-### Intra-specific differences and the urban fabric
+### Intra-specific differences
 
 Generally, the ten investigated species and cultivars showed consistent responses in sign, but differed substantially in magnitude, modulated by age.
-*Aesculum hippocastanum* , *Tilia intermedia*, *T. intermedia 'pallida'* and showed greatest absolute growth at high temperatures at younger (30-35) as well as older ages (60-65).
+*Aesculum hippocastanum* , *Tilia intermedia*, *T. intermedia* 'pallida' and showed greatest absolute growth at high temperatures at younger (30-35) as well as older ages (60-65).
 The drought tolerance and performance under high temperatures is well documented for the latter two and features in tree selection frameworks, such as species climate matrices in @roloff2009 and in @brune2016, where several studies investigating (hydro)climatic sensitivity were summarized.
@@
 
-- Urban fabric: LCZ density 
-- and sensitive trees in hot spots ..**Increased growth rates.. live fast, die young - high turnover! requires focused planting and/or maintenance to counter [@smith2019].**
-
 ## Utility, transferrability and limitations of approach
@@
 In addition, studies on climatic suitability and physiological responses are difficult to compare due to local acclimation of species to controls on growth and performance, as well as different methodological protocols [e.g., see @brune2016].
-Thus, the approach followed here, with its comparatively simple data requirements, can be implemented in a variety of settings to gain initial and/or guiding insights on tree performance in a specific urban region, and also serve as a benchmark to compare other empirical estimates or mechanistic modelswith.
+Thus, the approach followed here, with its comparatively simple data requirements, can be implemented in a variety of settings to gain initial and/or guiding insights on tree performance in a specific urban region, and also serve as a benchmark to compare other empirical estimates or mechanistic models with.
 It may even allow estimating tree performance under future conditions to some extent (e.g., considering increase in temperature alone for this study), but is of course contingent on data coverage (e.g., demographics, environmental gradients).
@@
 While their interpretation follows first principles and may also allow anticipating the performance of a given tree under a range of increased temperatures, insights derived from mechanistic understanding are indispensable.
-This is because a change in one environmental condition, e.g., temperature, may also entail alterations to other drivers, and ultimately, affect physiological processes in ways not yet captured in existing data.
+This is because a change in one environmental condition, e.g., temperature or pest abundance, may also entail alterations to other drivers, and ultimately, affect physiological processes in ways not yet captured in existing data.
 The data provided in Berlin's tree inventory required extensive quality controlling, and it is likely that inaccurate (diameter, age) or interpolated (diameter) observations were including in the analyses.
@@
 This not only allows identifying species-specific and intra-specific (age-dependent) sensitivities to urban conditions, but can also inform how age cohorts may respond as temperatures at a given location shift along the present continuum in the future, while accounting for the impacts of additional urban conditions (e.g., management, disturbance, urban fabric).
+Notably, increased growth rates at higher temperatures, as reported here, also entail greater demographic turnover through mortality [@smith2019].
+young age groups for most species showed greater absolute growth with higher temperatures, .
+<!-- Indeed, growing season temperatures in the Berlin region are predicted to rise for the remainder of this century. -->
+Thus, as the majority of individuals were situated at the upper end of the temperature continuum, high temperature sensitivity may entail additional management due to faster turn-over - subsequent replacement by more appropriate species.
+<!-- - ultimately, this will be a trade-off between tolerance of urban conditions, management goals (e.g., ecosystem services) and longevity. -->
+<!-- Assessments such as ours here may further aid in selecting appropriate species to achieve these goals. -->
 The insights derived from this study, and potentially others following this approach in the future, can thus add another dimension to tree selection approaches implemented down to individual trees and locations.

### 2022-01-14 analysis/paper/paper.Rmd
write write
first draft, abstract missing

@@
 
-* null model high, but for predictive purposes more is better
-Models including (excess) heat measures were on average better than those without.
+<!-- * null model high, but for predictive purposes more is better -->
+<!-- Models including (excess) heat measures were on average better than those without. -->
 
 The null models containing only age to explain diameter captured a large portion of deviance.
-
-Including additional covariates thus resulted only in comparatively small buincreases of predictive skill
+Including additional covariates thus resulted only in comparatively small increases of predictive skill.
 The inclusion of temperature and temperature-age interactions improved the predictive skill, with models including the LandSat-derived UHI measures performing best on average.
@@
 This indicated the importance of both covariates, with $LCZ6_{\over{150~m}}$ also represented in the best-performing model, corroborating its impact on stem diameter development.
-As $LCZ6_{\over{300~m}}$ only marginally increased performance compared to the null model (age only), the reference scale for the impact of the urban fabric (i.e., LCZ6) has a limit at $<300~m$.
+As $LCZ6_{\over{300~m}}$ only marginally increased performance compared to the null model (age only), the reference scale for the impact of the urban fabric (i.e., LCZ6) likely has a limit at $<300~m$.
 Nutrient availability, similar to $LCZ6_{\over{300~m}}$, also only improved model performance marginally.
@@
 
-1. Temperature, environmental and urban controls on tree growth
-
-	+ Intra specific differences
-	  + management
-	  + root vs. canopy (huber value?)
-	  +  accelerated recent growth
-	  + comparison young vs. old -> increased CO2, perhaps limitation of space for time,
-	     as lower conductance under high CO2 may result in more water savings (check WUE
-	     in old vs. young trees? future direction).[@dusenge2019] [@mccarthy2011]
-	+ Sensitive trees in hot spots
-	+ Temporal dynamics  
-	+ Validity and limitations
-
-	[@bussotti2014]
-
-2. Implications:  
-  + Policy / planning:  
-    - tree species
-	  - Increase temporal coverage and accuracy  
-		- which trees to prioritize for measurement campaigns  
-		- Supplement species matrix etc. with these analyses  
-		- *Landsat temp best on average - easily accessible/maintainable as compared to model outputs
-* $lcz6_{150}$ vs. $lcz6_{300}$ - scale of impact (further investigation, note in discussion) 
-        - planting bed area .. more data!
-  + Open data important --> drives innovation
+<!-- 1. Temperature, environmental and urban controls on tree growth   -->
+<!-- 	+ Intra specific differences -->
+<!-- 	  + management -->
+<!-- 	  + root vs. canopy (huber value?) -->
+<!-- 	  +  accelerated recent growth -->
+<!-- 	  + comparison young vs. old -> increased CO2, perhaps limitation of space for time, -->
+<!-- 	     as lower conductance under high CO2 may result in more water savings (check WUE -->
+<!-- 	     in old vs. young trees? future direction).[@dusenge2019] [@mccarthy2011] -->
+<!-- 	+ Sensitive trees in hot spots -->
+<!-- 	+ Temporal dynamics   -->
+<!-- 	+ Validity and limitations   -->
+<!-- 	[@bussotti2014] -->
+<!-- 2. Implications:   -->
+<!--   + Policy / planning:   -->
+<!--     - tree species -->
+<!-- 	  - Increase temporal coverage and accuracy   -->
+<!-- 		- which trees to prioritize for measurement campaigns   -->
+<!-- 		- Supplement species matrix etc. with these analyses   -->
+<!-- 		- *Landsat temp best on average - easily accessible/maintainable as compared to model outputs -->
+<!-- * $lcz6_{150}$ vs. $lcz6_{300}$ - scale of impact (further investigation, note in discussion)  -->
+<!--         - planting bed area .. more data! -->
+<!--   + Open data important, drives innovation -->
 
@@
 *Platanus acerifolia* and *Acer platanoides* are considered *suitable* and *very suitable* under @roloff2009, yet show the greatest decline with temperatures here.
-Indeed, several several studies reviewed in @brune2016 highlight *A. platanoides* as drought-tolerant.
+Indeed, several studies reviewed in @brune2016 highlight *A. platanoides* as drought-tolerant.
 Such contrasting indications are also reported for *A. pseudoplatanus*, and even report *A. hippocastanum* as only moderately tolerant, despite performing best here. 
@@
 This is because a change in one environmental condition, e.g., temperature or pest abundance, may also entail alterations to other drivers, and ultimately, affect physiological processes in ways not yet captured in existing data.
-The data provided in Berlin's tree inventory required extensive quality controlling, and it is likely that inaccurate (diameter, age) or interpolated (diameter) observations were including in the analyses.
-However, the minimum requirement of 10000 observations per species, and the focus on average diameter responses across age groups decreased potential impacts such observations may have had to a degree that was considered satisfactory, while uncertainty is communicated effectively (i.e., larger confidence bands for *T.
+Furthermore, the data provided in Berlin's tree inventory required extensive quality controlling, and it is likely that inaccurate (diameter, age) or interpolated (diameter) observations were including in the analyses.
+However, the minimum requirement of 10000 observations per species, and the focus on average diameter responses across age groups decreased potential impacts of such observations may have had to a degree that was considered satisfactory, while uncertainty is communicated effectively (i.e., larger confidence bands for *T.
 intermedia 'Pallida'*).
@@
 Notably, increased growth rates at higher temperatures, as reported here, also entail greater demographic turnover through mortality [@smith2019].
-young age groups for most species showed greater absolute growth with higher temperatures, .
+<!-- young age groups for most species showed greater absolute growth with higher temperatures, . -->
 <!-- Indeed, growing season temperatures in the Berlin region are predicted to rise for the remainder of this century. -->
@@
 <!-- Assessments such as ours here may further aid in selecting appropriate species to achieve these goals. -->
-The insights derived from this study, and potentially others following this approach in the future, can thus add another dimension to tree selection approaches implemented down to individual trees and locations.
+The insights derived from this study, and potentially others following this approach in the future, can therefore add another dimension to tree selection approaches implemented down to individual trees and locations.
 It thus complements quantitative and qualitative tools, such as the climate matching tool, ecological site classification or species-climate matrix [see @hirons2019; @roloff2009; @brune2016; and references therein].
@@
 Urban tree databases typically represent a single time point (e.g., see https://opentrees.org), despite repeated inventories at regular and/or irregular intervals within cities or their districts.
-Storing and providing multi-temporal inventories, rather than over-writing existing records, would give planners and researchers a wealth of additional information on individual tree performance, which – even at low temporal coverage – would allow verifying space-for-time substitutions, or make them superfluous.
+Storing and providing multi-temporal inventories, rather than over-writing existing records, would give planners and researchers a wealth of additional information on individual tree performance, which – even at low temporal coverage – would increase confidence in space-for-time substitutions, or make them superfluous.
 We also urge authorities within and across cities to enhance and ideally standardize their (meta) data collection and provisioning to facilitate data quality control, increase confidence, and enable synthesis studies across districts and regions [@ossola2020].
@@
 However, there are financial and logistic trade-offs for data collection of the above, with contrasting needs across stakeholder groups [e.g., academics and authorities, @ostberg2013] where data that supports meeting long-term objectives should be prioritized.
-Less abundant species which are candidates for future plantings could be assessed with higher-quality standards regarding points measured above, to build a more reliable data source even at potentially lower coverage demographically and spatially.
+Less abundant species which are candidates for future plantings could be assessed with higher-quality standards regarding the above, to build a more reliable data source even at potentially lower coverage demographically and spatially.
 Managers and planners also should be aware of potential scale-limit of the urban fabric (e.g., $LCZ6_{\over{150~m}}~vs.~LCZ6_{\over{300~m}}$) on tree performance when altering surroundings or planting in a given area; efforts should be directed toward identifying the most pertinent environmental covariates and how their influence scales in space.
@@
 
-## Conclusions
-
-- novel application of open data to infer sensitivity to urban heat and other pertinent environmental drivers
-- identified age-shift and most tolerant species
-- provide policy-relevant suggestions for authorities and practicioners
-- open several avenues for further inquiry:
-  + multi-temporal
-  + increase covariate coverage to see if predictable performance alteration (planting bed!)
-  + less abundant species --> potential to identify suitable trees in a specific urban context (e.g., bespoke for Berlin)
+## Conclusion
+
+We showcased the application of a GAM-driven analyses to identify the sensitivity of urban tree growth to environmental and urban drivers in a space-for-time framework. 
+The extensive data coverage (tree inventory and covariates) allowed us to account for location-specific factors and identify the most influences on growth potential (i.e., maximum diameter) for individual species.
+The identified age-dependent shift of temperature sensitivity, which was fairly consistent across species, as well as inter-specific differences thereof, will inform and guide further research. 
+Notably, increasing the confidence in and coverage of data collection on less abundant species is key to maximize the potential of the approach developed here.
+Lastly, together with recommendations for tree selection and infrastructure management, we hope to contribute to the design of best-practices for adaptation and mitigation using street trees with the flexible yet simple analyses approach (regarding data requirements) taken here.
+
+<!-- - novel application of open data to infer sensitivity to urban heat and other pertinent environmental driver -->
+<!-- - identified age-shift and most tolerant species -->
+<!-- - provide policy-relevant suggestions for authorities and practicioners -->
+<!-- - open several avenues for further inquiry: -->
+<!--   + multi-temporal -->
+<!--   + increase covariate coverage to see if predictable performance alteration (planting bed!) -->
+<!--   + less abundant species --> 
+<!-- potential to identify suitable trees in a specific urban context (e.g., bespoke for Berlin) -->
 
 # Acknowledgements
+
+AGH and IH were supported through the Helmholtz-Climate-Initiative (HI-CAM), funded by the Helmholtz Initiative and Networking Fund; the authors are responsible for the content of this publication.
 

### 2022-01-14 R/plan.R

@@ -1028,14 +1028,14 @@ plan <- drake::drake_plan(
         output_format = bookdown::html_document2(),
         quiet = TRUE
     ),
-    # paper_word = rmarkdown::render(
-    #     knitr_in("./analysis/paper/paper.Rmd"),
-    #     output_dir = "./analysis/paper/",
-    #     output_file = file_out("paper_knit.docx"),
-    #     # output_file = "./paper_knit.html",
-    #     output_format = bookdown::word_document2(),
-    #     quiet = TRUE
-    # ),
+    paper_word = rmarkdown::render(
+        knitr_in("./analysis/paper/paper.Rmd"),
+        output_dir = "./analysis/paper/",
+        output_file = file_out("paper_knit.docx"),
+        # output_file = "./paper_knit.html",
+        output_format = bookdown::word_document2(),
+        quiet = TRUE
+    ),
     # paper_pdf = rmarkdown::render(
     #     knitr_in("./analysis/paper/paper.Rmd"),
     #     output_dir = "./analysis/paper/",

### 2022-03-18 analysis/paper/paper.Rmd
adjusting intro

@@
 
-Urban trees show a tendency for enhanced growth rates and/or productivity compared to rural conspecifics [e.g., @briber2015; @obrien2012], which is typically attributed to increased temperatures [@jia2018; @pretzsch2017], yet feature a broad range of effect sizes and signs (i.e., reduced growth) specific to species and location.
+Several urban trees have been shown to have tendencies of enhanced growth rates and/or productivity compared to rural conspecifics [e.g., @briber2015; @obrien2012], which is typically attributed to increased temperatures [@jia2018; @pretzsch2017], however, a broad range of effect sizes and signs (i.e., reduced growth) specific to species and location are also commonly reported.
 @zhao2016 showed that productivity rates, as a proxy for growth, increased within urban clusters as urbanization intensifies using remotely sensed vegetation indices.
@@
 <!-- [@briber2015; @obrien2012] … generally increased growth -->
-Contrastingly,  @gillner2014 highlight decreased growth for *Acer* species  (*A.
-platanoides* and *pseudoplatanus*), *Platanus x hispanica* and *Quercus rubra* with higher summer temperatures of the preceding year, especially when compounded with drought, in another German metropolis (Dresden).
-@quigley2004 identified absolute growth potential decreased for species between rural and urban conspecifics, yet assessments were
-<!-- lacked spatially-explicit effect size estimates across the urban-rural space and  -->
-limited to comparatively small sample sizes per group ($n_{total}~=~230$ divided in 15 species, 3 groups and 2 locations).
-@pretzsch2017 inferred enhanced growth in recent decades and across urban locations spanning several latitudes, including Berlin - however, only 145 individuals of one species (*T.
+<!-- Contrastingly,  @gillner2014 highlight decreased growth for *Acer* species  (*A.
+platanoides* and *pseudoplatanus*), *Platanus x hispanica* and *Quercus rubra* with higher summer temperatures of the preceding year, especially when compounded with drought, in another major German city (Dresden).
+-->
+<!-- @quigley2004 identified absolute growth potential decreased for species between rural and urban conspecifics, yet assessments were -->
+<!-- <!-- lacked spatially-explicit effect size estimates across the urban-rural space and  --> -->
+<!-- limited to comparatively small sample sizes per group ($n_{total}~=~230$ divided in 15 species, 3 groups and 2 locations). -->
+<!-- @pretzsch2017 inferred enhanced growth in recent decades and across urban locations spanning several latitudes, including Berlin - however, only 145 individuals of one species (*T.
 cordata*) were assessed there.
-As mentioned previously, climate-growth relationships can vary substantially between species, and in fact, @quigley2004 and @pretzsch2017 report contrasting results regarding average tree diameter, i.e.
+-->
+<!-- As mentioned previously, climate-growth relationships can vary substantially between species, and in fact, @quigley2004 and @pretzsch2017 report contrasting results regarding average tree diameter, i.e.
 smaller or larger for urban $vs.$ rural trees of same age.
+-->
 Similarly, for Berlin, @dahlhausen2018, identified enhanced growth in highly urbanized environments (using basal area increments of a large sample of 252 trees) for *T.
 cordata*, the most abundant tree of the city, which they attributed to the UHI effect, while intermediate development intensity was adverse for tree growth.
+By contrast,  @gillner2014 highlight decreased growth for *Acer* species  (*A.
+platanoides* and *pseudoplatanus*), *Platanus x hispanica* and *Quercus rubra* with higher summer temperatures of the preceding year, especially when compounded with drought, in another major German city (Dresden).
 These differences in growth trends may result from contrasting species-specific responses to increased temperatures, but are indeed affected by other (time-varying) factors and stochastic processes, such as water availability, pollution and road-salt loading, structural impedance by infrastructure, or management, etc.
 [@rhoades1999; @quigley2004; @randrup2001; @pauleit2002].
-Further, the variability in responses may require that assessments are developed for a specific region, because well-understood tree characteristics [e.g., see @brune2016;@roloff2009], could be strongly modulated predictably due to management, planting practices, or other environmental controls; for example, if drought hardiness is related to extensive root networks, restricted soil volumes available to street trees will render a species more vulnerable to water stress.
+The potential for (predictable) variability in responses is a strong indication that assessments ought to be done specifically for a given urban area.
+<!-- , because well-understood tree characteristics [e.g., see @brune2016;@roloff2009] could be strongly - but predictably - modulated by management and/or other environmental controls; for example, if drought hardiness is related to extensive root networks, restricted soil volumes available to street trees will render a species more vulnerable to water stress.
+-->
+<!-- Further, the variability in responses may require that assessments are developed for a specific region, because well-understood tree characteristics [e.g., see @brune2016;@roloff2009], could be strongly modulated predictably due to management, planting practices, or other environmental controls; for example, if drought hardiness is related to extensive root networks, restricted soil volumes available to street trees will render a species more vulnerable to water stress.
+-->
 <!-- **expand on location-specific factors that can affect growth** -->
@@
 -->
-Space-for-time substitutions and time series comparisons between and within locations are a common approach (cf.
+<!-- Space-for-time substitutions and time series comparisons between and within locations are a common approach (cf.
 studies above) to generate inferences in observational (rather than treatment-control) studies, where manipulations are costly or logistically unfeasible due to time and/or financial constraints.
-However, they require  accounting for confounding factors specific to trees' environments, such as street characteristics, development intensity, available soil volume, etc.
+-->
+<!-- However, they require  accounting for confounding factors specific to trees' environments, such as street characteristics, development intensity, available soil volume, etc. -->
+
+Several of the aforementioned studies applied these approaches to quantify temperature and excess heat on growth, they typically compare trees grouped using qualitative or summary descriptors of sampling sites, disregarding the spatio-temporal variability in location-specific factors noted above.
 While several of the aforementioned studies applied these approaches to quantify temperature and excess heat on growth, they typically compare trees grouped using qualitative or summary descriptors of sampling sites, disregarding the spatio-temporal variability in location-specific factors noted above.
@@
 This can be  exacerbated by a lack of co-located environmental variables (i.e.
-measured in situ) at pertinent spatial scales, for instance, as noted by @wohlfahrt2019 for air temperature and tree leaf phenology, which may lead to incorrect inferences and interpretations for the role of climate change on growth/productivity when applying space-for-time substitutions.
+measured *in situ*) at pertinent spatial scales, for instance, as noted by @wohlfahrt2019 for air temperature and tree leaf phenology, which may lead to incorrect inferences and interpretations for the role of climate change on growth/productivity.
 It is thus likely that the varying and even contrasting growth responses observed for urban trees across and within studies are at least modulated by some confounding factors, making the attribution to a single driver, such as excess heat, more difficult and possibly less accurate.
@@
 <!-- This work constitutes a case study for Berlin, the results are a valuable contribution toward Berlin's current and future management of its tree stock.  -->
-Our results are a contribution toward Berlin's current and future management of its tree stock and may help drive adaptation to climate change.
-Despite being a case study for a single city, we believe our work may provide a flexible approach for other cities with available or growing inventories, as well as ancillary environmental data, and may also inform the use of other planning tools, such as species-climate matrices [@roloff2009] regarding temperature sensitivity.
+<!-- Our results are a contribution toward Berlin's current and future management of its tree stock and may help drive adaptation to climate change. -->
+<!-- Despite being a case study for a single city, we believe our work may provide a flexible approach for other cities with available or growing inventories, as well as ancillary environmental data, and may also inform the use of other planning tools, such as species-climate matrices [@roloff2009] regarding temperature sensitivity.
+-->
 <!-- This is because the high temporal and spatial variability of controls on growth, and thus variability in responses, require that assessments are developed for a specific region, because well-understood tree characteristics [e.g., see @brune2016;@roloff2009], could be strongly modulated  in predicable ways at a given location or time due to management, planting practices, or other environmental controls; for example, if drought hardiness is related to extensive root networks, small soil volumes available to street trees could render a species vulnerable to water stress.
@@
 Notably, increasing the confidence in and coverage of data collection on less abundant species is key to maximize the potential of the approach developed here.
-Lastly, together with recommendations for tree selection and infrastructure management, we hope to contribute to the design of best-practices for adaptation and mitigation using street trees with the flexible yet simple analyses approach (regarding data requirements) taken here.
+<!-- Lastly, together with recommendations for tree selection and infrastructure management, we hope to contribute to the design of best-practices for adaptation and mitigation using street trees with the flexible yet simple analyses approach (regarding data requirements) taken here.
+-->
+Our results are a contribution toward Berlin's current and future management of its tree stock and may help drive adaptation to climate change.
+Despite being a case study for a single city, we believe our work may provide a flexible approach for other cities with available or growing inventories, as well as ancillary environmental data, and may also inform the use of other planning tools, such as species-climate matrices [@roloff2009] regarding temperature sensitivity.
 

### 2022-03-18 analysis/paper/paper.Rmd
tighten intro

@@
 
-Several urban trees have been shown to have tendencies of enhanced growth rates and/or productivity compared to rural conspecifics [e.g., @briber2015; @obrien2012], which is typically attributed to increased temperatures [@jia2018; @pretzsch2017], however, a broad range of effect sizes and signs (i.e., reduced growth) specific to species and location are also commonly reported.
-@zhao2016 showed that productivity rates, as a proxy for growth, increased within urban clusters as urbanization intensifies using remotely sensed vegetation indices.
-Further, @moser-reischl2019a identified positive associations between air temperature and radial growth for two species (total of 20 individuals) commonly selected by urban planners (*Tilia  cordata* MilL., *Rubinia pseudoacacia*) in Munich.
+Several urban tree species have been shown to have tendencies of enhanced growth rates and/or productivity compared to rural conspecifics [e.g., @briber2015; @obrien2012], which is typically attributed to increased temperatures [@jia2018; @pretzsch2017], however, a broad range of effect sizes and signs (i.e., reduced growth) specific to species and location are also commonly reported alongside.
+For instance, @zhao2016 showed that productivity rates, as a proxy for growth, increased within urban clusters as urbanization intensifies using remotely sensed vegetation indices.
+Furthermore, @moser-reischl2019a identified positive associations between air temperature and radial growth for two species commonly selected by urban planners (*Tilia  cordata* MilL., *Rubinia pseudoacacia*) in Munich.
 <!-- [@briber2015; @obrien2012] … generally increased growth -->
@@
 -->
-Similarly, for Berlin, @dahlhausen2018, identified enhanced growth in highly urbanized environments (using basal area increments of a large sample of 252 trees) for *T.
+Similarly, for Berlin, @dahlhausen2018, identified enhanced growth in highly urbanized environments for *T.
 cordata*, the most abundant tree of the city, which they attributed to the UHI effect, while intermediate development intensity was adverse for tree growth.
@@
 [@rhoades1999; @quigley2004; @randrup2001; @pauleit2002].
-The potential for (predictable) variability in responses is a strong indication that assessments ought to be done specifically for a given urban area.
+<!-- The potential for (predictable) variability in responses is a strong indication that assessments ought to be done specifically for a given urban area. -->
 <!-- , because well-understood tree characteristics [e.g., see @brune2016;@roloff2009] could be strongly - but predictably - modulated by management and/or other environmental controls; for example, if drought hardiness is related to extensive root networks, restricted soil volumes available to street trees will render a species more vulnerable to water stress.
@@
 <!-- However, they require  accounting for confounding factors specific to trees' environments, such as street characteristics, development intensity, available soil volume, etc. -->
-
-Several of the aforementioned studies applied these approaches to quantify temperature and excess heat on growth, they typically compare trees grouped using qualitative or summary descriptors of sampling sites, disregarding the spatio-temporal variability in location-specific factors noted above.
-While several of the aforementioned studies applied these approaches to quantify temperature and excess heat on growth, they typically compare trees grouped using qualitative or summary descriptors of sampling sites, disregarding the spatio-temporal variability in location-specific factors noted above.
-This can hinder the extrapolation from individual sampling sites toward predicting effects across entire urban areas and tree stocks, especially when studies rely on labor-intensive methods, which are limited logistically by sampling effort, reducing sample sizes and coverage of species and space.
-This can be  exacerbated by a lack of co-located environmental variables (i.e.
-measured *in situ*) at pertinent spatial scales, for instance, as noted by @wohlfahrt2019 for air temperature and tree leaf phenology, which may lead to incorrect inferences and interpretations for the role of climate change on growth/productivity.
+<!-- While several of the aforementioned studies applied these approaches to quantify temperature and excess heat on growth, they typically compare trees grouped using qualitative or summary descriptors of sampling sites, disregarding the spatio-temporal variability in location-specific factors noted above.
+-->
+Several of the aforementioned studies applied space-for-time substitutions or time series comparisons to quantify temperature and excess heat on growth for a given region.
+However, they typically compared trees grouped using qualitative or summary descriptors of sampling sites, disregarding the spatial and/or temporal variability in location-specific factors noted above.
+This can hinder the extrapolation from individual sampling sites toward predicting effects across entire urban areas and tree stocks, especially when studies rely on labor-intensive methods, 
+<!-- which are limited logistically by sampling effort,  -->
+reducing sample sizes and coverage of species and space.
+A lack of co-located environmental variables (i.e.
+measured *in situ*) at pertinent spatial scales can further exacerbate these limitations, for instance, as noted by @wohlfahrt2019 for air temperature and tree leaf phenology, which may lead to incorrect inferences and interpretations for the role of climate change on growth/productivity.
 It is thus likely that the varying and even contrasting growth responses observed for urban trees across and within studies are at least modulated by some confounding factors, making the attribution to a single driver, such as excess heat, more difficult and possibly less accurate.
@@
 Berlin, as one of the greenest cities in Europe, provides an openly accessible tree inventory, with spatio-temporal environmental data sets relevant to tree growth. 
-It features a total of 650000 individuals covering 94 genera and at least 600 species and/or cultivars, listing information on location, stem diameter (at breast height; $DBH$), and stem height, amongst other variables, for the majority of street and park trees.
+It features nearly 650000 individuals covering 94 genera and some 600 species and/or cultivars, listing information on location, stem diameter (at breast height; $DBH$), and stem height, among other variables, for the majority of street and park trees.
 For this study, our objective was to assess the impact of excess urban heat, i.e.

### 2022-05-14 analysis/paper/paper.Rmd
update paper

@@
 
-## General approach: space-for-time analyses
-
-<!-- Species-specifc growth responses in relation across Berlin were assessed in relat -->
-We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of (excess) heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; LandSat-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section$~$\@ref(sec:stat-gam) for details).
-Covariates were extracted at 150 and 300$~m$ to infer the impact of reference scale of the urban fabric on tree growth. 
-From all tested models the most suitable (i.e., parsimonious with highest explanatory) was employed for further analyses.
+<!-- ## General approach: space-for-time analyses -->
+
+<!-- <!-- Species-specifc growth responses in relation across Berlin were assessed in relat --> -->
+<!-- We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of (excess) heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; LandSat-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section$~$\@ref(sec:stat-gam) for details).
+-->
+<!-- Covariates were extracted at 150 and 300$~m$ to infer the impact of reference scale of the urban fabric on tree growth.  -->
+<!-- From all tested models the most suitable (i.e., parsimonious with highest explanatory) was employed for further analyses. -->
 
@@
 
+## General approach: space-for-time analyses
+
+<!-- Species-specifc growth responses in relation across Berlin were assessed in relat -->
+We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of (excess) heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; LandSat-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section$~$\@ref(sec:stat-gam) for details).
+Covariates were extracted at 150 and 300$~m$ to infer the impact of reference scale of the urban fabric on tree growth. 
+From all tested models the most suitable (i.e., parsimonious with highest explanatory) was employed for further analyses.
+
 ## Dendrochronological sampling {#sec:methods-dendro}
@@
 Where any two comparable models had a $| \Delta AIC| > 10$, we considered the model with the lower score more suitable. 
-Future research may focus on collecting additional data (e.g., increasing the coverage on planting bed area) and subsequently deriving species-specific smooths for ancillary environmental covariates.
+<!-- Future research may focus on collecting additional data (e.g., increasing the coverage on planting bed area) and subsequently deriving species-specific smooths for ancillary environmental covariates.
+-->
 We chose to carry out the analysis in its current form rather than on considerably smaller but comparable sample sizes across models, to identify the strongest relationships in the existing data.
@@
 The inclusion of temperature and temperature-age interactions improved the predictive skill, with models including the LandSat-derived UHI measures performing best on average.
-Note that non-temperature models containing planting bed area and $LCZ6_{\over{150~m}}$ alone outperformed several of those including other temperature measures and covariate combinations.
-This indicated the importance of both covariates, with $LCZ6_{\over{150~m}}$ also represented in the best-performing model, corroborating its impact on stem diameter development.
-As $LCZ6_{\over{300~m}}$ only marginally increased performance compared to the null model (age only), the reference scale for the impact of the urban fabric (i.e., LCZ6) likely has a limit at $<300~m$.
-Nutrient availability, similar to $LCZ6_{\over{300~m}}$, also only improved model performance marginally.
+Note that non-temperature models containing planting bed area and $LCZ6_{\frac{}{150~m}}$ alone outperformed several of those including other temperature measures and covariate combinations.
+This indicated the importance of both covariates, with $LCZ6_{\frac{}{150~m}}$ also represented in the best-performing model, corroborating its impact on stem diameter development.
+As $LCZ6_{\frac{}{300~m}}$ only marginally increased performance compared to the null model (age only), the reference scale for the impact of the urban fabric (i.e., LCZ6) likely has a limit at $<300~m$.
+Nutrient availability, similar to $LCZ6_{\frac{}{300~m}}$, also only improved model performance marginally.
 Building height, regardless of its reference scale (i.e., $150$ $vs$ $300~m$), as another proxy for urban fabric, showed minimal to no improvement compared to the age-only null model.
@@
 (ref:heat-effect) Age-group averaged relationships between the LandSat-derived UHI and stem diameter for the ten most abundant species from the best model (i.e., Equation$~$\@ref(eq:gam-full) with LCZ6 as covariate).
-Note that estimates were derived with $LCZ6_{\over{150~m}} = 0.5$, and that y-axis scales differ between panel rows.
+Note that estimates were derived with $LCZ6_{\frac{}{150~m}} = 0.5$, and that y-axis scales differ between panel rows.
 The colored lines and bands are the averaged GAM fit, and gray lines are least-squares estimates fitted to these averages.
@@
 
-The chosen model also highlighted that, on average, trees growing in less urbanized environments (i.e., more open to mid-rise urban land cover; $LCZ6_{\over{150~m}}$) achieve somewhat greater diameters.
-Note, that planting bed area, similar to $LCZ6_{\over{150~m}}$, was identified as a covariate increasing explained deviance, both in models with and without a temperature measure, and represents an opportunity for further investigation, especially following additional data collection.
+The chosen model also highlighted that, on average, trees growing in less urbanized environments (i.e., more open to mid-rise urban land cover; $LCZ6_{\frac{}{150~m}}$) achieve somewhat greater diameters.
+Note, that planting bed area, similar to $LCZ6_{\frac{}{150~m}}$, was identified as a covariate increasing explained deviance, both in models with and without a temperature measure, and represents an opportunity for further investigation, especially following additional data collection.
 
@@
 
-## Utility, transferrability and limitations of approach
+## Utility, transferability and limitations of approach
 
@@
 Furthermore, the implementation of GAM models here with `mgcv::bam()` is computationally efficient [@wood2017a], making model testing, selection and final application a rapid process even for abundant observations (here $n >150000$ and processing times of $t < 60~s$ on a desktop computer for the selected model).
+Future research may focus on collecting additional data (e.g., increasing the coverage on planting bed area) and subsequently deriving species-specific smooths for ancillary environmental covariates.
 
@@
 
-### Tree selection
-
-Tree species selection for urban forests depends on several often competing factors -  such as aesthetic appeal and ecosystem service provision [e.g., @roman2020] - and has been focused on identifying trees suitable for current and future conditions for the last decades with increasingly strong evidence basis [e.g., @hirons2019].
+<!-- ### Tree selection -->
+
+Tree species selection for urban forests and urban greening depends on several often competing factors -  such as aesthetic appeal and ecosystem service provision [e.g., @roman2020] - and has been focused on identifying trees suitable for current and future conditions for the last decades with increasingly strong evidence basis [e.g., @hirons2019].
 Here, biogeographical approaches aim to match natural (i.e., non-urban) species distributions within current bioclimatic envelopes with future conditions at a planting site [e.g., @broadmeadow2005].
@@
 <!-- Indeed, growing season temperatures in the Berlin region are predicted to rise for the remainder of this century. -->
-Thus, as the majority of individuals were situated at the upper end of the temperature continuum, high temperature sensitivity may entail additional management due to faster turn-over - subsequent replacement by more appropriate species.
+Thus, as the majority of individuals were situated at the upper end of the temperature continuum, high temperature sensitivity may entail additional management due to faster turn-over and potentially subsequent replacement by more appropriate species.
 <!-- - ultimately, this will be a trade-off between tolerance of urban conditions, management goals (e.g., ecosystem services) and longevity. -->
@@
 
-### Green infrastructure management, planning and open data
-
-The availability and quality of green infrastructure data, such as street tree inventories, and environmental covariates, determine the confidence in any inference of infrastructure performance in relation to its context (e.g., temperature, planting bed area).
-This is fundamental to inform management and planning to ensure that current and future targets are met. 
-Indeed, the execution of this study was possible due to the readily accessible and rich data sources provided by Berlin’s senate.
-However, few additions or alterations to data collection and provision schemes would have greatly  enhanced the utility and applicability of the data sets applied here. 
-We acknowledge the financial and logistic constraints that authorities are subject to, yet the suggestions below are likely valuable for this and other urban green infrastructure initiatives, as several of them are easily implemented.
-
-Urban tree databases typically represent a single time point (e.g., see https://opentrees.org), despite repeated inventories at regular and/or irregular intervals within cities or their districts.
-Storing and providing multi-temporal inventories, rather than over-writing existing records, would give planners and researchers a wealth of additional information on individual tree performance, which – even at low temporal coverage – would increase confidence in space-for-time substitutions, or make them superfluous.
-We also urge authorities within and across cities to enhance and ideally standardize their (meta) data collection and provisioning to facilitate data quality control, increase confidence, and enable synthesis studies across districts and regions [@ossola2020].
-Key information includes (1) location, absolute age and planting date, (2) management regimes (e.g., planting regime, crown pruning, irrigation), (3) data collection protocol and body (e.g., actual measurement $vs.$ estimate or interpolation), as well as (4) tree provenance to account for potential differences in ecotype and/or acclimation [@watkins2020].
-However, there are financial and logistic trade-offs for data collection of the above, with contrasting needs across stakeholder groups [e.g., academics and authorities, @ostberg2013] where data that supports meeting long-term objectives should be prioritized.
-Less abundant species which are candidates for future plantings could be assessed with higher-quality standards regarding the above, to build a more reliable data source even at potentially lower coverage demographically and spatially.
-Managers and planners also should be aware of potential scale-limit of the urban fabric (e.g., $LCZ6_{\over{150~m}}~vs.~LCZ6_{\over{300~m}}$) on tree performance when altering surroundings or planting in a given area; efforts should be directed toward identifying the most pertinent environmental covariates and how their influence scales in space.
+<!--  ### Green infrastructure management, planning and open data -->
+
+<!-- The availability and quality of green infrastructure data, such as street tree inventories, and environmental covariates, determine the confidence in any inference of infrastructure performance in relation to its context (e.g., temperature, planting bed area).
+-->
+<!-- This is fundamental to inform management and planning to ensure that current and future targets are met.  -->
+<!-- Indeed, the execution of this study was possible due to the readily accessible and rich data sources provided by Berlin’s senate. -->
+<!-- However, few additions or alterations to data collection and provision schemes would have greatly  enhanced the utility and applicability of the data sets applied here.  -->
+<!-- We acknowledge the financial and logistic constraints that authorities are subject to, yet the suggestions below are likely valuable for this and other urban green infrastructure initiatives, as several of them are easily implemented.
+-->
+
+Urban tree databases may only represent a single point in time (e.g., see https://opentrees.org), despite repeated inventories at regular and/or irregular intervals within cities or their districts.
+However, storing and providing multi-temporal inventories, rather than over-writing existing records, would give planners and researchers a wealth of additional information on individual tree performance, which – even at low temporal coverage – would increase confidence in space-for-time substitutions, or make them superfluous.
+In this regard, it is recommended to enhance and standardize (meta) data collection and provisioning to facilitate data quality control, thereby increasing confidence in the data, and enabling synthesis studies across districts and cities [@ossola2020].
+Key information should include (1) location, absolute age and planting date, (2) management regimes (e.g., planting regime, crown pruning, irrigation), (3) data collection protocol and body (e.g., actual measurement $vs.$ estimate or interpolation), as well as (4) tree provenance to account for potential differences in ecotype and/or acclimation [@watkins2020].
+<!-- However, there are financial and logistic trade-offs for data collection of the above, with contrasting needs across stakeholder groups [e.g., academics and authorities, @ostberg2013] where data that supports meeting long-term objectives should be prioritized.
+-->
+<!-- Less abundant species which are candidates for future plantings could be assessed with higher-quality standards regarding the above, to build a more reliable data source even at potentially lower coverage demographically and spatially.
+-->
+In addition, managers and planners also should be aware of potential scale-limit of the urban fabric (e.g., $LCZ6_{\frac{}{150~m}}~vs.~LCZ6_{\frac{}{300~m}}$) on tree performance when altering surroundings or planting in a given area; efforts should be directed toward identifying the most pertinent environmental covariates and how their influence scales in space.
 

### 2022-05-14 analysis/paper/paper.Rmd
add AIC to results

@@
 Including additional covariates thus resulted only in comparatively small increases of predictive skill.
-The inclusion of temperature and temperature-age interactions improved the predictive skill, with models including the LandSat-derived UHI measures performing best on average.
+The inclusion of temperature (additive structure) and temperature-age interactions improved the predictive skill, with interactions reaching higher explained deviance (i.e., skill).
+Models including the LandSat-derived UHI measures performed best on average, where models containing $LCZ6_{\frac{}{150~m}}$ or all covariates (saturated model) did best (Figure$~$\@ref(fig:fig-model-deviance)), which also holds true for other temperature/UHI measures.
 Note that non-temperature models containing planting bed area and $LCZ6_{\frac{}{150~m}}$ alone outperformed several of those including other temperature measures and covariate combinations.
@@
 Building height, regardless of its reference scale (i.e., $150$ $vs$ $300~m$), as another proxy for urban fabric, showed minimal to no improvement compared to the age-only null model.
+
+```{r mod-aic, echo=FALSE, cache = TRUE}
+
+mod_lcz6_tensor <- readRDS("./analysis/data/models/stat/filtered/mI_spatial_age_x_temp_by_species_lcz6_reBEZIRK_var-day_2007.Rds")
+mod_lcz6_add <- readRDS("./analysis/data/models/stat/filtered/mI_spatial_age_ADD_temp_by_species_lcz6_reBEZIRK_var-day_2007.Rds")
+
+mod_assessment <- rbind(broom::glance(mod_lcz6_add), broom::glance(mod_lcz6_tensor)) %>% 
+  cbind(data.frame(mod = c("add", "tensor")))
+
+aic_tens = mod_assessment$AIC[mod_assessment$mod == "tensor"]
+aic_add = mod_assessment$AIC[mod_assessment$mod == "add"]
+
+```
+
+Differences in sample sizes between model structures due to covariate and/or temperature data coverage were considerable (e.g., planting bed area $vs.$ soil nutrient models for UrbClim UHI).
+Thus, we considered the model with the overall greatest predictive skill and largest sample for further investigations, namely $LCZ6_{\frac{}{150~m}}$ for LandSat derived UHI.
+For this variable combination, the model structure with age and temperature interaction also had a lower AIC ($AIC~=~8.95 \cdot 10^5$) compared to the additive one ($AIC~=~8.98 \cdot 10^5$).
+<!-- For this variable combination, the model structure with age and temperature interaction also had a lower AIC ($AIC~=~`r format(aic_tens, digits = 3, scientific = TRUE)`$) compared to the additive one ($AIC~=~`r format(aic_add, digits = 3, scientific = TRUE)`$).
+-->
 

### 2022-05-22 analysis/paper/paper.Rmd
added methods formulas

@@
 
-(ref:temp-overview) Temp overviews; rows are landsat, berlin environmental atlas, and UrbClim, columns are morning, day/afternoon, night.
-**ADD Labels for Data source and time of day at left and bottom**
+(ref:temp-overview) Overview of temperature datasets used for assessing modelling impact on tree growth, with data sources in rows and averaged times in columns.
+Note the varying spatial coverage and resolution.
 
@@
 Models were implemented with `mgcv::bam()` [@wood2017a;@li2020a] and readers are referred to the detailed implementation in the supplemental material code.
-
 Considering all combinations of (excess) heat measures and covariates (with point, as well as 150/300$~m$ extractions), a total of `r nrow(drake::readd(bam_dbh_filtered))` models were applied.
-We selected the model with the highest explanatory power, based on residual deviance and observed vs. predicted fit, with comparatively largest sample size for final analyses.
-From this model we derived age and species dependent $DBH$ averages across a temperature measure from predicted values in 5-year age groups starting at 30, 45, 60, 75, 90.
+With these models, we derived age and species dependent $DBH$ averages across a temperature measure from predicted values in 5-year age groups starting at 30, 45, 60, 75, 90
+
+Model selection was based on explained deviance (similar to explained variance) and the Aikake's information criteria (AIC).
+The best model, considered having the highest explanatory power (highest explained deviance) and simplest structure (lowest AIC), was chosen used for final analyses.
+Explained deviance assesses model fit in generalized linear modelling similar to explained variance in ordinary least squares modelling, but relies on a generalized form of residual, i.e., deviance residuals [see @wood2017].
+Deviance for the Gamma distribution is calculated as
+
+\begin{equation}
+D(y, \hat{\mu}) = 2 \cdot \left\{ \frac{y−\hat{\mu}}{\hat{\mu}}  − log(\frac{y}{\hat{\mu}}) \right\}
+
+(\#eq:gam-deviance)
+\end{equation}
+
+where $y$ is the observed and $\hat{\mu}$ the expected value.
 
 As the spatial extent and coverage varied between temperature and ancillary data, more complex models (and specifically those including planting bed area and LandSat temperatures) typically also had fewer total observations.
-While this prevented a full comparison with information-based model selection criteria, such as Aikake's (AIC), the appropriateness of models that differed only in their implementation of the temperature-age interaction (i.e., $f_j(temp_i)$ $vs.$ $f_j(temp_i, age_i)$) could be assessed.
+While this prevented a full comparison with information-based model selection criteria like AIC, the appropriateness of models that differed only in their implementation of the temperature-age interaction (i.e., $f_j(temp_i)$ $vs.$ $f_j(temp_i, age_i)$) could be assessed.
 For this reason, the suite of developed models presented above were limited to comparatively simple structures (i.e., few terms, interactions and restricted number of basis functions), reducing the potential for choosing over-fitted models without formal comparison.
@@
 
-$$AIC= −2log(L) \cdot 2k$$
+\begin{equation}
+AIC= −2log(L) \cdot 2k
+(\#eq:aic)
+\end{equation}
 
@@
 This allowed us to highlight the utility of the approach *per se* and for Berlin in particular.
-**Measures used to asses goodness-of-fit were residual deviance, RSME and MAE (add formuals)**
+Lastly, we also report the root mean squared error ($RSME = \sqrt{\frac{1}{n}  \sum_{i=1}^{n}{(y_i - \hat{y}_i)^2}}$ and mean absolute error ($MAE = \frac{1}{n}|y_i - \hat{y}_i|$ for the final model.
 

### 2022-05-29 analysis/paper/paper.Rmd
updating paper

@@
 <!-- @quigley2004 identified absolute growth potential decreased for species between rural and urban conspecifics, yet assessments were -->
-<!-- <!-- lacked spatially-explicit effect size estimates across the urban-rural space and  --> -->
+<!-- <!-- lacked spatially-explicit effect size estimates across the urban-rural space and  --> 
 <!-- limited to comparatively small sample sizes per group ($n_{total}~=~230$ divided in 15 species, 3 groups and 2 locations). -->
@@
 Berlin, as one of the greenest cities in Europe, provides an openly accessible tree inventory, with spatio-temporal environmental data sets relevant to tree growth. 
-It features nearly 650000 individuals covering 94 genera and some 600 species and/or cultivars, listing information on location, stem diameter (at breast height; $DBH$), and stem height, among other variables, for the majority of street and park trees.
+It features nearly 700000 individuals covering 94 genera and some 600 species and/or cultivars, listing information on location, stem diameter (at breast height; $DBH$), and stem height, among other variables, for the majority of street and park trees.
 For this study, our objective was to assess the impact of excess urban heat, i.e.
@@
 caption <- "Data description used for maps/visualizations and analyses.
-Resolution and radius are provided in $m$, the latter is the buffer in which data was averaged around each trees.
+Resolution and radius are in $m$, the latter is the buffer in which data was averaged around each tree.
 A zero-radius refers to a point extraction from categorical and location specific data.
@@
 Note that only street trees in urban, not rural areas or within green spaces, were considered here, but individual trees may grow along streets adjacent to green spaces and parks of varying sizes.
-Implausible observations, likely from erroneous data entry, were removed.
+Implausible observations, possibly from erroneous data entry, were removed.
 Additional manual data processing for quality control was done with a bespoke software  `datacleanr` by Hurley *et al*.

### 2022-06-06 analysis/paper/paper.Rmd
updating writing

@@
 Implausible observations, possibly from erroneous data entry, were removed.
-Additional manual data processing for quality control was done with a bespoke software  `datacleanr` by Hurley *et al*.
-(submitted), where obvious outliers or clearly interpolated data were removed; the latter was deemed necessary, as several observations in multiple city districts were derived by linear relationships (i.e., straight-line), which do not capture the ontogenic growth dynamics of trees, and leave no variation related to variables other than age.
+Additional manual data processing for quality control was done with a bespoke software  `datacleanr` by [@hurley2022], where obvious outliers or clearly interpolated data were removed.
+The latter was deemed necessary, as several observations in multiple city districts were derived by linear relationships (i.e., straight-line), which do not capture the ontogenic growth dynamics of trees, and leave no variation related to variables other than age.
 All of these operations were recorded, and can be viewed and reproduced via the supplementary code.
@@
 Temperature and UHI data were summarized temporally either by the provider or manually to provide a characteristic representation of heat loading during summer at different times (morning, afternoon/day, night), from which tree averages (radius of 150$~m$) were calculated (Figure$~$\@ref(fig:fig-temp-overview)).
-Two urban air temperature and one surface UHI data set were tested as explanatory variables in GAM models.
+Two urban air temperature and one surface UHI data set were tested as explanatory variables in generalized additive models (see Section \@ref(sec:stat-gam)).
 The air temperatures from the Berlin environmental atlas (EnvAt) are processed model outputs [FITNAH-3D; cf.
@@
 
-To contextualize tree growth patterns between age groups derived from Berlin's inventory data, we drew upon a recently established data set from Schneider *et al.* (in review), who sampled several common tree species across a rural-urban gradient.
+To contextualize tree growth patterns between age groups derived from Berlin's inventory data, we drew upon a recently established data set from [@schneider2021], who sampled several common tree species across a rural-urban gradient.
 For our purposes, we grouped trees sampled in parks, green spaces and along streets into a single urban category, and focused analyses on these.
@@
 
-We assessed trends in annual growth dynamics of urban trees across the 20$^{th}$ century for 1920-1960 and 1961-2001 [similar grouping as  @pretzsch2017 and @dahlhausen2018] with a hierarchical GAM implemented in `mgcv::gamm()` to leverage auto-correlation structures made available through the package `nlme` [@pinheiro2021].
+We assessed trends in annual growth dynamics of urban trees across the 20$^{th}$ century for 1920-1960 and 1961-2001 [similar grouping as  @pretzsch2017;@dahlhausen2018] with a hierarchical GAM implemented in `mgcv::gamm()` to leverage auto-correlation structures made available through the package `nlme` [@pinheiro2021].
 Annual growth was modeled as:
@@
 
-The diameter ($DBH$) of the ten most abundant species were modeled using GAMs as follows:
+We focused the analyses on the ten most abundant species after data quality control with a minimum of 10000 observations to ensure greatest possible spatial coverage and to increase confidence in estimates, totalling `r drake::readd(n_total_obs)` trees.
+The diameter ($DBH$) of these species was modeled using GAMs as follows:
 
@@
 This allowed us to highlight the utility of the approach *per se* and for Berlin in particular.
-Lastly, we also report the root mean squared error ($RSME = \sqrt{\frac{1}{n}  \sum_{i=1}^{n}{(y_i - \hat{y}_i)^2}}$ and mean absolute error ($MAE = \frac{1}{n}|y_i - \hat{y}_i|$ for the final model.
+Lastly, we also report the root mean squared error ($RSME = \sqrt{\frac{1}{n}  \sum_{i=1}^{n}{(y_i - \hat{y}_i)^2}}$) and mean absolute error ($MAE = \frac{1}{n}|y_i - \hat{y}_i|$) for the final model.
 
@@
 
-**Add table with ANOVA for interaction, non-interaction and MAE/RSME**
+**Add text for figure 3.3!**
 

### 2022-06-06 R/plan.R

@@ -1004,6 +1004,14 @@ plan <- drake::drake_plan(
 
     overview_table = make_overview_table(full_data_set_clean),
 
+   n_total_obs = nrow(model_df_stat_filtered),
+
+   n_total_species_obs = model_df_stat_filtered %>%
+       dplyr::group_by(species_corrected) %>%
+       dplyr::count(sort = TRUE) %>%
+       dplyr::rename('Species' = 1) %>%
+       mutate(Species = paste0('*',Species, '*')),
+
 
    # Generate table of genera age distribution
 

### 2022-06-06 analysis/paper/paper.Rmd
abs meth missing

@@
 We focused the analyses on the ten most abundant species after data quality control with a minimum of 10000 observations to ensure greatest possible spatial coverage and to increase confidence in estimates, totalling `r drake::readd(n_total_obs)` trees.
+
+```{r model-samples}
+
+drake::loadd(n_total_species_obs)
+
+caption <- "Ten Most abundant species and total sample size (n) for which GAM models were developed.
+Note that ancillary data availability across variables determined final sample sizes in individual models."
+caption_ov_short <- caption
+
+if (grepl("docx", out_format) ){
+  table_ds <- n_total_species_obs %>% 
+  knitr::kable(booktabs = TRUE,
+
+             caption = caption,
+             caption.short = caption,
+             format = "pipe",
+             escape = FALSE) 
+} else if (knitr::is_latex_output() ) {
+  table_ds <- n_total_species_obs %>% 
+  knitr::kable(booktabs = TRUE,
+
+             caption = caption,
+             caption.short = caption,
+             format = "latex",
+             # col.names = c("Subject/Relevance",
+             #              "Desc.",
+             #              "Type",
+             #              "Obsv. (n)",
+             #              "Resolution (m)",
+             #              "Source")
+             escape = FALSE)
+
+} else {
+table_ds <- n_total_species_obs %>% 
+  knitr::kable(booktabs = TRUE,
+               format = "html",
+
+             caption = caption,
+             caption.short = caption_ov_short,
+             # col.names = c("Subject/Relevance",
+             #              "Desc.",
+             #              "Type",
+             #              "Obsv. (n)",
+             #              "Resolution (m)",
+             #              "Source")
+             escape = FALSE) 
+}
+
+table_ds %>% 
+  kableExtra::kable_styling(full_width = TRUE, 
+                            font_size = 11,
+                            latex_options = c("hold_position")) %>% 
+  kableExtra::row_spec(0, bold = TRUE, font_size = 12)
+
+```
+
 The diameter ($DBH$) of these species was modeled using GAMs as follows:
@@
 
-Further, we implemented the interaction between temperature and age (i.e., $f_j(temp, age)$ as tensor smooths [@wood2006] to account for the different variable scales (i.e., units); all models were also tested without this interaction using a thin plate regression spline smooth for temperature (not shown in equations above).
+Further, we implemented the interaction between temperature and age (i.e., $f_j(temp, age)$) as tensor smooths [@wood2006] to account for the different variable scales (i.e., units); all models were also tested without this interaction using a thin plate regression spline smooth for temperature (not shown in equations above).
 The functions $f_j$ are for species-specific smooths [i.e., with individual smoothness penalties and functional shapes as detailed by @pedersen2019].
@@
 
-**Add text for figure 3.3!**
-
-(ref:model-performance) Predicted $vs.$ observed data for the best model according to explained deviance and 'local' AIC comparison, based on the LandSat-derived UHI measure [@chakraborty2019] for average summer conditions (2007) and LCZ6 [open to mid-rise,  @stewart2012].
-Hexagons and colors represent x-y bins (i.e., in the observed-predicted space) and their counts, and the red line is a least-squares fit.
+The final model, comprising a temperature-age interaction, the covariate $LCZ6_{\frac{}{150~m}}$ and LandSat derived UHI following Equation$~$\@ref(eq:gam-full), showed good agreement between predicted and observed $DBH$ for the majority of observations (Figure$~$\@ref(fig:fig-model-performance); $R^2~=~0.79$) resulting in a fairly low MAE of $6.03~cm$, and a RMSE of $7.7\cdot 10^{-7}$.
+While some prediction errors are large (spread in Figure$~$\@ref(fig:fig-model-performance)), the predictions are unbiased (see correspondence of 1:1 and least-squares fit in Figure$~$\@ref(fig:fig-model-performance)).
+Thus, we consider using species and age-group averages as appropriate for identifying and discussing relationships between $DBH$ and other covariates.
+
+(ref:model-performance) Predicted $vs.$ observed data for the best model according to explained deviance and AIC comparison, based on the LandSat-derived UHI measure [@chakraborty2019] for average summer conditions (2007) and LCZ6 [open to mid-rise,  @stewart2012].
+Hexagons and colors represent x-y bins (i.e., in the observed-predicted space) and their counts (N), and the red line is a least-squares fit.
 The model captures the mean response well, as indicated by lighters colors along dashed 1:1 line, and its close approximation by the least squares fit, as well as model metrics (see text label in plot).
-However, the generally large scatter around this mean response indicates that predictions on individual-level would be inappropriate.
 

### 2022-06-11 analysis/paper/paper.Rmd
abstract

@@
 abstract: |
+
+The manifold ecological and societal benefits urban trees provide depend on their health and performance. 
+Excess heat common for cities (i.e., Urban Heat Island; UHI) combined with other urban conditions, affects tree physiological processes with outcomes ranging from enhanced growth to branch die-back and even mortality.
+Thus, assessing the effect of increased temperatures on urban trees is necessary for understanding and adapting to current and expected climatic and socio-economic conditions – as well as their interactions - in this century.
+The growing conditions of urban trees vary substantially in space and time, jointly controlled by local climate and the urban fabric. 
+However, to date, studies assessing growth dynamics have primarily been based on remote sensing (typically with productivity as a growth proxy) focusing on community rather than individual level, or relied on labor-intensive methods that limit sample sizes and thus spatial (and temporal) coverage.
+Here, we present an approach to assess the growth potential of tree species in relation to the UHI in a large metropolis, Berlin (Germany), accounting for location-specific factors down to individual tree level for an extensive inventory data set comprising nearly 700000 trees.
+Out of a suite of generalized additive models based on different UHI or temperature data sets and ancillary environmental/urban data, we identified the best variable combinations to predict diameter at breast height while accounting for age, management effects, and the spatial dependence of observations (autocorrelation) for the ten most abundant street trees.
+
   **Populate..We analzed tree growth (potential) and its relationship with the Urban Heat Island (UHI) effect, as well as other characteristics of the urban fabric, for the ten most abundant street tree species/cultivars in Berlin using open data.
@@
 
-<!-- <!-- Species-specifc growth responses in relation across Berlin were assessed in relat --> -->
+<!-- <!-- Species-specifc growth responses in relation across Berlin were assessed in relat --> 
 <!-- We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of (excess) heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; LandSat-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section$~$\@ref(sec:stat-gam) for details).
@@
 Covariates were extracted at 150 and 300$~m$ to infer the impact of reference scale of the urban fabric on tree growth. 
-From all tested models the most suitable (i.e., parsimonious with highest explanatory) was employed for further analyses.
+From all tested models the most suitable (i.e., parsimonious with highest explanatory power) was employed for further analyses.
 
@@
 
-Further, we implemented the interaction between temperature and age (i.e., $f_j(temp, age)$) as tensor smooths [@wood2006] to account for the different variable scales (i.e., units); all models were also tested without this interaction using a thin plate regression spline smooth for temperature (not shown in equations above).
+Furthermore, we implemented the interaction between temperature and age (i.e., $f_j(temp, age)$) as tensor smooths [@wood2006] to account for the different variable scales (i.e., units); all models were also tested without this interaction using a thin plate regression spline smooth for temperature (not shown in equations above).
 The functions $f_j$ are for species-specific smooths [i.e., with individual smoothness penalties and functional shapes as detailed by @pedersen2019].
@@
 This allowed us to highlight the utility of the approach *per se* and for Berlin in particular.
-Lastly, we also report the root mean squared error ($RSME = \sqrt{\frac{1}{n}  \sum_{i=1}^{n}{(y_i - \hat{y}_i)^2}}$) and mean absolute error ($MAE = \frac{1}{n}|y_i - \hat{y}_i|$) for the final model.
+Lastly, we also report the root mean squared error ($RMSE = \sqrt{\frac{1}{n}  \sum_{i=1}^{n}{(y_i - \hat{y}_i)^2}}$) and mean absolute error ($MAE = \frac{1}{n}|y_i - \hat{y}_i|$) for the final model.
 
@@
 
-### Colophon
-
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
-
-```{r colophon, cache = FALSE}
-# which R packages and versions?
-if ("devtools" %in% installed.packages()) devtools::session_info()
-```
-
-The current Git commit details are:
-
-```{r}
-# what commit is this file at? 
-if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
-```
+<!-- ### Colophon -->
+
+<!-- This report was generated on `r Sys.time()` using the following computational environment and dependencies:  -->
+
+<!-- ```{r colophon, cache = FALSE} -->
+<!-- # which R packages and versions? -->
+<!-- if ("devtools" %in% installed.packages()) devtools::session_info() -->
+<!-- ``` -->
+
+<!-- The current Git commit details are: -->
+
+<!-- ```{r} -->
+<!-- # what commit is this file at?  -->
+<!-- if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())   -->
+<!-- ``` -->

### 2022-08-18 analysis/paper/paper.Rmd
abstract

@@
       toc: true
+      number_sections: false
     bookdown::word_document2:
+      number_sections: true
       fig_caption: yes
@@
       toc: yes
+      number_sections: true
 header-includes:
@@
 abstract: |
-
-The manifold ecological and societal benefits urban trees provide depend on their health and performance. 
-Excess heat common for cities (i.e., Urban Heat Island; UHI) combined with other urban conditions, affects tree physiological processes with outcomes ranging from enhanced growth to branch die-back and even mortality.
-Thus, assessing the effect of increased temperatures on urban trees is necessary for understanding and adapting to current and expected climatic and socio-economic conditions – as well as their interactions - in this century.
-The growing conditions of urban trees vary substantially in space and time, jointly controlled by local climate and the urban fabric. 
-However, to date, studies assessing growth dynamics have primarily been based on remote sensing (typically with productivity as a growth proxy) focusing on community rather than individual level, or relied on labor-intensive methods that limit sample sizes and thus spatial (and temporal) coverage.
-Here, we present an approach to assess the growth potential of tree species in relation to the UHI in a large metropolis, Berlin (Germany), accounting for location-specific factors down to individual tree level for an extensive inventory data set comprising nearly 700000 trees.
-Out of a suite of generalized additive models based on different UHI or temperature data sets and ancillary environmental/urban data, we identified the best variable combinations to predict diameter at breast height while accounting for age, management effects, and the spatial dependence of observations (autocorrelation) for the ten most abundant street trees.
-
-  **Populate..We analzed tree growth (potential) and its relationship with the Urban Heat Island (UHI) effect, as well as other characteristics of the urban fabric, for the ten most abundant street tree species/cultivars in Berlin using open data.
-We identify the most heat-sensitive species, give recommendations for future tree selections and provide guidelines to for local authorities for the prioitiziation of data acquisition (i.e., inventories) as well as data provision.**
+  The socio-ecological benefits that urban trees provide depend on their health and performance.
+Excess heat (i.e., Urban Heat Island; UHI) and other urban conditions affect tree physiology with outcomes from enhanced growth to mortality.
+Resilient urban forests in the face of climate change require species-specific understanding of growth responses.
+However, to date, studies assessing growth dynamics were primarily based on remote sensing of communities rather than individuals, or relied on labor-intensive methods that can limit the spatial coverage necessary to account for highly variable urban growing conditions.
+Here, we analyze growth dynamics of common urban tree species over time and across space for Berlin (Germany), a large metropolis, combining dendroecological (temporal) and inventory assessments (spatial).
+First, we show that annual increments increased across the 20th century for early (i.e., young) growth.
+Second, we use an approach relying on open inventory data to identify growth potential (diameter) in relation to excess heat across Berlin while accounting for age, potential management effects, and the urban fabric (e.g., planting area; building density, height; available soil nutrients) by applying generalized additive models for the ten most abundant species.
+Our analyses showed that younger trees may benefit from increased temperatures, while older individuals feature lower growth at greater UHI magnitudes.
+Furthermore, we show that planting area as well as building density modulate growth responses to temperature.
+Lastly, we discuss implications for the management of urban trees in the context of climate change mitigation, considering that younger trees are predominantly located at UHI “hot spots” and will undergo the observed age-dependent shift in temperature-growth sensitivity.
+By relying on increasingly available open data (inventories, urban/environmental data) our approach here is or will be readily transferable to other urban regions, allowing for species-specific growth assessments, while accounting for urban conditions at an individual level.
+
 keywords: |
@@
 <!-- The inclusion of independent, tree-level growth data, however, is paramount as it allows:   -->
-<!-- 1) validating the publicly-available Senate dataset ($DBH$) while providing a more reliable estimate of tree age,   -->
+<!-- 1) validating the publicly-available Senate data set ($DBH$) while providing a more reliable estimate of tree age,   -->
 <!-- 2) applying the model with annual basal area as a response (enabling incorporating effects of varying climate/UHI intensities over time), and   -->
@@
 
-(ref:temp-overview) Overview of temperature datasets used for assessing modelling impact on tree growth, with data sources in rows and averaged times in columns.
-Note the varying spatial coverage and resolution.
+(ref:temp-overview) Overview of data sets used for assessing the relationship between temperature and tree growth, with data sources in rows and averaged time intervals in columns.
+Note the varying spatial coverage and resolution as noted in Table \@ref(tab:tab-dsources).
 
@@
 
-caption <- "Overview of urban sampling locations and respective tree species coverage. Individual trees were sampled two or tree times to obtain a mean-tree ring width series."
+caption <- "Overview of urban sampling locations and species sample sizes (n). Individual trees were sampled two or tree times to obtain a mean-tree ring width series."
 caption_ov_short <- caption
@@
 These were compared to a sub-set of models without a spatial component (cf. Equation$~$\@ref(eq:gam-nocovariate), and Figure$~$\@ref(fig:fig-model-deviance) but not further discussed there).
-We also tested a suite of models without the spatial smooth for comparison
-**what's the impact of remaining auto correlation? inflated errors, larger coefficients?**
+We also tested a suite of models without the spatial smooth for comparison.
 
@@
 (ref:model-deviance) Overview of model fit, expressed as explained deviance, for all tested models.
-The Panels distinguish between models without (top) and with (bottom) spatial smooths to account for autocorrelation of residuals.
-Colored sub-sections define models without and with the applied (excess) heat measures.
-Symbols identify models with an interaction between temperature and age (diamond), and without (circle), while colors highlight included covariates and their reference radius (i.e.
-150/300$~m$, but also note x-axis labels), and size indicates the model's observations.
-Generally, including temperature (interactions) improve model fits above the null model (age only), and excluding planting bed area and LCZ6 cover, also above other non-temperature models.
-The LandSat-derived UHI measures provide the best fit on average, with the model including a temperature-age interaction and LCZ6 cover scoring highest.
+The Panels distinguish between models without (top) and with (bottom) spatial smooths to account for autocorrelated residuals.
+Colored areas reflect (inclusion of) temperature and (excess) heat measures.
+Symbols identify models with an interaction between temperature and age (diamond), and without (circle), while symbol colors highlight  covariates and their reference radius (i.e.
+150/300$~m$, but also note x-axis labels).
+Symbol size indicates the model's included observations (i.e., sample size).
+Generally, including temperature (interactions) improve model fits above the null model (age only) and other non-temperature models.
+However, note the importance ofplanting bed area and LCZ6 cover is apparent even without including temperature.
+The LandSat-derived UHI measures provide the best fit on average, with the model including a temperature-age interaction and LCZ6 cover scoring highest overall.
 
@@
 Hexagons and colors represent x-y bins (i.e., in the observed-predicted space) and their counts (N), and the red line is a least-squares fit.
-The model captures the mean response well, as indicated by lighters colors along dashed 1:1 line, and its close approximation by the least squares fit, as well as model metrics (see text label in plot).
+The model captures the mean response well, as indicated by lighter colors along the dashed 1:1 line, and its close approximation by the least squares fit, as well as model metrics (see text label in figure).
 

### 2022-08-18 analysis/paper/paper.Rmd
abstract done

@@
       fig_caption: yes
-      toc: true
+      toc: false
       number_sections: false
+      global_numbering: true
     bookdown::word_document2:
-      number_sections: true
+      number_sections: false
       fig_caption: yes
@@
       - --lua-filter=../templates/pagebreak.lua
+      global_numbering: true
     bookdown::pdf_document2:
@@
       toc: yes
-      number_sections: true
+      number_sections: false
+      global_numbering: true
 header-includes:
@@
 
-### Temperature/UHI data
+### Temperature/UHI data {#sec:tempdata}
 
 Temperature and UHI data were summarized temporally either by the provider or manually to provide a characteristic representation of heat loading during summer at different times (morning, afternoon/day, night), from which tree averages (radius of 150$~m$) were calculated (Figure$~$\@ref(fig:fig-temp-overview)).
-Two urban air temperature and one surface UHI data set were tested as explanatory variables in generalized additive models (see Section \@ref(sec:stat-gam)).
+Two urban air temperature and one surface UHI data set were tested as explanatory variables in generalized additive models (see Section [GAMs](#sec:stat-gam)).
 The air temperatures from the Berlin environmental atlas (EnvAt) are processed model outputs [FITNAH-3D; cf.
@@
 <!-- Species-specifc growth responses in relation across Berlin were assessed in relat -->
-We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of (excess) heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; LandSat-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section$~$\@ref(sec:stat-gam) for details).
+We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of (excess) heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; LandSat-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section [GAMs](#sec:stat-gam) for details).
 Covariates were extracted at 150 and 300$~m$ to infer the impact of reference scale of the urban fabric on tree growth. 
@@
 
-where $g()$ is a log-link for $DBH \sim Gamma$, and $i$, $j$ are indices for observations and species, respectively, and $t$ refers to an (excess) heat measure from UrbClim, Berlin EnvAt or LandSat at different times (morning, afternoon/day, night; cf.
-Section$~$\@ref(sec:dsources)); $c$ is a species-dependent intercept, while $tau$ represents a matrix of random effect coefficients for $k$ districts to account for differing management regimes across the city.
+where $g()$ is a log-link for $DBH \sim Gamma$, and $i$, $j$ are indices for observations and species, respectively, and $t$ refers to an (excess) heat measure from UrbClim, Berlin EnvAt or LandSat at different times (morning, afternoon/day, night; see Section [Temperature/UHI data](#sec:tempdata); $c$ is a species-dependent intercept, while $tau$ represents a matrix of random effect coefficients for $k$ districts to account for differing management regimes across the city.
 A global spatial smooth $f(x_i,y_i)$ (representing projected coordinates in UTM) via a Gaussian process [cf.
@@
 
-(ref:cambial-effect) Rates for annual incremental growth differ between recent and earlier times, as predicted by a hierarchical GAM (see Section$~$\@ref(sec:methods-dendro)).
+(ref:cambial-effect) Rates for annual incremental growth differ between recent and earlier times, as predicted by a hierarchical GAM (see Section [Dendrochronological analyses](#sec:methods-dendro).
 Thick lines and bands are for mean and 95$~\%$ confidence interval of all annual predictions across a time group (fine lines).
@@
 LCZ6 cover was derived as a weighted average within a radius of 150$~m$ around each tree.
-Note the identical shape, but difference in intercept as a result of the model specification with a global smooth the non-temperature covariate for all species (see Section$~$\@ref(sec:methods-gam-heat)); the UHI magnitude was fixed at $3^\circ C$ where most trees are present across age groups.
+Note the identical shape, but difference in intercept as a result of the model specification with a global smooth the non-temperature covariate for all species (see Section [Stem diameter model development and selection](#sec:methods-gam-heat); the UHI magnitude was fixed at $3^\circ C$ where most trees are present across age groups.
 The model identified a steady increase of average stem diameter with open to mid-rise urban land cover.
@@
 
+# Appendices
+Information too detailed to be included in the main text, for instance a list of areas sampled, may be presented in appendices which are to be placed before the References.
+Appendices should be as short and succinct as possible.
+
 <!-- <!-- The following line inserts a page break  --> 
-<!-- `r pgbreak()` -->
+`r pgbreak()`
 

### 2022-08-18 R/plan.R

@@ -1033,7 +1033,8 @@ plan <- drake::drake_plan(
         knitr_in("./analysis/paper/paper.Rmd"),
         output_dir = "./analysis/paper/",
         output_file = file_out("paper_knit.html"),
-        output_format = bookdown::html_document2(),
+        # output_format = bookdown::html_document2(),
+        output_format = "bookdown::html_document2",
         quiet = TRUE
     ),
     paper_word = rmarkdown::render(

### 2022-08-18 analysis/paper/paper.Rmd
moving captions to end

@@
 - \usepackage[normalem]{ulem}
+
 bibliography: 
@@
 
-(ref:study-area) Berlin's generalized land-use derived from @suvkberlin2019 and location within the European context (inset).
-
-```{r fig-study-area, fig.cap="(ref:study-area)",fig.width = 12, fig.height = 14, out.width="50%"}
-
-knitr::include_graphics(here::here("analysis/figures/map_00_studyarea.png"))
-```
+**Fig. \@ref(fig:fig-study_area)**
 
@@
 
-```{r dsources-prep, include=FALSE}
-
-options(knitr.kable.NA = "")
-
-dsources_path <- "analysis/data/tables/data_sources.xlsx"
-dsources <- readxl::read_excel(dsources_path, na = "NA")
-dsources$Accessed <- format(dsources$Accessed, "%b '%y")
-
-```
-
-```{r tab-dsources, eval = TRUE}
-
-caption <- "Data description used for maps/visualizations and analyses.
-Resolution and radius are in $m$, the latter is the buffer in which data was averaged around each tree.
-A zero-radius refers to a point extraction from categorical and location specific data.
-Polygons with radius data were rasterized to a resolution of $5~m$"
-caption_ov_short <- "Data descriptions"
-
-dsource <- dsources[-which(dsources$File == 'Wintertime_gridded_UHI_data.zip'), ]
-
-dsources <- dsources %>%
-    dplyr::mutate(Category = forcats::fct_relevel(Category, 'trees', 'temperature', 'landuse', 'ancillary', 'gov')) %>% 
-    arrange(Category) %>%
-  dplyr::select(Name = `Name-eng`, Accessed, Type, Unit = `Unit-latex`, Resolution, Radius, Source = `Source-short`, Reference = TextRef) %>%
-  # dplyr::filter(Name != "UHI Berlin",
-  dplyr::filter(Name != "Park Trees",
-                Name != "Riparian Trees")
-
-if (grepl("docx", out_format)) {
-    table_ds <- dsources %>% 
+**Table Table$~$\@ref(tab:tab-dsources)**
+
+### Street trees
+
+Berlin’s open data provided tree inventories including species, age, location, and circumference which was transformed into diameter.
+Note that only street trees in urban, not rural areas or within green spaces, were considered here, but individual trees may grow along streets adjacent to green spaces and parks of varying sizes.
+Implausible observations, possibly from erroneous data entry, were removed.
+Additional manual data processing for quality control was done with a bespoke software  `datacleanr` by [@hurley2022], where obvious outliers or clearly interpolated data were removed.
+The latter was deemed necessary, as several observations in multiple city districts were derived by linear relationships (i.e., straight-line), which do not capture the ontogenic growth dynamics of trees, and leave no variation related to variables other than age.
+All of these operations were recorded, and can be viewed and reproduced via the supplementary code.
+Lastly, observations with unlikely diameter-age combinations were identified via the residuals of a generalized linear model between diameter and age with a Gamma log-link distribution: if individual residuals exceeded seven times the median absolute deviation of all residuals, they were removed.
+The median absolute deviation (MAD) is comparable to the inter-quartile range, yet more robust to outliers:
+
+\begin{equation}
+  MAD = median(\|X_i - median(X)\|) 
+  (\#eq:mad)
+\end{equation}
+
+This approach is considered conservative (see supplementary information), yet all analyses were carried forward with the unfiltered and filtered data - no considerable differences were found, thus subsequent sections are based on the filtered data.
+<!-- which were processed with the bespoke software datacleanr by Hurley et al., submitted), -->
+<!-- Characteristics of street trees and their respective environments were collated from the city's tree inventory, as well as various elements of the Berlin Environment Atlas, which are openly available via https://daten.berlin.de and curated by the city's government.
+-->
+<!-- The data used from the inventory includes species, location, age, and diameter at breast height (dbh). -->
+Table$~$\@ref(tab:tab-age) shows the binned distribution of genera across age classes.
+Final samples applied in models were smaller, following the availability of ancillary data for a given observation, and limited to a maximum age of 125 years to increase confidence in reported values, and ultimately model estimates.
+
+**Table \@ref(tab:tab-age)**
+
+### Temperature/UHI data {#sec:tempdata}
+
+Temperature and UHI data were summarized temporally either by the provider or manually to provide a characteristic representation of heat loading during summer at different times (morning, afternoon/day, night), from which tree averages (radius of 150$~m$) were calculated (Figure$~$\@ref(fig:fig-temp-overview)).
+Two urban air temperature and one surface UHI data set were tested as explanatory variables in generalized additive models (see Section [GAMs](#sec:stat-gam)).
+The air temperatures from the Berlin environmental atlas (EnvAt) are processed model outputs [FITNAH-3D; cf.
+@gross1994] that are representations of typical summer conditions at 0400, 1400 and 2200 hours; these data are provided at city block basis (spatial polygons), from which weighted averages were extracted after rasterizing ($5~m$ resolution).
+UrbClim air temperatures are hourly model outputs [100$~m$ resolution, @deridder2015] based on ERA5 re-analyses data (ECMWF) for which observations from the hottest month  available (June, 2011) were averaged to hours equivalent to Berlin Environmental Atlas (referred to as Berlin EnvAt) data by using a window of $\pm~1~$hour (i.e., 0300 to 0500, etc.).
+Subsequently, a land-use and land-cover mask (CORINE; European Union, Copernicus Land Monitoring Service 2018, European Environment Agency) was used to define urban and rural/forested areas.
+Using this mask was deemed reasonable as Berlin's built-up area has not changed markedly over the past 50 years, i.e., about 52 to 61$\%$ [@mohamed2017]. 
+The urban heat loading was then calculated as
+
+\begin{equation}
+ UHI_{x,y} = T_{Air_{2m}~x,y} - \overline{T_{Air_{2m}~Rural}},
+  (\#eq:uhi)
+\end{equation}
+
+where $T$ is temperature ($^\circ C$) $x$ and $y$ define an urban grid cell.
+The LandSat-derived surface UHI data set by @chakraborty2019 (referred to as LandSat) estimates its measure in a similar fashion and the reader is referred to the detailed description therein; note this data set provides day and night-time averaged UHI estimates at 500$~m$ resolution, which were extracted for the hottest summer in this record (2007).
+
+**Fig. \@ref(fig:fig-temp-overview)
+
+### Ancillary environmental data
+
+Following the general approach described above, four ancillary covariates next to a temperature measure were employed in models; these were chosen due to their availability at high spatial resolution and coverage, and/or because their influence on growth was previously identified in literature or their likely impact could be deduced using ecophysiological principles.
+We included planting bed area and the sum of exchangeable basic cation as a proxy for soil nutrient availability (point extractions), as well as the proportional coverage of local climate zone 6 (LCZ6; open mid-rise urban land cover, see @demuzere2019 and @stewart2012 for details) and adjacent building height (rasterized to $5~m$ resolution).
+The latter was chosen as an increase reflects a transition away from densely urbanized areas and had the highest coverage (i.e., within [0-1]) for the processed tree inventory.
+
+<!-- surface area of tree planting beds ("Baumscheibe"), as well as general characteristics of the urban environment like building height. -->
+<!-- It also includes spatially-explicit model outputs for local climate simulations representative of typical conditions, such as air temperature (2 m) on a 2015 summer's day at different hours (0400, 1400, 2200).
+-->
+<!-- These model outputs were preferred over *in situ* measurements as **more representative based on averaged conditions, physiologically more relevant than surface urban heat island products**. -->
+<!-- These outputs are available at city block or planning unit level, and were used as a proxy for the general temperature distribution influenced by urbanization across the investigated growth period across the previous 80 years (see section X for details).
+-->
+<!-- This was deemed reasonable as Berlin's built-up area has not changed markedly over the past 50 years, i.e., about 52 to 61$\%$ [@mohamed2017].  -->
+<!-- Further insight into the role of the urban environment was assessed by including local climate zone classes derived by WUDAPT. -->
+<!-- These describe the proportion of urban area with a specific cover, such as dense high rises, sealed surfaces, etc.  -->
+<!-- For this study, the LCZ 6, representing dense mid-rise built up cover was used as a proxy for the degree of urbanization, and is the most frequent LCZ for Berlin [@fenner2017] (found ref in . -->
+<!-- Use percent sealed surface instead? -->
+
+## General approach: space-for-time analyses
+
+<!-- Species-specifc growth responses in relation across Berlin were assessed in relat -->
+We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of (excess) heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; LandSat-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section [GAMs](#sec:stat-gam) for details).
+Covariates were extracted at 150 and 300$~m$ to infer the impact of reference scale of the urban fabric on tree growth. 
+From all tested models the most suitable (i.e., parsimonious with highest explanatory power) was employed for further analyses.
+
+## Dendrochronological sampling {#sec:methods-dendro}
+
+To contextualize tree growth patterns between age groups derived from Berlin's inventory data, we drew upon a recently established data set from [@schneider2021], who sampled several common tree species across a rural-urban gradient.
+For our purposes, we grouped trees sampled in parks, green spaces and along streets into a single urban category, and focused analyses on these.
+Two to three cores were extracted at breast height from each tree.
+These were then prepared using standard dendro-ecological methods (i.e., mounting, sanding, measuring), and cross-dated with TSAP-Win\textsuperscript{TM} and COFECHA [@holmes1986], producing mean tree series of incremental growth.
+Additionally, cambial age of each increment was established by counting years from the inner most ring at the pith ($a = 0$) outward; on tangentially bored cores, missing rings to the pith were estimated.
+
+**Table \@ref(tab:tab-dendro-sites)**
+
+## Statistical Analyses {#sec:stat-gam}
+
+### GAMs
+
+We applied hierarchical generalized additive models (GAM) to estimate the relationship of several covariates with stem diamater growth ($DBH$).
+GAMs, as an extension of generalized linear models [@wood2017], allow modeling response variables as parametric and non-parametric combinations of smoothed explanatory covariates, and can assume non-normal response distributions.
+These smooths are constructed by summation of base functions of varying complexity and form, analogous to scatterplot smoothing [@hastie2017], which provides a high degree of flexibility, ideal for fitting ecosystem dynamics which are rarely linear [@pedersen2019], or correctly represented with deterministic functional forms (e.g.
+quadratic equations).
+In general, a GAM can be written as:
+
+\begin{equation}
+E (Y)~=~g^{-1}\left( \beta_0 + \sum_{i = 1}^{n} f_i (x_i) \right),
+(\#eq:gam-general)
+\end{equation}
+
+and
+
+\begin{equation}
+y~=~E (Y) + \epsilon,
+(\#eq:gam-expectation)
+\end{equation}
+
+where $Y$ is taken from an appropriate distribution and corresponding link function $g$, $\beta_0$ is the intercept and $f_i$ represents a smooth function of a predictor [@pedersen2019], and $\epsilon \sim \mathcal{N}(0, \sigma ^2)$.
+Nested data structures (e.g., city districts) can be accounted for by introducing random effects, while spatial dependence between observations can be accounted for by constructing smoothing functions with, for instance, northings and eastings [cf.
+@wood2017].
+All models were implemented in `R` [@rcoreteam2021] using functions available in the package `mgcv` [@wood2017].
+
+### Dendrochronological analyses
+
+We assessed trends in annual growth dynamics of urban trees across the 20$^{th}$ century for 1920-1960 and 1961-2001 [similar grouping as  @pretzsch2017;@dahlhausen2018] with a hierarchical GAM implemented in `mgcv::gamm()` to leverage auto-correlation structures made available through the package `nlme` [@pinheiro2021].
+Annual growth was modeled as:
+
+\begin{equation}
+g(\Delta r_i) =  f(year_i) + f_j(cambial~age_i)  + c_{i,j} + \tau_{i,k} + e_i 
+(\#eq:gam-cambial)
+\end{equation}
+
+where $g()$ is a log-link for $\Delta r \sim Gamma$, $\Delta r$ is the annual radial increment for observation $i$. 
+A global temporal (by year) and time-dependent ($j$, $\leq 1960$ or $>1960$) trend in cambial age were implemented with thin plate regression splines (default smoothing function in `mgcv`); $c_j$ is a time-group dependent intercept, while $tau$ represents a matrix of random effect coefficients for $k$ species to account for differences in growth patterns, and $e_i = \phi e_{i−1} + \epsilon_i$.
+A $3^{rd}$-order autocorrelation-moving average (ARMA) correlation structure was applied (i.e., $\phi(3,1)$) to account for the dependency of $\Delta r$ across years for each tree, as is frequently the case for tree growth [e.g., see @Fritts1989]; the detailed implementation for this model is given in the supplemental material code.
+$\Delta r$ was then derived for a range of cambial ages, and averaged for both time groups, allowing a comparison of recent to earlier growth.
+We acknowledge that tree cores obtained at breast height do not represent absolute tree age.
+However, here they serve as a proxy for growth between young ($\gtrapprox 1960$) and older individuals to contextualize growth patterns inferred from the larger-scale tree inventory.
+
+### Stem diameter model development and selection {#sec:methods-gam-heat}
+
+We focused the analyses on the ten most abundant species after data quality control with a minimum of 10000 observations to ensure greatest possible spatial coverage and to increase confidence in estimates, totalling `r drake::readd(n_total_obs)` trees.
+
+```{r model-samples}
+
+drake::loadd(n_total_species_obs)
+
+caption <- "Ten Most abundant species and total sample size (n) for which GAM models were developed.
+Note that ancillary data availability across variables determined final sample sizes in individual models."
+caption_ov_short <- caption
+
+if (grepl("docx", out_format) ){
+  table_ds <- n_total_species_obs %>% 
   knitr::kable(booktabs = TRUE,
@@
              format = "pipe",
-                     # col.names = c("Subject/Relevance",
-                     #              "Desc.",
-                     #              "Type",
-                     #              "Obsv. (n)",
-                     #              "Resolution (m)",
-                     #              "Source")
-                     escape = FALSE) %>% 
-        kableExtra::landscape()
-} else if(knitr::is_latex_output() ){
-    table_ds <- dsources %>% 
+             escape = FALSE) 
+} else if (knitr::is_latex_output() ) {
+  table_ds <- n_total_species_obs %>% 
   knitr::kable(booktabs = TRUE,
@@
              #              "Source")
-                     escape = FALSE) %>% 
-        kableExtra::landscape()
+             escape = FALSE)
 
 } else {
-    table_ds <- dsources %>% 
+table_ds <- n_total_species_obs %>% 
   knitr::kable(booktabs = TRUE,
@@
 
-# dsources %>%
-#   
-#   knitr::kable(caption = caption, escape = FALSE) %>%
-#   kableExtra::kable_styling(full_width = TRUE,
-#                             bootstrap_options = c("striped", "hover", "condensed", "responsive"))
-
-# flextable::set_flextable_defaults(padding.left = 0)
-# 
-# dsources %>%
-#   dplyr::select(`Name-eng`, Accessed, Type, Source, TextRef) %>%
-#   flextable::flextable() %>%
-#   flextable::autofit()
-
-```
-
-### Street trees
-
-Berlin’s open data provided tree inventories including species, age, location, and circumference which was transformed into diameter.
-Note that only street trees in urban, not rural areas or within green spaces, were considered here, but individual trees may grow along streets adjacent to green spaces and parks of varying sizes.
-Implausible observations, possibly from erroneous data entry, were removed.
-Additional manual data processing for quality control was done with a bespoke software  `datacleanr` by [@hurley2022], where obvious outliers or clearly interpolated data were removed.
-The latter was deemed necessary, as several observations in multiple city districts were derived by linear relationships (i.e., straight-line), which do not capture the ontogenic growth dynamics of trees, and leave no variation related to variables other than age.
-All of these operations were recorded, and can be viewed and reproduced via the supplementary code.
-Lastly, observations with unlikely diameter-age combinations were identified via the residuals of a generalized linear model between diameter and age with a Gamma log-link distribution: if individual residuals exceeded seven times the median absolute deviation of all residuals, they were removed.
-The median absolute deviation (MAD) is comparable to the inter-quartile range, yet more robust to outliers:
-
-\begin{equation}
-  MAD = median(\|X_i - median(X)\|) 
-  (\#eq:mad)
-\end{equation}
-
-This approach is considered conservative (see supplementary information), yet all analyses were carried forward with the unfiltered and filtered data - no considerable differences were found, thus subsequent sections are based on the filtered data.
-<!-- which were processed with the bespoke software datacleanr by Hurley et al., submitted), -->
-<!-- Characteristics of street trees and their respective environments were collated from the city's tree inventory, as well as various elements of the Berlin Environment Atlas, which are openly available via https://daten.berlin.de and curated by the city's government.
--->
-<!-- The data used from the inventory includes species, location, age, and diameter at breast height (dbh). -->
-Table$~$\@ref(tab:tab-age) shows the binned distribution of genera across age classes.
-Final samples applied in models were smaller, following the availability of ancillary data for a given observation, and limited to a maximum age of 125 years to increase confidence in reported values, and ultimately model estimates.
-
-```{r tab-age}
-
-drake::loadd(age_tables)
-
-age_tables <- age_tables %>% 
-  dplyr::arrange(desc(n_total))
-colnames(age_tables)[c(1,7,8, 9)] <- c("Genera", "150+", "Total (n)", "Missing (n)")
-
-age_tables <- dplyr::bind_rows(age_tables[age_tables$Genera!="Other",],
-                        age_tables[age_tables$Genera=="Other",])
-
-add_row <- as.data.frame(matrix(c(NA, colSums(age_tables[,-1])), nrow = 1))
-add_row[1,1] <- "Marg. Totals"
-colnames(add_row) <- colnames(age_tables)
-
-age_tables <- dplyr::bind_rows(age_tables,
-                    add_row)
-
-caption_trees <- "Binned age-distribution for street tree genera in Berlin data set, and entries missing age information. "
-caption_trees_short <- "Binned age-distribution for genera in Berlin data set."
-
-if (grepl("docx", out_format)) {
-    table_dat_req_kable <- age_tables %>% 
-        knitr::kable(booktabs = TRUE,
-                     format = "pipe",
-                     caption = caption_trees,
-                     caption.short = caption_trees_short
-        ) %>% 
-        kableExtra::row_spec(13, hline_after = TRUE) 
-} else if(knitr::is_latex_output() ){
-    table_dat_req_kable <- age_tables %>% 
-        knitr::kable(booktabs = TRUE,
-                     format = "latex",
-                     caption = caption_trees,
-                     caption.short = caption_trees_short
-        ) %>% 
-        kableExtra::row_spec(13, hline_after = TRUE)
-
-} else {
-    table_dat_req_kable <- age_tables %>% 
-        knitr::kable(booktabs = TRUE,
-                     format = "html",
-                     caption = caption_trees,
-                     caption.short = caption_trees_short
-        ) %>% 
-        kableExtra::row_spec(13, extra_css = "border-bottom: 1px solid")
-
-    # %>% 
-#   kableExtra::kable_styling(full_width = TRUE, 
-#                             font_size = 10) %>% 
-#   kableExtra::column_spec(1, italic = TRUE) %>%
-#   kableExtra::row_spec(0, bold = TRUE) %>% 
-#     kableExtra::row_spec(13, hline_after = TRUE)
-
-}
-
-   table_dat_req_kable %>% 
-  kableExtra::kable_styling(full_width = TRUE, 
-                            font_size = 11) %>% 
-  kableExtra::column_spec(1, italic = TRUE) %>%  
-  kableExtra::row_spec(0, italic = FALSE, bold = TRUE, font_size = 12)
-
-```
-
-### Temperature/UHI data {#sec:tempdata}
-
-Temperature and UHI data were summarized temporally either by the provider or manually to provide a characteristic representation of heat loading during summer at different times (morning, afternoon/day, night), from which tree averages (radius of 150$~m$) were calculated (Figure$~$\@ref(fig:fig-temp-overview)).
-Two urban air temperature and one surface UHI data set were tested as explanatory variables in generalized additive models (see Section [GAMs](#sec:stat-gam)).
-The air temperatures from the Berlin environmental atlas (EnvAt) are processed model outputs [FITNAH-3D; cf.
-@gross1994] that are representations of typical summer conditions at 0400, 1400 and 2200 hours; these data are provided at city block basis (spatial polygons), from which weighted averages were extracted after rasterizing ($5~m$ resolution).
-UrbClim air temperatures are hourly model outputs [100$~m$ resolution, @deridder2015] based on ERA5 re-analyses data (ECMWF) for which observations from the hottest month  available (June, 2011) were averaged to hours equivalent to Berlin Environmental Atlas (referred to as Berlin EnvAt) data by using a window of $\pm~1~$hour (i.e., 0300 to 0500, etc.).
-Subsequently, a land-use and land-cover mask (CORINE; European Union, Copernicus Land Monitoring Service 2018, European Environment Agency) was used to define urban and rural/forested areas.
-Using this mask was deemed reasonable as Berlin's built-up area has not changed markedly over the past 50 years, i.e., about 52 to 61$\%$ [@mohamed2017]. 
-The urban heat loading was then calculated as
-
-\begin{equation}
- UHI_{x,y} = T_{Air_{2m}~x,y} - \overline{T_{Air_{2m}~Rural}},
-  (\#eq:uhi)
-\end{equation}
-
-where $T$ is temperature ($^\circ C$) $x$ and $y$ define an urban grid cell.
-The LandSat-derived surface UHI data set by @chakraborty2019 (referred to as LandSat) estimates its measure in a similar fashion and the reader is referred to the detailed description therein; note this data set provides day and night-time averaged UHI estimates at 500$~m$ resolution, which were extracted for the hottest summer in this record (2007).
-
-(ref:temp-overview) Overview of data sets used for assessing the relationship between temperature and tree growth, with data sources in rows and averaged time intervals in columns.
-Note the varying spatial coverage and resolution as noted in Table \@ref(tab:tab-dsources).
-
-```{r fig-temp-overview, fig.cap="(ref:temp-overview)",fig.width = 12, fig.height = 12, out.width="100%"}
-
-knitr::include_graphics(here::here("./analysis/figures/fig_temp_maps.png"))
-```
-
-### Ancillary environmental data
-
-Following the general approach described above, four ancillary covariates next to a temperature measure were employed in models; these were chosen due to their availability at high spatial resolution and coverage, and/or because their influence on growth was previously identified in literature or their likely impact could be deduced using ecophysiological principles.
-We included planting bed area and the sum of exchangeable basic cation as a proxy for soil nutrient availability (point extractions), as well as the proportional coverage of local climate zone 6 (LCZ6; open mid-rise urban land cover, see @demuzere2019 and @stewart2012 for details) and adjacent building height (rasterized to $5~m$ resolution).
-The latter was chosen as an increase reflects a transition away from densely urbanized areas and had the highest coverage (i.e., within [0-1]) for the processed tree inventory.
-
-<!-- surface area of tree planting beds ("Baumscheibe"), as well as general characteristics of the urban environment like building height. -->
-<!-- It also includes spatially-explicit model outputs for local climate simulations representative of typical conditions, such as air temperature (2 m) on a 2015 summer's day at different hours (0400, 1400, 2200).
--->
-<!-- These model outputs were preferred over *in situ* measurements as **more representative based on averaged conditions, physiologically more relevant than surface urban heat island products**. -->
-<!-- These outputs are available at city block or planning unit level, and were used as a proxy for the general temperature distribution influenced by urbanization across the investigated growth period across the previous 80 years (see section X for details).
--->
-<!-- This was deemed reasonable as Berlin's built-up area has not changed markedly over the past 50 years, i.e., about 52 to 61$\%$ [@mohamed2017].  -->
-<!-- Further insight into the role of the urban environment was assessed by including local climate zone classes derived by WUDAPT. -->
-<!-- These describe the proportion of urban area with a specific cover, such as dense high rises, sealed surfaces, etc.  -->
-<!-- For this study, the LCZ 6, representing dense mid-rise built up cover was used as a proxy for the degree of urbanization, and is the most frequent LCZ for Berlin [@fenner2017] (found ref in . -->
-<!-- Use percent sealed surface instead? -->
-
-## General approach: space-for-time analyses
-
-<!-- Species-specifc growth responses in relation across Berlin were assessed in relat -->
-We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of (excess) heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; LandSat-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section [GAMs](#sec:stat-gam) for details).
-Covariates were extracted at 150 and 300$~m$ to infer the impact of reference scale of the urban fabric on tree growth. 
-From all tested models the most suitable (i.e., parsimonious with highest explanatory power) was employed for further analyses.
-
-## Dendrochronological sampling {#sec:methods-dendro}
-
-To contextualize tree growth patterns between age groups derived from Berlin's inventory data, we drew upon a recently established data set from [@schneider2021], who sampled several common tree species across a rural-urban gradient.
-For our purposes, we grouped trees sampled in parks, green spaces and along streets into a single urban category, and focused analyses on these.
-Two to three cores were extracted at breast height from each tree.
-These were then prepared using standard dendro-ecological methods (i.e., mounting, sanding, measuring), and cross-dated with TSAP-Win\textsuperscript{TM} and COFECHA [@holmes1986], producing mean tree series of incremental growth.
-Additionally, cambial age of each increment was established by counting years from the inner most ring at the pith ($a = 0$) outward; on tangentially bored cores, missing rings to the pith were estimated.
-
-```{r tab-dendro-sites}
-
-drake::loadd(series_long)
-drake::loadd(species_lookup)
-
-sampling_site_overview <-     series_long %>% 
-        dplyr::filter(grepl("urban", x = site_type)) %>% 
-        dplyr::select(sample_location, species, tree_id) %>% 
-        dplyr::group_by(sample_location, species) %>%
-        dplyr::summarize(tree_id = unique(tree_id)) %>% 
-        # mutate(site_type = gsub("[-].*$", "", site_type)) %>% 
-        dplyr::tally() %>% 
-    dplyr::mutate(sample_location = stringr::str_replace(sample_location, "[-][a-zA-Zä]+$", "") %>% 
-               stringr::str_replace(pattern = "-StEi", ""))
-
-spec_matched <- species_lookup$Species[match(sampling_site_overview$species, species_lookup$Code)]
-
-sampling_site_overview$species_long <- spec_matched
-
-sampling_site_overview <- sampling_site_overview %>% 
-    dplyr::select(Location = sample_location, Species = species_long, n)
-
-caption <- "Overview of urban sampling locations and species sample sizes (n). Individual trees were sampled two or tree times to obtain a mean-tree ring width series."
-caption_ov_short <- caption
-
-if (grepl("docx", out_format) ){
-  table_ds <- sampling_site_overview %>% 
-  knitr::kable(booktabs = TRUE,
-
-             caption = caption,
-             caption.short = caption,
-             format = "pipe",
-             escape = FALSE) 
-} else if (knitr::is_latex_output() ) {
-  table_ds <- sampling_site_overview %>% 
-  knitr::kable(booktabs = TRUE,
-
-             caption = caption,
-             caption.short = caption,
-             format = "latex",
-             # col.names = c("Subject/Relevance",
-             #              "Desc.",
-             #              "Type",
-             #              "Obsv. (n)",
-             #              "Resolution (m)",
-             #              "Source")
-             escape = FALSE)
-
-} else {
-table_ds <- sampling_site_overview %>% 
-  knitr::kable(booktabs = TRUE,
-               format = "html",
-
-             caption = caption,
-             caption.short = caption_ov_short,
-             # col.names = c("Subject/Relevance",
-             #              "Desc.",
-             #              "Type",
-             #              "Obsv. (n)",
-             #              "Resolution (m)",
-             #              "Source")
-             escape = FALSE) 
-}
-
-table_ds %>% 
-  kableExtra::kable_styling(full_width = TRUE, 
-                            font_size = 11,
-                            latex_options = c("hold_position")) %>% 
-  kableExtra::row_spec(0, bold = TRUE, font_size = 12)
-
-```
-
-## Statistical Analyses {#sec:stat-gam}
-
-### GAMs
-
-We applied hierarchical generalized additive models (GAM) to estimate the relationship of several covariates with stem diamater growth ($DBH$).
-GAMs, as an extension of generalized linear models [@wood2017], allow modeling response variables as parametric and non-parametric combinations of smoothed explanatory covariates, and can assume non-normal response distributions.
-These smooths are constructed by summation of base functions of varying complexity and form, analogous to scatterplot smoothing [@hastie2017], which provides a high degree of flexibility, ideal for fitting ecosystem dynamics which are rarely linear [@pedersen2019], or correctly represented with deterministic functional forms (e.g.
-quadratic equations).
-In general, a GAM can be written as:
-
-\begin{equation}
-E (Y)~=~g^{-1}\left( \beta_0 + \sum_{i = 1}^{n} f_i (x_i) \right),
-(\#eq:gam-general)
-\end{equation}
-
-and
-
-\begin{equation}
-y~=~E (Y) + \epsilon,
-(\#eq:gam-expectation)
-\end{equation}
-
-where $Y$ is taken from an appropriate distribution and corresponding link function $g$, $\beta_0$ is the intercept and $f_i$ represents a smooth function of a predictor [@pedersen2019], and $\epsilon \sim \mathcal{N}(0, \sigma ^2)$.
-Nested data structures (e.g., city districts) can be accounted for by introducing random effects, while spatial dependence between observations can be accounted for by constructing smoothing functions with, for instance, northings and eastings [cf.
-@wood2017].
-All models were implemented in `R` [@rcoreteam2021] using functions available in the package `mgcv` [@wood2017].
-
-### Dendrochronological analyses
-
-We assessed trends in annual growth dynamics of urban trees across the 20$^{th}$ century for 1920-1960 and 1961-2001 [similar grouping as  @pretzsch2017;@dahlhausen2018] with a hierarchical GAM implemented in `mgcv::gamm()` to leverage auto-correlation structures made available through the package `nlme` [@pinheiro2021].
-Annual growth was modeled as:
-
-\begin{equation}
-g(\Delta r_i) =  f(year_i) + f_j(cambial~age_i)  + c_{i,j} + \tau_{i,k} + e_i 
-(\#eq:gam-cambial)
-\end{equation}
-
-where $g()$ is a log-link for $\Delta r \sim Gamma$, $\Delta r$ is the annual radial increment for observation $i$. 
-A global temporal (by year) and time-dependent ($j$, $\leq 1960$ or $>1960$) trend in cambial age were implemented with thin plate regression splines (default smoothing function in `mgcv`); $c_j$ is a time-group dependent intercept, while $tau$ represents a matrix of random effect coefficients for $k$ species to account for differences in growth patterns, and $e_i = \phi e_{i−1} + \epsilon_i$.
-A $3^{rd}$-order autocorrelation-moving average (ARMA) correlation structure was applied (i.e., $\phi(3,1)$) to account for the dependency of $\Delta r$ across years for each tree, as is frequently the case for tree growth [e.g., see @Fritts1989]; the detailed implementation for this model is given in the supplemental material code.
-$\Delta r$ was then derived for a range of cambial ages, and averaged for both time groups, allowing a comparison of recent to earlier growth.
-We acknowledge that tree cores obtained at breast height do not represent absolute tree age.
-However, here they serve as a proxy for growth between young ($\gtrapprox 1960$) and older individuals to contextualize growth patterns inferred from the larger-scale tree inventory.
-
-### Stem diameter model development and selection {#sec:methods-gam-heat}
-
-We focused the analyses on the ten most abundant species after data quality control with a minimum of 10000 observations to ensure greatest possible spatial coverage and to increase confidence in estimates, totalling `r drake::readd(n_total_obs)` trees.
-
-```{r model-samples}
-
-drake::loadd(n_total_species_obs)
-
-caption <- "Ten Most abundant species and total sample size (n) for which GAM models were developed.
-Note that ancillary data availability across variables determined final sample sizes in individual models."
-caption_ov_short <- caption
-
-if (grepl("docx", out_format) ){
-  table_ds <- n_total_species_obs %>% 
-  knitr::kable(booktabs = TRUE,
-
-             caption = caption,
-             caption.short = caption,
-             format = "pipe",
-             escape = FALSE) 
-} else if (knitr::is_latex_output() ) {
-  table_ds <- n_total_species_obs %>% 
-  knitr::kable(booktabs = TRUE,
-
-             caption = caption,
-             caption.short = caption,
-             format = "latex",
-             # col.names = c("Subject/Relevance",
-             #              "Desc.",
-             #              "Type",
-             #              "Obsv. (n)",
-             #              "Resolution (m)",
-             #              "Source")
-             escape = FALSE)
-
-} else {
-table_ds <- n_total_species_obs %>% 
-  knitr::kable(booktabs = TRUE,
-               format = "html",
-
-             caption = caption,
-             caption.short = caption_ov_short,
-             # col.names = c("Subject/Relevance",
-             #              "Desc.",
-             #              "Type",
-             #              "Obsv. (n)",
-             #              "Resolution (m)",
-             #              "Source")
-             escape = FALSE) 
-}
-
-table_ds %>% 
-  kableExtra::kable_styling(full_width = TRUE, 
-                            font_size = 11,
-                            latex_options = c("hold_position")) %>% 
-  kableExtra::row_spec(0, bold = TRUE, font_size = 12)
-
 ```
@@
 
-(ref:cambial-effect) Rates for annual incremental growth differ between recent and earlier times, as predicted by a hierarchical GAM (see Section [Dendrochronological analyses](#sec:methods-dendro).
-Thick lines and bands are for mean and 95$~\%$ confidence interval of all annual predictions across a time group (fine lines).
-On average, early stages of recent growth (up to approximately 22 years) exceed that of earlier periods discernibly.
-
-```{r fig-cambial-effect, fig.cap="(ref:cambial-effect)", out.width="50%", include=TRUE, eval=TRUE}
-
-knitr::include_graphics(here::here("analysis/figures/fig-biwi-growth.png"))
-
-```
+**Figure \@ref(fig:fig-cambial-effect)**
 
@@
 
-# Appendices
-Information too detailed to be included in the main text, for instance a list of areas sampled, may be presented in appendices which are to be placed before the References.
-Appendices should be as short and succinct as possible.
+# Appendix {-}
 
@@
 
-<!-- ### Colophon -->
-
-<!-- This report was generated on `r Sys.time()` using the following computational environment and dependencies:  -->
-
-<!-- ```{r colophon, cache = FALSE} -->
-<!-- # which R packages and versions? -->
-<!-- if ("devtools" %in% installed.packages()) devtools::session_info() -->
-<!-- ``` -->
-
-<!-- The current Git commit details are: -->
-
-<!-- ```{r} -->
-<!-- # what commit is this file at?  -->
-<!-- if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())   -->
-<!-- ``` -->
+# Figures
+
+## Legends
+
+**Figure \@ref(fig:fig-study-area)** Berlin's generalized land-use derived from @suvkberlin2019 and location within the European context (inset).
+
+**Figure \@ref(fig:fig-temp-overview)** Overview of data sets used for assessing the relationship between temperature and tree growth, with data sources in rows and averaged time intervals in columns.
+Note the varying spatial coverage and resolution as noted in Table \@ref(tab:tab-dsources).
+
+**Figure \@ref(fig:fig-cambial-effect)** Rates for annual incremental growth differ between recent and earlier times, as predicted by a hierarchical GAM (see Section [Dendrochronological analyses](#sec:methods-dendro).
+Thick lines and bands are for mean and 95$~\%$ confidence interval of all annual predictions across a time group (fine lines).
+On average, early stages of recent growth (up to approximately 22 years) exceed that of earlier periods discernibly.
+
+`r pgbreak()`
+
+## Visualization
+
+```{r fig-study-area, fig.cap = "PLACEHOLDER", fig.width = 12, fig.height = 14, out.width="50%"}
+
+knitr::include_graphics(here::here("analysis/figures/map_00_studyarea.png"))
+```
+
+```{r fig-temp-overview, fig.cap="PLACEHOLDER",fig.width = 12, fig.height = 12, out.width="100%"}
+
+knitr::include_graphics(here::here("./analysis/figures/fig_temp_maps.png"))
+```
+
+```{r fig-cambial-effect, fig.cap="PLACEHOLDER", out.width="50%", include=TRUE, eval=TRUE}
+
+knitr::include_graphics(here::here("analysis/figures/fig-biwi-growth.png"))
+
+```
+
+`r pgbreak()`
+
+# Tables
+
+```{r dsources-prep, include=FALSE}
+
+options(knitr.kable.NA = "")
+
+dsources_path <- "analysis/data/tables/data_sources.xlsx"
+dsources <- readxl::read_excel(dsources_path, na = "NA")
+dsources$Accessed <- format(dsources$Accessed, "%b '%y")
+
+```
+
+```{r tab-dsources, eval = TRUE}
+
+caption <- "Data description used for maps/visualizations and analyses.
+Resolution and radius are in $m$, the latter is the buffer in which data was averaged around each tree.
+A zero-radius refers to a point extraction from categorical and location specific data.
+Polygons with radius data were rasterized to a resolution of $5~m$"
+caption_ov_short <- "Data descriptions"
+
+dsource <- dsources[-which(dsources$File == 'Wintertime_gridded_UHI_data.zip'), ]
+
+dsources <- dsources %>%
+    dplyr::mutate(Category = forcats::fct_relevel(Category, 'trees', 'temperature', 'landuse', 'ancillary', 'gov')) %>% 
+    arrange(Category) %>%
+  dplyr::select(Name = `Name-eng`, Accessed, Type, Unit = `Unit-latex`, Resolution, Radius, Source = `Source-short`, Reference = TextRef) %>%
+  # dplyr::filter(Name != "UHI Berlin",
+  dplyr::filter(Name != "Park Trees",
+                Name != "Riparian Trees")
+
+if (grepl("docx", out_format)) {
+    table_ds <- dsources %>% 
+        knitr::kable(booktabs = TRUE,
+
+                     caption = caption,
+                     caption.short = caption,
+                     format = "pipe",
+                     # col.names = c("Subject/Relevance",
+                     #              "Desc.",
+                     #              "Type",
+                     #              "Obsv. (n)",
+                     #              "Resolution (m)",
+                     #              "Source")
+                     escape = FALSE) %>% 
+        kableExtra::landscape()
+} else if(knitr::is_latex_output() ){
+    table_ds <- dsources %>% 
+        knitr::kable(booktabs = TRUE,
+
+                     caption = caption,
+                     caption.short = caption,
+                     format = "latex",
+                     # col.names = c("Subject/Relevance",
+                     #              "Desc.",
+                     #              "Type",
+                     #              "Obsv. (n)",
+                     #              "Resolution (m)",
+                     #              "Source")
+                     escape = FALSE) %>% 
+        kableExtra::landscape()
+
+} else {
+    table_ds <- dsources %>% 
+        knitr::kable(booktabs = TRUE,
+                     format = "html",
+
+                     caption = caption,
+                     caption.short = caption_ov_short,
+                     # col.names = c("Subject/Relevance",
+                     #              "Desc.",
+                     #              "Type",
+                     #              "Obsv. (n)",
+                     #              "Resolution (m)",
+                     #              "Source")
+                     escape = FALSE) 
+}
+
+table_ds %>% 
+  kableExtra::kable_styling(full_width = TRUE, 
+                            font_size = 11,
+                            latex_options = c("hold_position")) %>% 
+  kableExtra::row_spec(0, bold = TRUE, font_size = 12)
+
+# dsources %>%
+#   
+#   knitr::kable(caption = caption, escape = FALSE) %>%
+#   kableExtra::kable_styling(full_width = TRUE,
+#                             bootstrap_options = c("striped", "hover", "condensed", "responsive"))
+
+# flextable::set_flextable_defaults(padding.left = 0)
+# 
+# dsources %>%
+#   dplyr::select(`Name-eng`, Accessed, Type, Source, TextRef) %>%
+#   flextable::flextable() %>%
+#   flextable::autofit()
+
+```
+
+```{r tab-age}
+
+drake::loadd(age_tables)
+
+age_tables <- age_tables %>% 
+  dplyr::arrange(desc(n_total))
+colnames(age_tables)[c(1,7,8, 9)] <- c("Genera", "150+", "Total (n)", "Missing (n)")
+
+age_tables <- dplyr::bind_rows(age_tables[age_tables$Genera!="Other",],
+                        age_tables[age_tables$Genera=="Other",])
+
+add_row <- as.data.frame(matrix(c(NA, colSums(age_tables[,-1])), nrow = 1))
+add_row[1,1] <- "Marg. Totals"
+colnames(add_row) <- colnames(age_tables)
+
+age_tables <- dplyr::bind_rows(age_tables,
+                    add_row)
+
+caption_trees <- "Binned age-distribution for street tree genera in Berlin data set, and entries missing age information. "
+caption_trees_short <- "Binned age-distribution for genera in Berlin data set."
+
+if (grepl("docx", out_format)) {
+    table_dat_req_kable <- age_tables %>% 
+        knitr::kable(booktabs = TRUE,
+                     format = "pipe",
+                     caption = caption_trees,
+                     caption.short = caption_trees_short
+        ) %>% 
+        kableExtra::row_spec(13, hline_after = TRUE) 
+} else if(knitr::is_latex_output() ){
+    table_dat_req_kable <- age_tables %>% 
+        knitr::kable(booktabs = TRUE,
+                     format = "latex",
+                     caption = caption_trees,
+                     caption.short = caption_trees_short
+        ) %>% 
+        kableExtra::row_spec(13, hline_after = TRUE)
+
+} else {
+    table_dat_req_kable <- age_tables %>% 
+        knitr::kable(booktabs = TRUE,
+                     format = "html",
+                     caption = caption_trees,
+                     caption.short = caption_trees_short
+        ) %>% 
+        kableExtra::row_spec(13, extra_css = "border-bottom: 1px solid")
+
+    # %>% 
+#   kableExtra::kable_styling(full_width = TRUE, 
+#                             font_size = 10) %>% 
+#   kableExtra::column_spec(1, italic = TRUE) %>%
+#   kableExtra::row_spec(0, bold = TRUE) %>% 
+#     kableExtra::row_spec(13, hline_after = TRUE)
+
+}
+
+   table_dat_req_kable %>% 
+  kableExtra::kable_styling(full_width = TRUE, 
+                            font_size = 11) %>% 
+  kableExtra::column_spec(1, italic = TRUE) %>%  
+  kableExtra::row_spec(0, italic = FALSE, bold = TRUE, font_size = 12)
+
+```
+
+```{r tab-dendro-sites}
+
+drake::loadd(series_long)
+drake::loadd(species_lookup)
+
+sampling_site_overview <-     series_long %>% 
+        dplyr::filter(grepl("urban", x = site_type)) %>% 
+        dplyr::select(sample_location, species, tree_id) %>% 
+        dplyr::group_by(sample_location, species) %>%
+        dplyr::summarize(tree_id = unique(tree_id)) %>% 
+        # mutate(site_type = gsub("[-].*$", "", site_type)) %>% 
+        dplyr::tally() %>% 
+    dplyr::mutate(sample_location = stringr::str_replace(sample_location, "[-][a-zA-Zä]+$", "") %>% 
+               stringr::str_replace(pattern = "-StEi", ""))
+
+spec_matched <- species_lookup$Species[match(sampling_site_overview$species, species_lookup$Code)]
+
+sampling_site_overview$species_long <- spec_matched
+
+sampling_site_overview <- sampling_site_overview %>% 
+    dplyr::select(Location = sample_location, Species = species_long, n)
+
+caption <- "Overview of urban sampling locations and species sample sizes (n). Individual trees were sampled two or tree times to obtain a mean-tree ring width series."
+caption_ov_short <- caption
+
+if (grepl("docx", out_format) ){
+  table_ds <- sampling_site_overview %>% 
+  knitr::kable(booktabs = TRUE,
+
+             caption = caption,
+             caption.short = caption,
+             format = "pipe",
+             escape = FALSE) 
+} else if (knitr::is_latex_output() ) {
+  table_ds <- sampling_site_overview %>% 
+  knitr::kable(booktabs = TRUE,
+
+             caption = caption,
+             caption.short = caption,
+             format = "latex",
+             # col.names = c("Subject/Relevance",
+             #              "Desc.",
+             #              "Type",
+             #              "Obsv. (n)",
+             #              "Resolution (m)",
+             #              "Source")
+             escape = FALSE)
+
+} else {
+table_ds <- sampling_site_overview %>% 
+  knitr::kable(booktabs = TRUE,
+               format = "html",
+
+             caption = caption,
+             caption.short = caption_ov_short,
+             # col.names = c("Subject/Relevance",
+             #              "Desc.",
+             #              "Type",
+             #              "Obsv. (n)",
+             #              "Resolution (m)",
+             #              "Source")
+             escape = FALSE) 
+}
+
+table_ds %>% 
+  kableExtra::kable_styling(full_width = TRUE, 
+                            font_size = 11,
+                            latex_options = c("hold_position")) %>% 
+  kableExtra::row_spec(0, bold = TRUE, font_size = 12)
+
+```

### 2022-08-18 R/plan.R

@@ -546,14 +546,18 @@ plan <- drake::drake_plan(
 
     # Plotting --------------------------------
 
+   journal_fig_width_single = 8.2,
+   journal_fig_width_double = 17,
+   journal_fig_width_mid = 8.2 * 1.5,
+
 
    ### map: study area ------
 
    plot_study_area_map = make_map_study_area(blu = berlin_lu,
                                              berlin_poly = berlin_polygons,
-                                             path_out = drake::file_out("./analysis/figures/map_00_studyarea.png"),
-                                             height = 5.5,
-                                             width = 7,
+                                             path_out = drake::file_out("./analysis/figures/map_00_studyarea.tiff"),
+                                             height = journal_fig_width_single * 0.7857143,
+                                             width = journal_fig_width_single,
                                              dpi = 300),
 
 
@@ -1037,14 +1041,14 @@ plan <- drake::drake_plan(
         output_format = "bookdown::html_document2",
         quiet = TRUE
     ),
-    paper_word = rmarkdown::render(
-        knitr_in("./analysis/paper/paper.Rmd"),
-        output_dir = "./analysis/paper/",
-        output_file = file_out("paper_knit.docx"),
-        # output_file = "./paper_knit.html",
-        output_format = bookdown::word_document2(),
-        quiet = TRUE
-    ),
+    # paper_word = rmarkdown::render(
+    #     knitr_in("./analysis/paper/paper.Rmd"),
+    #     output_dir = "./analysis/paper/",
+    #     output_file = file_out("paper_knit.docx"),
+    #     # output_file = "./paper_knit.html",
+    #     output_format = bookdown::word_document2(),
+    #     quiet = TRUE
+    # ),
     # paper_pdf = rmarkdown::render(
     #     knitr_in("./analysis/paper/paper.Rmd"),
     #     output_dir = "./analysis/paper/",

### 2022-08-18 analysis/paper/paper.Rmd
moving figures

@@
 
-(ref:model-deviance) Overview of model fit, expressed as explained deviance, for all tested models.
-The Panels distinguish between models without (top) and with (bottom) spatial smooths to account for autocorrelated residuals.
-Colored areas reflect (inclusion of) temperature and (excess) heat measures.
-Symbols identify models with an interaction between temperature and age (diamond), and without (circle), while symbol colors highlight  covariates and their reference radius (i.e.
-150/300$~m$, but also note x-axis labels).
-Symbol size indicates the model's included observations (i.e., sample size).
-Generally, including temperature (interactions) improve model fits above the null model (age only) and other non-temperature models.
-However, note the importance ofplanting bed area and LCZ6 cover is apparent even without including temperature.
-The LandSat-derived UHI measures provide the best fit on average, with the model including a temperature-age interaction and LCZ6 cover scoring highest overall.
-
-```{r fig-model-deviance, fig.cap="(ref:model-deviance)", out.width="100%", include=TRUE, eval=TRUE}
-
-knitr::include_graphics(here::here("analysis/figures/fig_model_deviance.png"))
-
-```
+**Figure \@ref(fig:fig-model-deviance)**
 
@@
 
-(ref:model-performance) Predicted $vs.$ observed data for the best model according to explained deviance and AIC comparison, based on the LandSat-derived UHI measure [@chakraborty2019] for average summer conditions (2007) and LCZ6 [open to mid-rise,  @stewart2012].
-Hexagons and colors represent x-y bins (i.e., in the observed-predicted space) and their counts (N), and the red line is a least-squares fit.
-The model captures the mean response well, as indicated by lighter colors along the dashed 1:1 line, and its close approximation by the least squares fit, as well as model metrics (see text label in figure).
-
-```{r fig-model-performance, fig.cap="(ref:model-performance)", out.width="50%", include=TRUE, eval=TRUE}
-
-knitr::include_graphics(here::here("analysis/figures/fig-gam-dbh_temp-day2007_lcz6_obs_pred.png"))
-
-```
+**Figure$~$\@ref(fig:fig-model-performance)**
 
@@
 
+**Figure \@ref(fig:fig-model-deviance)** Overview of model fit, expressed as explained deviance, for all tested models.
+The Panels distinguish between models without (top) and with (bottom) spatial smooths to account for autocorrelated residuals.
+Colored areas reflect (inclusion of) temperature and (excess) heat measures.
+Symbols identify models with an interaction between temperature and age (diamond), and without (circle), while symbol colors highlight  covariates and their reference radius (i.e.
+150/300$~m$, but also note x-axis labels).
+Symbol size indicates the model's included observations (i.e., sample size).
+Generally, including temperature (interactions) improve model fits above the null model (age only) and other non-temperature models.
+However, note the importance ofplanting bed area and LCZ6 cover is apparent even without including temperature.
+The LandSat-derived UHI measures provide the best fit on average, with the model including a temperature-age interaction and LCZ6 cover scoring highest overall.
+
+**Figure \@ref(fig:fig-model-performance)** Predicted $vs.$ observed data for the best model according to explained deviance and AIC comparison, based on the LandSat-derived UHI measure [@chakraborty2019] for average summer conditions (2007) and LCZ6 [open to mid-rise,  @stewart2012].
+Hexagons and colors represent x-y bins (i.e., in the observed-predicted space) and their counts (N), and the red line is a least-squares fit.
+The model captures the mean response well, as indicated by lighter colors along the dashed 1:1 line, and its close approximation by the least squares fit, as well as model metrics (see text label in figure).
+
 `r pgbreak()`
@@
 knitr::include_graphics(here::here("analysis/figures/fig-biwi-growth.png"))
+
+```
+
+```{r fig-model-deviance, fig.cap="PLACEHOLDER", out.width="100%", include=TRUE, eval=TRUE}
+
+knitr::include_graphics(here::here("analysis/figures/fig_model_deviance.png"))
+
+```
+
+```{r fig-model-performance, fig.cap="(ref:model-performance)", out.width="50%", include=TRUE, eval=TRUE}
+
+knitr::include_graphics(here::here("analysis/figures/fig-gam-dbh_temp-day2007_lcz6_obs_pred.png"))
 

### 2022-08-18 analysis/paper/paper.Rmd
moved figs

@@
 
-We focused the analyses on the ten most abundant species after data quality control with a minimum of 10000 observations to ensure greatest possible spatial coverage and to increase confidence in estimates, totalling `r drake::readd(n_total_obs)` trees.
-
-```{r model-samples}
-
-drake::loadd(n_total_species_obs)
-
-caption <- "Ten Most abundant species and total sample size (n) for which GAM models were developed.
-Note that ancillary data availability across variables determined final sample sizes in individual models."
-caption_ov_short <- caption
-
-if (grepl("docx", out_format) ){
-  table_ds <- n_total_species_obs %>% 
-  knitr::kable(booktabs = TRUE,
-
-             caption = caption,
-             caption.short = caption,
-             format = "pipe",
-             escape = FALSE) 
-} else if (knitr::is_latex_output() ) {
-  table_ds <- n_total_species_obs %>% 
-  knitr::kable(booktabs = TRUE,
-
-             caption = caption,
-             caption.short = caption,
-             format = "latex",
-             # col.names = c("Subject/Relevance",
-             #              "Desc.",
-             #              "Type",
-             #              "Obsv. (n)",
-             #              "Resolution (m)",
-             #              "Source")
-             escape = FALSE)
-
-} else {
-table_ds <- n_total_species_obs %>% 
-  knitr::kable(booktabs = TRUE,
-               format = "html",
-
-             caption = caption,
-             caption.short = caption_ov_short,
-             # col.names = c("Subject/Relevance",
-             #              "Desc.",
-             #              "Type",
-             #              "Obsv. (n)",
-             #              "Resolution (m)",
-             #              "Source")
-             escape = FALSE) 
-}
-
-table_ds %>% 
-  kableExtra::kable_styling(full_width = TRUE, 
-                            font_size = 11,
-                            latex_options = c("hold_position")) %>% 
-  kableExtra::row_spec(0, bold = TRUE, font_size = 12)
-
-```
+We focused the analyses on the ten most abundant species after data quality control with a minimum of 10000 observations to ensure greatest possible spatial coverage and to increase confidence in estimates, totaling `r drake::readd(n_total_obs)` trees, with Table$~$\@ref(tab:tab-model-samples) showing species-specific sample sizes.
 
@@
 
-Relationships derived via the GAM in Equation$~$\@ref(eq:gam-full) with LCZ6 as covariate show fairly consistent responses across age groups considering trajectories, yet they differ somewhat in shape (Figure$~$\@ref(fig:fig-heat-effect)).
-Generally, younger age groups (30-35) showed increased growth, or little to no impact (*Acer platanoides*, *Acer pseudoplatanus*), with stronger UHI magnitude.
-At older ages, this trajectory gradually shifted toward decreasing stem diameter at higher temperatures.
-Notably, *Platanus acerifolia* showed an earlier reversal of it at age group 45-50 as compared to other species. 
+Relationships derived via the GAM in Equation$~$\@ref(eq:gam-full) with LCZ6 as covariate show fairly consistent responses across age groups considering trajectories, yet they differ somewhat in shape (Figure$~$\@ref(fig:fig-heat-effect)) and considerably in absolute magnitude (i.e., slope in Fig.$\@ref(fig:fig-temp-sens)).
+Generally, younger age groups (30-35) showed increased growth (strongest for *A.
+hippocastanum*, *T.
+intermedia*, *T.
+intermedia* 'Pallida'), or little to no impact (*Acer platanoides*, *Acer pseudoplatanus*), with stronger UHI magnitude.
+At older ages, this trajectory gradually shifted toward decreasing stem diameter at higher temperatures, with *T.
+euchlora*, *A.
+pseudoplatanus* and *P.
+acerifolia* having the most negative relationships.
+Notably, *Platanus acerifolia* showed an earlier reversal of sensitivity at age group 45-50 as compared to other species.
 This may be due to the stronger upward curvature towards the estimates' lower temperature limit (see respective panel in Figure$~$\@ref(fig:fig-heat-effect)).
 Although the uncertainty around this edge is comparatively well-constrained, the low number of observations at the end of the continuum does call for caution for further interpretation.
+
 The distribution of individual street trees along the UHI continuum, evident in the density estimates in Figure$~$\@ref(fig:fig-heat-effect), show that the majority of species investigated here are subjected to intermediate to high UHI levels.
@@
 
-(ref:heat-effect) Age-group averaged relationships between the LandSat-derived UHI and stem diameter for the ten most abundant species from the best model (i.e., Equation$~$\@ref(eq:gam-full) with LCZ6 as covariate).
-Note that estimates were derived with $LCZ6_{\frac{}{150~m}} = 0.5$, and that y-axis scales differ between panel rows.
-The colored lines and bands are the averaged GAM fit, and gray lines are least-squares estimates fitted to these averages.
-Estimates were only generated within temperature ranges that a given species experienced.
-Density plots (scaled to match 100$~cm$ on the y-axis) show the distribution of individual trees along the urban heat continuum, aiding both with interpreting uncertainties (band width, "wiggliness") and assessing species' exposure to urban heat.
-A general shift from increased or stable growth with temperature at younger ages to a decrease for older trees is apparent.
-
-```{r fig-heat-effect, fig.cap="(ref:heat-effect)", out.width="100%", include=TRUE, eval=TRUE}
-
-knitr::include_graphics(here::here("analysis/figures/fig-gam-dbh_temp-day2007_lcz6.png"))
-
-```
-
-(ref:temp-sens) Temperature sensitivity of diameter growth centered on species-specific means for comparability and to aid interpretation of inter-specific differences.
-Lines and bands are from ordinary least squares fits on the GAM estimates in Fig.$~$\@ref(fig:fig-heat-effect) (A) and the corresponding coefficients (slopes, B).
-
-```{r fig-temp-sens, fig.cap="(ref:temp-sens)", out.width="100%", include=TRUE, eval=TRUE}
-
-knitr::include_graphics(here::here("analysis/figures/fig_model_day2007_lcz6_growthsens.png"))
-
-```
-
-The chosen model also highlighted that, on average, trees growing in less urbanized environments (i.e., more open to mid-rise urban land cover; $LCZ6_{\frac{}{150~m}}$) achieve somewhat greater diameters.
+**Figure$~$\@ref(fig:fig-heat-effect)**
+
+**Figure \@ref(fig:fig-temp-sens)**
+
+The chosen model also highlighted that, on average, trees growing in less urbanized environments (i.e., more open to mid-rise urban land cover; $LCZ6_{\frac{}{150~m}}$) achieve greater diameters (Fig.$~$\@ref(fig:fig-lcz6-effect)).
 Note, that planting bed area, similar to $LCZ6_{\frac{}{150~m}}$, was identified as a covariate increasing explained deviance, both in models with and without a temperature measure, and represents an opportunity for further investigation, especially following additional data collection.
 
-(ref:lcz6-effect) Age-group averaged relationships between the LCZ6 and stem diameter for two $Tilia$ genera from the best model (i.e., Equation$~$\@ref(eq:gam-full) with LandSat-derived UHI as excess heat measure).
-LCZ6 cover was derived as a weighted average within a radius of 150$~m$ around each tree.
-Note the identical shape, but difference in intercept as a result of the model specification with a global smooth the non-temperature covariate for all species (see Section [Stem diameter model development and selection](#sec:methods-gam-heat); the UHI magnitude was fixed at $3^\circ C$ where most trees are present across age groups.
-The model identified a steady increase of average stem diameter with open to mid-rise urban land cover.
-
-```{r fig-lcz6-effect, fig.cap="(ref:lcz6-effect)", out.width="50%", include=TRUE, eval=TRUE}
-
-knitr::include_graphics(here::here("analysis/figures/fig-gam-dbh_temp-day2007_xvar-lcz6_tilia.png"))
-
-```
+**Figure$~$\@ref(fig:fig-lcz6-effect)**
 
@@
 
+**Figure \@ref(fig:fig-heat-effect)** Age-group averaged relationships between the LandSat-derived UHI and stem diameter for the ten most abundant species from the best model (i.e., Equation$~$\@ref(eq:gam-full) with LCZ6 as covariate).
+Note that estimates were derived with $LCZ6_{\frac{}{150~m}} = 0.5$, and that y-axis scales differ between panel rows.
+The colored lines and bands are the averaged GAM fit, and gray lines are least-squares estimates fitted to these averages.
+Estimates were only generated within temperature ranges that a given species experienced.
+Density plots (scaled to match 100$~cm$ on the y-axis) show the distribution of individual trees along the urban heat continuum, aiding both with interpreting uncertainties (band width, "wiggliness") and assessing species' exposure to urban heat.
+A general shift from increased or stable growth with temperature at younger ages to a decrease for older trees is apparent.
+
+**Figure \@ref(fig:fig-temp-sens)** Temperature sensitivity of diameter growth centered on species-specific means for comparability and to aid interpretation of inter-specific differences.
+Lines and bands are from ordinary least squares fits on the GAM estimates in Fig.$~$\@ref(fig:fig-heat-effect) (A) and the corresponding coefficients (slopes, B).
+
+**Figure \@ref(fig:fig-lcz6-effect)** Age-group averaged relationships between the LCZ6 and stem diameter for two $Tilia$ genera from the best model (i.e., Equation$~$\@ref(eq:gam-full) with LandSat-derived UHI as excess heat measure).
+LCZ6 cover was derived as a weighted average within a radius of 150$~m$ around each tree.
+Note the identical shape, but difference in intercept as a result of the model specification with a global smooth the non-temperature covariate for all species (see Section [Stem diameter model development and selection](#sec:methods-gam-heat); the UHI magnitude was fixed at $3^\circ C$ where most trees are present across age groups.
+The model identified a steady increase of average stem diameter with open to mid-rise urban land cover.
+
 `r pgbreak()`
@@
 
-```{r fig-model-performance, fig.cap="(ref:model-performance)", out.width="50%", include=TRUE, eval=TRUE}
+```{r fig-model-performance, fig.cap="PLACEHOLDER", out.width="50%", include=TRUE, eval=TRUE}
 
 knitr::include_graphics(here::here("analysis/figures/fig-gam-dbh_temp-day2007_lcz6_obs_pred.png"))
+
+```
+
+```{r fig-heat-effect, fig.cap="PLACEHOLDER", out.width="100%", include=TRUE, eval=TRUE}
+
+knitr::include_graphics(here::here("analysis/figures/fig-gam-dbh_temp-day2007_lcz6.png"))
+
+```
+
+```{r fig-temp-sens, fig.cap="PLACEHOLDER", out.width="100%", include=TRUE, eval=TRUE}
+
+knitr::include_graphics(here::here("analysis/figures/fig_model_day2007_lcz6_growthsens.png"))
+
+```
+
+```{r fig-lcz6-effect, fig.cap="PLACEHOLDER", out.width="50%", include=TRUE, eval=TRUE}
+
+knitr::include_graphics(here::here("analysis/figures/fig-gam-dbh_temp-day2007_xvar-lcz6_tilia.png"))
 
@@
 ```
+
+```{r tab-model-samples}
+
+drake::loadd(n_total_species_obs)
+
+caption <- "Ten Most abundant species and total sample size (n) for which GAM models were developed.
+Note that ancillary data availability across variables determined final sample sizes in individual models."
+caption_ov_short <- caption
+
+if (grepl("docx", out_format) ){
+  table_ds <- n_total_species_obs %>% 
+  knitr::kable(booktabs = TRUE,
+
+             caption = caption,
+             caption.short = caption,
+             format = "pipe",
+             escape = FALSE) 
+} else if (knitr::is_latex_output() ) {
+  table_ds <- n_total_species_obs %>% 
+  knitr::kable(booktabs = TRUE,
+
+             caption = caption,
+             caption.short = caption,
+             format = "latex",
+             # col.names = c("Subject/Relevance",
+             #              "Desc.",
+             #              "Type",
+             #              "Obsv. (n)",
+             #              "Resolution (m)",
+             #              "Source")
+             escape = FALSE)
+
+} else {
+table_ds <- n_total_species_obs %>% 
+  knitr::kable(booktabs = TRUE,
+               format = "html",
+
+             caption = caption,
+             caption.short = caption_ov_short,
+             # col.names = c("Subject/Relevance",
+             #              "Desc.",
+             #              "Type",
+             #              "Obsv. (n)",
+             #              "Resolution (m)",
+             #              "Source")
+             escape = FALSE) 
+}
+
+table_ds %>% 
+  kableExtra::kable_styling(full_width = TRUE, 
+                            font_size = 11,
+                            latex_options = c("hold_position")) %>% 
+  kableExtra::row_spec(0, bold = TRUE, font_size = 12)
+
+```

### 2022-08-19 analysis/paper/paper.Rmd
resized figures

@@
 
-**Fig. \@ref(fig:fig-study_area)**
+**Fig. \@ref(fig:fig-study-area) **
 
@@
 
-**Table Table$~$\@ref(tab:tab-dsources)**
+**Table$~$\@ref(tab:tab-dsources)**
 
@@
 
-**Fig. \@ref(fig:fig-temp-overview)
+**Fig. \@ref(fig:fig-temp-overview)**
 
@@
 
-Relationships derived via the GAM in Equation$~$\@ref(eq:gam-full) with LCZ6 as covariate show fairly consistent responses across age groups considering trajectories, yet they differ somewhat in shape (Figure$~$\@ref(fig:fig-heat-effect)) and considerably in absolute magnitude (i.e., slope in Fig.$\@ref(fig:fig-temp-sens)).
+Relationships derived via the GAM in Equation$~$\@ref(eq:gam-full) with LCZ6 as covariate show fairly consistent responses across age groups considering trajectories, yet they differ somewhat in shape (Figure$~$\@ref(fig:fig-heat-effect)) and considerably in absolute magnitude (i.e., slope in Fig.$~$\@ref(fig:fig-temp-sens)).
 Generally, younger age groups (30-35) showed increased growth (strongest for *A.
@@
 Furthermore, the data provided in Berlin's tree inventory required extensive quality controlling, and it is likely that inaccurate (diameter, age) or interpolated (diameter) observations were including in the analyses.
-However, the minimum requirement of 10000 observations per species, and the focus on average diameter responses across age groups decreased potential impacts of such observations may have had to a degree that was considered satisfactory, while uncertainty is communicated effectively (i.e., larger confidence bands for *T.
+However, the minimum requirement of 10000 observations per species, and the focus on average diameter responses across age groups decreased potential impacts of such observations to a degree that was considered satisfactory, while uncertainty is communicated effectively and transparently (i.e., larger confidence bands for *T.
 intermedia 'Pallida'*).

### 2022-08-19 R/functions.R

@@ -4178,7 +4178,7 @@ make_ranef_plot <- function(model_out,
 #' @export
 #'
 #' @examples
-make_map_study_area <- function(blu, berlin_poly, path_out, height, width, dpi){
+make_map_study_area <- function(blu, berlin_poly, path_out, height, width, dpi, base_size = 10){
 
 
 
@@ -4204,7 +4204,6 @@ make_map_study_area <- function(blu, berlin_poly, path_out, height, width, dpi){
 
 
 
-    base_size <- 16
 
 
 
@@ -4235,7 +4234,8 @@ make_map_study_area <- function(blu, berlin_poly, path_out, height, width, dpi){
                          # geom_sf(data = countries  %>%
                          #             dplyr::filter(admin != 'Russia'),
                          ggplot2::aes(fill = admin == 'Germany'),
-                         color = "gray20", show.legend = FALSE) +
+                         color = "gray20", show.legend = FALSE,
+                         size = 0.25) +
         ggplot2::geom_sf(data = berlin_loc, fill = "firebrick1", shape = 23, size = 2) +
         ggplot2::coord_sf(xlim = c(-10,45), ylim = c(34,71), crs = 4326) +
         # geom_sf(data = countries  %>%
@@ -4270,7 +4270,7 @@ make_map_study_area <- function(blu, berlin_poly, path_out, height, width, dpi){
         ggplot2::scale_color_manual(values = c("greenspace" = "seagreen4",
                                                "urbanized" = "beige",
                                                "water" = "steelblue1")) +
-        ggplot2::coord_sf(xlim = c(370000, 425000.2), ylim = c(5795000    , 5837259.4 )) +
+        ggplot2::coord_sf(xlim = c(370000, 425000.2), ylim = c(5791000    , 5837259.4 )) +
         ggplot2::scale_y_continuous(breaks = seq(52.35, 52.65, by = 0.1)) +
         ggplot2::theme_minimal(
             # base_family = "Roboto Condensed",
@@ -4292,7 +4292,8 @@ make_map_study_area <- function(blu, berlin_poly, path_out, height, width, dpi){
                     height = height,
                     width = width,
                     dpi = dpi,
-                    units = 'cm')
+                    units = 'cm',
+                    bg = 'white')
 
 
 
@@ -4512,7 +4513,7 @@ make_deviance_plot <- function(deviance_list,
                                 "bold(Berlin~EnvAt)"
                             ),
[…]

### 2022-08-19 R/plan.R

@@ -610,7 +610,7 @@ plan <- drake::drake_plan(
                           envat = berlin_heat_model_2015,
                           urbclim = raster::stack(uhi_urbclim),
                           berlin_poly = berlin_polygons,
-                          base_size = 18,
+                          base_size = 8,
                           file = "./analysis/figures/fig_temp_maps.png",
                           dpi = 300,
                           height = journal_fig_width_double,
@@ -684,7 +684,7 @@ plan <- drake::drake_plan(
 
    plot_gam_deviance = make_deviance_plot(
        deviance_list = gam_deviances,
-       base_size = 18,
+       base_size = 10,
        file = drake::file_out("./analysis/figures/fig_model_deviance.png"),
        height =  0.8333333 * journal_fig_width_double,
        width = journal_fig_width_double,
@@ -721,9 +721,9 @@ plan <- drake::drake_plan(
        age_filter = NULL,
        age_expression = age_expr,
        prediction_range = "within",
-       base_size = 18,
+       base_size = 12,
        file = drake::file_out("./analysis/figures/fig-gam-dbh_temp-day2007_lcz6.png"),
-       height = 0.6111111 * journal_fig_width_double,
+       height = 0.75 * journal_fig_width_double,
        width = journal_fig_width_double,
        dpi = 300),
 
@@ -829,7 +829,7 @@ plan <- drake::drake_plan(
                                  # species_filter = c("Tilia cordata","Platanus acerifolia"),
                                  age_expression = age_expr,
                                  prediction_range = "within",
-                                 base_size = 18,
+                                 base_size = 10,
                                  x_label = expression('Proportional Cover - LCZ6'[bar(150~m)]),
                                  file = drake::file_out("./analysis/figures/fig-gam-dbh_temp-day2007_xvar-lcz6_tilia.png"),
                                  height = 0.875 * journal_fig_width_single,
@@ -866,6 +866,7 @@ plan <- drake::drake_plan(
                                             file = drake::file_out("./analysis/figures/fig-gam-dbh_temp-day2007_lcz6_obs_pred.png"),
                                             height = 0.875 * journal_fig_width_single,
                                             width = journal_fig_width_single,
+                                            base_size = 10,
                                             dpi = 300),
 
 
@@ -881,7 +882,7 @@ plan <- drake::drake_plan(
                                                 # species_filter = c("Tilia cordata","Platanus acerifolia"),
                                                 age_expression = age_expr,
[…]

### 2022-09-29 analysis/paper/paper.Rmd
cont submission

@@
 abstract: |
-  The socio-ecological benefits that urban trees provide depend on their health and performance.
+  **Abstract** The socio-ecological benefits that urban trees provide depend on their health and performance.
 Excess heat (i.e., Urban Heat Island; UHI) and other urban conditions affect tree physiology with outcomes from enhanced growth to mortality.
@@
 
+`r pgbreak()`
+
 # Introduction {#sec:intro}
 
-The manifold ecological and societal benefits urban trees provide [e.g., @roy2012] depend critically on their health and performance. 
-For instance, trees alter local energy budgets [@grimmond1996 ; @hertel2019] through shading and transpiration [@gillner2015;@endlicher2016], and therefore can reduce ambient temperatures, infrastructure power-consumption and (human) thermal discomfort  [e.g.
+<!-- The manifold ecological and societal benefits urban trees provide [e.g., @roy2012] depend on their health and performance.  -->
+<!-- For instance, trees alter local energy budgets [@grimmond1996 ; @hertel2019] through shading and transpiration [@gillner2015;@endlicher2016], and therefore can reduce ambient temperatures, infrastructure power-consumption and (human) thermal discomfort  [e.g.
 @gulyas2006; @akbari2001; @mayer1987; @hoyano1988].
-However, excess heat common for cities [i.e., Urban Heat Island, UHI, @oke1982], combined with other urban conditions, affects tree physiological functioning with outcomes ranging from enhanced growth to early senescence, branch die-back, and even mortality [e.g., @hilbert2019;@au2018;@gillner2014].
+-->
+<!-- However,  -->
+Excess heat common for cities [i.e., Urban Heat Island, UHI, @oke1982], combined with other urban conditions, affects tree physiological functioning with outcomes ranging from enhanced growth to early senescence, branch die-back, and even mortality [e.g., @hilbert2019;@au2018;@gillner2014].
 Thus, assessing the effect of increased temperatures on trees, as part of urban green infrastructure, is instrumental for understanding as well as adapting to current and expected conditions in this century [@ward2007], especially considering ever more urbanized societies  and the potential for UHI effects to compound with more frequent atmospheric drought [@roloff2009;@brune2016;@norton2015].
@@
 
-Increased air temperatures due to UHIs can affect tree growth through altering several physiological processes across plant organs directly or indirectly [@dusenge2019].
+**Delete**? Increased air temperatures due to UHIs can affect tree growth through altering several physiological processes across plant organs directly or indirectly [@dusenge2019].
 Generally, reaction times at cellular level increase with temperature up to a maximum, after which a drop in enzymatic activity results in a species-dependent optimum curve  [@arcus2016; @parent2010].
@@
 
-Temperature and UHI data were summarized temporally either by the provider or manually to provide a characteristic representation of heat loading during summer at different times (morning, afternoon/day, night), from which tree averages (radius of 150$~m$) were calculated (Figure$~$\@ref(fig:fig-temp-overview)).
+Temperature and UHI data were summarized temporally either by the provider or manually to provide a characteristic representation of heat loading during summer at different times (morning, afternoon/day, night), from which tree averages (radius of 150$~m$) were calculated (Fig.$~$\@ref(fig:fig-temp-overview)).
 Two urban air temperature and one surface UHI data set were tested as explanatory variables in generalized additive models (see Section [GAMs](#sec:stat-gam)).
@@
 242 in @wood2017] was included to account for the spatial structure of observations, which reduced auto-correlation of model residuals considerably (see supplemental information).
-These were compared to a sub-set of models without a spatial component (cf. Equation$~$\@ref(eq:gam-nocovariate), and Figure$~$\@ref(fig:fig-model-deviance) but not further discussed there).
+These were compared to a sub-set of models without a spatial component (cf. Equation$~$\@ref(eq:gam-nocovariate), and Fig.$~$\@ref(fig:fig-model-deviance) but not further discussed there).
 We also tested a suite of models without the spatial smooth for comparison.
@@
 
-Recently established annual, incremental growth is on average greater as compared to earlier times (i.e., prior 1960; Figure$~$\@ref(fig:fig-cambial-effect)).
+Recently established annual, incremental growth is on average greater as compared to earlier times (i.e., prior 1960; Fig.$~$\@ref(fig:fig-cambial-effect)).
 The contrast is strongest in the first 30 years of cambial development, with clear indication that averages are statistically different up to approximately 22 years (cf.
@@
 This may be due to inaccurately estimated cambial ages, explaining the largely monotonic decrease for recent growth due to missing the typical (but not always present) initial rise and fall in pith-near stages.
-This would inadvertently create a left-shift in Figure$~$\@ref(fig:fig-cambial-effect) for recent growth.
+This would inadvertently create a left-shift in Fig.$~$\@ref(fig:fig-cambial-effect) for recent growth.
 Assuming that to be the case, the actual difference in average rates would be greater than presented here.
 
-**Figure \@ref(fig:fig-cambial-effect)**
+**Fig. \@ref(fig:fig-cambial-effect)**
 
@@
 The inclusion of temperature (additive structure) and temperature-age interactions improved the predictive skill, with interactions reaching higher explained deviance (i.e., skill).
-Models including the LandSat-derived UHI measures performed best on average, where models containing $LCZ6_{\frac{}{150~m}}$ or all covariates (saturated model) did best (Figure$~$\@ref(fig:fig-model-deviance)), which also holds true for other temperature/UHI measures.
+Models including the LandSat-derived UHI measures performed best on average, where models containing $LCZ6_{\frac{}{150~m}}$ or all covariates (saturated model) did best (Fig.$~$\@ref(fig:fig-model-deviance)), which also holds true for other temperature/UHI measures.
 Note that non-temperature models containing planting bed area and $LCZ6_{\frac{}{150~m}}$ alone outperformed several of those including other temperature measures and covariate combinations.
@@
 
-**Figure \@ref(fig:fig-model-deviance)**
-
-The final model, comprising a temperature-age interaction, the covariate $LCZ6_{\frac{}{150~m}}$ and LandSat derived UHI following Equation$~$\@ref(eq:gam-full), showed good agreement between predicted and observed $DBH$ for the majority of observations (Figure$~$\@ref(fig:fig-model-performance); $R^2~=~0.79$) resulting in a fairly low MAE of $6.03~cm$, and a RMSE of $7.7\cdot 10^{-7}$.
-While some prediction errors are large (spread in Figure$~$\@ref(fig:fig-model-performance)), the predictions are unbiased (see correspondence of 1:1 and least-squares fit in Figure$~$\@ref(fig:fig-model-performance)).
+**Fig. \@ref(fig:fig-model-deviance)**
+
+The final model, comprising a temperature-age interaction, the covariate $LCZ6_{\frac{}{150~m}}$ and LandSat derived UHI following Equation$~$\@ref(eq:gam-full), showed good agreement between predicted and observed $DBH$ for the majority of observations (Fig.$~$\@ref(fig:fig-model-performance); $R^2~=~0.79$) resulting in a fairly low MAE of $6.03~cm$, and a RMSE of $7.7\cdot 10^{-7}$.
+While some prediction errors are large (spread in Fig.$~$\@ref(fig:fig-model-performance)), the predictions are unbiased (see correspondence of 1:1 and least-squares fit in Fig.$~$\@ref(fig:fig-model-performance)).
 Thus, we consider using species and age-group averages as appropriate for identifying and discussing relationships between $DBH$ and other covariates.
 
-**Figure$~$\@ref(fig:fig-model-performance)**
+**Fig.$~$\@ref(fig:fig-model-performance)**
 
@@
 
-Relationships derived via the GAM in Equation$~$\@ref(eq:gam-full) with LCZ6 as covariate show fairly consistent responses across age groups considering trajectories, yet they differ somewhat in shape (Figure$~$\@ref(fig:fig-heat-effect)) and considerably in absolute magnitude (i.e., slope in Fig.$~$\@ref(fig:fig-temp-sens)).
+Relationships derived via the GAM in Equation$~$\@ref(eq:gam-full) with LCZ6 as covariate show fairly consistent responses across age groups considering trajectories, yet they differ somewhat in shape (Fig.$~$\@ref(fig:fig-heat-effect)) and considerably in absolute magnitude (i.e., slope in Fig.$~$\@ref(fig:fig-temp-sens)).
 Generally, younger age groups (30-35) showed increased growth (strongest for *A.
@@
 Notably, *Platanus acerifolia* showed an earlier reversal of sensitivity at age group 45-50 as compared to other species.
-This may be due to the stronger upward curvature towards the estimates' lower temperature limit (see respective panel in Figure$~$\@ref(fig:fig-heat-effect)).
+This may be due to the stronger upward curvature towards the estimates' lower temperature limit (see respective panel in Fig.$~$\@ref(fig:fig-heat-effect)).
 Although the uncertainty around this edge is comparatively well-constrained, the low number of observations at the end of the continuum does call for caution for further interpretation.
 
-The distribution of individual street trees along the UHI continuum, evident in the density estimates in Figure$~$\@ref(fig:fig-heat-effect), show that the majority of species investigated here are subjected to intermediate to high UHI levels.
+The distribution of individual street trees along the UHI continuum, evident in the density estimates in Fig.$~$\@ref(fig:fig-heat-effect), show that the majority of species investigated here are subjected to intermediate to high UHI levels.
 Trees in younger age groups are typically further toward the upper end of this continuum.
@@
 
-**Figure$~$\@ref(fig:fig-heat-effect)**
-
-**Figure \@ref(fig:fig-temp-sens)**
+**Fig.$~$\@ref(fig:fig-heat-effect)**
+
+**Fig. \@ref(fig:fig-temp-sens)**
 
@@
 
-**Figure$~$\@ref(fig:fig-lcz6-effect)**
+**Fig.$~$\@ref(fig:fig-lcz6-effect)**
 
@@
 Our work extends this approach by identifying the heat space (i.e., UHI range) occupied by several key species within a city.
-This not only allows identifying species-specific and intra-specific (age-dependent) sensitivities to urban conditions, but can also inform how age cohorts may respond as temperatures at a given location shift along the present continuum in the future, while accounting for the impacts of additional urban conditions (e.g., management, disturbance, urban fabric).
+This not only allows identifying species-specific and intra-specific (age-dependent) sensitivities to urban conditions, but can also inform how age cohorts may respond as temperatures at a given location shift along the present continuum through time, while accounting for the impacts of additional urban conditions (e.g., management, disturbance, urban fabric).
 Notably, increased growth rates at higher temperatures, as reported here, also entail greater demographic turnover through mortality [@smith2019].
@@
 Thus, as the majority of individuals were situated at the upper end of the temperature continuum, high temperature sensitivity may entail additional management due to faster turn-over and potentially subsequent replacement by more appropriate species.
+In addition, our approaach can allow managers and planners to identify potential scale-limits of the urban fabric (e.g., $LCZ6_{\frac{}{150~m}}~vs.~LCZ6_{\frac{}{300~m}}$) on tree performance when altering surroundings or planting in a given area; efforts should be directed toward identifying the most pertinent environmental covariates and how their influence scales in space.
 <!-- - ultimately, this will be a trade-off between tolerance of urban conditions, management goals (e.g., ecosystem services) and longevity. -->
@@
 
-Urban tree databases may only represent a single point in time (e.g., see https://opentrees.org), despite repeated inventories at regular and/or irregular intervals within cities or their districts.
-However, storing and providing multi-temporal inventories, rather than over-writing existing records, would give planners and researchers a wealth of additional information on individual tree performance, which – even at low temporal coverage – would increase confidence in space-for-time substitutions, or make them superfluous.
-In this regard, it is recommended to enhance and standardize (meta) data collection and provisioning to facilitate data quality control, thereby increasing confidence in the data, and enabling synthesis studies across districts and cities [@ossola2020].
-Key information should include (1) location, absolute age and planting date, (2) management regimes (e.g., planting regime, crown pruning, irrigation), (3) data collection protocol and body (e.g., actual measurement $vs.$ estimate or interpolation), as well as (4) tree provenance to account for potential differences in ecotype and/or acclimation [@watkins2020].
+<!-- Urban tree databases may only represent a single point in time (e.g., see https://opentrees.org), despite repeated inventories at regular and/or irregular intervals within cities or their districts.
+-->
+<!-- However, storing and providing multi-temporal inventories, rather than over-writing existing records, would give planners and researchers a wealth of additional information on individual tree performance, which – even at low temporal coverage – would increase confidence in space-for-time substitutions, or make them superfluous.
+-->
+<!-- In this regard, it is recommended to enhance and standardize (meta) data collection and provisioning to facilitate data quality control, thereby increasing confidence in the data, and enabling synthesis studies across districts and cities [@ossola2020].
+-->
+<!-- Key information should include (1) location, absolute age and planting date, (2) management regimes (e.g., planting regime, crown pruning, irrigation), (3) data collection protocol and body (e.g., actual measurement $vs.$ estimate or interpolation), as well as (4) tree provenance to account for potential differences in ecotype and/or acclimation [@watkins2020].
+-->
 <!-- However, there are financial and logistic trade-offs for data collection of the above, with contrasting needs across stakeholder groups [e.g., academics and authorities, @ostberg2013] where data that supports meeting long-term objectives should be prioritized.
@@
 -->
-In addition, managers and planners also should be aware of potential scale-limit of the urban fabric (e.g., $LCZ6_{\frac{}{150~m}}~vs.~LCZ6_{\frac{}{300~m}}$) on tree performance when altering surroundings or planting in a given area; efforts should be directed toward identifying the most pertinent environmental covariates and how their influence scales in space.
 
 Finally, open data was key for this study.
-We expressedly urge authorities to implement FAIR (findable, accessible, inter-operable and reusable) data principles to enable work similar to ours, as well as the creation of larger-scale databases and/or execution of synoptic studies [e.g., @ossola2020].
+We expressedly urge authorities to implement FAIR (findable, accessible, inter-operable and reusable) data principles to enable work similar to ours, as well as the creation of larger-scale, multi-temporal databases and/or execution of synoptic studies [e.g., @ossola2020].
+This also includes the collection and porivison of standardized (meta) data for quality control and to improve statistical inference.
 
@@
 
-# Appendix {-}
+# Author Contributions
 
@@
 
-**Figure \@ref(fig:fig-study-area)** Berlin's generalized land-use derived from @suvkberlin2019 and location within the European context (inset).
-
-**Figure \@ref(fig:fig-temp-overview)** Overview of data sets used for assessing the relationship between temperature and tree growth, with data sources in rows and averaged time intervals in columns.
+**Fig. \@ref(fig:fig-study-area)** Berlin's generalized land-use derived from @suvkberlin2019 and location within the European context (inset).
+
+**Fig.
+\@ref(fig:fig-temp-overview)** Overview of data sets used for assessing the relationship between temperature and tree growth, with data sources in rows and averaged time intervals in columns.
 Note the varying spatial coverage and resolution as noted in Table \@ref(tab:tab-dsources).
 
-**Figure \@ref(fig:fig-cambial-effect)** Rates for annual incremental growth differ between recent and earlier times, as predicted by a hierarchical GAM (see Section [Dendrochronological analyses](#sec:methods-dendro).
+**Fig.
+\@ref(fig:fig-cambial-effect)** Rates for annual incremental growth differ between recent and earlier times, as predicted by a hierarchical GAM (see Section [Dendrochronological analyses](#sec:methods-dendro).
 Thick lines and bands are for mean and 95$~\%$ confidence interval of all annual predictions across a time group (fine lines).
@@
 
-**Figure \@ref(fig:fig-model-deviance)** Overview of model fit, expressed as explained deviance, for all tested models.
+**Fig.
+\@ref(fig:fig-model-deviance)** Overview of model fit, expressed as explained deviance, for all tested models.
 The Panels distinguish between models without (top) and with (bottom) spatial smooths to account for autocorrelated residuals.
@@
 
-**Figure \@ref(fig:fig-model-performance)** Predicted $vs.$ observed data for the best model according to explained deviance and AIC comparison, based on the LandSat-derived UHI measure [@chakraborty2019] for average summer conditions (2007) and LCZ6 [open to mid-rise,  @stewart2012].
+**Fig.
+\@ref(fig:fig-model-performance)** Predicted $vs.$ observed data for the best model according to explained deviance and AIC comparison, based on the LandSat-derived UHI measure [@chakraborty2019] for average summer conditions (2007) and LCZ6 [open to mid-rise,  @stewart2012].
 Hexagons and colors represent x-y bins (i.e., in the observed-predicted space) and their counts (N), and the red line is a least-squares fit.
@@
 
-**Figure \@ref(fig:fig-heat-effect)** Age-group averaged relationships between the LandSat-derived UHI and stem diameter for the ten most abundant species from the best model (i.e., Equation$~$\@ref(eq:gam-full) with LCZ6 as covariate).
+**Fig.
+\@ref(fig:fig-heat-effect)** Age-group averaged relationships between the LandSat-derived UHI and stem diameter for the ten most abundant species from the best model (i.e., Equation$~$\@ref(eq:gam-full) with LCZ6 as covariate).
 Note that estimates were derived with $LCZ6_{\frac{}{150~m}} = 0.5$, and that y-axis scales differ between panel rows.
@@
 
-**Figure \@ref(fig:fig-temp-sens)** Temperature sensitivity of diameter growth centered on species-specific means for comparability and to aid interpretation of inter-specific differences.
+**Fig.
+\@ref(fig:fig-temp-sens)** Temperature sensitivity of diameter growth centered on species-specific means for comparability and to aid interpretation of inter-specific differences.
 Lines and bands are from ordinary least squares fits on the GAM estimates in Fig.$~$\@ref(fig:fig-heat-effect) (A) and the corresponding coefficients (slopes, B).
 
-**Figure \@ref(fig:fig-lcz6-effect)** Age-group averaged relationships between the LCZ6 and stem diameter for two $Tilia$ genera from the best model (i.e., Equation$~$\@ref(eq:gam-full) with LandSat-derived UHI as excess heat measure).
+**Fig.
+\@ref(fig:fig-lcz6-effect)** Age-group averaged relationships between the LCZ6 and stem diameter for two $Tilia$ genera from the best model (i.e., Equation$~$\@ref(eq:gam-full) with LandSat-derived UHI as excess heat measure).
 LCZ6 cover was derived as a weighted average within a radius of 150$~m$ around each tree.

### 2022-09-29 R/functions.R

@@ -3842,7 +3842,7 @@ FFGGHH
         plot_layout(design = layout, heights = 1/3)
 
 
-    png(filename = file, width = width, height = height, res = dpi, units = "cm")
+    tiff(filename = file, width = width, height = height, res = dpi, units = "cm")
 
     gp <- grid::gpar(fontfamily = "Roboto Condensed", fontsize = base_size)
 
@@ -4655,7 +4655,8 @@ make_deviance_plot <- function(deviance_list,
                     dpi = dpi,
                     height = height,
                     width = width,
-                    units = 'cm')
+                    units = 'cm',
+                    bg = 'white')
 
     return(gplot)
 }
@@ -4942,7 +4943,8 @@ plot_dbh_temp_single_var <- function(pred_list,
                     dpi = dpi,
                     height = height,
                     width = width,
-                    units = 'cm')
+                    units = 'cm',
+                    bg = 'white')
 
 
 }
@@ -5218,7 +5220,8 @@ plot_dbh_temp_single_var_flex <- function(pred_list,
                     dpi = dpi,
                     height = height,
                     width = width,
-                    units = 'cm')
+                    units = 'cm',
+                    bg = 'white')
 
 
 }
@@ -5330,7 +5333,8 @@ plot_obs_predicted_model <- function(path_model,
                     dpi = dpi,
                     height = height,
                     width = width,
-                    units = 'cm')
+                    units = 'cm',
+                    bg = 'white')
 
     return(gplot)
 
@@ -5551,7 +5555,8 @@ plot_dbh_temp_comparison <- function(
            width = width,
            height = height,
            dpi = dpi,
-           units = "cm")
+           units = "cm",
+           bg = 'white')
 
     return(list(plot = gplot, est_plot = estimate_plot, lms = lms, df = plot_data, estimates = zero_val))
 
@@ -5750,7 +5755,8 @@ make_biwi_plot <- function(biwi_preds,
                     dpi = dpi,
                     height = height,
                     width = width,
-                    units = 'cm')
+                    units = 'cm',
+                    bg = 'white')
 
     return(file)
 

### 2022-09-29 R/plan.R

@@ -551,14 +551,159 @@ plan <- drake::drake_plan(
    journal_fig_width_mid = 8.2 * 1.5,
 
 
+   # UFUG Submission --------------------------
+
+   base_path_fig = "./analysis/figures/ufug_submission/",
+
    ### map: study area ------
 
    plot_study_area_map = make_map_study_area(blu = berlin_lu,
                                              berlin_poly = berlin_polygons,
-                                             path_out = drake::file_out("./analysis/figures/map_00_studyarea.tiff"),
+                                             path_out = paste0(base_path_fig, "01_single_map_studyarea.tiff"),
                                              height = journal_fig_width_single * 0.7857143,
                                              width = journal_fig_width_single,
                                              dpi = 300),
+
+
+   ### map: temp/UHI overview ----------------
+
+   plot_uhi_overview_maps = plot_temp_maps(landsat = uhi_stacks,
+                                           envat = berlin_heat_model_2015,
+                                           urbclim = raster::stack(uhi_urbclim),
+                                           berlin_poly = berlin_polygons,
+                                           base_size = 8,
+                                           file = paste0(base_path_fig, "02_double_map_temp.tiff"),
+                                           dpi = 300,
+                                           height = journal_fig_width_double,
+                                           width = journal_fig_width_double),
+
+
+
+   ### BIWI: cambial age vs. annual growth -------------------
+
+   species_lookup = read.csv(drake::file_in("analysis/data/tables/species_lookup.csv"), header = TRUE),
+
+
+
+   series_long = prep_rwl_data(path_meta_cores = drake::file_in("./analysis/data/raw_data/biwi/BIWi_INV_20190123.xlsx"),
+                               path_meta_trees = drake::file_in("./analysis/data/raw_data/biwi/BIWi_INV_20190123.xlsx"),
+                               path_meta_sites = drake::file_in("./analysis/data/raw_data/biwi/BIWi_INV_20190123.xlsx"),
+                               path_dir_fh = drake::file_in("./analysis/data/raw_data/biwi/rwl/")),
+
+   biwi_preds = biwi_mod_predict(series_long),
+
+   biwi_plot = make_biwi_plot(biwi_preds,
+                              file = paste0(base_path_fig, "03_single_biwi_annual_growth.tiff"),
+                              height = journal_fig_width_single,
+                              width = journal_fig_width_single,
[…]

### 2022-10-11 analysis/paper/author_statement_credit.Rmd
reading submission

---
title: "CRediT author statement"
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: true
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage[normalem]{ulem}
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style

---

Following https://www.elsevier.com/authors/policies-and-guidelines/credit-author-statement

Alexander Hurley: Conceptualization, Methodology, Data curation, Formal analysis, Visualization, Writing - Original Draft.
Ingo Heinreich: Resources, Investigation, Project administration, Funding acquisition, Writing - Review & Editing

### 2022-10-11 analysis/paper/paper.Rmd
reading submission

@@
 -->
-Several of the aforementioned studies applied space-for-time substitutions or time series comparisons to quantify temperature and excess heat on growth for a given region.
+Multiple of the aforementioned studies applied space-for-time substitutions or time series comparisons to quantify temperature and excess heat on growth for a given region.
 However, they typically compared trees grouped using qualitative or summary descriptors of sampling sites, disregarding the spatial and/or temporal variability in location-specific factors noted above.
@@
 
-This approach is considered conservative (see supplementary information), yet all analyses were carried forward with the unfiltered and filtered data - no considerable differences were found, thus subsequent sections are based on the filtered data.
+Similar cut-off approaches are frequently applied in allometric and trait databases [e.g., @kattge2020].
+We considered this approach (i.e., $MAD \cdot 7$) conservative, and necessary given the presence of clearly erroneous data.
+The approach and impact thereof are visualized in the supplementary information (Section Data quality control).
 <!-- which were processed with the bespoke software datacleanr by Hurley et al., submitted), -->
@@
 
-## Conclusion
+# Conclusion
 
@@
 AGH and IH were supported through the Helmholtz-Climate-Initiative (HI-CAM), funded by the Helmholtz Initiative and Networking Fund; the authors are responsible for the content of this publication.
-
-# Author Contributions
+We thank Prof Dr Christoph Schneider for providing dendrochronological data and pertinent discussion on its use.
 

### 2022-10-11 analysis/paper/supplemental_information.Rmd
adjusting eqs
reading submission

@@
 author:
-  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^a$, ...
+  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^a$
   - $^a$ *Geoforschungszentrum Potsdam, Section 4.3, Germany*
@@
       toc: true
-    bookdown::word_document2:
-      fig_caption: yes
-      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
-      pandoc_args:
-      - --lua-filter=../templates/scholarly-metadata.lua
-      - --lua-filter=../templates/author-info-blocks.lua
-      - --lua-filter=../templates/pagebreak.lua
-    bookdown::pdf_document2:
-      fig_caption: yes
-      toc: yes
 header-includes:
@@
 csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
-abstract: |
-  **This document outlines the rationale for an analysis of tree growth (potential) and its relationship with the Urban Heat Island (UHI) effect in Berlin using an extensive, publicly available data set.
-It introduces preliminary results and provides an outlook for up-coming and potential work.**
-keywords: |
-  Uferbaeume; Anlagenbaeume; Strassenbaeume; 
-highlights: |
-  We will see. 
+
 ---
@@
 
-# Overview
+# Code and reproducibility
 
-- Upload full code with DOI
-- datacleanr imgs
-- diagnostic lm imgs
-- spatial comparison (Moran's I)
+The code to reproduce the paper and its analyses, as well as this supplement, are found at [https://github.com/the-Hull/berlin.trees/](), with permanently archived versions at **DOI**.
+The project relies on the workflow management tool `drake`, where all analyses steps are maintained within the file `berlin.trees/R/plan.R`, with necessary functions in `berlin.trees/R/functions.R`.
+The code downloads most necessary files to execute the analyses.
+However, some files were too large for tracking with version control (i.e., `git` on GitHub); thus, should any issues arise with executing the code, please contact the corresponding author.
 
-<!-- # Acknowledgements -->
+Further information on `drake` is available at:  
+- William Michael Landau, (2018). The drake R package: a pipeline toolkit for
+  reproducibility and high-performance computing. **Journal of Open Source Software, 3(21)**,
+  550, https://doi.org/10.21105/joss.00550
 
-<!-- <!-- The following line inserts a page break  --> 
-<!-- `r pgbreak()` -->
+# Data quality control for street tree inventory
 
-# References 
-<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
-<div id="refs"></div>
+Manual data processing for quality control was done with a bespoke software  `datacleanr` by [@hurley2022], where obvious outliers or clearly interpolated data were removed.
+The latter was deemed necessary, as several observations in multiple city districts were derived by linear relationships (i.e., straight-line), which do not capture the ontogenic growth dynamics of trees, and leave no variation related to variables other than age.
+All of these operations were recorded, and can be viewed in the repository [https://github.com/the-Hull/berlin.trees/]() and reproduced via the supplementary code:
 
-`r pgbreak()`
+- exploratory and diagnostic plots for age and diameter relationships: `berlin.trees/analysis/figures/eda/`
+- `datacleanr` cleaning operations: `berlin.trees/analysis/data/raw_data/tree_splits`
 
-### Colophon
+```{r fig-diag-dcr, fig.cap = "Manual data inspection and cleaning with the R package `datacleanr`.
+This example for the genus *Tilia* in the district Spandau clearly shows linearly interpolated age-diameter observations.
+These were manually removed with utmost diligence to retain non-interpolated data.
+Note, that the supplementary code includes files to view and reproduce the results of manual cleaning in *analysis/data/raw_data/tree_splits/*.", fig.width = 14, fig.height = 14, out.width="100%"}
 
-This report was generated on `r Sys.time()` using the following computational environment and dependencies:
-
-```{r colophon, cache = FALSE}
-# which R packages and versions?
-if ("devtools" %in% installed.packages()) devtools::session_info()
+knitr::include_graphics(here::here("analysis/figures/diagnostic_00_datacleanr-Tilia-district_spandau.png"))
 ```
 
-The current Git commit details are:
+Lastly, observations with unlikely diameter-age combinations were identified via the residuals of a generalized linear model between diameter and age with a Gamma log-link distribution: if individual residuals exceeded seven times the median absolute deviation of all residuals, they were removed.
+The median absolute deviation (MAD) is comparable to the inter-quartile range, yet more robust to outliers:
 
-```{r}
-# what commit is this file at? 
-if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
+\begin{equation}
+  MAD = median(\|X_i - median(X)\|) 
+  (\#eq:mad)
+\end{equation}
+
+```{r fig-diag-lm, fig.cap = "Observations with unlikely diameter-age combinations ($dbh_{cm}~vs.~STANDALTER$) were identified via the residuals of a generalized linear model between diameter and age with a Gamma log-link distribution for every species included in the study.
+Where individual residuals exceeded seven times the median absolute deviation of all residuals, they were removed.
+The line is the fitted model, and colors represent retained or removed observations (see legend).
+See main text (Section Street Trees) for further explanation.", fig.width = 14, fig.height = 14, out.width="100%"}
+
+knitr::include_graphics(here::here("analysis/figures/diagnostic_01_model_data.png"))
 ```
+
+# Spatial auto-correlation
+
+Exploratory analyses of (initial) model resiudals showed spatial auto-correlation was present in the structure of the inventory data (identified by correlation of x and y coordinates with residuals).
+Note that this was expected given that trees planted on the same road (and/or neighborhood, district) may share several features, such as planting time, species, adjacent urban fabric. 
+Thus, we included model features (i.e., dependent variables) in the model formulation to account for spatial auto-correlation. 
+This was achieved with isometric smoothing splines with basis derived from both x and y co-ordinates (i.e., interaction terms in the generalized additive model). 
+We used Morans' I to calcualte spatial dependency with the function `Moran.I()` from the `R` package `ape`.
+
+$$
+I = \frac{n}{S_0} \cdot \frac{\sum_{i=1}^{n} \sum_{j=1}^{n} w_{ij}(y_{i} - \bar{y})(y_{j} - \bar{y})} { \sum_{i=1}^{n} (y_{i} - \bar{y})^2}
+$$
+where $y_{i}$ are individual observations, $w_{ij}$ are distance weights, $n$ are the totall number of observations, and $S_{0} = \sum_{i=1}^{n} \cdot \sum_{j=1}^{n} w_{ij}$.
+
+Details on `ape` can be found in:
+
+- *Paradis E. & Schliep K. 2019. ape 5.0: an environment for modern phylogenetics and evolutionary analyses in R. **Bioinformatics 35**: 526-528.*
+
+Due to computational limitations, we split the data set into several groups using a spatial grid from the Berlin authorities, based on $2 \cdot 2~km$ cells, found at:  
+- [https://fbinter.stadt-berlin.de/fb/atom//DOP/Blattschnitt2x2km.gif](),
+
+with further details at:  
+- [https://fbinter.stadt-berlin.de/fb/berlin/service_intern.jsp?id=a_luftbild2010_cir@senstadt&type=FEED]() (accessed at 2022-09-29),
+
+with the resulting splits below:
+
+```{r fig-grid, fig.cap = "Grid for splitting observations into discrete groups.
+Regions are based on $2 \\cdot 2~km$ grid cells and grouped based on Berlin's data acquisition scheme for aerial images", echo = FALSE, out.width='50%'}
+
+grid <- sf::st_read(here::here("analysis/data/raw_data/spatial_ancillary/grid_2x2.geojson"), quiet = TRUE)
+
+library(ggplot2)
+ggplot() +
+  geom_sf(data = grid, aes(fill = grid_region, color = grid_region)) +
+  scale_fill_brewer(palette = 'Set3') +
+  scale_color_brewer(palette = 'Set3') +
+  theme_minimal() +
+  labs(title = "Gridded Berlin regions (2x2 km)", color = "Grid Region", fill = "Grid Region")
+
+```
+
+We observed a considerable reduction of Moran's I (i.e., spatial auto-correlation) by including spatial terms in the model structure (i.e., x and y coordinates). 
+Given the large sample sizes available for this study we considered this approach (i.e., spatial terms) satisfactory to ensure the inference of model coefficients was not violated due to dependence of residuals.
+
+```{r fig-moran, echo = FALSE, fig.cap = "Comparison of Moran's I with and without spatial terms included in the model structure as dependent variable (bar color) between two model structures (additive on left, and interaction structures for temperature on the right).", out.width='100%'}
+
+knitr::include_graphics(here::here("analysis/figures/fig-gam-spatial_moran_comparison_blocks.png"))
+
+```

### 2022-10-11 R/plan.R

@@ -553,7 +553,7 @@ plan <- drake::drake_plan(
 
    # UFUG Submission --------------------------
 
-   base_path_fig = "./analysis/figures/ufug_submission/",
+   base_path_fig = "./analysis/figures/ufug_submission/hurley_heinrich_",
 
    ### map: study area ------
 
@@ -1094,22 +1094,22 @@ plan <- drake::drake_plan(
 
 
     # Reporting ------------------------------
-    paper_html = rmarkdown::render(
-        knitr_in("./analysis/paper/paper.Rmd"),
-        output_dir = "./analysis/paper/",
-        output_file = file_out("paper_knit.html"),
-        # output_format = bookdown::html_document2(),
-        output_format = "bookdown::html_document2",
-        quiet = TRUE
-    ),
-    paper_word = rmarkdown::render(
-        knitr_in("./analysis/paper/paper.Rmd"),
-        output_dir = "./analysis/paper/",
-        output_file = file_out("paper_knit.docx"),
-        # output_file = "./paper_knit.html",
-        output_format = "bookdown::word_document2",
-        quiet = TRUE
-    ),
+    # paper_html = rmarkdown::render(
+    #     knitr_in("./analysis/paper/paper.Rmd"),
+    #     output_dir = "./analysis/paper/",
+    #     output_file = file_out("paper_knit.html"),
+    #     # output_format = bookdown::html_document2(),
+    #     output_format = "bookdown::html_document2",
+    #     quiet = TRUE
+    # ),
+    # paper_word = rmarkdown::render(
+    #     knitr_in("./analysis/paper/paper.Rmd"),
+    #     output_dir = "./analysis/paper/",
+    #     output_file = file_out("paper_knit.docx"),
+    #     # output_file = "./paper_knit.html",
+    #     output_format = "bookdown::word_document2",
+    #     quiet = TRUE
+    # ),
     # paper_pdf = rmarkdown::render(
     #     knitr_in("./analysis/paper/paper.Rmd"),
     #     output_dir = "./analysis/paper/",
@@ -1123,9 +1123,19 @@ plan <- drake::drake_plan(
        knitr_in("./analysis/paper/supplemental_information.Rmd"),
        output_dir = "./analysis/paper/",
        output_file = file_out("supplemental_information.html"),
-       output_format = bookdown::html_document2(),
+       output_format = "bookdown::html_document2",
+       quiet = TRUE
+   ),
+
[…]

### 2022-10-11 R/temp_gam.R

@@ -1,1840 +1,0 @@
-# exploratory GAM analyses
-
-
-library(drake)
-library(sf)
-library(dplyr)
-library(mgcv)
-
-# load data ---------------------------------------------------------------
-
-loadd(full_data_set_clean_with_UHI_covariates)
-geo_recover <- readRDS("./analysis/data/raw_data/tree_splits/berlin_trees_GEOBACKUP.RDS")
-
-full_data_set_clean_with_UHI_covariates <- cbind(full_data_set_clean_with_UHI_covariates, berlin_id = geo_recover$berlin_id) %>%
-    tidyr::unite("old_id", KENNZEICH, GISID)
-
-
-
-split_files <- list.files(path = "./analysis/data/raw_data/tree_splits/",
-                          pattern = "cleaned.Rds$",
-                          full.names = TRUE)
-
-
-cleaned_data_join <- lapply(split_files,
-                            readRDS) %>%
-    dplyr::bind_rows() %>%
-    dplyr::ungroup() %>%
-    tidyr::unite("check_id", KENNZEICH, GISID) %>%
-    dplyr::filter(is.na(.annotation)) %>%
-    dplyr::select(berlin_id, check_id) %>%
-    dplyr::left_join(full_data_set_clean_with_UHI_covariates, by = "berlin_id") %>%
-    tidyr::drop_na(STANDALTER, dbh_cm)
-# %>% sf::st_set_geometry(.$geometry)
-
-
-# cleaned_data_join <- purrr::map2(
-#     cleaned_data,
-#     full_data_set_clean_with_UHI_covariates %>%
-#         select(ID, STANDORTNR, KENNZEICH, geometry, match_id) %>%
-#         split(., f = .$BEZIRK),
-#     ~left_join(.x,
-#                .y,
-#                by = c(".dcrkey" = "match_id"))) %>%
-#     dplyr::bind_rows()
-
-# make model df -----------------------------------------------------------
-
-mdf <- cleaned_data_join %>%
-    dplyr::select(provenance,
-                  ART_BOT,
-                  BEZIRK = BEZIRK,
-                  gattung_short,
-                  STANDALTER,
-                  dbh_cm,
-                  day_2007,
-                  T2M22HMEA,
-                  T2M14HMEA,
-                  T2M04HMEA,
-                  geometry,
-                  species_corrected,
-                  baumsch_flaeche_m2,
-                  soil_nutrients_swertstu,
[…]

### 2022-10-11 analysis/scripts/01_get_tree_data.R

@@ -1,185 +1,0 @@
-library(httr)
-library(xml2)
-library(dplyr)
-library(ckanr)
-library(ggplot2)
-
-
-
-# set base url
-ckanr_setup(url = "https://datenregister.berlin.de")
-
-# pull all packages from api, limit to 2500 (currentl ~ 2100 pckgs)
-p_list <- package_list(as = "table", url = "https://datenregister.berlin.de/", limit = 2500)
-
-
-# check for tree-related data sets
--
-
-# extract target urls for first packages (street trees)
-anlagen_id <- package_results[[1]]$resources[[2]]$id
-anlagen_url <- resource_show(id = anlagen_id)$url
-
-# request data via GET (REST API)
-
-
-json_path <- file.path(here::here(),
-                       "analysis",
-                       "data",
-                       "raw_data",
-                       paste0(package_results[[1]]$title,
-                              ".json"))
-
-
-# n <- 10
-query <- list(service = "WFS",
-              request = "GetFeature",
-              version = "2.0.0",
-              TypeNames = "s_wfs_baumbestand_an",
-              # count = n,
-              outputFormat = 'application/geo+json')
-
-
-# extract content
-
-httr::GET(anlagen_url, query = query, httr::write_disk(json_path, overwrite = TRUE))
-trees_sf <- sf::read_sf(json_path)
-
-## Alternative approach via loading data as geojson string directly into memory
-# result <- httr::GET(anlagen_url, query = query, httr::write_disk(json_path, overwrite = TRUE))
-# tree_data <- httr::content(result, type = 'text')
-# trees_sf <- sf::read_sf(tree_data)
-
-
-
-
-
-
-
-# get berlin district shapes
-berlin_poly <-  sf::read_sf("https://raw.githubusercontent.com/m-hoerz/berlin-shapes/master/berliner-bezirke.geojson")
-
-
-many_trees.plot <- trees_sf %>%
-
-    mutate(STAMMUMFG = as.numeric(STAMMUMFG),
-           GATTUNG = forcats::fct_infreq(as.factor(GATTUNG)) ,
-           gattung_short = forcats::fct_lump(GATTUNG,10) %>% forcats::fct_rev(),
-           bezirk_num = as.numeric(as.factor(BEZIRK))) %>%
-    filter(STAMMUMFG > 62.83185) %>%
-    ggplot() +
-    ggplot2::geom_sf(inherit.aes = FALSE,
-                     data = berlin_poly,
-                     fill = "transparent",
[…]

### 2022-10-11 analysis/scripts/02_get_tree_data_full.R

@@ -1,114 +1,0 @@
-
-
-
-# set base url
-ckanr_setup(url = "https://datenregister.berlin.de")
-
-# pull all packages from api, limit to 2500 (currentl ~ 2100 pckgs)
-p_list <- package_list(as = "table", url = "https://datenregister.berlin.de/", limit = 2500)
-
-
-# get tree-related data sets
-baum_index <- grep("baum", p_list)
-baum_packages <- p_list[baum_index]
-
-package_drop <- grep("baumscheibe", baum_packages)
-baum_packages <- baum_packages[-package_drop]
-
-# download meta-data for resources in each package
-package_results <- lapply(baum_packages, package_show)
-
-# extract only resources with WFS Service data
-wfs_subset <- purrr::map(package_results, "resources") %>%
-    purrr::map_depth(2, ~purrr::has_element(.x, "WFS Service"))
-
-wfs_baum_resources <- purrr::map(package_results, "resources") %>%
-    purrr::flatten() %>%
-   purrr::keep(.p = unlist(wfs_subset))
-
-
-# get names from pertinent packages,
-# could call API again, but potentially costly
-wfs_baum_packages <- baum_packages[unlist(
-    purrr::map(wfs_subset,
-               function(x) any(x == TRUE)))]
-
-
-# TEST IF COLLECTING RIGHT DATA
-# identical(wfs_baum_packages %>%
-#     purrr::map(package_show) %>%
-#     purrr::map("id"),
-#
-#     wfs_baum_resources %>%
-#         purrr::map("package_id") )
-# TEST TRUE
-
-
-
-# get urls for WFS service
-wfs_urls <- wfs_baum_resources %>%
-            purrr::map("id") %>%
-    purrr::map(resource_show) %>%
-    purrr::map_chr("url")
-
-
-# set up paths to write results to disk
-json_paths <- wfs_baum_packages %>%
-        purrr::map(package_show) %>%
-        purrr::map_chr("title") %>%
-    purrr::map_chr(function(x){
-
-        file.path(here::here(),
-                  "analysis",
-                  "data",
-                  "raw_data",
-                  paste0(x,
-                         ".json"))
-    })
-
-
-# get feature names for GET request via WFS urls
-feature_names <- purrr::map(wfs_urls, function(x){
-
-    capabilities <- httr::GET(x)
-    xml_capabilities <- content(capabilities, type = "text/xml")
-
-    feature_name <- xml_contents(
-        xml_child(
-            xml_child(
[…]

### 2022-10-11 analysis/scripts/03_dump.R

@@ -1,563 +1,0 @@
-library(drake)
-library(ggplot2)
-library(dplyr)
-loadd(full_data_set_clean)
-loadd(extract_uhi_values_to_list)
-
-
-
-#' Title
-#'
-#' @param full_df
-#' @param extract_uhi
-#'
-#' @return
-#' @export
-#'
-#' @examples
-make_test_data_set <- function(full_df = full_data_set_clean,
-                               extract_uhi = extract_uhi_values_to_list){
-
-
-    test_set <- cbind(as.data.frame(full_df),
-                      extract_uhi$Summertime_gridded_UHI_data$day)
-
-
-    test_set <- test_set %>%
-        # dplyr::filter(provenance == "s_wfs_baumbestand") %>%
-        dplyr::mutate(STANDALTER = as.numeric(STANDALTER),
-                      age_group = cut(STANDALTER, breaks = seq(0, 280, 40))) %>%
-        mutate(ART_BOT = ifelse(is.na(ART_BOT),
-                                paste(gattung_short, "spec."),
-                                ART_BOT))
-
-
-    return(test_set)
-}
-
-
-
-test_set <- berlin.trees::make_test_data_set(full_df = full_data_set_clean,
-                                             extract_uhi = extract_uhi_values_to_list)
-
-
-library(berlin.trees)
-
-
-#' Apply models (lme4)
-#'
-#' @param full_df cleaned data set
-#' @param extract_uhi extracted UHI data
-#' @param model_list list of models to apply
-#'
-#' @return a list containing lme4 model outputs
-#' @export
-#'
-#' @import furrr
-#' @import dplyr
-#' @import lme4
-#'
-apply_models <- function(test_df,
-         model_list = model_list,
-         n_top_species = 6){
-
-
-
-
-    apply_model_full <- function(.model, .data){
-        .model(.data)
-    }
-
-    top_species <- test_df %>%
-        group_by(gattung_short) %>%
-        count(ART_BOT) %>%
-        arrange(desc(n), .by_group = TRUE) %>%
-        top_n(n_top_species)
-
-
-    future::plan(future::multiprocess)
-    model_out <- furrr::future_map(model_list,
-                                   apply_model_full,
-                                   .data = test_df %>%
-                                       filter(STANDALTER < 350,
[…]

### 2022-10-11 analysis/scripts/03_tree_dbh_eda_cleaning.R

@@ -1,103 +1,0 @@
-library(ggplot2)
-library(drake)
-library(data.table)
-loadd(full_data_set_clean)
-
-full_data_set_clean <- full_data_set_clean %>%
-    mutate(STANDALTER = as.numeric(as.character(STANDALTER)))
-
-min_dbh <- 1
-max_dbh <- 400
-max_age <- 350
-min_age <- 15
-
-filter_expression <- expression(dbh_cm > min_dbh &
-                                    dbh_cm < max_dbh &
-                                    STANDALTER < max_age &
-                                    STANDALTER > min_age)
-
-top_spec <- full_data_set_clean %>%
-    as.data.frame() %>%
-    select(ART_BOT, gattung_short, dbh_cm, STANDALTER) %>%
-    filter(eval(filter_expression)) %>%
-    dplyr::count(ART_BOT) %>%
-
-    arrange(-n) %>%
-    filter(n > 1000) %>%
-    mutate(gattung_short = stringr::str_extract(ART_BOT, "^([\\w]+)"))
-
-
-bezirke <- unique(full_data_set_clean$BEZIRK)
-
-
-purrr::walk(bezirke,
-            function(bz){
-
-                g_plot <- full_data_set_clean %>%
-                    filter(eval(filter_expression),
-                           ART_BOT %in% top_spec$ART_BOT,
-                           BEZIRK == bz) %>%
-                    ggplot(aes(x = STANDALTER,
-                               y = dbh_cm,
-                               color = NAMENR)) +
-                    geom_point(alpha = 0.3) +
-                    geom_smooth(method = "lm") +
-
-                    # ggforce::facet_grid_paginate(~ART_BOT, ncol = 5, nrow = 5)
-                    facet_wrap(~ART_BOT, nrow = 10, ncol = 6, scales = "free")
-
-                # np <- ggforce::n_pages(g_plot)
-
-                ggsave(paste0("./analysis/figures/eda/00_eda_dbh_age_", bz, ".png"),
-                       g_plot,
-                       width = 100,
-                       height = 100,
-                       units = "cm",
-                       dpi = 300)
-
-
-
-            })
-
-g_plot <- full_data_set_clean %>%
-    filter(eval(filter_expression),
-           ART_BOT %in% top_spec$ART_BOT) %>%
-    ggplot(aes(x = STANDALTER,
[…]

### 2022-10-11 analysis/scripts/04_urban_tree_db_EDA.R

@@ -1,30 +1,0 @@
-# EDA for  Urban tree database -------------------------------------------
-
-# # for source
-# https://data.nal.usda.gov/dataset/urban-tree-database
-# # for overview
-# https://www.fs.usda.gov/rds/archive/catalog/RDS-2016-0005
-# # for meta
-# https://www.fs.usda.gov/rds/archive/products/RDS-2016-0005/_fileindex_RDS-2016-0005.html
-# # for download
-# https://www.fs.usda.gov/rds/archive/products/RDS-2016-0005/RDS-2016-0005.zip
-
-
-# download ----------------------------------------------------------------
-
-
-
-url <- "https://www.fs.usda.gov/rds/archive/products/RDS-2016-0005/RDS-2016-0005.zip"
-
-
-udb <- curl::curl_download(url,
-    destfile = tempfile(fileext = ".zip")
-)
-
-td <- tempdir()
-
-udb_unzip <- unzip(udb, exdir = td)
-
-
-udb_trees <- udb_unzip[grepl("Raw_tree_data", x = udb_unzip)]
-udb_data <- read.csv(udb_trees, header = TRUE, stringsAsFactors = FALSE)

### 2022-10-11 analysis/scripts/05_berlin_trees_GAMM_EDA.R

@@ -1,58 +1,0 @@
-
-# EDA for GAMM application ------------------------------------------------
-
-
-library(dplyr)
-library(drake)
-library(data.table)
-library(mgcv)
-
-loadd(model_df)
-# loadd(full_data_set_clean)
-
-proj <- "+proj=utm +zone=33 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"
-
-tilia <- data.table(model_df)[gattung_short == "Tilia", .(id,
-                                              provenance,
-                                              gattung_short,
-                                              ART_BOT,
-                                              BAUMHOEHE,
-                                              STANDALTER,
-                                              dbh_cm,
-                                              geometry,
-                                              day_2017)
-                              ][ , ":="(x_coord = sf::st_coordinates(geometry)[ ,1],
-                                    y_coord = sf::st_coordinates(geometry)[ ,2],
-                                    age = as.numeric(STANDALTER))]
-
-
-top_spec <- tilia[order(ART_BOT),.N , by = ART_BOT
-                  ][order(-N)][c(1,2,3,5,6), ART_BOT]
-
-tilia <- tilia[ART_BOT %in% top_spec & age < 200 & dbh_cm < 200
-               ][, `:=`(ART_BOT = as.factor(ART_BOT),
-                        provenance = as.factor(provenance))]
-
-
-
-tilia_gam <- mgcv::bam(dbh_cm ~ s(x_coord, y_coord, k = 12) +
-                           # s(day_2017, as.factor(ART_BOT), k = c(30, 6),  bs = "fs") +
-                           s(age, k  = 60) +
-                           provenance,
-                       data = tilia,
-                       method = "REML")
-
-
-tilia_gam <- mgcv::bam(dbh_cm ~
-                           # provenance +
-                           ART_BOT +
-                           s(day_2017, k = 120, by = ART_BOT) +
-                           s(age, k  = 120, by = ART_BOT),
-                       data = tilia[provenance == "s_wfs_baumbestand"],
-                       method = "REML",
-                       nthreads = 4)
-
-mgcv::gam.check(tilia_gam)
-summary(tilia_gam)
-
-

### 2022-10-11 analysis/scripts/06_baumscheiben_testing.R

@@ -1,293 +1,0 @@
-library(future)
-library(dplyr)
-library(ggplot2)
-library(sf)
-
-drake::loadd(full_data_set_clean)
-drake::loadd(baumscheiben_in_lists)
-
-
-
-
-
-
-
-# check baumscheiben ------------------------------------------------------
-glimpse(baumscheiben_in_lists)
-
-
-# rough plot
-baumscheiben_in_lists$s_Baumscheibe[1:5000, ] %>%
-    ggplot() +
-    geom_sf(color = "gray20")
-
-
-
-
-# test on subset
-
-baumsch <- baumscheiben_in_lists$s_Baumscheibe[1:15000, ] %>%
-    sf::st_set_crs(st_crs(full_data_set_clean))
-
-
-baumsch_buff <- sf::st_buffer(x = sf::st_centroid(baumsch),
-                              dist = 15)%>%
-    sf::st_set_crs(st_crs(full_data_set_clean))
-
-
-baumsch_centroid <- sf::st_centroid(baumsch)
-trees_in_buffs <- sf::st_intersects(baumsch_buff, full_data_set_clean)
-
-
-plan(multiprocess(workers = availableCores() - 1))
-
-
-options(future.globals.maxSize = 600 * 1024 ^ 2)
-
-tree_idx <- furrr::future_map(seq_along(trees_in_buffs),
-                              function(x){
-
-                                  idx <- which.min(st_distance(full_data_set_clean[trees_in_buffs[[x]], ],
-                                                               baumsch_centroid[x, ]))
-                                  out <- trees_in_buffs[[x]][idx]
-                                  out <- ifelse(length(out) == 0,
-                                                NA,
-                                                out)
-                              })
-
-
-baumsch$tree_idx <- tree_idx
-
-
-
-baumsch <- baumscheiben_in_lists$s_Baumscheibe[1:1000, ] %>%
-    sf::st_set_crs(st_crs(full_data_set_clean))
-
-
-baumsch_buff <- sf::st_buffer(x = sf::st_centroid(baumsch),
-                              dist = 15)%>%
-    sf::st_set_crs(st_crs(full_data_set_clean))
-
-
-baumsch_centroid <- sf::st_centroid(baumsch)
-trees_in_buffs <- sf::st_intersects(baumsch_buff, full_data_set_clean)
-
-
-
-tree_idx <- sapply(seq_along(trees_in_buffs),
-                   function(x){
-
[…]

### 2022-10-11 analysis/scripts/07_LCZ_exploration.R

@@ -1,110 +1,0 @@
-loadd(bounding_box)
-loadd(berlin_polygons)
-loadd(full_data_set_clean)
-loadd(baumscheiben_in_lists)
-
-
-tifpath <- "./analysis/data/raw_data/WUDAPT_LCZ.geotiff"
-
-rgdal::GDALinfo(tifpath)
-
-
-lcz_europe <- raster::raster(tifpath)
-
-raster::extent(lcz_europe)
-lcz_proj <- raster::crs(lcz_europe)
-
-berlin_polygons <- sf::st_transform(berlin_polygons,
-                                    crs = lcz_proj)
-berlin_polygons %>% raster::extent()
-
-# gdaluti::gdalbuildvrt(gdalfile = "yourraster.tif",
-#              output.vrt = "tmp.vrt",
-#              te = st_bbox(shp_layerbuffer1))
-#
-# tempraster <- raster("tmp.vrt")
-
-
-
-crop_raster <- raster::crop(lcz_europe,
-                            sf::st_union(berlin_polygons) %>% sf::st_centroid() %>%
-                                sf::st_buffer(35000)  %>% sf::st_as_sf())
-
-
-lcz_stars <- stars::st_as_stars(crop_raster)
-
-colourCount = length(unique(raster::values(crop_raster)))
-getPalette = colorRampPalette(RColorBrewer::brewer.pal(9, "Set1"))
-
-plot(lcz_stars, col = viridisLite::cividis(14))
-plot(lcz_stars, col =getPalette(colourCount))
-
-map_grid <- sf::st_graticule(berlin_polygons)
-ggplot() +
-    stars::geom_stars(data = lcz_stars,
-                      aes(x = x,
-                          y = y,
-                          fill = as.factor(WUDAPT_LCZ))) +
-    geom_sf(data = berlin_polygons,
-            color = "black",
-            fill = NA,
-            size = 0.5) +
-    geom_sf(data = map_grid, col = "gray20", alpha = .4) +
-    theme(panel.background = element_rect(fill = NA))
-
-
-
-# lcz_values_by_district <- raster::extract(crop_raster,
-#                                           sf::st_make_valid(berlin_polygons)) %>%
-# lapply(., function(x)prop.table(table(x))) %>%
-#     setNames(berlin_polygons$NAMGEM) %>% dplyr::bind_rows(.id = "bezirk")
-
-
-lcz_values_by_district <- raster::extract(crop_raster,
-                                          berlin_polygons) %>%
[…]

### 2022-10-11 analysis/scripts/07_extract_raster_vals.R

@@ -1,90 +1,0 @@
-
-#' Extract values from raster based on (point) sf
-#'
-#' @param sf_data sf dframe, e.g. berlin trees
-#' @param lcz_raster, raster
-#' @param buff_dist, buffer around points / polygons
-#' @param method character, "simple" (for categorical) or "bilinear"
-#'
-#' @return dframe, 1 row per sf geometry, with proportions of coverage
-#' @export
-#'
-assess_relative_cover <- function(sf_data, lcz_raster, buff_dist = 100, method = "simple"){
-
-    # ensure both objects have same projection
-
-    if(!sf::st_crs(sf_data)$wkt ==
-       raster::wkt(lcz_raster)){
-
-        #
-        #
-        # lcz_raster <- raster::projectRaster(lcz_raster,
-        #                       crs = sf::st_crs(full_data_set_clean)$wkt %>% raster::crs())
-        #
-        sf_data <- sf::st_transform(sf_data,
-                                    crs = raster::crs(lcz_raster))
-        message("adjusted CRS")
-    }
-
-    plot(lcz_raster)
-    plot(sf_data, add = TRUE, cex = 1, col = "black")
-
-
-    # extract values (proportionally) by tree buffer
-    # raster::beginCluster(n = 10, type='SOCK')
-
-    # extracted_vals <- raster::extract(lcz_raster, sf_data, method = "simple", buffer = buff_dist)
-    extracted_vals <- dplyr::bind_rows(lapply(raster::extract(lcz_raster, sf_data, method = method, buffer = buff_dist),
-                             function(x)prop.table(table(x))))
-    # raster::endCluster()
-
-}
-
-#' Split df in chunks for mapping/looping
-#'
-#' @param dframe df, nrow > 0
-#' @param cut_size numeric, number of rows in eaech chunk
-#'
-#' @return list of dfs with nrow <= cut_size
-#' @export
-#'
-split_by_n <- function(dframe, cut_size){
-
-    if(cut_size >= nrow(dframe)){
-        message("no splitting done")
-        return(dframe)
-    }
-
-    # make unique id
-    cuts <- base::cut(1:nrow(dframe),
-                      breaks = seq(from = 0, to = nrow(dframe) +1,
-                                   by = cut_size))
-    # make lists
-    dframe_lists <- base::split(dframe, f = as.factor(cuts))
-
-    return(dframe_lists)
-
-
-}
-
-library(future)
-future::plan(multisession)
-
-
-library(microbenchmark)
-
-
[…]

### 2022-10-11 analysis/scripts/08_bai_explo.R

@@ -1,1255 +1,0 @@
-# process BIWI dendro data
-# data from: Prof Christoph Schneider; Sebastian Schneider (MSc)
-
-# libraries ---------------------------------------------------------------
-
-library(dplR)
-library(dplyr)
-
-
-
-# functions ---------------------------------------------------------------
-
-
-
-predict_rw_to_bark <- function(meta, series_names, rwl){
-
-    `%nin%` <- Negate(`%in%`)
-
-    rwl$year <- as.numeric(row.names(rwl))
-
-    out_list <- list()
-
-    for(sn in series_names){
-        # grab plot id and subset data
-        current_plot <- substr(sn, 1, 5)
-
-
-
-        model_df <- rwl[ ,grepl(pattern = sprintf("(%s)|(%s)", "year", current_plot), x = colnames(rwl))]
-        # remove year and response for averaging
-        year_col <- which(colnames(model_df) == "year")
-        drop_columns <- c(year_col,
-                          grep(pattern = sn, x = colnames(model_df)))
-
-        print(drop_columns)
-
-        model_df <- data.frame(year = as.numeric(model_df$year),
-                               mean_rwl = rowMeans(
-                                   model_df[ , -drop_columns],
-                                   na.rm = TRUE),
-                               response = model_df[ , sn],
-                               # define weights based on non-NA entries per plot
-                               weights = 1 + ncol( model_df[ , -drop_columns]) - (apply(model_df[ , -drop_columns],
-                                                                                        MARGIN = 1,
-                                                                                        function(x){sum(is.na(x))})
-                               ))
-
-        print(colnames(model_df))
-        print(model_df$weights)
-
-
-        # form <- paste(sprintf("%s ~ year + ", sn), paste(colnames(model_df)[colnames(model_df) %nin% c(sn, "year") ], collapse = " + "))
-        # form <- sprintf("%s ~ year + mean_rwl", sn)
-        form <- as.formula("response ~ year + mean_rwl")
-        # print(form)
-
-        mod <- glm(formula = form,
-                   data = model_df,
-                   na.action = "na.omit",
[…]

### 2022-10-11 analysis/paper/paper.Rmd
rendering

@@
 
-Several urban tree species have been shown to have tendencies of enhanced growth rates and/or productivity compared to rural conspecifics [e.g., @briber2015; @obrien2012], which is typically attributed to increased temperatures [@jia2018; @pretzsch2017], however, a broad range of effect sizes and signs (i.e., reduced growth) specific to species and location are also commonly reported alongside.
+Several urban tree species have been shown to have tendencies of enhanced growth rates and/or productivity compared to rural conspecifics [e.g., @briber2015; @obrien2012], which is typically attributed to increased temperatures [@jia2018; @pretzsch2017], however, a broad range of effect sizes and signs (i.e., reduced growth) specific to species and location are also commonly reported alongside [e.g., @berland2020].
 For instance, @zhao2016 showed that productivity rates, as a proxy for growth, increased within urban clusters as urbanization intensifies using remotely sensed vegetation indices.
@@
 D(y, \hat{\mu}) = 2 \cdot \left\{ \frac{y−\hat{\mu}}{\hat{\mu}}  − log(\frac{y}{\hat{\mu}}) \right\}
-
 (\#eq:gam-deviance)

### 2022-10-11 analysis/paper/supplemental_information.Rmd
rendering

@@
       toc: true
+toc-title: "Table of Contents"
 header-includes:
@@
 ```
+
+# References

### 2022-10-11 R/plan.R

@@ -1094,22 +1094,22 @@ plan <- drake::drake_plan(
 
 
     # Reporting ------------------------------
-    # paper_html = rmarkdown::render(
-    #     knitr_in("./analysis/paper/paper.Rmd"),
-    #     output_dir = "./analysis/paper/",
-    #     output_file = file_out("paper_knit.html"),
-    #     # output_format = bookdown::html_document2(),
-    #     output_format = "bookdown::html_document2",
-    #     quiet = TRUE
-    # ),
-    # paper_word = rmarkdown::render(
-    #     knitr_in("./analysis/paper/paper.Rmd"),
-    #     output_dir = "./analysis/paper/",
-    #     output_file = file_out("paper_knit.docx"),
-    #     # output_file = "./paper_knit.html",
-    #     output_format = "bookdown::word_document2",
-    #     quiet = TRUE
-    # ),
+    paper_html = rmarkdown::render(
+        knitr_in("./analysis/paper/paper.Rmd"),
+        output_dir = "./analysis/paper/",
+        output_file = file_out("paper_knit.html"),
+        # output_format = bookdown::html_document2(),
+        output_format = "bookdown::html_document2",
+        quiet = TRUE
+    ),
+    paper_word = rmarkdown::render(
+        knitr_in("./analysis/paper/paper.Rmd"),
+        output_dir = "./analysis/paper/",
+        output_file = file_out("paper_knit.docx"),
+        # output_file = "./paper_knit.html",
+        output_format = "bookdown::word_document2",
+        quiet = TRUE
+    ),
     # paper_pdf = rmarkdown::render(
     #     knitr_in("./analysis/paper/paper.Rmd"),
     #     output_dir = "./analysis/paper/",

### 2022-10-17 README.Rmd
submission release

--- 2019-12-11 README.Rmd
@@
 )
-# Please put your title here to include it in the file below.
-Title <- "Title of your paper goes here"
 ```
@@
 
-[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)
+[![DOI](https://zenodo.org/badge/227480623.svg)](https://zenodo.org/badge/latestdoi/227480623)
 
-This repository contains the data and code for our paper:
+ <img src="analysis/figures/map_03_uhi.png" align="right" height = 275px, width = 275px />
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
+Repository for **Assessing the impact of urban heating on tree growth in Berlin with open inventory and environmental data**
 
-Our pre-print is online here:
+## Purpose
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+This repository contains an analysis of tree growth (potential) and its relationship with the Urban Heat Island (UHI) effect in Berlin using an extensive, publicly available data set. 
+The paper can be viewed/downloaded here as an [html file](./analysis/paper/paper_knit.html), or as a [docx file](./analysis/paper/Hurley-et-al_UFUG_berlin-trees.docx).
 
-### How to cite
+## Approach
 
-Please cite this compendium as:
+The folder structure uses the principal of a [**research compendium**](https://research-compendium.science/) following concepts developed by [Ben Marwick and others](https://www.tandfonline.com/doi/abs/10.1080/00031305.2017.1375986?journalCode=utas20).
+That means all analyses and the final write-up are organized in a (fairly) coherent manner, encouraging reproducibility.
 
-> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+A workflow-management software [**`drake`**](https://github.com/ropensci/drake) is implemented to run (and re-run) all analyses steps, ultimately producing the manuscript, supporting information, etc.
+`drake` is extensively documented ([see book](https://books.ropensci.org/drake/)), and is highly recommended.
+Note, however, that it is now superseded by the `targets` package.
 
-### How to download or install
+### Repository structure
 
-You can download the compendium as a zip from from this URL: </archive/master.zip>
+The repo structure follows that of a typical `R` package.
+There are:
 
-Or you can install this compendium as an R package, berlin.trees, from GitHub with:
+- high-level files (this readme, `NAMESPACE`, `DESCRIPTION`) relating to typical `R` package/compendium development
+- high-level files relating to `drake` and its implementation (`make.R`, `_drake.R`, `plan.R`)
+- high-level files that I use to get going with a package/compendium (`.Rprofile`, `_setup.R`)
 
-```{r gh-installation, eval = FALSE}
-# install.packages("devtools")
-remotes::install_github("/")
-```
+- `R` folder for functions (these are used in `plan.R` to drive the analyses)
+- `analysis` that contains raw and derived data, figures and the manuscript in different formats
 
-### Licenses
+### Using `drake`
 
-**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)
+#### The `plan.R`
 
-**Code :** See the [DESCRIPTION](DESCRIPTION) file
+`plan.R` is structured following the entire analyses from downloading tabular and spatial data, cleaning it, producing stats and plots, and then compiling the manuscript using functions.
+![](./functions.png)
+Every analyses step is named, and that name can then be used in a sub-sequent step (this is how drake manages dependencies).
 
-**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse
+### Workflow
 
-### Contributions
-
-We welcome contributions from everyone.
-Before you get started, please see our [contributor guidelines](CONTRIBUTING.md).
-Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md).
-By participating in this project you agree to abide by its terms.
+Note, that I've added some descriptions to the steps in `plan.R`, but typically I rely on the function help that I generate in `R/functions.R`.

### 2022-10-17 analysis/paper/paper.Rmd
submission release

@@
   - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^{a,b}$, ...
-  - $^a$ *Geoforschungszentrum Potsdam, Section 4.3, Potsdam, Germany*
+  - $^a$ *Geoforschungszentrum Potsdam, Climate Dynamics and Landscape Evolution, Potsdam, Germany*
   - $^b$ *Deutsches Archäologisches Institut, Dendroclimatology, Berlin, Germany*
+  - $^c$ *Geography Department, Humboldt-Universität zu Berlin, Berlin, Germany*
   - "* Corresponding author: hurley@gfz-potsdam.de"
@@
 
-**Delete**? Increased air temperatures due to UHIs can affect tree growth through altering several physiological processes across plant organs directly or indirectly [@dusenge2019].
+Increased air temperatures due to UHIs can affect tree growth through altering several physiological processes across plant organs directly or indirectly [@dusenge2019].
 Generally, reaction times at cellular level increase with temperature up to a maximum, after which a drop in enzymatic activity results in a species-dependent optimum curve  [@arcus2016; @parent2010].
@@
 The latter was deemed necessary, as several observations in multiple city districts were derived by linear relationships (i.e., straight-line), which do not capture the ontogenic growth dynamics of trees, and leave no variation related to variables other than age.
-All of these operations were recorded, and can be viewed and reproduced via the supplementary code.
+All of these operations were recorded, and can be viewed and reproduced via the code linked in the supplementary material S1.
 Lastly, observations with unlikely diameter-age combinations were identified via the residuals of a generalized linear model between diameter and age with a Gamma log-link distribution: if individual residuals exceeded seven times the median absolute deviation of all residuals, they were removed.
@@
 
-Similar cut-off approaches are frequently applied in allometric and trait databases [e.g., @kattge2020].
+Similar cut-off approaches are frequently applied in allometric and trait databases.
 We considered this approach (i.e., $MAD \cdot 7$) conservative, and necessary given the presence of clearly erroneous data.
-The approach and impact thereof are visualized in the supplementary information (Section Data quality control).
+The approach and impact thereof are visualized in the supplementary material S1 (Section Data quality control).
 <!-- which were processed with the bespoke software datacleanr by Hurley et al., submitted), -->
@@
 A global temporal (by year) and time-dependent ($j$, $\leq 1960$ or $>1960$) trend in cambial age were implemented with thin plate regression splines (default smoothing function in `mgcv`); $c_j$ is a time-group dependent intercept, while $tau$ represents a matrix of random effect coefficients for $k$ species to account for differences in growth patterns, and $e_i = \phi e_{i−1} + \epsilon_i$.
-A $3^{rd}$-order autocorrelation-moving average (ARMA) correlation structure was applied (i.e., $\phi(3,1)$) to account for the dependency of $\Delta r$ across years for each tree, as is frequently the case for tree growth [e.g., see @Fritts1989]; the detailed implementation for this model is given in the supplemental material code.
+A $3^{rd}$-order autocorrelation-moving average (ARMA) correlation structure was applied (i.e., $\phi(3,1)$) to account for the dependency of $\Delta r$ across years for each tree, as is frequently the case for tree growth [e.g., see @Fritts1989]; the detailed implementation for this model is given in code linked to in the supplemental material S1.
 $\Delta r$ was then derived for a range of cambial ages, and averaged for both time groups, allowing a comparison of recent to earlier growth.
@@
 Note, that Equation$~$\@ref(eq:gam-age) was considered as the appropriate null model for interpretations, and a model including all covariates and the temperature-age interaction was tested as a saturated model.
-Models were implemented with `mgcv::bam()` [@wood2017a;@li2020a] and readers are referred to the detailed implementation in the supplemental material code.
+Models were implemented with `mgcv::bam()` [@wood2017a;@li2020a] and readers are referred to the detailed implementation in the code linked to in the supplemental material.
 Considering all combinations of (excess) heat measures and covariates (with point, as well as 150/300$~m$ extractions), a total of `r nrow(drake::readd(bam_dbh_filtered))` models were applied.
@@
 The drought tolerance and performance under high temperatures is well documented for the latter two and features in tree selection frameworks, such as species climate matrices in @roloff2009 and in @brune2016, where several studies investigating (hydro)climatic sensitivity were summarized.
-Contrastingly, *Acer pseudoplatanus* showed decreased absolute growth with higher temperatures; this corresponds with their classification as *not very suitable* under [@roloff2009].
-*Platanus acerifolia* and *Acer platanoides* are considered *suitable* and *very suitable* under @roloff2009, yet show the greatest decline with temperatures here.
+Contrastingly, *Acer pseudoplatanus* showed decreased absolute growth with higher temperatures; this corresponds with their classification as *not very suitable* according to @roloff2009.
+*Platanus acerifolia* and *Acer platanoides* are considered *suitable* and *very suitable* (@roloff2009), yet show the greatest decline with temperatures here.
 Indeed, several studies reviewed in @brune2016 highlight *A. platanoides* as drought-tolerant.
@@
 We showcased the application of a GAM-driven analyses to identify the sensitivity of urban tree growth to environmental and urban drivers in a space-for-time framework. 
-The extensive data coverage (tree inventory and covariates) allowed us to account for location-specific factors and identify the most influences on growth potential (i.e., maximum diameter) for individual species.
+The extensive data coverage (tree inventory and covariates) allowed us to account for location-specific influences and identify the factors with strongest impacts on growth potentials (i.e., maximum diameter) for individual species.
 The identified age-dependent shift of temperature sensitivity, which was fairly consistent across species, as well as inter-specific differences thereof, will inform and guide further research. 
@@
 AGH and IH were supported through the Helmholtz-Climate-Initiative (HI-CAM), funded by the Helmholtz Initiative and Networking Fund; the authors are responsible for the content of this publication.
-We thank Prof Dr Christoph Schneider for providing dendrochronological data and pertinent discussion on its use.
+We thank Prof Dr Christoph Schneider for providing dendrochronological data and pertinent discussion on its use, as well as Berlin's senate for providing FAIR data and support for its use.
+
+# Supplementary material
+
+**File S1**: html document outlining 1) code and reproducibility for this study, 2) data quality control, and 3) handling of spatial autocorrelation.
 

### 2022-10-17 analysis/paper/supplemental_information.Rmd
submission release

@@
 
-Exploratory analyses of (initial) model resiudals showed spatial auto-correlation was present in the structure of the inventory data (identified by correlation of x and y coordinates with residuals).
+Exploratory analyses of (initial) model residuals showed spatial auto-correlation was present in the structure of the inventory data (identified by correlation of x and y coordinates with residuals).
 Note that this was expected given that trees planted on the same road (and/or neighborhood, district) may share several features, such as planting time, species, adjacent urban fabric. 

### 2022-10-17 R/plan.R

@@ -1105,7 +1105,7 @@ plan <- drake::drake_plan(
     paper_word = rmarkdown::render(
         knitr_in("./analysis/paper/paper.Rmd"),
         output_dir = "./analysis/paper/",
-        output_file = file_out("paper_knit.docx"),
+        output_file = file_out("Hurley-et-al_UFUG_berlin-trees.docx"),
         # output_file = "./paper_knit.html",
         output_format = "bookdown::word_document2",
         quiet = TRUE
@@ -1122,7 +1122,7 @@ plan <- drake::drake_plan(
    si_html = rmarkdown::render(
        knitr_in("./analysis/paper/supplemental_information.Rmd"),
        output_dir = "./analysis/paper/",
-       output_file = file_out("supplemental_information.html"),
+       output_file = file_out("Hurley-et-al_UFUG_berlin-trees_SI.html"),
        output_format = "bookdown::html_document2",
        quiet = TRUE
    ),

### 2023-09-24 analysis/paper/paper.Rmd
update repo - revisions

@@
 author:
-  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^{a,b}$, ...
+  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^{a,b,c}$ \\
   - $^a$ *Geoforschungszentrum Potsdam, Climate Dynamics and Landscape Evolution, Potsdam, Germany*
@@
 where $g()$ is a log-link for $\Delta r \sim Gamma$, $\Delta r$ is the annual radial increment for observation $i$. 
-A global temporal (by year) and time-dependent ($j$, $\leq 1960$ or $>1960$) trend in cambial age were implemented with thin plate regression splines (default smoothing function in `mgcv`); $c_j$ is a time-group dependent intercept, while $tau$ represents a matrix of random effect coefficients for $k$ species to account for differences in growth patterns, and $e_i = \phi e_{i−1} + \epsilon_i$.
-A $3^{rd}$-order autocorrelation-moving average (ARMA) correlation structure was applied (i.e., $\phi(3,1)$) to account for the dependency of $\Delta r$ across years for each tree, as is frequently the case for tree growth [e.g., see @Fritts1989]; the detailed implementation for this model is given in code linked to in the supplemental material S1.
+A global temporal (by year) and time-dependent ($j$, $\leq 1960$ or $>1960$) trend in cambial age were implemented with thin plate regression splines (default smoothing function in `mgcv`); $c_j$ is a time-group dependent intercept, while $\tau$ represents a matrix of random effect coefficients for $k$ species to account for differences in growth patterns, and $e_i = \phi e_{i−1} + \epsilon_i$.
+A $3^{rd}$-order autocorrelation-moving average (ARMA) correlation structure was applied (i.e., $\phi(3,1)$) to account for the dependency of $\Delta r$ across years for each tree, as is frequently the case for tree growth [e.g., see @Fritts1989]; the detailed implementation for this model is given in the code linked to in the supplemental material S1.
 $\Delta r$ was then derived for a range of cambial ages, and averaged for both time groups, allowing a comparison of recent to earlier growth.
@@
 
-where $g()$ is a log-link for $DBH \sim Gamma$, and $i$, $j$ are indices for observations and species, respectively, and $t$ refers to an (excess) heat measure from UrbClim, Berlin EnvAt or LandSat at different times (morning, afternoon/day, night; see Section [Temperature/UHI data](#sec:tempdata); $c$ is a species-dependent intercept, while $tau$ represents a matrix of random effect coefficients for $k$ districts to account for differing management regimes across the city.
+where $g()$ is a log-link for $DBH \sim Gamma$, and $i$, $j$ are indices for observations and species, respectively, and $t$ refers to an (excess) heat measure from UrbClim, Berlin EnvAt or LandSat at different times (morning, afternoon/day, night; see Section [Temperature/UHI data](#sec:tempdata); $c$ is a species-dependent intercept, while $\tau$ represents a matrix of random effect coefficients for $k$ districts to account for differing management regimes across the city.
 A global spatial smooth $f(x_i,y_i)$ (representing projected coordinates in UTM) via a Gaussian process [cf.
@@
 We expressedly urge authorities to implement FAIR (findable, accessible, inter-operable and reusable) data principles to enable work similar to ours, as well as the creation of larger-scale, multi-temporal databases and/or execution of synoptic studies [e.g., @ossola2020].
-This also includes the collection and porivison of standardized (meta) data for quality control and to improve statistical inference.
+This also includes the collection and provision of standardized (meta) data for quality control and to improve statistical inference.
 

### 2023-09-24 analysis/paper/supplemental_information.Rmd
update repo - revisions

@@
 author:
-  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^a$
+  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^{a,b,c}$
   - $^a$ *Geoforschungszentrum Potsdam, Section 4.3, Germany*
+  - $^b$ *Deutsches Archäologisches Institut, Dendroclimatology, Berlin, Germany*
+  - $^c$ *Geography Department, Humboldt-Universität zu Berlin, Berlin, Germany*
   - "* Corresponding author: hurley@gfz-potsdam.de"
@@
 
-The code to reproduce the paper and its analyses, as well as this supplement, are found at [https://github.com/the-Hull/berlin.trees/](), with permanently archived versions at **DOI**.
-The project relies on the workflow management tool `drake`, where all analyses steps are maintained within the file `berlin.trees/R/plan.R`, with necessary functions in `berlin.trees/R/functions.R`.
+The code to reproduce the paper and its analyses, as well as this supplement, are found at [https://github.com/the-Hull/berlin.trees/](), with a **permanently archived version at [https://doi.org/10.5281/zenodo.7217031]()**.
+The project relies on the workflow management tool `drake`, where all analyses steps are maintained within the file `berlin.trees/R/plan.R`, with functions representing actions/steps in `berlin.trees/R/functions.R`.
 The code downloads most necessary files to execute the analyses.
@@
 - William Michael Landau, (2018). The drake R package: a pipeline toolkit for
-  reproducibility and high-performance computing. **Journal of Open Source Software, 3(21)**,
-  550, https://doi.org/10.21105/joss.00550
+  reproducibility and high-performance computing. **Journal of Open Source Software, 3(21)**, 550, https://doi.org/10.21105/joss.00550
 

### 2023-09-24 R/functions.R

@@ -3584,6 +3584,7 @@ plot_temp_maps <- function(landsat,
     rn[1] <- floor(rn[1])
     rn[2] <- ceiling(rn[2])
 
+
     p_envat_night  <- ggplot2::ggplot() +
 
 
@@ -3669,7 +3670,7 @@ plot_temp_maps <- function(landsat,
         ggplot2::coord_sf(xlim = xlim, ylim = ylim) +
 
         ggplot2::scale_fill_distiller(palette = "RdBu",
-                                      # rescaler = mid_rescaler(),
+                                      rescaler = mid_rescaler(),
                                       na.value = "transparent")   +
 
 
@@ -3726,7 +3727,7 @@ plot_temp_maps <- function(landsat,
         ggplot2::coord_sf(xlim = xlim, ylim = ylim) +
 
         ggplot2::scale_fill_distiller(palette = "RdBu",
-                                      # rescaler = mid_rescaler(),
+                                      rescaler = mid_rescaler(),
                                       na.value = "transparent")   +
 
         # ggplot2::scale_fill_viridis_c(option = "plasma", na.value = "transparent",
@@ -3780,7 +3781,7 @@ plot_temp_maps <- function(landsat,
         ggplot2::coord_sf(xlim = xlim, ylim = ylim) +
 
         ggplot2::scale_fill_distiller(palette = "RdBu",
-                                      # rescaler = mid_rescaler(),
+                                      rescaler = mid_rescaler(),
                                       na.value = "transparent")   +
 
         # ggplot2::scale_fill_viridis_c(option = "plasma", na.value = "transparent",
@@ -3839,7 +3840,8 @@ FFGGHH
         p_list$p_urbclim_night +
 
 
-        plot_layout(design = layout, heights = 1/3)
+        plot_layout(design = layout, heights = 1/3) +
+        theme(plot.margin = margin(r = 13))
 
 
     tiff(filename = file, width = width, height = height, res = dpi, units = "cm")
@@ -3849,7 +3851,7 @@ FFGGHH
     print(p_map)
     grid::grid.text(gp = gp, "UrbClim", x = 0.03, y = 1/3 * 0.5, rot = 90)
     grid::grid.text(gp = gp, "Berlin EnvAT", x = 0.03, y = 1/3 * 0.5 + 1/3, rot = 90)
[…]

### 2023-09-24 R/plan.R

@@ -641,7 +641,7 @@ plan <- drake::drake_plan(
        age_expression = age_expr,
        prediction_range = "within",
        base_size = 12,
-       file = paste0(base_path_fig, "06_double_gam-dbh_temp-day2007_lcz6.png"),
+       file = paste0(base_path_fig, "06_double_gam-dbh_temp-day2007_lcz6.tiff"),
        height = 0.75 * journal_fig_width_double,
        width = journal_fig_width_double,
        dpi = 300),

### 2023-09-24 analysis/paper/paper.Rmd
update paper

@@
 <!-- <!-- Species-specifc growth responses in relation across Berlin were assessed in relat --> 
-<!-- We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of (excess) heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; LandSat-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section$~$\@ref(sec:stat-gam) for details).
+<!-- We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of (excess) heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; MODIS-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section$~$\@ref(sec:stat-gam) for details).
 -->
@@
 where $T$ is temperature ($^\circ C$) $x$ and $y$ define an urban grid cell.
-The LandSat-derived surface UHI data set by @chakraborty2019 (referred to as LandSat) estimates its measure in a similar fashion and the reader is referred to the detailed description therein; note this data set provides day and night-time averaged UHI estimates at 500$~m$ resolution, which were extracted for the hottest summer in this record (2007).
+The MODIS-derived surface UHI data set by @chakraborty2019 (referred to as MODIS) estimates its measure in a similar fashion and the reader is referred to the detailed description therein; note this data set provides day and night-time averaged UHI estimates at 500$~m$ resolution, which were extracted for the hottest summer in this record (2007).
 
@@
 <!-- Species-specifc growth responses in relation across Berlin were assessed in relat -->
-We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of (excess) heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; LandSat-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section [GAMs](#sec:stat-gam) for details).
+We modeled the stem diameter ($DBH$) of Berlin’s ten most abundant species (contingent to ancillary data availability) in relationship to their location, age, a measure of (excess) heat [UrbClim by @deridder2015; Berlin Environmental Atlas models; MODIS-derived surface urban heat island by @chakraborty2019], and additional environmental covariates with generalized additive models (GAMs, see Section [GAMs](#sec:stat-gam) for details).
 Covariates were extracted at 150 and 300$~m$ to infer the impact of reference scale of the urban fabric on tree growth. 
@@
 
-where $g()$ is a log-link for $DBH \sim Gamma$, and $i$, $j$ are indices for observations and species, respectively, and $t$ refers to an (excess) heat measure from UrbClim, Berlin EnvAt or LandSat at different times (morning, afternoon/day, night; see Section [Temperature/UHI data](#sec:tempdata); $c$ is a species-dependent intercept, while $\tau$ represents a matrix of random effect coefficients for $k$ districts to account for differing management regimes across the city.
+where $g()$ is a log-link for $DBH \sim Gamma$, and $i$, $j$ are indices for observations and species, respectively, and $t$ refers to an (excess) heat measure from UrbClim, Berlin EnvAt or MODIS at different times (morning, afternoon/day, night; see Section [Temperature/UHI data](#sec:tempdata); $c$ is a species-dependent intercept, while $\tau$ represents a matrix of random effect coefficients for $k$ districts to account for differing management regimes across the city.
 A global spatial smooth $f(x_i,y_i)$ (representing projected coordinates in UTM) via a Gaussian process [cf.
@@
 
-As the spatial extent and coverage varied between temperature and ancillary data, more complex models (and specifically those including planting bed area and LandSat temperatures) typically also had fewer total observations.
+As the spatial extent and coverage varied between temperature and ancillary data, more complex models (and specifically those including planting bed area and MODIS temperatures) typically also had fewer total observations.
 While this prevented a full comparison with information-based model selection criteria like AIC, the appropriateness of models that differed only in their implementation of the temperature-age interaction (i.e., $f_j(temp_i)$ $vs.$ $f_j(temp_i, age_i)$) could be assessed.
@@
 The inclusion of temperature (additive structure) and temperature-age interactions improved the predictive skill, with interactions reaching higher explained deviance (i.e., skill).
-Models including the LandSat-derived UHI measures performed best on average, where models containing $LCZ6_{\frac{}{150~m}}$ or all covariates (saturated model) did best (Fig.$~$\@ref(fig:fig-model-deviance)), which also holds true for other temperature/UHI measures.
+Models including the MODIS-derived UHI measures performed best on average, where models containing $LCZ6_{\frac{}{150~m}}$ or all covariates (saturated model) did best (Fig.$~$\@ref(fig:fig-model-deviance)), which also holds true for other temperature/UHI measures.
 Note that non-temperature models containing planting bed area and $LCZ6_{\frac{}{150~m}}$ alone outperformed several of those including other temperature measures and covariate combinations.
@@
 Differences in sample sizes between model structures due to covariate and/or temperature data coverage were considerable (e.g., planting bed area $vs.$ soil nutrient models for UrbClim UHI).
-Thus, we considered the model with the overall greatest predictive skill and largest sample for further investigations, namely $LCZ6_{\frac{}{150~m}}$ for LandSat derived UHI.
+Thus, we considered the model with the overall greatest predictive skill and largest sample for further investigations, namely $LCZ6_{\frac{}{150~m}}$ for MODIS derived UHI.
 For this variable combination, the model structure with age and temperature interaction also had a lower AIC ($AIC~=~8.95 \cdot 10^5$) compared to the additive one ($AIC~=~8.98 \cdot 10^5$).
@@
 
-The final model, comprising a temperature-age interaction, the covariate $LCZ6_{\frac{}{150~m}}$ and LandSat derived UHI following Equation$~$\@ref(eq:gam-full), showed good agreement between predicted and observed $DBH$ for the majority of observations (Fig.$~$\@ref(fig:fig-model-performance); $R^2~=~0.79$) resulting in a fairly low MAE of $6.03~cm$, and a RMSE of $7.7\cdot 10^{-7}$.
+The final model, comprising a temperature-age interaction, the covariate $LCZ6_{\frac{}{150~m}}$ and MODIS derived UHI following Equation$~$\@ref(eq:gam-full), showed good agreement between predicted and observed $DBH$ for the majority of observations (Fig.$~$\@ref(fig:fig-model-performance); $R^2~=~0.79$) resulting in a fairly low MAE of $6.03~cm$, and a RMSE of $7.7\cdot 10^{-7}$.
 While some prediction errors are large (spread in Fig.$~$\@ref(fig:fig-model-performance)), the predictions are unbiased (see correspondence of 1:1 and least-squares fit in Fig.$~$\@ref(fig:fig-model-performance)).
@@
 <!-- 		- Supplement species matrix etc. with these analyses   -->
-<!-- 		- *Landsat temp best on average - easily accessible/maintainable as compared to model outputs -->
+<!-- 		- *MODIS temp best on average - easily accessible/maintainable as compared to model outputs -->
 <!-- * $lcz6_{150}$ vs. $lcz6_{300}$ - scale of impact (further investigation, note in discussion)  -->
@@
 However, note the importance ofplanting bed area and LCZ6 cover is apparent even without including temperature.
-The LandSat-derived UHI measures provide the best fit on average, with the model including a temperature-age interaction and LCZ6 cover scoring highest overall.
+The MODIS-derived UHI measures provide the best fit on average, with the model including a temperature-age interaction and LCZ6 cover scoring highest overall.
 
 **Fig.
-\@ref(fig:fig-model-performance)** Predicted $vs.$ observed data for the best model according to explained deviance and AIC comparison, based on the LandSat-derived UHI measure [@chakraborty2019] for average summer conditions (2007) and LCZ6 [open to mid-rise,  @stewart2012].
+\@ref(fig:fig-model-performance)** Predicted $vs.$ observed data for the best model according to explained deviance and AIC comparison, based on the MODIS-derived UHI measure [@chakraborty2019] for average summer conditions (2007) and LCZ6 [open to mid-rise,  @stewart2012].
 Hexagons and colors represent x-y bins (i.e., in the observed-predicted space) and their counts (N), and the red line is a least-squares fit.
@@
 **Fig.
-\@ref(fig:fig-heat-effect)** Age-group averaged relationships between the LandSat-derived UHI and stem diameter for the ten most abundant species from the best model (i.e., Equation$~$\@ref(eq:gam-full) with LCZ6 as covariate).
+\@ref(fig:fig-heat-effect)** Age-group averaged relationships between the MODIS-derived UHI and stem diameter for the ten most abundant species from the best model (i.e., Equation$~$\@ref(eq:gam-full) with LCZ6 as covariate).
 Note that estimates were derived with $LCZ6_{\frac{}{150~m}} = 0.5$, and that y-axis scales differ between panel rows.
@@
 **Fig.
-\@ref(fig:fig-lcz6-effect)** Age-group averaged relationships between the LCZ6 and stem diameter for two $Tilia$ genera from the best model (i.e., Equation$~$\@ref(eq:gam-full) with LandSat-derived UHI as excess heat measure).
+\@ref(fig:fig-lcz6-effect)** Age-group averaged relationships between the LCZ6 and stem diameter for two $Tilia$ genera from the best model (i.e., Equation$~$\@ref(eq:gam-full) with MODIS-derived UHI as excess heat measure).
 LCZ6 cover was derived as a weighted average within a radius of 150$~m$ around each tree.
